---
title: "02_2026_analysis"
format: html
---
### Load Libraries

```{r, warning=FALSE}
# --- Load libraries ---
library(terra)
library(tidyverse)
library(sf)
library(sp)
library(gstat)
library(randomForest)
library(FNN)
library(purrr)
library(RColorBrewer)
library(ggrepel)
library(patchwork)
library(blockCV)
library(leaflet)
library(tidytext)
library(viridis)
```

### Load Data and Prepare Datasets

```{r, warning=FALSE}

# --- Load data ---
covariate_stack <- rast("../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m.tif")
soil_covariate_data_scaled_SP <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/soil_covariate_data_scaled_SP.csv")
soil_covariate_data_scaled_C  <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/soil_covariate_data_scaled_C.csv")

# Load carbon + nitrogen dataset
soil_with_CN <- read_csv("../LOCAL DATA DO NOT PUSH/carbon/soil_data_with_CN.csv") %>%
  rename(c = leco_carbon_average,
         n = leco_nitrogen_average)

# Split by sampling type
SP_soil_with_CN <- soil_with_CN %>% filter(type == "SPID")
C_soil_with_CN  <- soil_with_CN %>% filter(type == "CID")

# Merge CN values into soil covariate tables by x,y coordinate
soil_SP_full <- soil_covariate_data_scaled_SP %>%
  left_join(SP_soil_with_CN %>% select(x,y,c,n), by=c("x","y"))

soil_C_full <- soil_covariate_data_scaled_C %>%
  left_join(C_soil_with_CN %>% select(x,y,c,n), by=c("x","y"))

# Define soil variable names using actual dataset column names
soil_vars <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")

# Define covariates by removing coordinates + soil variables
covariates <- setdiff(names(soil_SP_full), c("x","y", soil_vars))

# Convert to sf projected
soil_sf_SP <- st_as_sf(soil_SP_full, coords = c("x","y"), crs = 4326) %>% st_transform(32617)
soil_sf_C  <- st_as_sf(soil_C_full, coords = c("x","y"), crs = 4326) %>% st_transform(32617)


# #rename variables:
# rename_soils <- function(x) {
#   x %>%
#     rename(
#       lbc        = lbc_1,
#       lbc_eq     = lb_ceq,
#       ph         = p_h_2,
#       calcium    = ca,
#       potassium  = k,
#       magnesium  = mg,
#       manganese  = mn,
#       phosphorus = p,
#       zinc       = zn,
#       carbon     = c,
#       nitrogen   = n
#     )
# }
# 
# soil_sf_SP <- rename_soils(soil_sf_SP)
# soil_sf_C  <- rename_soils(soil_sf_C)
# 
# soil_vars_clean <- c("lbc","lbc_eq","ph","calcium","potassium","magnesium","manganese","phosphorus","zinc","carbon","nitrogen")
```

### UMAP test

```{r}
library(umap)
# Prepare data for UMAP (only soil variables, no coordinates)
soil_umap_SP <- soil_sf_SP %>%
  st_drop_geometry() %>%
  select(all_of(soil_vars)) %>%
  drop_na()

soil_umap_C <- soil_sf_C %>%
  st_drop_geometry() %>%
  select(all_of(soil_vars)) %>%
  drop_na()

# Run UMAP separately on SP and Composite datasets
set.seed(626)  # for reproducibility
umap_SP <- umap(soil_umap_SP)
umap_C <- umap(soil_umap_C)

#plot UMAP results
umap_df_SP <- as.data.frame(umap_SP$layout) %>% mutate(dataset = "Single Point")
umap_df_C <- as.data.frame(umap_C$layout) %>% mutate(dataset = "Composite")
umap_df <- bind_rows(umap_df_SP, umap_df_C)
ggplot(umap_df, aes(x = V1, y = V2, color = dataset)) +
  geom_point(alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  labs(title = "UMAP of Soil Nutrient Profiles",
       x = "UMAP Dimension 1", y = "UMAP Dimension 2", color = "Dataset") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

#cluster:
library(dbscan)
set.seed(626)
dbscan_SP <- dbscan(umap_SP$layout, eps = 0.5, minPts = 5)
dbscan_C  <- dbscan(umap_C$layout, eps = 0.5, minPts = 5)
umap_df_SP$cluster <- as.factor(dbscan_SP$cluster)
umap_df_C$cluster  <- as.factor(dbscan_C$cluster)
ggplot(umap_df_SP, aes(x = V1, y = V2, color = cluster)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_d() +
  theme_minimal(base_size = 14) +
  labs(title = "UMAP Clusters of Single Point Soil Nutrient Profiles",
       x = "UMAP Dimension 1", y = "UMAP Dimension 2", color = "Cluster") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggplot(umap_df_C, aes(x = V1, y = V2, color = cluster)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_d() +
  theme_minimal(base_size = 14) +
  labs(title = "UMAP Clusters of Composite Soil Nutrient Profiles",
       x = "UMAP Dimension 1", y = "UMAP Dimension 2", color = "Cluster") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

## Explore cluster characteristics (e.g., mean nutrient values per cluster)
cluster_summary_SP <- soil_umap_SP %>%
  mutate(cluster = as.factor(dbscan_SP$cluster)) %>%
  group_by(cluster) %>%
  summarise(across(all_of(soil_vars), mean, na.rm = TRUE))
cluster_summary_C <- soil_umap_C %>%
  mutate(cluster = as.factor(dbscan_C$cluster)) %>%
  group_by(cluster) %>%
  summarise(across(all_of(soil_vars), mean, na.rm = TRUE))  
  
print(cluster_summary_SP)
print(cluster_summary_C)

## Note: UMAP and clustering results can be sensitive to parameters (e.g., eps for DBSCAN), so you may want to experiment with those for better separation or more meaningful clusters.

#how can i use this to predict properties?

```


### Plot Composite vs Single Point with Outlier Labels

```{r, warning=FALSE}

# 1) Select only nutrient columns + coordinates
SP_vals <- soil_SP_full %>% select(x, y, all_of(soil_vars)) %>% mutate(dataset = "SP")
C_vals  <- soil_C_full %>% select(x, y, all_of(soil_vars)) %>% mutate(dataset = "Composite")

# 2) Combine and reshape to long format
soil_long <- bind_rows(SP_vals, C_vals) %>%
  pivot_longer(cols = all_of(soil_vars), names_to = "variable", values_to = "value")

# 3) Pivot wide: SP vs Composite in one row per point and nutrient
soil_wide <- soil_long %>%
  pivot_wider(names_from = dataset, values_from = value) %>%
  drop_na(SP, Composite)   # keeps only nutrients where both exist

  
# --- Attach PointLabel to merged dataset ---
# We join from SP_soil_with_CN since PointLabel is stable there
soil_wide_loc <- soil_wide %>%
  left_join(SP_soil_with_CN %>% select(x, y, PointLabel), by = c("x","y"))

# If you want cleaner facet labels again:
soil_wide_loc <- soil_wide_loc %>%
  mutate(variable = recode(variable,
                           "lbc_1" = "LBC1",
                           "lb_ceq" = "LBCeq",
                           "p_h_2" = "pH",
                           "ca" = "Ca",
                           "k" = "K",
                           "mg" = "Mg",
                           "mn" = "Mn",
                           "p" = "P",
                           "zn" = "Zn",
                           "c" = "C",
                           "n" = "N"))
# --- Compute residuals and identify outliers ---
soil_outliers <- soil_wide_loc %>%
  group_by(variable) %>%
  mutate(
    residual = SP - Composite,
    z = (residual - mean(residual, na.rm = TRUE)) / sd(residual, na.rm = TRUE),
    outlier = abs(z) > 2   # threshold
  )

# ✔ Check how many outliers per nutrient (optional)
soil_outliers %>% count(variable, outlier)

soil_no_outliers <- soil_outliers %>%
  filter(outlier == FALSE)

# --- Plot with 1:1 line and label only outliers ---
ggplot(soil_outliers, aes(x = Composite, y = SP)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  geom_text_repel(
    data = ~ dplyr::filter(.x, outlier == TRUE),
    aes(label = PointLabel),
    size = 2,
    fontface = "bold",
    max.overlaps = Inf,
    min.segment.length = 0
  ) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Composite vs Single Point with Outlier Labels",
    x = "Composite",
    y = "Single Point"
  ) +
  theme(strip.text = element_text(face = "bold"))
```

### Plot without outliers and R² values

```{r, warning=FALSE}

# --- Compute R² for each nutrient ---
r2_labels <- soil_no_outliers %>%
  filter(!is.na(SP), !is.na(Composite), !is.na(variable)) %>%
  group_by(variable) %>%
  summarise(
    r2 = cor(SP, Composite, use = "complete.obs")^2,
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 2)))

# --- Remove any NA variables before plotting ---
soil_no_outliers <- soil_no_outliers %>%
  filter(!is.na(variable), variable != "NA")

# --- Ensure custom facet order ---
var_order <- c("pH", "Ca", "K", "Mg", "Mn", "P", "Zn", "C", "N")
soil_no_outliers$variable <- factor(soil_no_outliers$variable, levels = var_order)
r2_labels$variable <- factor(r2_labels$variable, levels = var_order)

# --- Plot ---
ggplot(soil_no_outliers, aes(x = Composite, y = SP)) +
  geom_point(size = 2, alpha = 0.7, color = "#4C72B0") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "red", linewidth = 0.9) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.9) +
  geom_text(
    data = r2_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.1, vjust = 1.4,
    size = 3.3, fontface = "bold"
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Composite vs Single Point Nutrient Values",
    x = "Composite (ppm)",
    y = "Single Point (ppm)"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# ggsave("../outputs/figures/composite_vs_singlepoint_no_outliers_with_R2.png", width = 16, height = 9, units = "in", dpi = 300)

```

### Density Plots Comparing Distributions

```{r, warning=FALSE eval=FALSE}
# ---- Prepare long-format dataset for density plots ----

# Select only the soil variables + dataset label
soil_SP_long <- soil_SP_full %>%
  mutate(dataset = "Single Point") %>%
  select(dataset, all_of(soil_vars)) %>%
  pivot_longer(cols = all_of(soil_vars), names_to = "variable", values_to = "value")

soil_C_long <- soil_C_full %>%
  mutate(dataset = "Composite") %>%
  select(dataset, all_of(soil_vars)) %>%
  pivot_longer(cols = all_of(soil_vars), names_to = "variable", values_to = "value")


# Optional: nicer variable names for plotting
soil_long_all <- soil_long_all %>%
  mutate(variable = recode(variable,
                           "lbc_1" = "LBC1",
                           "lb_ceq" = "LBCeq",
                           "p_h_2" = "pH",
                           "ca" = "Ca",
                           "k" = "K",
                           "mg" = "Mg",
                           "mn" = "Mn",
                           "p" = "P",
                           "zn" = "Zn",
                           "c" = "C",
                           "n" = "N"))
#plot all nutrients with SP on x axis and C on y axis

# Function to make each density plot with nutrient name as title
make_density <- function(var_name){
  ggplot(filter(soil_long_all, variable == var_name),
         aes(x = value, fill = dataset)) +
    geom_density(alpha = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    theme_minimal(base_size = 14) +
    labs(title = var_name, x = NULL, y = "Density", fill = "Sampling Method") +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "bottom"
    )
}

# Create each plot
p_pH <- make_density("pH")
p_Ca <- make_density("Ca")
p_K  <- make_density("K")
p_Mg <- make_density("Mg")
p_Mn <- make_density("Mn")
p_P  <- make_density("P")
p_Zn <- make_density("Zn")
p_C  <- make_density("C")
p_N  <- make_density("N")

# Arrange layout (Zn centered)
final_plot <- 
  (p_Ca | p_K | p_Mg) /
  (p_Mn | p_P | p_pH) /
  (p_C | p_Zn | p_N) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

# Add main title
# final_plot + 
#   plot_annotation(
#     title = "Soil Nutrient Distributions: Single Point vs Composite Sampling",
#     theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
#   )

#again but log scale all but pH

final_plot_log <-
  (
    (p_Ca + scale_x_log10() | p_K + scale_x_log10() | p_Mg + scale_x_log10()) /
    (p_Mn + scale_x_log10() | p_P + scale_x_log10() | p_pH) /
    (p_C + scale_x_log10()  | p_Zn + scale_x_log10() | p_N + scale_x_log10())
  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Soil Nutrient Distributions (Log Scale): Single Point vs Composite Sampling",
    theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  ) &
  theme(legend.position = "bottom")

final_plot_log

```
### Spatial Bootstrapping and Model Comparison

```{r, warning=FALSE}

# --- Spatial blocking function ---
generate_spatial_bootstrap_block <- function(sf_data,
                                             n_boot = 50,
                                             train_frac = 0.7,
                                             theRange = 50,     # in meters; adjust based on variogram range
                                             n_folds = 5,       # number of spatial blocks
                                             seed = 123) {
  set.seed(seed)

  # Run blockCV spatial blocking to assign folds based on variogram range
  sb <- spatialBlock(speciesData = sf_data,
                     theRange = theRange,       # block size = variogram range
                     k = n_folds,               # number of folds (blocks)
                     selection = "random",      # random assignment of folds
                     showBlocks = FALSE,
                     progress = FALSE)

  # Extract fold assignments
  sf_data$fold <- sb$foldID

  # --- Build list of bootstrapped train/test splits ---
  boot_list <- vector("list", n_boot)
  for (i in seq_len(n_boot)) {
    # randomly pick some folds for training (e.g., 70%)
    n_train_folds <- ceiling(train_frac * n_folds)
    train_folds <- sample(unique(sf_data$fold), n_train_folds)

    df <- sf_data %>%
      mutate(set = ifelse(fold %in% train_folds, "train", "test"))

    boot_list[[i]] <- df
  }

  return(boot_list)
}


# --- Safe R² function ---
safe_cor2 <- function(x, y){
  if(length(x)==0 || length(y)==0) return(NA)
  if(all(is.na(x)) || all(is.na(y))) return(NA)
  cor(x, y, use="complete.obs")^2
}
  #uses correlation with complete.obs to avoid NA issues... but I am not sure if this is the best way to handle it.

# --- Generate bootstraps ---
# Estimate an average variogram range first if you have it
# e.g., from your variogram fits:
# mean_range <- mean(fit_variograms$range, na.rm = TRUE)
# Otherwise start with 50 m as a field-scale approximation

boot_SP <- generate_spatial_bootstrap_block(soil_sf_SP,
                                            n_boot = 50,
                                            train_frac = 0.7,
                                            theRange = 50,
                                            n_folds = 5)

boot_C  <- generate_spatial_bootstrap_block(soil_sf_C,
                                            n_boot = 50,
                                            train_frac = 0.7,
                                            theRange = 50,
                                            n_folds = 5)


# --- Random Forest function ---
run_RF <- function(boot_list, covariates, soil_vars){
  results <- list()
  for(var in soil_vars){
    boot_results <- map(boot_list, function(df){
      train_df <- st_drop_geometry(df %>% filter(set=="train")) %>% select(all_of(c(covariates,var))) %>% na.omit()
      test_df  <- st_drop_geometry(df %>% filter(set=="test")) %>% select(all_of(c(covariates,var))) %>% na.omit()
      if(nrow(train_df) < 5 | nrow(test_df) < 5) return(list(RMSE=NA,R2=NA))
      rf_model <- randomForest(x = train_df[,covariates], y = train_df[[var]], ntree = 500)
      pred <- predict(rf_model, newdata = test_df[,covariates])
      list(model = rf_model, pred = pred,
           RMSE = sqrt(mean((pred - test_df[[var]])^2)),
           R2 = safe_cor2(pred, test_df[[var]]))
    })
    results[[var]] <- boot_results
  }
  return(results)
}

# --- Ordinary Kriging function with fallback ---
# --- Ordinary Kriging function with fallback (always returns R² even if only nugget) ---
run_OK_safe <- function(boot_list, soil_vars) {
  results <- list()

  for (var in soil_vars) {
    boot_results <- map(boot_list, function(df) {
      train_sf <- df %>% filter(set == "train")
      test_sf  <- df %>% filter(set == "test")

      train_vals <- train_sf[[var]]
      test_vals  <- test_sf[[var]]

      # --- skip if too few usable data points ---
      if (all(is.na(train_vals)) || length(unique(train_vals)) <= 1) {
        pred <- rep(mean(train_vals, na.rm = TRUE), nrow(test_sf))
        rmse <- sqrt(mean((pred - test_vals)^2, na.rm = TRUE))
        r2   <- ifelse(var(pred, na.rm = TRUE) > 0 && var(test_vals, na.rm = TRUE) > 0,
                       suppressWarnings(cor(pred, test_vals, use = "complete.obs")^2),
                       0)
        return(list(pred = pred, RMSE = rmse, R2 = r2))
      }

      # --- variogram and model fit (fallback to nugget if needed) ---
      vgm_model <- tryCatch({
        v_emp <- variogram(as.formula(paste(var, "~ 1")), train_sf)
        fit.variogram(v_emp, vgm(psill = var(train_vals, na.rm = TRUE),
                                 model = "Exp",
                                 nugget = 0.1,
                                 range = max(v_emp$dist, na.rm = TRUE)/3))
      }, error = function(e) NULL)

      if (is.null(vgm_model)) {
        vgm_model <- vgm(psill = var(train_vals, na.rm = TRUE),
                         model = "Nug", nugget = 0.1)
      }

      # --- try kriging prediction ---
      ok_pred <- tryCatch({
        krige(as.formula(paste(var, "~ 1")), train_sf, test_sf, model = vgm_model)
      }, error = function(e) NULL)

      # --- fallback if krige fails ---
      pred <- if (!is.null(ok_pred) && "var1.pred" %in% names(ok_pred)) {
        ok_pred$var1.pred
      } else {
        rep(mean(train_vals, na.rm = TRUE), nrow(test_sf))
      }

      # --- safe metrics (handles no complete pairs) ---
      rmse <- sqrt(mean((pred - test_vals)^2, na.rm = TRUE))
      r2 <- tryCatch({
        if (var(pred, na.rm = TRUE) == 0 || var(test_vals, na.rm = TRUE) == 0) 0
        else cor(pred, test_vals, use = "complete.obs")^2
      }, error = function(e) 0)

      list(pred = pred, RMSE = rmse, R2 = r2)
    })
    results[[var]] <- boot_results
  }
  return(results)
}


# --- Regression Kriging function with fallback ---
run_RK_safe <- function(boot_list, covariates, soil_vars){
  results <- list()
  for(var in soil_vars){
    boot_results <- map(boot_list, function(df){
      train_sf <- df %>% filter(set=="train")
      test_sf  <- df %>% filter(set=="test")
      
      # Remove rows with NA in predictors or target
      train_df <- st_drop_geometry(train_sf) %>% select(all_of(c(covariates,var))) %>% na.omit()
      test_df  <- st_drop_geometry(test_sf) %>% select(all_of(c(covariates,var))) %>% na.omit()
      
      # Skip if too few points
      if(nrow(train_df)<5 | nrow(test_df)<5) return(list(RMSE=NA,R2=NA))
      
      # Match sf objects to filtered rows
      train_sf <- train_sf[rownames(train_df), ]
      test_sf  <- test_sf[rownames(test_df), ]
      
      # Trend model (RF)
      rf_model <- randomForest(x=train_df[,covariates], y=train_df[[var]], ntree=500)
      trend_train <- predict(rf_model, newdata=train_df[,covariates])
      trend_test  <- predict(rf_model, newdata=test_df[,covariates])
      
      # Residuals
      residuals_train <- train_df[[var]] - trend_train
      
      # Make a new sf object with only the filtered residuals
      train_res_sf <- train_sf %>% mutate(residuals=residuals_train)
      
      # Variogram for residuals
      vgm_model <- tryCatch(
        variogram(residuals~1, train_res_sf) %>%
          fit.variogram(vgm(psill=var(residuals_train, na.rm=TRUE), model="Exp", nugget=0.1, range=100)),
        error=function(e) NULL
      )
      if(is.null(vgm_model)) vgm_model <- vgm(psill=var(residuals_train, na.rm=TRUE), model="Exp", nugget=0.1, range=100)
      
      # Krige residuals
      rk_pred <- tryCatch(
        krige(residuals~1, train_res_sf, test_sf, model=vgm_model),
        error=function(e) NULL
      )
      # fallback if krige fails or lengths mismatch
      if(is.null(rk_pred) || length(rk_pred$var1.pred)!=nrow(test_sf)){
        rk_pred <- data.frame(var1.pred=rep(0,nrow(test_sf)))
      }
      
      # Combine trend + residuals
      final_pred <- trend_test + rk_pred$var1.pred
      rmse <- sqrt(mean((final_pred - test_df[[var]])^2, na.rm=TRUE))
      r2   <- safe_cor2(final_pred, test_df[[var]])
      
      list(RMSE=rmse, R2=r2)
    })
    results[[var]] <- boot_results
  }
  return(results)
}

# --- Run all methods ---
RF_results_SP <- run_RF(boot_SP, covariates, soil_vars)
RF_results_C  <- run_RF(boot_C, covariates, soil_vars)
OK_results_SP <- run_OK_safe(boot_SP, soil_vars)
OK_results_C  <- run_OK_safe(boot_C, soil_vars)
RK_results_SP <- run_RK_safe(boot_SP, covariates, soil_vars)
RK_results_C  <- run_RK_safe(boot_C, covariates, soil_vars)

# --- Function to extract metrics ---
extract_metrics <- function(results_list, dataset_name){
  map_dfr(names(results_list), function(var){
    res <- results_list[[var]]
    tibble(variable=var, RMSE=map_dbl(res,"RMSE"), R2=map_dbl(res,"R2"))
  }) %>% mutate(dataset=dataset_name)
}

rf_metrics_all <- bind_rows(extract_metrics(RF_results_SP,"Single Point"),
                            extract_metrics(RF_results_C,"Composite"))
ok_metrics_all <- bind_rows(extract_metrics(OK_results_SP,"Single Point"),
                            extract_metrics(OK_results_C,"Composite"))
rk_metrics_all <- bind_rows(extract_metrics(RK_results_SP,"Single Point"),
                            extract_metrics(RK_results_C,"Composite"))

# --- Combine all for plotting ---
combined_metrics <- bind_rows(rf_metrics_all %>% mutate(method="RF"),
                              ok_metrics_all %>% mutate(method="OK"),
                              rk_metrics_all %>% mutate(method="RK")) %>%
  pivot_longer(cols=c(RMSE,R2), names_to="metric", values_to="value")

# --- Plot ---

ggplot(combined_metrics, aes(x=variable, y=value, fill=dataset)) +
  geom_violin(trim=FALSE, alpha=0.6) +
  geom_boxplot(width=0.1, position=position_dodge(width=0.9), outlier.shape=NA) +
  theme_minimal(base_size=13) +
  labs(title="Distribution of RMSE and R² across nutrients - All Methods",
       x="Nutrient", y="Value") +
  scale_fill_brewer(palette="Set2") +
  facet_grid(metric~method, scales="free_y")

```

```{r, warning=FALSE}
#better plot: Just R2 for OK RF and RK. Remove LBC1 and LBCeq
#rename variables for better readability
filtered_metrics <- combined_metrics %>%
  filter(metric=="R2", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(variable = recode(variable,
                           "p_h_2" = "pH",
                           "ca" = "Ca",
                           "k" = "K",
                           "mg" = "Mg",
                           "mn" = "Mn",
                           "p" = "P",
                           "zn" = "Zn",
                           "c" = "C",
                           "n" = "N"))

# Separate RMSE and R2 for plotting
filtered_metrics_RMSE <- combined_metrics %>%
  filter(metric == "RMSE", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(
    variable = recode(variable, "p_h_2"="pH","ca"="Ca","k"="K","mg"="Mg","mn"="Mn","p"="P","zn"="Zn","c"="C","n"="N")
  )

filtered_metrics_R2 <- combined_metrics %>%
  filter(metric == "R2", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(
    variable = recode(variable, "p_h_2"="pH","ca"="Ca","k"="K","mg"="Mg","mn"="Mn","p"="P","zn"="Zn","c"="C","n"="N")
  )
# Consistent colors for methods
method_colors <- c(RF = "#1b9e77", OK = "#d95f02", RK = "#7570b3")

# Factor levels to control ordering
filtered_metrics_RMSE <- filtered_metrics_RMSE %>%
  mutate(
    method  = factor(method, levels = c("RF","OK","RK")),
    dataset = factor(dataset, levels = c("Single Point","Composite"))
  )

# If you don't have `filtered_metrics_R2` yet, build it from your combined_metrics
filtered_metrics_R2 <- combined_metrics %>%
  filter(metric == "R2", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(
    variable = recode(variable, "p_h_2"="pH","ca"="Ca","k"="K","mg"="Mg","mn"="Mn","p"="P","zn"="Zn","c"="C","n"="N"),
    method   = factor(method, levels = c("RF","OK","RK")),
    dataset  = factor(dataset, levels = c("Single Point","Composite"))
  )

# RMSE plot
p_rmse <- ggplot(filtered_metrics_RMSE, aes(x = dataset, y = value)) +
  geom_violin(aes(fill = method),
              position = position_dodge(width = 0.85),
              width = 0.8, trim = FALSE, alpha = 0.6) +
  # Box outlines fully colored by method (fill = NA so outlines/whiskers show)
  geom_boxplot(aes(color = method, group = interaction(dataset, method)),
               position = position_dodge(width = 0.85),
               width = 0.14, outlier.shape = NA, fill = NA, size = 0.7) +
  # Mean marker
  stat_summary(aes(color = method),
               fun = mean, geom = "point",
               position = position_dodge(width = 0.85),
               size = 2.5, shape = 18) +
  facet_wrap(~ variable, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = method_colors, name = "Method") +
  scale_color_manual(values = method_colors, name = "Method") +
  labs(title = "RMSE by nutrient, method, and dataset",
       x = "Dataset", y = "RMSE") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
print(p_rmse)

# R2 plot
p_r2 <- ggplot(filtered_metrics_R2, aes(x = dataset, y = value)) +
  geom_violin(aes(fill = method),
              position = position_dodge(width = 0.85),
              width = 0.8, trim = FALSE, alpha = 0.6) +
  geom_boxplot(aes(color = method, group = interaction(dataset, method)),
               position = position_dodge(width = 0.85),
               width = 0.14, outlier.shape = NA, fill = NA, size = 0.7) +
  stat_summary(aes(color = method),
               fun = mean, geom = "point",
               position = position_dodge(width = 0.85),
               size = 2.5, shape = 18) +
  facet_wrap(~ variable, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = method_colors, name = "Method") +
  scale_color_manual(values = method_colors, name = "Method") +
  labs(title = "R² by nutrient, method, and dataset",
       x = "Dataset", y = "R²") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))



make_panel <- function(df_one_metric, nut_label, ylab = "RMSE/R²") {
  ggplot(df_one_metric, aes(x = dataset, y = value)) +
    geom_violin(aes(fill = method),
                position = position_dodge(width = 0.85),
                width = 0.8, trim = FALSE, alpha = 0.6) +
    geom_boxplot(aes(color = method, group = interaction(dataset, method)),
                 position = position_dodge(width = 0.85),
                 width = 0.14, outlier.shape = NA, fill = NA, size = 0.7) +
    stat_summary(aes(color = method),
                 fun = mean, geom = "point",
                 position = position_dodge(width = 0.85),
                 size = 2.5, shape = 18) +
    scale_fill_manual(values = method_colors, name = "Method") +
    scale_color_manual(values = method_colors, name = "Method") +
    labs(title = nut_label, x = "Dataset", y = ylab) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
}

# Build per-nutrient panels for RMSE
vars_rmse <- levels(factor(filtered_metrics_RMSE$variable))
panels_rmse <- lapply(vars_rmse, function(v) {
  dfv <- filtered_metrics_RMSE %>% filter(variable == v)
  make_panel(dfv, v, ylab = "RMSE")
})

# If last row has a single panel (e.g., total facets %% 3 == 1), center it by adding spacers
ncol <- 3
rem <- length(panels_rmse) %% ncol
if (rem == 1) {
  # add one spacer before and one after the last panel to center it
  panels_rmse <- append(panels_rmse, list(plot_spacer()), after = length(panels_rmse) - 1)
  panels_rmse <- append(panels_rmse, list(plot_spacer()))
} else if (rem == 2) {
  # add one spacer after to visually center two panels (optional)
  panels_rmse <- append(panels_rmse, list(plot_spacer()))
}

p_rmse_centered <- wrap_plots(panels_rmse, ncol = ncol, guides = "collect") &
  theme(legend.position = "top")

# If you want a shared legend, add it manually from one base plot; for brevity we keep legends off here.

# Build per-nutrient panels for R2
vars_r2 <- levels(factor(filtered_metrics_R2$variable))
panels_r2 <- lapply(vars_r2, function(v) {
  dfv <- filtered_metrics_R2 %>% filter(variable == v)
  make_panel(dfv, v, ylab = "R²")
})
rem2 <- length(panels_r2) %% ncol
if (rem2 == 1) {
  panels_r2 <- append(panels_r2, list(plot_spacer()), after = length(panels_r2) - 1)
  panels_r2 <- append(panels_r2, list(plot_spacer()))
} else if (rem2 == 2) {
  panels_r2 <- append(panels_r2, list(plot_spacer()))
}
p_r2_centered <- wrap_plots(panels_r2, ncol = ncol, guides = "collect") &
  theme(legend.position = "top")
p_r2_centered

```

### Metric Extraction Helpers

```{r, warning=FALSE}

safe_cor2 <- function(x,y){
  if(length(x)==0 || length(y)==0) return(NA_real_)
  if(all(is.na(x)) || all(is.na(y))) return(NA_real_)
  cc <- suppressWarnings(cor(x,y,use="complete.obs"))
  if(is.na(cc)) NA_real_ else cc^2
}

# helper to rebuild test observed vector in same order as how predictions were produced in your functions
# df_boot: an sf tibble that has column 'set' with "train"/"test" (like your generate_spatial_bootstrap output)
# var: target variable name
# returns: test_df (data.frame) after the na.omit used in your run functions, plus row_ids to map back to original rows
build_test_df_from_boot <- function(df_boot, var, covariates){
  # df_boot is sf with geometry; use st_drop_geometry in same manner as your run functions
  df <- sf::st_drop_geometry(df_boot)
  test_df_full <- df %>% filter(set == "test") %>% mutate(.orig_row = row_number())
  # select covariates and var and remove rows with NA similarly to run_RF/run_RK
  sel_cols <- c(covariates, var)
  test_sel <- test_df_full %>% select(all_of(sel_cols), .orig_row)
  # na.omit - keep track of original row indices
  complete_idx <- stats::complete.cases(test_sel)
  test_sel2 <- test_sel[complete_idx, , drop = FALSE]
  # test_sel2 has covariates, var and .orig_row
  return(list(test_df = test_sel2, omitted = !complete_idx))
}

# generic function to compute metrics from prediction and observed vectors
compute_metrics_vecs <- function(obs, pred){
  ok <- !is.na(obs) & !is.na(pred)
  n <- sum(ok)
  if(n==0) return(tibble(n = 0, RMSE = NA_real_, MAE = NA_real_, Bias = NA_real_, R2_cor = NA_real_))
  err <- pred[ok] - obs[ok]
  tibble(
    n = n,
    RMSE = sqrt(mean(err^2)),
    MAE = mean(abs(err)),
    Bias = mean(err),
    R2_cor = safe_cor2(obs[ok], pred[ok])
  )
}

# main extractor that returns a list: predictions_long (tibble), per_boot_metrics (tibble)
extract_metrics_for_method <- function(results_list, boot_list, method_name, covariates, soil_vars){
  preds_long <- list()
  per_boot_rows <- list()
  n_nut <- length(soil_vars)
  for(nut in soil_vars){
    res_boots <- results_list[[nut]]
    # iterate boots; results_list[[nut]] should be a list of length n_boot
    for(bi in seq_along(res_boots)){
      res <- res_boots[[bi]]
      boot_sf <- boot_list[[bi]]
      # build test_df in same logic as your functions
      test_build <- build_test_df_from_boot(boot_sf, nut, covariates)
      test_df <- test_build$test_df
      # try to extract pred vector from res (narrow cases)
      pred_vec <- NULL
      obs_vec <- NULL
      # If res already stored pred and maybe observed, prefer them
      if(!is.null(res$pred) && !is.null(res$observed)){
        pred_vec <- res$pred
        obs_vec  <- res$observed
      } else if(!is.null(res$pred) && is.data.frame(test_df) && nrow(test_df) == length(res$pred)){
        # assume ordering matches the test_df after na.omit (this is how your run_RF created pred)
        pred_vec <- res$pred
        obs_vec  <- test_df[[nut]]
      } else if(!is.null(res$model) && inherits(res$model, "randomForest") && is.data.frame(test_df)){
        # RF stored model object; predict fresh on test covariates (should match original pred)
        # ensure columns in test_df align
        predictor_matrix <- test_df[, covariates, drop = FALSE]
        pred_vec <- tryCatch(predict(res$model, newdata = predictor_matrix), error = function(e) NULL)
        obs_vec  <- test_df[[nut]]
      } else {
        # As a fallback: attempt to recompute prediction using simple approach depending on method
        if(method_name == "RF"){
          # try to fit a quick RF on training partition of this boot and predict
          train_df <- sf::st_drop_geometry(boot_sf) %>% filter(set=="train") %>% select(all_of(c(covariates, nut))) %>% na.omit()
          test_df2  <- test_df
          if(nrow(train_df) >= 3 && nrow(test_df2) >= 1){
            rf_mod <- tryCatch(randomForest::randomForest(x = train_df[, covariates, drop = FALSE], y = train_df[[nut]], ntree = 200), error = function(e) NULL)
            if(!is.null(rf_mod)){
              pred_vec <- tryCatch(predict(rf_mod, newdata = test_df2[, covariates, drop = FALSE]), error = function(e) NULL)
              obs_vec  <- test_df2[[nut]]
            }
          }
        } else if(method_name == "OK"){
          # run ordinary kriging on this boot
          train_sf <- boot_sf %>% filter(set=="train")
          test_sf  <- boot_sf %>% filter(set=="test")
          # keep only rows with non-NA target
          if(sum(!is.na(train_sf[[nut]])) >= 3 && nrow(test_sf) >= 1){
            # ensure train_sf/test_sf are sf objects with projected CRS (your boot list used sf transformed earlier)
            v_emp <- tryCatch(variogram(as.formula(paste(nut,"~1")), train_sf), error = function(e) NULL)
            vgm_model <- tryCatch(fit.variogram(v_emp, model = vgm(psill = var(train_sf[[nut]], na.rm=TRUE), model = "Exp", nugget = 0.1, range = max(v_emp$dist, na.rm=TRUE)/3)), error = function(e) NULL)
            if(is.null(vgm_model)) vgm_model <- tryCatch(vgm(psill = var(train_sf[[nut]], na.rm=TRUE), model="Exp", nugget=0.1, range=100), error = function(e) NULL)
            ok_pred <- tryCatch(krige(as.formula(paste(nut,"~1")), train_sf, test_sf, model = vgm_model), error = function(e) NULL)
            if(!is.null(ok_pred) && "var1.pred" %in% names(ok_pred)){
              pred_vec <- ok_pred$var1.pred
              # align obs: remove NA rows similarly to how you computed predictions
              # build test_df2 as earlier
              obs_build <- build_test_df_from_boot(boot_sf, nut, covariates)
              obs_vec <- obs_build$test_df[[nut]]
            }
          }
        } else if(method_name == "RK"){
          # recompute RK: RF trend on train + krige residuals
          train_df <- sf::st_drop_geometry(boot_sf) %>% filter(set=="train") %>% select(all_of(c(covariates, nut))) %>% na.omit()
          test_df2  <- test_df
          if(nrow(train_df) >= 5 && nrow(test_df2) >= 1){
            train_sf_filtered <- boot_sf %>% filter(set=="train")
            # align train_sf_filtered to non-NA rows in train_df
            # We assume original row ordering remains consistent enough; best to join by coordinates if needed
            # Fit RF trend
            rf_mod <- tryCatch(randomForest::randomForest(x = train_df[, covariates, drop = FALSE], y = train_df[[nut]], ntree = 200), error = function(e) NULL)
            if(!is.null(rf_mod)){
              trend_train <- predict(rf_mod, newdata = train_df[, covariates, drop = FALSE])
              resid_train <- train_df[[nut]] - trend_train
              # attach residuals to train_sf_filtered: subset to rows matching train_df by coordinates
              # This is a heuristic mapping: select rows in train_sf_filtered with values matching train_df rows
              # Safer approach: if train_df had rownames carrying original row IDs, use that. If not, we proceed best-effort.
              train_sf_filtered2 <- train_sf_filtered
              # assign residuals in order of non-NA rows
              non_na_idx <- which(!is.na(train_sf_filtered2[[nut]]))
              assign_len <- min(length(resid_train), length(non_na_idx))
              if(assign_len > 0) train_sf_filtered2$resid <- NA_real_
              train_sf_filtered2$resid[non_na_idx[1:assign_len]] <- resid_train[1:assign_len]
              # fit variogram and krige residuals
              v_emp_r <- tryCatch(variogram(resid~1, train_sf_filtered2), error = function(e) NULL)
              vgm_model <- tryCatch(fit.variogram(v_emp_r, model = vgm(psill = var(resid_train, na.rm=TRUE), model="Exp", nugget=0.1, range = max(v_emp_r$dist, na.rm=TRUE)/3)), error = function(e) NULL)
              if(is.null(vgm_model)) vgm_model <- tryCatch(vgm(psill = var(resid_train, na.rm=TRUE), model="Exp", nugget=0.1, range=100), error = function(e) NULL)
              test_sf <- boot_sf %>% filter(set=="test")
              kr_res <- tryCatch(krige(resid~1, train_sf_filtered2, test_sf, model = vgm_model), error = function(e) NULL)
              if(!is.null(kr_res) && "var1.pred" %in% names(kr_res)){
                trend_test <- predict(rf_mod, newdata = test_df2[, covariates, drop = FALSE])
                pred_vec <- trend_test + kr_res$var1.pred
                obs_vec <- test_df2[[nut]]
              }
            }
          }
        }
      } # end fallback block

      # Now compute metrics for this boot if pred_vec and obs_vec exist; align lengths
      if(!is.null(pred_vec) && !is.null(obs_vec) && length(pred_vec) == length(obs_vec)){
        metrics_row <- compute_metrics_vecs(obs_vec, pred_vec)
        per_boot_rows[[length(per_boot_rows) + 1]] <- tibble(
          nutrient = nut, method = method_name, boot = bi,
          n_test = metrics_row$n, RMSE = metrics_row$RMSE, MAE = metrics_row$MAE, Bias = metrics_row$Bias, R2_cor = metrics_row$R2_cor
        )
        # store predictions with coordinates by mapping test_df .orig_row back to original geometry from boot_sf
        # test_df came from build_test_df_from_boot and has .orig_row -> position within test partition
        # We will include coordinates from boot_sf for the test rows used
        # Rebuild a matching sf test subset (after na.omit): 
        # Use the same logic as in build_test_df_from_boot to identify which test rows used
        test_sf_full <- boot_sf %>% filter(set == "test")
        # find rows in test_sf_full that correspond to the kept ones: use complete.cases on selected columns
        test_sel_full <- sf::st_drop_geometry(test_sf_full)[, c(covariates, nut), drop = FALSE]
        keep_mask <- stats::complete.cases(test_sel_full)
        if(sum(keep_mask) > 0){
          test_sf_kept <- test_sf_full[keep_mask, ]
          # Now ensure lengths match pred_vec
          if(nrow(test_sf_kept) == length(pred_vec)){
            preds_long[[length(preds_long) + 1]] <- tibble(
              nutrient = nut, method = method_name, boot = bi,
              x = as.numeric(sf::st_coordinates(test_sf_kept)[,1]),
              y = as.numeric(sf::st_coordinates(test_sf_kept)[,2]),
              observed = obs_vec,
              predicted = pred_vec
            )
          } else {
            # fallback: store without coords
            preds_long[[length(preds_long) + 1]] <- tibble(
              nutrient = nut, method = method_name, boot = bi,
              x = NA_real_, y = NA_real_,
              observed = obs_vec,
              predicted = pred_vec
            )
          }
        } else {
          preds_long[[length(preds_long) + 1]] <- tibble(
            nutrient = nut, method = method_name, boot = bi,
            x = NA_real_, y = NA_real_,
            observed = obs_vec,
            predicted = pred_vec
          )
        }
      } else {
        # record NA metrics if we couldn't get predictions
        per_boot_rows[[length(per_boot_rows) + 1]] <- tibble(
          nutrient = nut, method = method_name, boot = bi, n_test = 0,
          RMSE = NA_real_, MAE = NA_real_, Bias = NA_real_, R2_cor = NA_real_
        )
      }

    } # end boots loop
  } # end nutrients loop

  per_boot_metrics <- bind_rows(per_boot_rows)
  predictions_long <- bind_rows(preds_long)
  list(predictions_long = predictions_long, per_boot_metrics = per_boot_metrics)
}
#SP:
# Run extraction for Single Point boots (boot_SP) and Composite boots (boot_C)
message("Extracting metrics for Single Point (SP) boots...")
rf_SP_out <- extract_metrics_for_method(RF_results_SP, boot_SP, "RF", covariates, soil_vars)
ok_SP_out <- extract_metrics_for_method(OK_results_SP, boot_SP, "OK", covariates, soil_vars)
rk_SP_out <- extract_metrics_for_method(RK_results_SP, boot_SP, "RK", covariates, soil_vars)

# Combine SP predictions and metrics
predictions_long_SP <- bind_rows(rf_SP_out$predictions_long, ok_SP_out$predictions_long, rk_SP_out$predictions_long)
per_boot_metrics_SP <- bind_rows(rf_SP_out$per_boot_metrics, ok_SP_out$per_boot_metrics, rk_SP_out$per_boot_metrics)

# compute pooled metrics per nutrient + method from predictions_long_SP
pooled_metrics_SP <- predictions_long_SP %>%
  group_by(nutrient, method) %>%
  summarize(
    n_total = sum(!is.na(observed) & !is.na(predicted)),
    pooled_RMSE = ifelse(n_total==0, NA_real_, sqrt(mean((observed - predicted)^2, na.rm=TRUE))),
    pooled_MAE  = ifelse(n_total==0, NA_real_, mean(abs(observed - predicted), na.rm=TRUE)),
    pooled_Bias = ifelse(n_total==0, NA_real_, mean(predicted - observed, na.rm=TRUE)),
    pooled_R2_cor = ifelse(n_total <= 1, NA_real_, {cc <- suppressWarnings(cor(observed, predicted, use="complete.obs")); ifelse(is.na(cc), NA_real_, cc^2)}),
    .groups = "drop"
  )

# Save results
dir.create("../ouputs/spatial_cv_results", showWarnings = FALSE)
saveRDS(predictions_long_SP, file = "spatial_cv_results/predictions_long_SP.rds")
saveRDS(per_boot_metrics_SP, file = "spatial_cv_results/per_boot_metrics_SP.rds")
saveRDS(pooled_metrics_SP, file = "spatial_cv_results/pooled_metrics_SP.rds")

# Print short summaries
message("Per-boot metrics (first rows):")
print(head(per_boot_metrics_SP))
message("Pooled metrics (first rows):")
print(head(pooled_metrics_SP))

# Return invisibly
invisible(list(
  predictions_long_SP = predictions_long_SP,
  per_boot_metrics_SP = per_boot_metrics_SP,
  pooled_metrics_SP = pooled_metrics_SP
))

#C:
message("Extracting metrics for Composite (C) boots...")
rf_C_out <- extract_metrics_for_method(RF_results_C, boot_C, "RF", covariates, soil_vars)
ok_C_out <- extract_metrics_for_method(OK_results_C, boot_C, "OK", covariates, soil_vars)
rk_C_out <- extract_metrics_for_method(RK_results_C, boot_C, "RK", covariates, soil_vars)
# Combine C predictions and metrics
predictions_long_C <- bind_rows(rf_C_out$predictions_long, ok_C_out$predictions_long, rk_C_out$predictions_long)   
per_boot_metrics_C <- bind_rows(rf_C_out$per_boot_metrics, ok_C_out$per_boot_metrics, rk_C_out$per_boot_metrics)
# compute pooled metrics per nutrient + method from predictions_long_C
pooled_metrics_C <- predictions_long_C %>%
  group_by(nutrient, method) %>%
  summarize(
    n_total = sum(!is.na(observed) & !is.na(predicted)),
    pooled_RMSE = ifelse(n_total==0, NA_real_, sqrt(mean((observed - predicted)^2, na.rm=TRUE))),
    pooled_MAE  = ifelse(n_total==0, NA_real_, mean(abs(observed - predicted), na.rm=TRUE)),
    pooled_Bias = ifelse(n_total==0, NA_real_, mean(predicted - observed, na.rm=TRUE)),
    pooled_R2_cor = ifelse(n_total <= 1, NA_real_, {cc <- suppressWarnings(cor(observed, predicted, use="complete.obs")); ifelse(is.na(cc), NA_real_, cc^2)}),
    .groups = "drop"
  )
# Save results                                    

saveRDS(predictions_long_C, file = "spatial_cv_results/predictions_long_C.rds")
saveRDS(per_boot_metrics_C, file = "spatial_cv_results/per_boot_metrics_C.rds")
saveRDS(pooled_metrics_C, file = "spatial_cv_results/pooled_metrics_C.rds")
# Print short summaries
message("Per-boot metrics (first rows):")
print(head(per_boot_metrics_C))                
message("Pooled metrics (first rows):")
print(head(pooled_metrics_C))
# Return invisibly          
invisible(list(
  predictions_long_C = predictions_long_C,
  per_boot_metrics_C = per_boot_metrics_C,
  pooled_metrics_C = pooled_metrics_C
))           
                                       
```

### Statistical Testing Between Methods

```{r, warning=FALSE}
prepare_paired_data <- function(results_list, dataset_name, method_name){
  map_dfr(names(results_list), function(var){
    res <- results_list[[var]]
    tibble(variable=var, RMSE=map_dbl(res,"RMSE"), R2=map_dbl(res,"R2")) %>%
      mutate(dataset=dataset_name, method=method_name)
  })
}
rf_paired <- bind_rows(prepare_paired_data(RF_results_SP,"Single Point","RF"),
                       prepare_paired_data(RF_results_C,"Composite","RF"))
ok_paired <- bind_rows(prepare_paired_data(OK_results_SP,"Single Point","OK"),
                       prepare_paired_data(OK_results_C,"Composite","OK"))
rk_paired <- bind_rows(prepare_paired_data(RK_results_SP,"Single Point","RK"),
                       prepare_paired_data(RK_results_C,"Composite","RK"))

all_paired <- bind_rows(rf_paired, ok_paired, rk_paired)

# Paired tests between methods for each variable and dataset
test_results <- map_dfr(unique(all_paired$variable), function(var){
  map_dfr(unique(all_paired$dataset), function(ds){
    subset_data <- all_paired %>% filter(variable==var, dataset==ds)
    rf_data <- subset_data %>% filter(method=="RF")
    ok_data <- subset_data %>% filter(method=="OK")
    rk_data <- subset_data %>% filter(method=="RK")
    # RF vs OK
    rf_ok_test <- wilcox.test(rf_data$R2, ok_data$R2, paired=TRUE, alternative="two.sided")
    # RF vs RK
    rf_rk_test <- wilcox.test(rf_data$R2, rk_data$R2, paired=TRUE, alternative="two.sided")
    # OK vs RK
    ok_rk_test <- wilcox.test(ok_data$R2, rk_data$R2, paired=TRUE, alternative="two.sided")
    tibble(variable=var, dataset=ds,
           RF_vs_OK_p=rf_ok_test$p.value,
           RF_vs_RK_p=rf_rk_test$p.value,
           OK_vs_RK_p=ok_rk_test$p.value)
  })
})
#create formatted results with significance <0.05 marked with *
formatted_results <- test_results %>%
  #sort by lowest p-value in RF_vs_OK
  arrange(RF_vs_OK_p) %>%
  mutate(RF_vs_OK = ifelse(RF_vs_OK_p < 0.05, paste0(round(RF_vs_OK_p,4), "*"), round(RF_vs_OK_p,4)),
         RF_vs_RK = ifelse(RF_vs_RK_p < 0.05, paste0(round(RF_vs_RK_p,4), "*"), round(RF_vs_RK_p,4)),
         OK_vs_RK = ifelse(OK_vs_RK_p < 0.05, paste0(round(OK_vs_RK_p,4), "*"), round(OK_vs_RK_p,4))) %>%
  select(variable, dataset, RF_vs_OK, RF_vs_RK, OK_vs_RK)
formatted_results
```
### Covariate Name Matching

```{r}


# ------------------------------------------------------------
# 1) Define soil vars we do NOT want as predictors
# ------------------------------------------------------------
soil_vars_all <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")

# ------------------------------------------------------------
# 2) Build normalized-name matching between points and raster
# ------------------------------------------------------------
norm_name <- function(x){
  x %>%
    tolower() %>%
    str_replace_all("[^a-z0-9]+", "_") %>%  # collapse punctuation/spaces to _
    str_replace_all("_+", "_") %>%          # remove repeats
    str_replace_all("^_|_$", "")            # trim ends
}

stack_names <- names(covariate_stack)
sp_names <- names(st_drop_geometry(soil_sf_SP))

# candidates on points side = everything except soil vars + geometry-ish columns
sp_candidates <- setdiff(sp_names, c(soil_vars_all, "geometry", "set", "fold"))

# make lookup tables
stack_tbl <- tibble(stack = stack_names, stack_norm = norm_name(stack_names))
sp_tbl    <- tibble(sp = sp_candidates, sp_norm = norm_name(sp_candidates))

# join by normalized name
map_tbl <- sp_tbl %>%
  left_join(stack_tbl, by = c("sp_norm" = "stack_norm"))

# which point cols did not match any raster layer?
unmatched_sp <- map_tbl %>% filter(is.na(stack)) %>% pull(sp)

# which raster layers were not matched by any point col?
matched_stack <- na.omit(map_tbl$stack) %>% unique()
unmatched_stack <- setdiff(stack_names, matched_stack)

cat("Point covariate cols:", length(sp_candidates), "\n")
cat("Raster layers:", length(stack_names), "\n")
cat("Matched:", length(matched_stack), "\n")
cat("UNmatched point cols:", length(unmatched_sp), "\n")
cat("UNmatched raster layers:", length(unmatched_stack), "\n")

if (length(unmatched_sp) > 0) {
  cat("\n--- Unmatched point columns (first 30) ---\n")
  print(head(unmatched_sp, 30))
}
if (length(unmatched_stack) > 0) {
  cat("\n--- Unmatched raster layers (first 30) ---\n")
  print(head(unmatched_stack, 30))
}

# ------------------------------------------------------------
# 3) Actually rename point columns to match raster stack names
#    Only for those with a match, and avoid collisions.
# ------------------------------------------------------------
rename_vec <- map_tbl %>%
  filter(!is.na(stack)) %>%
  distinct(sp, stack) %>%
  # if multiple sp map to same stack (rare), keep first
  group_by(stack) %>% slice(1) %>% ungroup() %>%
  { setNames(.$stack, .$sp) }   # old -> new mapping (dplyr::rename expects new=old, so we’ll do base rename)

# base R rename for data.frames: names(df)[match(old, names(df))] <- new
apply_rename <- function(sf_obj, rename_vec){
  nm <- names(sf_obj)
  olds <- names(rename_vec)
  idx <- match(olds, nm)
  hit <- !is.na(idx)
  nm[idx[hit]] <- rename_vec[hit]
  names(sf_obj) <- nm
  sf_obj
}

soil_sf_SP <- apply_rename(soil_sf_SP, rename_vec)
soil_sf_C  <- apply_rename(soil_sf_C,  rename_vec)

# ------------------------------------------------------------
# 4) Now you can use ALL raster layers as covariates
#    (except soil variables if they exist as layers)
# ------------------------------------------------------------
covariates <- setdiff(names(covariate_stack), soil_vars_all)

# Final sanity
missing_now_SP <- setdiff(covariates, names(st_drop_geometry(soil_sf_SP)))
missing_now_C  <- setdiff(covariates, names(st_drop_geometry(soil_sf_C)))

cat("\nAfter renaming:\n")
cat("Missing in SP:", length(missing_now_SP), "\n")
cat("Missing in C :", length(missing_now_C), "\n")

if (length(missing_now_SP) > 0) {
  cat("\n--- Still missing in SP (first 30) ---\n")
  print(head(missing_now_SP, 30))
}

# Optional: reorder stack to covariates
covariate_stack <- covariate_stack[[covariates]]

```
### Final Map Prediction Using Best Boot Models

```{r}
# ============================================================
# Best-boot selection + FINAL mapping (RF / OK / RK) in ONE chunk
# ============================================================

# libs assumed loaded: terra, sf, dplyr, gstat, randomForest

# -----------------------------
# 0) Soil vars to map
# -----------------------------
soil_vars_to_map <- c("ca","k","mg","mn","p","zn","c","n","p_h_2")

# -----------------------------
# 1) Best boot table (high R2, low RMSE) — for reporting only
# -----------------------------
per_boot_metrics_all <- dplyr::bind_rows(
  per_boot_metrics_SP %>% dplyr::mutate(dataset = "SP"),
  per_boot_metrics_C  %>% dplyr::mutate(dataset = "C")
) %>%
  dplyr::rename(R2 = R2_cor) %>%
  dplyr::filter(is.finite(RMSE), is.finite(R2))

best_boot_tbl <- per_boot_metrics_all %>%
  dplyr::filter(nutrient %in% soil_vars_to_map, method %in% c("RF","RK","OK")) %>%
  dplyr::group_by(dataset, method, nutrient) %>%
  dplyr::mutate(
    zR2   = as.numeric(scale(R2)),
    zRMSE = as.numeric(scale(RMSE)),
    score = zR2 - zRMSE
  ) %>%
  dplyr::slice_max(order_by = score, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

print(best_boot_tbl)

# -----------------------------
# 2) Covariates = raster stack names minus soil vars
# -----------------------------
soil_vars_all <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")
covariates <- setdiff(names(covariate_stack), soil_vars_all)

stopifnot(all(covariates %in% names(sf::st_drop_geometry(soil_sf_SP))))
stopifnot(all(covariates %in% names(sf::st_drop_geometry(soil_sf_C))))

template_r <- covariate_stack[[1]]

# -----------------------------
# 3) Helpers
# -----------------------------
align_sf_to_raster <- function(soil_sf, r){
  crs_r <- terra::crs(r, proj = TRUE)
  if (is.na(crs_r) || crs_r == "") stop("Raster has no CRS.")
  sf::st_transform(soil_sf, crs_r)
}

# RF final map (safe predict)
predict_RF_final <- function(soil_sf, covariate_stack, covariates, nutrient, ntree = 500) {

  df <- sf::st_drop_geometry(soil_sf) %>%
    dplyr::select(dplyr::all_of(c(covariates, nutrient))) %>%
    stats::na.omit()

  if (nrow(df) < 10) stop("Too few training rows for RF: ", nutrient)

  rf_mod <- randomForest::randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nutrient]],
    ntree = ntree
  )

  pred_df <- terra::as.data.frame(covariate_stack[[covariates]], cells = TRUE, na.rm = FALSE)
  X <- pred_df[, covariates, drop = FALSE]
  keep <- stats::complete.cases(X)

  pred <- rep(NA_real_, nrow(pred_df))
  pred[keep] <- as.numeric(predict(rf_mod, newdata = as.data.frame(X[keep, , drop = FALSE])))

  out <- covariate_stack[[1]]
  vals <- rep(NA_real_, terra::ncell(out))
  vals[pred_df$cell] <- pred
  terra::values(out) <- vals
  out
}

# OK final map (CRS-safe + cell-index assignment)
predict_OK_final <- function(soil_sf, template_raster, nutrient, debug = TRUE) {

  # align to raster CRS
  train_sf <- sf::st_transform(soil_sf, terra::crs(template_raster, proj = TRUE))

  # drop NA response rows (CRITICAL for gstat)
  train_sf <- train_sf %>% dplyr::filter(!is.na(.data[[nutrient]]))

  if (debug) {
    cat("\n[OK DEBUG] nutrient:", nutrient, "\n")
    cat("  n rows (non-NA):", nrow(train_sf), "\n")
    cat("  unique values:", length(unique(train_sf[[nutrient]])), "\n")
    cat("  CRS:", sf::st_crs(train_sf)$wkt %>% substr(1, 60), "...\n")
  }

  vals <- train_sf[[nutrient]]
  if (length(vals) < 10 || length(unique(vals)) < 2)
    stop("Too few usable (non-NA) points for OK: ", nutrient)

  # empirical variogram
  v_emp <- tryCatch(
    gstat::variogram(stats::as.formula(paste(nutrient, "~ 1")), train_sf),
    error = function(e) stop("Variogram failed for OK ", nutrient, ": ", e$message)
  )

  if (nrow(v_emp) < 3) stop("Variogram has too few bins for OK: ", nutrient)

  init_range <- max(v_emp$dist, na.rm = TRUE) / 3
  init_psill <- stats::var(vals, na.rm = TRUE)

  v_fit <- tryCatch(
    gstat::fit.variogram(
      v_emp,
      gstat::vgm(psill = init_psill, model = "Exp",
                 nugget = 0.1 * init_psill, range = init_range)
    ),
    error = function(e) gstat::vgm(psill = init_psill, model = "Nug", nugget = 0.1 * init_psill)
  )

  # grid WITH cell index (prevents ordering artifacts)
  grid_df <- terra::as.data.frame(template_raster, cells = TRUE, xy = TRUE, na.rm = FALSE) %>%
    dplyr::select(cell, x, y)

  grd_sf <- sf::st_as_sf(
    grid_df, coords = c("x","y"),
    crs = terra::crs(template_raster, proj = TRUE),
    remove = FALSE
  )

  ok <- tryCatch(
    gstat::krige(stats::as.formula(paste(nutrient, "~ 1")), train_sf, grd_sf, model = v_fit),
    error = function(e) stop("Kriging failed for OK ", nutrient, ": ", e$message)
  )

  out <- template_raster
  vals_out <- rep(NA_real_, terra::ncell(out))
  vals_out[grid_df$cell] <- ok$var1.pred
  terra::values(out) <- vals_out

  out
}


# RK final map (RF trend + kriged residuals; CRS-safe + cell-index assignment)
predict_RK_final <- function(soil_sf, covariate_stack, covariates, nutrient, ntree = 500) {

  soil_sf_r <- align_sf_to_raster(soil_sf, covariate_stack[[1]])

  df0 <- sf::st_drop_geometry(soil_sf_r) %>%
    dplyr::select(dplyr::all_of(c(covariates, nutrient))) %>%
    dplyr::mutate(.row_id = dplyr::row_number())

  keep_train <- stats::complete.cases(df0[, c(covariates, nutrient)])
  df <- df0[keep_train, , drop = FALSE]

  if (nrow(df) < 10) stop("Too few training rows for RK: ", nutrient)

  rf_mod <- randomForest::randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nutrient]],
    ntree = ntree
  )

  trend_pts <- as.numeric(predict(rf_mod, newdata = df[, covariates, drop = FALSE]))
  resid <- df[[nutrient]] - trend_pts

  soil_sf2 <- soil_sf_r[df$.row_id, ]
  train_res_sf <- soil_sf2 %>% dplyr::mutate(resid = resid)

  v_emp <- tryCatch(gstat::variogram(resid ~ 1, train_res_sf), error = function(e) NULL)
  if (is.null(v_emp) || nrow(v_emp) < 3) {
    message("RK residual variogram failed for ", nutrient, " -> returning RF trend only")
    return(predict_RF_final(soil_sf_r, covariate_stack, covariates, nutrient, ntree = ntree))
  }

  init_range <- max(v_emp$dist, na.rm = TRUE) / 3
  init_psill <- stats::var(resid, na.rm = TRUE)

  v_fit <- tryCatch(
    gstat::fit.variogram(
      v_emp,
      gstat::vgm(psill = init_psill, model = "Exp",
                 nugget = 0.1 * init_psill, range = init_range)
    ),
    error = function(e) gstat::vgm(psill = init_psill, model = "Nug", nugget = 0.1 * init_psill)
  )

  # RF trend raster (safe predict)
  pred_df <- terra::as.data.frame(covariate_stack[[covariates]], cells = TRUE, na.rm = FALSE)
  X <- pred_df[, covariates, drop = FALSE]
  keep <- stats::complete.cases(X)

  trend_pred <- rep(NA_real_, nrow(pred_df))
  trend_pred[keep] <- as.numeric(predict(rf_mod, newdata = as.data.frame(X[keep, , drop = FALSE])))

  trend_r <- covariate_stack[[1]]
  vals_tr <- rep(NA_real_, terra::ncell(trend_r))
  vals_tr[pred_df$cell] <- trend_pred
  terra::values(trend_r) <- vals_tr

  # residual kriging to raster grid (cell-index assignment)
  grid_df <- terra::as.data.frame(trend_r, cells = TRUE, xy = TRUE, na.rm = FALSE) %>%
    dplyr::select(cell, x, y)

  grd_sf <- sf::st_as_sf(grid_df, coords = c("x","y"),
                         crs = terra::crs(trend_r, proj = TRUE),
                         remove = FALSE)

  kr_res <- gstat::krige(resid ~ 1, train_res_sf, grd_sf, model = v_fit)

  res_r <- trend_r
  vals_res <- rep(NA_real_, terra::ncell(res_r))
  vals_res[grid_df$cell] <- kr_res$var1.pred
  terra::values(res_r) <- vals_res

  trend_r + res_r
}

# -----------------------------
# 4) Write FINAL maps (train on all points) for ALL methods
# -----------------------------
out_dir <- "best_boot_maps"
dir.create(out_dir, showWarnings = FALSE)

for (ds in c("SP","C")) {

  soil_sf_use <- if (ds == "SP") soil_sf_SP else soil_sf_C

  for (nut in soil_vars_to_map) {

    for (meth in c("RF","OK","RK")) {

      cat("Mapping:", ds, nut, meth, "\n")

      r <- tryCatch({
        if (meth == "RF")      predict_RF_final(soil_sf_use, covariate_stack, covariates, nut)
        else if (meth == "OK") predict_OK_final(soil_sf_use, template_r, nut)
        else                   predict_RK_final(soil_sf_use, covariate_stack, covariates, nut)
      }, error = function(e) {
        message("FAILED: ", ds, " ", nut, " ", meth, " -> ", e$message)
        NULL
      })

      if (!is.null(r)) {
        fn <- file.path(out_dir, sprintf("%s_%s_%s_bestboot.tif", ds, nut, meth))
        terra::writeRaster(r, fn, overwrite = TRUE)
      }
    }
  }
}

cat("Done. Outputs in:", out_dir, "\n")

```

```{r}


# helper: make soil_sf match raster CRS
align_sf_to_raster <- function(soil_sf, r){
  crs_r <- terra::crs(r, proj = TRUE)
  if (is.na(crs_r) || crs_r == "") stop("Raster has no CRS.")
  sf::st_transform(soil_sf, crs_r)
}

# OK final map (CRS-safe + variogram fallback)
predict_OK_final <- function(soil_sf, template_raster, nutrient) {

  # align points to raster CRS
  train_sf <- sf::st_transform(soil_sf, terra::crs(template_raster, proj = TRUE))

  vals <- train_sf[[nutrient]]
  if (sum(!is.na(vals)) < 10 || length(unique(na.omit(vals))) < 2)
    stop("Too few points for OK: ", nutrient)

  # empirical variogram + fit
  v_emp <- gstat::variogram(stats::as.formula(paste(nutrient, "~ 1")), train_sf)

  init_range <- max(v_emp$dist, na.rm = TRUE) / 3
  init_psill <- stats::var(vals, na.rm = TRUE)

  v_fit <- tryCatch(
    gstat::fit.variogram(
      v_emp,
      gstat::vgm(psill = init_psill, model = "Exp",
                 nugget = 0.1 * init_psill, range = init_range)
    ),
    error = function(e) gstat::vgm(psill = init_psill, model = "Nug", nugget = 0.1 * init_psill)
  )

  # build raster grid WITH cell index (this prevents re-ordering artifacts)
  grid_df <- terra::as.data.frame(template_raster, cells = TRUE, xy = TRUE, na.rm = FALSE) |>
    dplyr::select(cell, x, y)

  grd_sf <- sf::st_as_sf(grid_df, coords = c("x","y"),
                         crs = terra::crs(template_raster, proj = TRUE), remove = FALSE)

  ok <- gstat::krige(stats::as.formula(paste(nutrient, "~ 1")),
                     train_sf, grd_sf, model = v_fit)

  # assign by cell index (robust)
  out <- template_raster
  vals_out <- rep(NA_real_, terra::ncell(out))
  vals_out[grid_df$cell] <- ok$var1.pred
  terra::values(out) <- vals_out

  out
}

# RK final map (CRS-safe + robust row alignment)
predict_RK_final <- function(soil_sf, covariate_stack, covariates, nutrient, ntree = 500) {

  # 1) align soil_sf to raster CRS (critical)
  soil_sf_r <- align_sf_to_raster(soil_sf, covariate_stack[[1]])

  # 2) training df (keep row ids for alignment)
  df0 <- sf::st_drop_geometry(soil_sf_r) %>%
    dplyr::select(dplyr::all_of(c(covariates, nutrient))) %>%
    dplyr::mutate(.row_id = dplyr::row_number())

  keep_train <- stats::complete.cases(df0[, c(covariates, nutrient)])
  df <- df0[keep_train, , drop = FALSE]

  if (nrow(df) < 10) stop("Too few training rows for RK: ", nutrient)

  # 3) RF trend on points
  rf_mod <- randomForest::randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nutrient]],
    ntree = ntree
  )

  trend_pts <- as.numeric(predict(rf_mod, newdata = df[, covariates, drop = FALSE]))
  resid <- df[[nutrient]] - trend_pts

  # 4) attach residuals back to matching sf rows (via .row_id)
  soil_sf2 <- soil_sf_r[df$.row_id, ]
  train_res_sf <- soil_sf2 %>% dplyr::mutate(resid = resid)

  # 5) variogram on residuals + fit
  v_emp <- tryCatch(gstat::variogram(resid ~ 1, train_res_sf), error = function(e) NULL)
  if (is.null(v_emp) || nrow(v_emp) < 3) {
    # if variogram fails, just return RF trend map
    message("RK residual variogram failed for ", nutrient, " -> returning RF trend only")
    return(predict_RF_final(soil_sf_r, covariate_stack, covariates, nutrient, ntree = ntree))
  }

  init_range <- max(v_emp$dist, na.rm = TRUE) / 3
  init_psill <- stats::var(resid, na.rm = TRUE)

  v_fit <- tryCatch(
    gstat::fit.variogram(
      v_emp,
      gstat::vgm(psill = init_psill, model = "Exp", nugget = 0.1 * init_psill, range = init_range)
    ),
    error = function(e) gstat::vgm(psill = init_psill, model = "Nug", nugget = 0.1 * init_psill)
  )

  # 6) RF trend raster (your safe predict logic)
  pred_df <- terra::as.data.frame(covariate_stack[[covariates]], cells = TRUE, na.rm = FALSE)
  X <- pred_df[, covariates, drop = FALSE]
  keep <- stats::complete.cases(X)

  trend_pred <- rep(NA_real_, nrow(pred_df))
  trend_pred[keep] <- as.numeric(predict(rf_mod, newdata = as.data.frame(X[keep, , drop = FALSE])))

  trend_r <- covariate_stack[[1]]
  vals <- rep(NA_real_, terra::ncell(trend_r))
  vals[pred_df$cell] <- trend_pred
  terra::values(trend_r) <- vals

  # 7) krige residuals to raster grid (grid CRS = raster CRS)
  grid_df <- terra::as.data.frame(trend_r, cells = TRUE, xy = TRUE, na.rm = FALSE) |>
  dplyr::select(cell, x, y)

    grd_sf <- sf::st_as_sf(grid_df, coords = c("x","y"),
                           crs = terra::crs(trend_r, proj = TRUE), remove = FALSE)
    
    kr_res <- gstat::krige(resid ~ 1, train_res_sf, grd_sf, model = v_fit)
    
    res_r <- trend_r
    vals_res <- rep(NA_real_, terra::ncell(res_r))
    vals_res[grid_df$cell] <- kr_res$var1.pred
    terra::values(res_r) <- vals_res
    
    trend_r + res_r

}
template_r <- covariate_stack[[1]]
out_dir <- "best_boot_maps"
dir.create(out_dir, showWarnings = FALSE)

for (ds in c("SP","C")) {
  soil_sf_use <- if (ds == "SP") soil_sf_SP else soil_sf_C

  for (nut in soil_vars_to_map) {

    # OK
    cat("Mapping:", ds, nut, "OK\n")
    r_ok <- tryCatch(predict_OK_final(soil_sf_use, template_r, nut), error = function(e){message(e$message); NULL})
    if (!is.null(r_ok)) terra::writeRaster(r_ok, file.path(out_dir, sprintf("%s_%s_OK_bestboot.tif", ds, nut)), overwrite=TRUE)

    # RK
    cat("Mapping:", ds, nut, "RK\n")
    r_rk <- tryCatch(predict_RK_final(soil_sf_use, covariate_stack, covariates, nut), error = function(e){message(e$message); NULL})
    if (!is.null(r_rk)) terra::writeRaster(r_rk, file.path(out_dir, sprintf("%s_%s_RK_bestboot.tif", ds, nut)), overwrite=TRUE)
  }
}

```

### Example Leaflet Plot of a Best Boot Map

```{r}
#plot

ca_ok <- rast("best_boot_maps/C_ca_OK_bestboot.tif")

# project to lat/long for leaflet
ca_ok_proj <- project(ca_ok, "EPSG:4326")

pal <- colorNumeric("viridis", values(ca_ok_proj), na.color = "transparent")

leaflet() %>%
  addTiles() %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite")  %>%
  addRasterImage(ca_ok_proj, colors = pal, opacity = 0.8) %>%
  addLegend(pal = pal, values = values(ca_ok_proj), title = "Ca (ppm)", opacity = 1)
  #basemap: satellite bottom layer

```
### Map Panels of Best Boot Maps


```{r}

in_dir  <- "best_boot_maps"
out_dir <- "map_panels"
dir.create(out_dir, showWarnings = FALSE)

methods   <- c("RF","OK","RK")
datasets  <- c("SP","C")
nutrients <- c("ca","k","mg","mn","p","zn","c","n","p_h_2")   # 9 nutrients

nut_labels <- c(ca="Ca", k="K", mg="Mg", mn="Mn", p="P", zn="Zn", c="Carbon", n="Nitrogen", p_h_2="pH")
`%||%` <- function(a, b) if (!is.null(a)) a else b

# locate file robustly (exact filename expected)
get_path <- function(ds, nut, meth){
  f <- file.path(in_dir, sprintf("%s_%s_%s_bestboot.tif", ds, nut, meth))
  if (file.exists(f)) return(f)
  return(NA_character_)
}

make_map_one <- function(r, title){
  df <- as.data.frame(r, xy = TRUE, na.rm = FALSE)
  names(df) <- c("x","y","pred")

  ggplot(df, aes(x, y, fill = pred)) +
    geom_raster() +
    coord_equal() +
    labs(title = title, x = NULL, y = NULL, fill = NULL) +
    scale_fill_viridis_c(na.value = "transparent") +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face="bold", hjust=0.5, size=11),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.position = "bottom",
      legend.key.width  = unit(2.0, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.text  = element_text(size = 8),
      plot.margin = ggplot2::margin(2, 2, 2, 2, unit = "pt")
    )
}

# Build one patchwork for a single (method, dataset)
make_panel <- function(meth, ds, ncol = 3){

  paths <- tibble(nutrient = nutrients) %>%
    mutate(path = map_chr(nutrient, ~get_path(ds, .x, meth))) %>%
    mutate(exists = !is.na(path))

  cat("\n---", meth, ds, "---\n")
  print(paths)

  plots <- list()

  for (i in seq_len(nrow(paths))) {
    nut <- paths$nutrient[i]
    f   <- paths$path[i]
    if (is.na(f)) next

    r <- tryCatch(rast(f), error = function(e) NULL)
    if (is.null(r)) next

    lab <- nut_labels[[nut]] %||% nut
    plots[[length(plots) + 1]] <- make_map_one(r, lab)
  }

  if (length(plots) == 0) return(NULL)

  wrap_plots(plots, ncol = ncol) +
    plot_annotation(
      title = paste0(meth, " — ", ds, " (Best-boot maps)")
    ) &
    theme(plot.title = element_text(face="bold", hjust=0.5, size=16))
}

# Make 6 patchwork plots
pw_RF_SP <- make_panel("RF", "SP")
pw_RF_C  <- make_panel("RF", "C")
pw_OK_SP <- make_panel("OK", "SP")
pw_OK_C  <- make_panel("OK", "C")
pw_RK_SP <- make_panel("RK", "SP")  # likely NULL until RK rasters exist
pw_RK_C  <- make_panel("RK", "C")

pw_RF_SP; pw_RF_C
pw_OK_SP; pw_OK_C
pw_RK_SP; pw_RK_C

# Save big enough so plots aren't tiny
save_pw <- function(pw, name){
  if (is.null(pw)) return(invisible(NULL))
  ggsave(file.path(out_dir, name),
         pw,
         width = 12,
         height = 8.5,   # good for 3x3 grid
         dpi = 300)
}

# save_pw(pw_RF_SP, "RF_SP.png")
# save_pw(pw_RF_C,  "RF_C.png")
# save_pw(pw_OK_SP, "OK_SP.png")
# save_pw(pw_OK_C,  "OK_C.png")
# save_pw(pw_RK_SP, "RK_SP.png")
# save_pw(pw_RK_C,  "RK_C.png")



```

### Variogram Diagnostics and VIP Analysis

```{r}

# -----------------------------
# Settings
# -----------------------------
soil_vars_to_use <- c("ca","k","mg","mn","p","zn","c","n","p_h_2")  # edit if needed
top_k_vip <- 6

template_r <- covariate_stack[[1]]
out_dir_vario <- "variogram_diagnostics"
out_dir_vip   <- "vip_outputs"
out_dir_vipmaps <- file.path(out_dir_vip, "vip_maps")
dir.create(out_dir_vario, showWarnings = FALSE, recursive = TRUE)
dir.create(out_dir_vip, showWarnings = FALSE, recursive = TRUE)
dir.create(out_dir_vipmaps, showWarnings = FALSE, recursive = TRUE)

# Covariates (from raster stack names, robust)
soil_vars_all <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")
covariates <- setdiff(names(covariate_stack), soil_vars_all)

stopifnot(all(covariates %in% names(st_drop_geometry(soil_sf_SP))))
stopifnot(all(covariates %in% names(st_drop_geometry(soil_sf_C))))

# -----------------------------
# Helpers
# -----------------------------
align_sf_to_raster <- function(soil_sf, r){
  crs_r <- terra::crs(r, proj = TRUE)
  if (is.na(crs_r) || crs_r == "") stop("Raster has no CRS.")
  sf::st_transform(soil_sf, crs_r)
}

fit_variogram_safe <- function(v_emp, vals){
  # reasonable starts + fallback
  init_range <- max(v_emp$dist, na.rm = TRUE)/3
  init_psill <- stats::var(vals, na.rm = TRUE)
  v_fit <- tryCatch(
    gstat::fit.variogram(
      v_emp,
      gstat::vgm(psill = init_psill, model = "Exp",
                 nugget = 0.1 * init_psill, range = init_range)
    ),
    error = function(e) gstat::vgm(psill = init_psill, model = "Nug", nugget = 0.1 * init_psill)
  )
  v_fit
}

save_variogram_plot <- function(v_emp, v_fit, title, out_png){
  # model line
  vline <- tryCatch({
    vl <- gstat::variogramLine(v_fit, maxdist = max(v_emp$dist, na.rm = TRUE), n = 200)
    as_tibble(vl)
  }, error = function(e) NULL)

  p <- ggplot(as_tibble(v_emp), aes(x = dist, y = gamma)) +
    geom_point(size = 2, alpha = 0.85) +
    labs(title = title, x = "Distance", y = "Semivariance") +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face="bold", hjust = 0.5))

  if (!is.null(vline)) {
    p <- p + geom_line(data = vline, aes(x = dist, y = gamma), linewidth = 1)
  }

  ggsave(out_png, p, width = 7.2, height = 5.0, dpi = 300)
  invisible(p)
}

get_vgm_params <- function(v_fit){
  # v_fit is a data.frame-like variogram model
  # handle cases where rownames are model types
  tibble(
    model  = paste(v_fit$model, collapse = "+"),
    nugget = sum(v_fit$psill[v_fit$model == "Nug"], na.rm = TRUE),
    sill   = sum(v_fit$psill, na.rm = TRUE),
    range  = suppressWarnings(max(v_fit$range, na.rm = TRUE))
  )
}

# -----------------------------
# 1) Variogram diagnostics: OK (nutrient ~ 1)
# -----------------------------
ok_param_rows <- list()

for (ds in c("SP","C")) {

  soil_sf_use <- if (ds == "SP") soil_sf_SP else soil_sf_C
  soil_sf_use <- align_sf_to_raster(soil_sf_use, template_r)

  for (nut in soil_vars_to_use) {

    # drop NA response rows (critical for gstat)
    train_sf <- soil_sf_use %>% filter(!is.na(.data[[nut]]))
    vals <- train_sf[[nut]]

    # skip if not enough data
    if (length(vals) < 10 || length(unique(vals)) < 2) {
      message("[OK VARIO] skipping ", ds, " ", nut, " (too few non-NA/unique)")
      next
    }

    v_emp <- tryCatch(
      gstat::variogram(stats::as.formula(paste(nut, "~ 1")), train_sf),
      error = function(e) NULL
    )
    if (is.null(v_emp) || nrow(v_emp) < 3) {
      message("[OK VARIO] skipping ", ds, " ", nut, " (variogram failed)")
      next
    }

    v_fit <- fit_variogram_safe(v_emp, vals)

    out_png <- file.path(out_dir_vario, sprintf("OK_%s_%s_variogram.png", ds, nut))
    save_variogram_plot(v_emp, v_fit,
                        title = paste0("OK variogram — ", ds, " — ", nut),
                        out_png = out_png)

    params <- get_vgm_params(v_fit) %>%
      mutate(dataset = ds, method = "OK", nutrient = nut)

    ok_param_rows[[length(ok_param_rows) + 1]] <- params

    # also save empirical variogram table
    write.csv(v_emp, file.path(out_dir_vario, sprintf("OK_%s_%s_empirical.csv", ds, nut)), row.names = FALSE)
  }
}

ok_params_tbl <- bind_rows(ok_param_rows)
write.csv(ok_params_tbl, file.path(out_dir_vario, "OK_variogram_params.csv"), row.names = FALSE)

# -----------------------------
# 2) Variogram diagnostics: RK residuals (resid ~ 1)
# -----------------------------
rk_param_rows <- list()

for (ds in c("SP","C")) {

  soil_sf_use <- if (ds == "SP") soil_sf_SP else soil_sf_C
  soil_sf_use <- align_sf_to_raster(soil_sf_use, template_r)

  for (nut in soil_vars_to_use) {

    # build training df for RF trend
    df0 <- st_drop_geometry(soil_sf_use) %>%
      select(all_of(c(covariates, nut))) %>%
      mutate(.row_id = row_number())

    keep <- complete.cases(df0[, c(covariates, nut)])
    df <- df0[keep, , drop = FALSE]

    if (nrow(df) < 15) {
      message("[RK VARIO] skipping ", ds, " ", nut, " (too few complete rows)")
      next
    }

    # fit RF trend (importance not needed here)
    rf_mod <- tryCatch(
      randomForest::randomForest(
        x = df[, covariates, drop = FALSE],
        y = df[[nut]],
        ntree = 500
      ),
      error = function(e) NULL
    )
    if (is.null(rf_mod)) {
      message("[RK VARIO] skipping ", ds, " ", nut, " (RF failed)")
      next
    }

    trend <- as.numeric(predict(rf_mod, newdata = df[, covariates, drop = FALSE]))
    resid <- df[[nut]] - trend

    # attach residuals to matching sf rows
    train_res_sf <- soil_sf_use[df$.row_id, ] %>% mutate(resid = resid)

    v_emp <- tryCatch(gstat::variogram(resid ~ 1, train_res_sf), error = function(e) NULL)
    if (is.null(v_emp) || nrow(v_emp) < 3) {
      message("[RK VARIO] skipping ", ds, " ", nut, " (resid variogram failed)")
      next
    }

    v_fit <- fit_variogram_safe(v_emp, resid)

    out_png <- file.path(out_dir_vario, sprintf("RKresid_%s_%s_variogram.png", ds, nut))
    save_variogram_plot(v_emp, v_fit,
                        title = paste0("RK residual variogram — ", ds, " — ", nut),
                        out_png = out_png)

    params <- get_vgm_params(v_fit) %>%
      mutate(dataset = ds, method = "RK_residual", nutrient = nut)

    rk_param_rows[[length(rk_param_rows) + 1]] <- params

    write.csv(v_emp, file.path(out_dir_vario, sprintf("RKresid_%s_%s_empirical.csv", ds, nut)), row.names = FALSE)
  }
}

rk_params_tbl <- bind_rows(rk_param_rows)
write.csv(rk_params_tbl, file.path(out_dir_vario, "RK_residual_variogram_params.csv"), row.names = FALSE)

# -----------------------------
# 3) VIP (RF importance) for SP and C, per nutrient
# -----------------------------
get_rf_vip <- function(soil_sf, covariates, nut, ntree = 500){
  df <- st_drop_geometry(soil_sf) %>%
    select(all_of(c(covariates, nut))) %>%
    na.omit()
  if (nrow(df) < 15) return(NULL)

  rf <- randomForest::randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nut]],
    ntree = ntree,
    importance = TRUE
  )

  imp <- randomForest::importance(rf)

  # regression: importance is matrix; keep whatever exists
  imp_tbl <- as.data.frame(imp) %>%
    tibble::rownames_to_column("covariate") %>%
    as_tibble()

  # pick a primary importance column
  # commonly: "%IncMSE" or "IncNodePurity"
  primary <- intersect(c("%IncMSE", "IncNodePurity"), names(imp_tbl))
  if (length(primary) == 0) {
    # fallback: first numeric column
    primary <- names(imp_tbl)[sapply(imp_tbl, is.numeric)][1]
  } else {
    primary <- primary[1]
  }

  imp_tbl %>%
    mutate(primary_metric = primary, importance_value = .data[[primary]]) %>%
    arrange(desc(importance_value))
}

vip_rows <- list()

for (ds in c("SP","C")) {
  soil_sf_use <- if (ds == "SP") soil_sf_SP else soil_sf_C

  for (nut in soil_vars_to_use) {
    vip_tbl <- get_rf_vip(soil_sf_use, covariates, nut)
    if (is.null(vip_tbl)) next

    vip_rows[[length(vip_rows) + 1]] <- vip_tbl %>%
      mutate(dataset = ds, nutrient = nut)
  }
}

vip_all <- bind_rows(vip_rows)
write.csv(vip_all, file.path(out_dir_vip, "vip_all.csv"), row.names = FALSE)

vip_top <- vip_all %>%
  group_by(dataset, nutrient) %>%
  slice_max(order_by = importance_value, n = top_k_vip, with_ties = FALSE) %>%
  ungroup()

write.csv(vip_top, file.path(out_dir_vip, sprintf("vip_top%d.csv", top_k_vip)), row.names = FALSE)

# -----------------------------
# 4) Plot VIP covariate maps (patchwork) — SP and C separately
#    For each nutrient, map its top_k covariates from vip_top
# -----------------------------
make_cov_map <- function(r, title){
  df <- as.data.frame(r, xy = TRUE, na.rm = FALSE)
  names(df) <- c("x","y","val")

  ggplot(df, aes(x, y, fill = val)) +
    geom_raster() +
    coord_equal() +
    labs(title = title, x = NULL, y = NULL, fill = NULL) +
    scale_fill_viridis_c(na.value = "transparent") +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face="bold", hjust=0.5, size=11),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.position = "bottom",
      legend.key.width  = unit(2.0, "cm"),
      legend.key.height = unit(0.25, "cm"),
      legend.text  = element_text(size = 8),
      plot.margin = ggplot2::margin(2, 2, 2, 2, unit = "pt")
    )
}

plot_vip_maps_one <- function(dataset_label, nutrient, vip_top_tbl, covariate_stack, top_k = 6, ncol = 3){
  sel <- vip_top_tbl %>%
    filter(dataset == dataset_label, nutrient == nutrient) %>%
    slice_head(n = top_k)

  if (nrow(sel) == 0) return(NULL)

  plots <- list()
  for (cv in sel$covariate) {
    if (!cv %in% names(covariate_stack)) next
    r <- covariate_stack[[cv]]
    plots[[length(plots) + 1]] <- make_cov_map(r, cv)
  }

  if (length(plots) == 0) return(NULL)

  wrap_plots(plots, ncol = ncol) +
    plot_annotation(title = paste0("Top ", top_k, " RF predictors — ", dataset_label, " — ", nutrient)) &
    theme(plot.title = element_text(face="bold", hjust=0.5, size=16))
}

# build and save VIP map panels (per dataset + nutrient)
for (ds in c("SP","C")) {
  for (nut in soil_vars_to_use) {

    pw <- plot_vip_maps_one(ds, nut, vip_top, covariate_stack, top_k = top_k_vip, ncol = 3)
    if (is.null(pw)) next

    out_png <- file.path(out_dir_vipmaps, sprintf("VIPmaps_%s_%s_top%d.png", ds, nut, top_k_vip))
    ggsave(out_png, pw, width = 12, height = 8.5, dpi = 300)
  }
}

cat("\nDone.\n",
    "Variograms saved in: ", out_dir_vario, "\n",
    "VIP tables saved in: ", out_dir_vip, "\n",
    "VIP map panels saved in: ", out_dir_vipmaps, "\n")
```

### VIP Analysis for RF and RK Models with SR Indices

```{r}
#VIPS for RF and RK models with SR indices:
#cov stack:
covariate_stack_idx <- terra::rast("../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m_all_with_SR_indices.tif")


# stack with indices + covariates you want to use
stack_use <- covariate_stack_idx

# soil variables you want to predict (keep these from the point table)
soil_vars_all <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")

extract_covs_to_points_nodupes <- function(soil_sf, rstack, soil_vars_all){

  # 1) align CRS to raster
  soil_sf_r <- st_transform(soil_sf, terra::crs(rstack, proj = TRUE))

  # 2) keep ONLY soil vars that exist + geometry (prevents duplicates later)
  keep_soils <- intersect(soil_vars_all, names(soil_sf_r))
  soil_sf_min <- soil_sf_r %>% dplyr::select(all_of(keep_soils), geometry)

  # 3) extract all raster layers to points
  ex <- terra::extract(rstack, terra::vect(soil_sf_min))

  # 4) bind extracted values back by row order (ID is 1..n in order)
  #    ex includes ID + raster columns
  ex <- ex %>% dplyr::select(-ID)

  out <- dplyr::bind_cols(soil_sf_min, ex)

  # 5) sanity check: extracted covariates should match raster names exactly
  stopifnot(all(names(rstack) %in% names(out)))

  out
}

soil_sf_SP2 <- extract_covs_to_points_nodupes(soil_sf_SP, stack_use, soil_vars_all)
soil_sf_C2  <- extract_covs_to_points_nodupes(soil_sf_C,  stack_use, soil_vars_all)

# covariates are exactly raster names minus soil variables (if any overlap)
covariates <- setdiff(names(stack_use), soil_vars_all)

# sanity checks (these should PASS now)
stopifnot(all(covariates %in% names(st_drop_geometry(soil_sf_SP2))))
stopifnot(all(covariates %in% names(st_drop_geometry(soil_sf_C2))))

cat("✅ Extraction complete with NO duplicate .x/.y columns\n")
cat("Covariates used:", length(covariates), "\n")
cat("First covariates:", paste(head(covariates, 12), collapse = ", "), "\n")

# -----------------------------
# 1) Fit RF + extract VIP (bar-graph-ready)
# -----------------------------
get_vip_one <- function(soil_sf, covariates, nutrient, ntree = 500, seed = 1){

  set.seed(seed)

  df <- st_drop_geometry(soil_sf) %>%
    select(all_of(c(covariates, nutrient))) %>%
    na.omit()

  if (nrow(df) < 15) return(NULL)

  rf <- randomForest::randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nutrient]],
    ntree = ntree,
    importance = TRUE
  )

  imp <- randomForest::importance(rf) %>% as.data.frame()
  imp <- tibble::rownames_to_column(imp, "covariate")

  # pick a primary importance column (prefer %IncMSE)
  primary <- intersect(c("%IncMSE", "IncNodePurity"), names(imp))
  primary <- if (length(primary) == 0) names(imp)[sapply(imp, is.numeric)][1] else primary[1]

  imp %>%
    transmute(
      covariate,
      importance_metric = primary,
      importance_value = .data[[primary]]
    ) %>%
    arrange(desc(importance_value))
}

# -----------------------------
# 2) Build VIP table for SP/C and RF/RK
#    (RK uses RF trend VIP; we label it "RK" so you can plot alongside RF.)
# -----------------------------
vip_rows <- list()

for (ds in c("SP","C")) {

  soil_sf_use <- if (ds == "SP") soil_sf_SP2 else soil_sf_C2

  for (nut in soil_vars_to_map) {

    vip_rf <- get_vip_one(soil_sf_use, covariates, nut, ntree = 500, seed = 1)

    if (!is.null(vip_rf)) {

      vip_rf <- vip_rf %>%
        mutate(dataset = ds, method = "RF", nutrient = nut)

      # RK “VIP” = same RF-trend VIP (because RK trend is RF)
      vip_rk <- vip_rf %>%
        mutate(method = "RK")

      vip_rows[[length(vip_rows) + 1]] <- vip_rf
      vip_rows[[length(vip_rows) + 1]] <- vip_rk
    }
  }
}

vip_all <- bind_rows(vip_rows) %>%
  group_by(dataset, method, nutrient) %>%
  mutate(rank = row_number()) %>%
  ungroup()

# Save bar-graph data
dir.create("vip_outputs", showWarnings = FALSE)
write.csv(vip_all, "vip_outputs/vip_all_RF_RK_SP_C.csv", row.names = FALSE)

# -----------------------------
# 3) Quick bar plot (top K predictors per nutrient)
# -----------------------------
top_k <- 10

vip_plot_dat <- vip_all %>%
  filter(rank <= top_k) %>%
  mutate(covariate = reorder_within(covariate, importance_value, interaction(dataset, method, nutrient)))

# helper for reorder_within
reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  new_x <- paste(x, within, sep = sep)
  stats::reorder(new_x, by, FUN = fun)
}
scale_x_reordered <- function(..., sep = "___") {
  ggplot2::scale_x_discrete(labels = function(x) sub(paste0(sep, ".*$"), "", x), ...)
}

p_vip <- ggplot(vip_plot_dat, aes(x = covariate, y = importance_value)) +
  geom_col() +
  coord_flip() +
  facet_grid(nutrient ~ dataset + method, scales = "free_y") +
  scale_x_reordered() +
  theme_minimal(base_size = 8) +
  labs(
    title = paste0("Top ", top_k, " VIP predictors (new indices)"),
    x = NULL,
    y = "Importance"
  )

# ggsave("vip_outputs/vip_top_predictors_RF_RK_SP_C.png", p_vip, width = 14, height = 10, dpi = 300)

#p_vip

```

### VIP Bar Plots with Shared X-Limits for SP vs Composite Comparison

```{r}

top_k <- 5
method_keep <- "RF"   # change to "RK" for RK plot

soil_vars_to_plot <- c("p_h_2","ca","k","mg","mn","p","zn","c","n")

nut_labs <- c(
  "p_h_2" = "pH",
  "ca"    = "Calcium (Ca)",
  "k"     = "Potassium (K)",
  "mg"    = "Magnesium (Mg)",
  "mn"    = "Manganese (Mn)",
  "p"     = "Phosphorus (P)",
  "zn"    = "Zinc (Zn)",
  "c"     = "Carbon (C)",
  "n"     = "Nitrogen (N)"
)

# 1) Top K per nutrient PER dataset (SP and C separately)
top_each_ds <- vip_all %>%
  filter(method == method_keep,
         dataset %in% c("SP","C"),
         nutrient %in% soil_vars_to_plot) %>%
  group_by(nutrient, dataset) %>%
  slice_max(order_by = importance_value, n = top_k, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    dataset_lab = factor(dataset, levels = c("SP","C"),
                         labels = c("Single Point (SP)", "Composite (C)")),
    nutrient = recode(nutrient, !!!nut_labs)
  )

# quick proof it’s top 5 per dataset:
# top_each_ds %>% count(nutrient, dataset_lab)

# 2) Per-nutrient x-limit (shared within nutrient so SP/C comparable)
xlim_tbl <- top_each_ds %>%
  group_by(nutrient) %>%
  summarize(xmax = max(importance_value, na.rm = TRUE), .groups = "drop") %>%
  mutate(xmax = xmax * 1.08)

make_one_nutrient_plot <- function(df_nut, title_txt, xmax){

  # dataset-specific y-key so each facet can have its own ordering
  df_nut <- df_nut %>%
    mutate(cov_plot = paste(dataset_lab, covariate, sep = "___")) %>%
    group_by(dataset_lab) %>%
    arrange(importance_value, .by_group = TRUE) %>%  # low->high (so high ends up top after flip-like reading)
    mutate(cov_plot = factor(cov_plot, levels = unique(cov_plot))) %>%
    ungroup()

  ggplot(df_nut, aes(x = importance_value, y = cov_plot, fill = dataset_lab)) +
    geom_col(width = 0.75, alpha = 0.95) +
    facet_wrap(~ dataset_lab, nrow = 1, scales = "free_y") +
    scale_x_continuous(limits = c(0, xmax), expand = expansion(mult = c(0, 0))) +
    # strip the dataset prefix from y labels
    scale_y_discrete(labels = function(x) sub("^.*___", "", x)) +
    labs(title = title_txt, x = "VIP", y = NULL) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
      strip.text = element_text(face = "bold", size = 9),
      axis.text.y = element_text(size = 8),
      axis.text.x = element_text(size = 8),
      legend.position = "none",
      panel.grid.major.y = element_blank()
    )
}

plots <- top_each_ds %>%
  group_split(nutrient) %>%
  lapply(function(d){
    nut <- unique(d$nutrient)
    xmax <- xlim_tbl$xmax[xlim_tbl$nutrient == nut][1]
    make_one_nutrient_plot(d, nut, xmax)
  })

final_patch <- wrap_plots(plots, ncol = 2) +
  plot_annotation(
    title = paste0("Top ", top_k, " VIP predictors per nutrient (", method_keep, "): SP vs Composite"),
    theme = theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14))
  )

final_patch
#save as 300 dpi png, with powerpoint slide size (11 in x 8.5 in)
ggsave("vip_outputs/vip_top5_RF_SP_vs_C.png", final_patch, width = 16, height = 9, dpi = 300)
# Negative importance → shuffling improves performance → predictor is actively harmful / noisy / redundant

```


### Validation of predictions at point locations

```{r}
# now onto validation. I have a dataset with validation points.

map_dir <- "C:/git/Perry_surface_analysis/code/best_boot_maps"
r <- rast(file.path(map_dir, "SP_c_RF_bestboot.tif"))  # raster is EPSG:26917

validation_data <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/perry_fp_Ap_0_20_with_coords.csv") %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y)
  )

# IMPORTANT: your x/y are Web Mercator meters
validation_sf_fix <- st_as_sf(validation_data, coords = c("x","y"), crs = 3857, remove = FALSE) %>%
  st_transform(terra::crs(r, proj = TRUE))  # -> 26917

# confirm coords now look like UTM-ish
xy <- st_coordinates(validation_sf_fix)
print(summary(xy[,1]))
print(summary(xy[,2]))

# check overlap
val_v_fix <- terra::vect(validation_sf_fix)
inside <- terra::relate(val_v_fix, r, "intersects")
cat("Points intersect raster extent:", sum(inside), "of", length(inside), "\n")

map_dir <- "C:/git/Perry_surface_analysis/code/best_boot_maps"
r <- rast(file.path(map_dir, "SP_c_RF_bestboot.tif"))  # raster is EPSG:26917

validation_data <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/perry_fp_Ap_0_20_with_coords.csv") %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y)
  )

# IMPORTANT: your x/y are Web Mercator meters
validation_sf_fix <- st_as_sf(validation_data, coords = c("x","y"), crs = 3857, remove = FALSE) %>%
  st_transform(terra::crs(r, proj = TRUE))  # -> 26917

# confirm coords now look like UTM-ish
xy <- st_coordinates(validation_sf_fix)
print(summary(xy[,1]))
print(summary(xy[,2]))

# check overlap
val_v_fix <- terra::vect(validation_sf_fix)
inside <- terra::relate(val_v_fix, r, "intersects")
cat("Points intersect raster extent:", sum(inside), "of", length(inside), "\n")


# -----------------------------
# SETTINGS
# -----------------------------
map_dir <- "C:/git/Perry_surface_analysis/code/best_boot_maps"
dataset_keep <- "SP"
soil_vars_to_validate <- c("ca","k","mg","mn","p","zn","c","n","p_h_2")
methods <- c("RF","OK","RK")

stopifnot(exists("validation_sf_fix"))
val_v <- terra::vect(validation_sf_fix)

# -----------------------------
# Index map files
# -----------------------------
map_files <- list.files(map_dir, pattern = "\\.tif$", full.names = TRUE)

map_index <- tibble(path = map_files, file = basename(map_files)) %>%
  mutate(
    dataset = str_match(file, "^(SP|C)_")[,2],
    method  = str_match(file, "_(RF|OK|RK)_bestboot\\.tif$")[,2],
    nutrient = str_match(file, "^(SP|C)_(.*)_(RF|OK|RK)_bestboot\\.tif$")[,3]
  ) %>%
  filter(!is.na(dataset), !is.na(method), !is.na(nutrient))

map_index_use <- map_index %>%
  filter(dataset == dataset_keep, method %in% methods, nutrient %in% soil_vars_to_validate)

# -----------------------------
# Extract predictions -> long table
# -----------------------------
extract_one <- function(nut, meth){
  if(!nut %in% names(validation_sf_fix)) {
    return(tibble(observed = numeric(0), predicted = numeric(0)))
  }

  obs <- as.numeric(validation_sf_fix[[nut]])

  row_match <- map_index_use %>% filter(nutrient == nut, method == meth)
  if(nrow(row_match) == 0){
    return(tibble(observed = obs, predicted = rep(NA_real_, length(obs))))
  }

  r <- terra::rast(row_match$path[1])
  val_v_use <- if(!terra::same.crs(r, val_v)) terra::project(val_v, terra::crs(r)) else val_v

  ex <- terra::extract(r, val_v_use)
  pred <- if(!is.null(ex) && nrow(ex) == length(obs)) as.numeric(ex[[2]]) else rep(NA_real_, length(obs))

  tibble(observed = obs, predicted = pred)
}

val_pred_long <- crossing(nutrient = soil_vars_to_validate, method = methods) %>%
  mutate(dat = map2(nutrient, method, extract_one)) %>%
  unnest(dat) %>%
  mutate(dataset = dataset_keep) %>%
  filter(is.finite(observed), is.finite(predicted))

stopifnot(nrow(val_pred_long) > 0)

# -----------------------------
# Metrics
# -----------------------------
safe_r2 <- function(o, p){
  ok <- is.finite(o) & is.finite(p)
  if(sum(ok) < 2) return(NA_real_)
  cc <- suppressWarnings(cor(o[ok], p[ok], use="complete.obs"))
  if(is.na(cc)) NA_real_ else cc^2
}

val_metrics <- val_pred_long %>%
  group_by(dataset, nutrient, method) %>%
  summarise(
    n = n(),
    RMSE = sqrt(mean((predicted - observed)^2)),
    R2   = safe_r2(observed, predicted),
    .groups = "drop"
  ) %>%
  mutate(
    lab = paste0("n=", n,
                 "\nRMSE=", format(round(RMSE, 3), nsmall = 3),
                 "\nR²=",   ifelse(is.na(R2), "NA", format(round(R2, 3), nsmall = 3)))
  )

print(val_metrics)

# -----------------------------
# Pretty nutrient labels
# -----------------------------
nut_labs <- c(ca="Ca", k="K", mg="Mg", mn="Mn", p="P", zn="Zn", c="C", n="N", p_h_2="pH")

make_one_nutrient_plot <- function(nut){

  df  <- val_pred_long %>% filter(nutrient == nut)
  met <- val_metrics    %>% filter(nutrient == nut)

  # ---- if NOTHING to facet, return a placeholder plot ----
  if(nrow(df) == 0){
    return(
      ggplot() +
        theme_void(base_size = 12) +
        labs(title = paste0("Validation: ", nut_labs[[nut]] %||% nut, " (", dataset_keep, ")")) +
        annotate("text", x = 0, y = 0, label = "No finite observed/predicted pairs\n(check extraction + NAs)", size = 4) +
        theme(plot.title = element_text(face = "bold", hjust = 0.5))
    )
  }

  # make sure method is a factor so facet order is stable
  df <- df %>% mutate(method = factor(method, levels = methods))
  met <- met %>% mutate(method = factor(method, levels = methods))

  # if a method is missing for this nutrient, facet_wrap is still OK as long as df has rows
  ggplot(df, aes(observed, predicted)) +
    geom_point(alpha = 0.85, size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
    facet_wrap(~ method, scales = "free") +
    theme_minimal(base_size = 12) +
    labs(
      title = paste0("Validation: ", nut_labs[[nut]] %||% nut, " (", dataset_keep, ")"),
      x = "Observed (validation)",
      y = "Predicted (map)"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold", hjust = 0.5)
    ) +
    geom_label(
      data = met,
      aes(x = -Inf, y = Inf, label = lab),
      inherit.aes = FALSE,
      hjust = -0.05, vjust = 1.1,
      size = 3,
      label.size = 0.2
    )
}

plots <- lapply(soil_vars_to_validate, make_one_nutrient_plot)

final_validation_plot <- wrap_plots(plots, ncol = 3)
print(final_validation_plot)

# Optional save:
ggsave("../ouputs/validation_SP_all_nutrients.png", final_validation_plot, width = 16, height = 9, dpi = 300)

#this is OVERFIT "best r2"
# | Nutrient | Best method   | Interpretation                               |
# | -------- | ------------- | -------------------------------------------- |
# | **P**    | OK            | Strong spatial signal, weak covariate signal |
# | **Mg**   | OK            | Moderate spatial coherence                   |
# | **Ca**   | OK            | Spatial smoothing beats ML                   |
# | **K**    | None          | Weak signal overall                          |
# | **Mn**   | None          | High noise / microscale                      |
# | **Zn**   | RF/RK (weak)  | Slight covariate signal                      |
# | **pH**   | None          | Depth + management driven                    |
# | **C, N** | Not validated | (and that’s okay to state explicitly)        |



```


### Validation: Standardized Errors and “How Far Off?” Plots

```{r}


# val_pred_long must have: nutrient, method, observed, predicted
# example methods: "RF","OK","RK"
stopifnot(all(c("nutrient","method","observed","predicted") %in% names(val_pred_long)))

# 1) Compute validation metrics from val_pred_long (so we don't rely on a separate table)
val_metrics <- val_pred_long %>%
  group_by(nutrient, method) %>%
  summarise(
    n = sum(is.finite(observed) & is.finite(predicted)),
    RMSE = sqrt(mean((predicted - observed)^2, na.rm = TRUE)),
    R2 = {cc <- suppressWarnings(cor(observed, predicted, use = "complete.obs")); ifelse(is.na(cc), NA_real_, cc^2)},
    sd_obs = sd(observed, na.rm = TRUE),
    .groups = "drop"
  )

# 2) Standardize errors: by RMSE (recommended) + by observed SD (optional)
val_std <- val_pred_long %>%
  left_join(val_metrics, by = c("nutrient","method")) %>%
  mutate(
    err = predicted - observed,
    z_rmse = err / RMSE,                     # "how many RMSEs off?"
    z_sd   = err / sd_obs,                   # "how many observed SDs off?"
    abs_z_rmse = abs(z_rmse)
  ) %>%
  filter(is.finite(z_rmse))

# 3) Nice labels (edit if you want)
nut_labs <- c(
  ca="Ca", k="K", mg="Mg", mn="Mn", p="P", zn="Zn", c="C", n="N", p_h_2="pH"
)

val_std <- val_std %>%
  mutate(
    nutrient_lab = ifelse(nutrient %in% names(nut_labs), nut_labs[nutrient], nutrient),
    method = factor(method, levels = c("OK","RF","RK"))
  )

# ---- PLOT 1: Distribution of standardized errors (by RMSE) ----
# Interpretation: 0 = perfect; +/-1 = off by 1 RMSE; +/-2 = off by 2 RMSE
p1 <- ggplot(val_std, aes(x = z_rmse, fill = method)) +
  geom_histogram(bins = 15, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 0, linewidth = 0.7) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", linewidth = 0.6) +
  geom_vline(xintercept = c(-2, 2), linetype = "dotted", linewidth = 0.6) +
  facet_wrap(~ nutrient_lab, scales = "free_y") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Validation: standardized prediction error (in RMSE units)",
    x = "(Predicted − Observed) / RMSE  [per nutrient + method]",
    y = "Count",
    fill = "Method"
  ) +
  theme(legend.position = "top")
print(p1)

# ---- PLOT 2: “How far off?” absolute standardized errors ----
# This one is super intuitive: lower is better.
p2 <- val_std %>%
  group_by(nutrient_lab, method) %>%
  summarise(
    n = dplyr::n(),
    median_abs = median(abs_z_rmse, na.rm = TRUE),
    mean_abs = mean(abs_z_rmse, na.rm = TRUE),
    p90_abs = quantile(abs_z_rmse, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = nutrient_lab, y = median_abs, color = method, group = method)) +
  geom_point(size = 2.6) +
  geom_line(linewidth = 0.7) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Validation: median absolute error in RMSE units (lower = better)",
    x = NULL,
    y = "Median |(Pred−Obs)/RMSE|",
    color = "Method"
  ) +
  theme(legend.position = "top")
print(p2)

# ---- SUMMARY TABLE: % within 1 RMSE and 2 RMSE ----
val_within <- val_std %>%
  group_by(nutrient_lab, method) %>%
  summarise(
    n = dplyr::n(),
    pct_within_1 = mean(abs_z_rmse <= 1, na.rm = TRUE) * 100,
    pct_within_2 = mean(abs_z_rmse <= 2, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(nutrient_lab, method)

print(val_within)

#nice table:
library(knitr)
val_within %>%
  kable(
    caption = "Validation: Percentage of predictions within 1 RMSE and 2 RMSE",
    col.names = c("Nutrient", "Method", "N", "% within 1 RMSE", "% within 2 RMSE"),
    digits = c(0,0,0,1,1)
  )
```


### MZ maps for P

```{r}


# -----------------------------
# 0) Settings you control
# -----------------------------
map_dir  <- "C:/git/Perry_surface_analysis/code/best_boot_maps"  # adjust if needed
dataset  <- "SP"
nutrient <- "p"

# IMPORTANT: set to the depth your P map represents
depth_cm <- 15      # <-- change to 20 if your P map is 0–20 cm
bd_gcm3  <- 1.33    # bulk density assumption (g/cm3)

# -----------------------------
# 1) Coastal Plain P thresholds (lbs/ac) from your table
# -----------------------------
# Low: 0–30, Medium: 31–60, High: 61–100, Very High: 100+
p_lbs_breaks <- c(0, 30, 60, 100, Inf)

# -----------------------------
# 2) Convert lbs/ac thresholds -> ppm thresholds for your depth + BD
#    ppm_to_lbsac(1 ppm) gives the lbs/ac per ppm at that depth.
# -----------------------------
ppm_to_lbsac_factor <- function(depth_cm, bulk_density_gcm3 = 1.33){
  area_m2 <- 4046.8564224
  depth_m <- depth_cm / 100
  bd_kgm3 <- bulk_density_gcm3 * 1000
  soil_mass_kg <- area_m2 * depth_m * bd_kgm3
  # 1 ppm = 1 mg/kg, so lbs/ac per ppm:
  (soil_mass_kg * 2.2046226218) / 1e6
}

lbs_per_ppm <- ppm_to_lbsac_factor(depth_cm, bd_gcm3)

# Convert lbs/ac breaks to ppm breaks (same number of breaks)
# (Inf stays Inf)
p_ppm_breaks <- p_lbs_breaks
p_ppm_breaks[is.finite(p_ppm_breaks)] <- p_ppm_breaks[is.finite(p_ppm_breaks)] / lbs_per_ppm

cat("Depth (cm):", depth_cm, " BD:", bd_gcm3, "\n")
cat("lbs/ac per ppm:", round(lbs_per_ppm, 4), "\n")
cat("P thresholds (lbs/ac):", paste(p_lbs_breaks, collapse=" | "), "\n")
cat("P thresholds (ppm)   :", paste(round(p_ppm_breaks, 2), collapse=" | "), "\n\n")

# -----------------------------
# 3) Helper to bin a ppm raster into categorical zones
# -----------------------------
classify_P_zones_ppm <- function(r_ppm, breaks_ppm, breaks_lbs){
  # Create a categorical raster with classes 1..4
  # zones: [0,b1], (b1,b2], (b2,b3], (b3,Inf]
  r_cat <- classify(r_ppm, rcl = matrix(c(
    -Inf, breaks_ppm[2], 1,
    breaks_ppm[2], breaks_ppm[3], 2,
    breaks_ppm[3], breaks_ppm[4], 3,
    breaks_ppm[4],  Inf,          4
  ), ncol = 3, byrow = TRUE), include.lowest = TRUE, right = TRUE)

  r_cat <- as.factor(r_cat)

  # Pretty labels include BOTH ppm & lbs/ac ranges
  lab <- c(
    paste0("Low (≤",   round(breaks_ppm[2],2), " ppm; ≤", breaks_lbs[2], " lb/ac)"),
    paste0("Medium (", round(breaks_ppm[2],2), "–", round(breaks_ppm[3],2), " ppm; ",
           breaks_lbs[2]+1, "–", breaks_lbs[3], " lb/ac)"),
    paste0("High (",   round(breaks_ppm[3],2), "–", round(breaks_ppm[4],2), " ppm; ",
           breaks_lbs[3]+1, "–", breaks_lbs[4], " lb/ac)"),
    paste0("Very High (>", round(breaks_ppm[4],2), " ppm; >", breaks_lbs[4], " lb/ac)")
  )

  levels(r_cat) <- data.frame(value = 1:4, zone = lab)
  r_cat
}

# -----------------------------
# 4) Load SP P rasters for RF / OK / RK and classify
# -----------------------------
methods <- c("RF","OK","RK")

p_maps <- lapply(methods, function(m){
  f <- file.path(map_dir, sprintf("%s_%s_%s_bestboot.tif", dataset, nutrient, m))
  if(!file.exists(f)) stop("Missing file: ", f)
  rast(f)
})
names(p_maps) <- methods

p_zones <- lapply(p_maps, classify_P_zones_ppm, breaks_ppm = p_ppm_breaks, breaks_lbs = p_lbs_breaks)

# -----------------------------
# 5) Plot side-by-side (not saving yet)
# -----------------------------
op <- par(no.readonly = TRUE)
par(mfrow = c(1,3), mar = c(3,3,4,7))

for(m in methods){
  plot(p_zones[[m]],
       main = sprintf("P Zones (%s %s)\nDepth=%scm  BD=%s", dataset, m, depth_cm, bd_gcm3),
       axes = FALSE)
}

par(op)
```

```{r}

# -----------------------------
# SETTINGS
# -----------------------------
map_dir  <- "C:/git/Perry_surface_analysis/code/best_boot_maps"
dataset  <- "SP"
nutrient <- "p"
methods  <- c("RF","OK","RK")

# Your UGA Coastal Plain thresholds (lbs/ac)
p_lbs_breaks <- c(0, 30, 60, 100, Inf)

# Choose ONE binning basis for EVERYTHING:
#   "ppm"  = assume raster+validation are ppm
#   "lbs"  = assume raster+validation should be treated as lbs/ac
bin_basis <- "lbs"   # <-- TRY "lbs" FIRST given your weird result

# Conversion settings (only used if we need ppm <-> lbs/ac)
depth_cm <- 15
bd_gcm3  <- 1.33

# -----------------------------
# 1) Read reference raster for CRS
# -----------------------------
r_ref <- rast(file.path(map_dir, "SP_c_RF_bestboot.tif"))  # EPSG:26917

# -----------------------------
# 2) Validation points (your fix: x/y are EPSG:3857)
# -----------------------------
validation_data <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/perry_fp_Ap_0_20_with_coords.csv") %>%
  mutate(x = as.numeric(x), y = as.numeric(y))

validation_sf_fix <- st_as_sf(validation_data, coords = c("x","y"), crs = 3857, remove = FALSE) %>%
  st_transform(terra::crs(r_ref, proj = TRUE))

stopifnot(nutrient %in% names(validation_sf_fix))
val_v <- vect(validation_sf_fix)

# overlap check
inside <- relate(val_v, r_ref, "intersects")
cat("Points intersect raster extent:", sum(inside), "of", length(inside), "\n")

# -----------------------------
# 3) Helpers
# -----------------------------
ppm_to_lbsac_factor <- function(depth_cm, bulk_density_gcm3 = 1.33){
  area_m2 <- 4046.8564224
  depth_m <- depth_cm / 100
  bd_kgm3 <- bulk_density_gcm3 * 1000
  soil_mass_kg <- area_m2 * depth_m * bd_kgm3
  (soil_mass_kg * 2.2046226218) / 1e6  # lbs/ac per ppm
}

lbs_per_ppm <- ppm_to_lbsac_factor(depth_cm, bd_gcm3)

# Binning function (returns 1..4)
bin_P <- function(x, breaks){
  as.integer(cut(x,
    breaks = c(-Inf, breaks[2], breaks[3], breaks[4], Inf),
    labels = 1:4, right = TRUE, include.lowest = TRUE
  ))
}

# -----------------------------
# 4) Extract RAW predicted values at points
# -----------------------------
# --- Extract RAW predicted values at points (FIXED) ---
extract_pred_vals <- function(meth){
  f <- file.path(map_dir, sprintf("%s_%s_%s_bestboot.tif", dataset, nutrient, meth))
  if(!file.exists(f)) stop("Missing raster: ", f)

  r <- terra::rast(f)

  val_use <- if(!terra::same.crs(r, val_v)) terra::project(val_v, terra::crs(r)) else val_v

  ex <- terra::extract(r, val_use)   # <-- force terra::extract
  as.numeric(ex[[2]])
}


obs_ppm <- as.numeric(validation_sf_fix[[nutrient]])

pred_tbl <- tibble(
  id = seq_along(obs_ppm),
  observed_ppm = obs_ppm,
  RF = extract_pred_vals("RF"),
  OK = extract_pred_vals("OK"),
  RK = extract_pred_vals("RK")
) %>%
  pivot_longer(cols = all_of(methods), names_to = "method", values_to = "pred_raw") %>%
  filter(is.finite(observed_ppm), is.finite(pred_raw))

# -----------------------------
# 5) DEBUG PRINTS: are rasters ppm or lbs?
# -----------------------------
cat("\n--- Observed P (validation) summary (ppm) ---\n")
print(summary(pred_tbl$observed_ppm))

cat("\n--- Predicted RAW P summary by method (whatever units raster is in) ---\n")
print(pred_tbl %>% group_by(method) %>% summarise(
  n = n(),
  min = min(pred_raw, na.rm=TRUE),
  q25 = quantile(pred_raw, 0.25, na.rm=TRUE),
  med = median(pred_raw, na.rm=TRUE),
  q75 = quantile(pred_raw, 0.75, na.rm=TRUE),
  max = max(pred_raw, na.rm=TRUE)
))

cat("\n--- Quick ratio check: median(pred)/median(obs) ---\n")
print(pred_tbl %>% group_by(method) %>% summarise(
  ratio_med = median(pred_raw, na.rm=TRUE) / median(observed_ppm, na.rm=TRUE)
))

# If ratio ~2 (or >>2), your raster is probably NOT ppm.

# -----------------------------
# 6) Put OBSERVED and PREDICTED on same binning basis
# -----------------------------
if(bin_basis == "ppm"){
  # interpret table thresholds as lbs/ac, convert thresholds -> ppm (your earlier approach)
  p_ppm_breaks <- p_lbs_breaks
  p_ppm_breaks[is.finite(p_ppm_breaks)] <- p_ppm_breaks[is.finite(p_ppm_breaks)] / lbs_per_ppm

  cat("\nUsing bin_basis = ppm\n")
  cat("lbs/ac per ppm:", round(lbs_per_ppm, 4), "\n")
  cat("P thresholds (ppm):", paste(round(p_ppm_breaks, 2), collapse=" | "), "\n")

  pred_tbl2 <- pred_tbl %>%
    mutate(
      obs_bin  = bin_P(observed_ppm, p_ppm_breaks),
      pred_bin = bin_P(pred_raw,      p_ppm_breaks)
    )

} else {
  # bin_basis == "lbs": convert observed ppm -> lbs/ac, and assume raster pred_raw is ALSO lbs/ac
  # (This is the most likely fix if your rasters were already converted or are “furrow-slice-like”.)
  cat("\nUsing bin_basis = lbs\n")
  cat("lbs/ac per ppm factor:", round(lbs_per_ppm, 4), "\n")

  pred_tbl2 <- pred_tbl %>%
    mutate(
      observed_lbs = observed_ppm * lbs_per_ppm,
      obs_bin  = bin_P(observed_lbs, p_lbs_breaks),
      pred_bin = bin_P(pred_raw,     p_lbs_breaks)
    )
}

# -----------------------------
# 7) Accuracy on bins (overall + within 1 bin)
# -----------------------------
mz_accuracy <- pred_tbl2 %>%
  group_by(method) %>%
  summarise(
    n = n(),
    accuracy = mean(obs_bin == pred_bin),
    pct_accuracy = 100 * accuracy,
    pct_within_1_bin = 100 * mean(abs(obs_bin - pred_bin) <= 1),
    .groups = "drop"
  )

cat("\n--- Bin accuracy ---\n")
print(mz_accuracy)

cat("\n--- Counts table (like a confusion matrix) ---\n")
print(pred_tbl2 %>% count(method, obs_bin, pred_bin) %>% arrange(method, obs_bin, pred_bin))

```


```{r}

# -----------------------------
# SETTINGS (same as your MZ chunk)
# -----------------------------
map_dir  <- "C:/git/Perry_surface_analysis/code/best_boot_maps"
dataset  <- "SP"
nutrient <- "p"
methods  <- c("RF","OK","RK")
P_0_15 <- read.csv("../LOCAL DATA DO NOT PUSH/Ap_hz_data/P_0_15_clean_test_02_10_2026.csv")  # must have Point, value_0_15
depth_cm <- 15
bd_gcm3  <- 1.33

# Coastal Plain P thresholds (lbs/ac) from your table
p_lbs_breaks <- c(0, 30, 60, 100, Inf)

ppm_to_lbsac_factor <- function(depth_cm, bulk_density_gcm3 = 1.33){
  area_m2 <- 4046.8564224
  depth_m <- depth_cm / 100
  bd_kgm3 <- bulk_density_gcm3 * 1000
  soil_mass_kg <- area_m2 * depth_m * bd_kgm3
  (soil_mass_kg * 2.2046226218) / 1e6
}
lbs_per_ppm <- ppm_to_lbsac_factor(depth_cm, bd_gcm3)

# Convert lbs/ac breaks -> ppm breaks (0–15 cm support)
p_ppm_breaks <- p_lbs_breaks
p_ppm_breaks[is.finite(p_ppm_breaks)] <- p_ppm_breaks[is.finite(p_ppm_breaks)] / lbs_per_ppm

cat("P breaks (lbs/ac):", paste(p_lbs_breaks, collapse=" | "), "\n")
cat("P breaks (ppm)   :", paste(round(p_ppm_breaks, 2), collapse=" | "), "\n\n")

# -----------------------------
# 1) Observed P for validation = depth-weighted 0–15 cm per Point
#    ASSUMES you already made P_0_15 with: Point, value_0_15
# -----------------------------
stopifnot(exists("validation_sf_fix"))
stopifnot(exists("P_0_15"))
stopifnot(all(c("Point","value_0_15") %in% names(P_0_15)))

#rename P_0_15 Point to match validation_sf_fix
P_0_15 <- P_0_15 %>%
  rename(PointLabel = Point)

val_pts <- validation_sf_fix %>%
  left_join(P_0_15, by = "PointLabel") %>%
  filter(is.finite(value_0_15))


cat("Validation points with depth-harmonized P (0–15):", nrow(val_pts), "\n")

# classify observed bins (1..4) using ppm thresholds
bin_from_ppm <- function(x, breaks_ppm){
  cut(x,
      breaks = breaks_ppm,
      include.lowest = TRUE,
      right = TRUE,
      labels = FALSE)
}

val_pts <- val_pts %>%
  mutate(obs_bin = bin_from_ppm(value_0_15, p_ppm_breaks))

# terra points object
val_v <- terra::vect(val_pts)

# -----------------------------
# 2) Make / load zone rasters for each method and extract predicted bin
#    IMPORTANT: classify zones based on ppm breaks (same as observed)
# -----------------------------
classify_P_zones_ppm <- function(r_ppm, breaks_ppm){
  # output values 1..4
  terra::classify(
    r_ppm,
    rcl = matrix(c(
      -Inf, breaks_ppm[2], 1,
      breaks_ppm[2], breaks_ppm[3], 2,
      breaks_ppm[3], breaks_ppm[4], 3,
      breaks_ppm[4],  Inf,          4
    ), ncol = 3, byrow = TRUE),
    include.lowest = TRUE,
    right = TRUE
  )
}

extract_pred_bin <- function(meth){

  f <- file.path(map_dir, sprintf("%s_%s_%s_bestboot.tif", dataset, nutrient, meth))
  stopifnot(file.exists(f))

  r_ppm <- terra::rast(f)

  # zone raster (1..4)
  r_zone <- classify_P_zones_ppm(r_ppm, p_ppm_breaks)

  # extract
  ex <- terra::extract(r_zone, val_v)
  pred_bin <- as.integer(ex[[2]])

  tibble(
    method = meth,
    Point = val_pts$Point,
    observed_ppm_0_15 = val_pts$value_0_15,
    obs_bin = val_pts$obs_bin,
    pred_bin = pred_bin
  )
}

mz_val_long <- bind_rows(lapply(methods, extract_pred_bin)) %>%
  filter(is.finite(obs_bin), is.finite(pred_bin))

stopifnot(nrow(mz_val_long) > 0)

# -----------------------------
# 3) Accuracy metrics on bins
# -----------------------------
mz_metrics <- mz_val_long %>%
  group_by(method) %>%
  summarise(
    n = n(),
    accuracy = mean(pred_bin == obs_bin),
    pct_accuracy = 100 * accuracy,
    pct_within_1_bin = 100 * mean(abs(pred_bin - obs_bin) <= 1),
    .groups = "drop"
  )

print(mz_metrics)

# confusion-ish counts
mz_counts <- mz_val_long %>%
  count(method, obs_bin, pred_bin) %>%
  arrange(method, obs_bin, pred_bin)

print(mz_counts)

# Optional: nice labels for bins
bin_labels <- tibble(
  bin = 1:4,
  label = c("Low", "Medium", "High", "Very High")
)

mz_counts_labeled <- mz_counts %>%
  left_join(bin_labels, by = c("obs_bin" = "bin")) %>%
  rename(obs_label = label) %>%
  left_join(bin_labels, by = c("pred_bin" = "bin")) %>%
  rename(pred_label = label)

print(mz_counts_labeled)
```





