---
title: "02_2026_analysis"
format: html
---
### Load Libraries

```{r, warning=FALSE}
# --- Load libraries ---
library(terra)
library(tidyverse)
library(sf)
library(sp)
library(gstat)
library(randomForest)
library(FNN)
library(purrr)
library(RColorBrewer)
library(ggrepel)
library(patchwork)
library(blockCV)
library(leaflet)
```

### Load Data and Prepare Datasets

```{r, warning=FALSE}

# --- Load data ---
covariate_stack <- rast("../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m.tif")
soil_covariate_data_scaled_SP <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/soil_covariate_data_scaled_SP.csv")
soil_covariate_data_scaled_C  <- read.csv("../LOCAL DATA DO NOT PUSH/covariates_and_soil_data/soil_covariate_data_scaled_C.csv")

# Load carbon + nitrogen dataset
soil_with_CN <- read_csv("../LOCAL DATA DO NOT PUSH/carbon/soil_data_with_CN.csv") %>%
  rename(c = leco_carbon_average,
         n = leco_nitrogen_average)

# Split by sampling type
SP_soil_with_CN <- soil_with_CN %>% filter(type == "SPID")
C_soil_with_CN  <- soil_with_CN %>% filter(type == "CID")

# Merge CN values into soil covariate tables by x,y coordinate
soil_SP_full <- soil_covariate_data_scaled_SP %>%
  left_join(SP_soil_with_CN %>% select(x,y,c,n), by=c("x","y"))

soil_C_full <- soil_covariate_data_scaled_C %>%
  left_join(C_soil_with_CN %>% select(x,y,c,n), by=c("x","y"))

# Define soil variable names using actual dataset column names
soil_vars <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")

# Define covariates by removing coordinates + soil variables
covariates <- setdiff(names(soil_SP_full), c("x","y", soil_vars))

# Convert to sf projected
soil_sf_SP <- st_as_sf(soil_SP_full, coords = c("x","y"), crs = 4326) %>% st_transform(32617)
soil_sf_C  <- st_as_sf(soil_C_full, coords = c("x","y"), crs = 4326) %>% st_transform(32617)


# #rename variables:
# rename_soils <- function(x) {
#   x %>%
#     rename(
#       lbc        = lbc_1,
#       lbc_eq     = lb_ceq,
#       ph         = p_h_2,
#       calcium    = ca,
#       potassium  = k,
#       magnesium  = mg,
#       manganese  = mn,
#       phosphorus = p,
#       zinc       = zn,
#       carbon     = c,
#       nitrogen   = n
#     )
# }
# 
# soil_sf_SP <- rename_soils(soil_sf_SP)
# soil_sf_C  <- rename_soils(soil_sf_C)
# 
# soil_vars_clean <- c("lbc","lbc_eq","ph","calcium","potassium","magnesium","manganese","phosphorus","zinc","carbon","nitrogen")
```


### Plot Composite vs Single Point with Outlier Labels

```{r, warning=FALSE}

# 1) Select only nutrient columns + coordinates
SP_vals <- soil_SP_full %>% select(x, y, all_of(soil_vars)) %>% mutate(dataset = "SP")
C_vals  <- soil_C_full %>% select(x, y, all_of(soil_vars)) %>% mutate(dataset = "Composite")

# 2) Combine and reshape to long format
soil_long <- bind_rows(SP_vals, C_vals) %>%
  pivot_longer(cols = all_of(soil_vars), names_to = "variable", values_to = "value")

# 3) Pivot wide: SP vs Composite in one row per point and nutrient
soil_wide <- soil_long %>%
  pivot_wider(names_from = dataset, values_from = value) %>%
  drop_na(SP, Composite)   # keeps only nutrients where both exist

  
# --- Attach PointLabel to merged dataset ---
# We join from SP_soil_with_CN since PointLabel is stable there
soil_wide_loc <- soil_wide %>%
  left_join(SP_soil_with_CN %>% select(x, y, PointLabel), by = c("x","y"))

# If you want cleaner facet labels again:
soil_wide_loc <- soil_wide_loc %>%
  mutate(variable = recode(variable,
                           "lbc_1" = "LBC1",
                           "lb_ceq" = "LBCeq",
                           "p_h_2" = "pH",
                           "ca" = "Ca",
                           "k" = "K",
                           "mg" = "Mg",
                           "mn" = "Mn",
                           "p" = "P",
                           "zn" = "Zn",
                           "c" = "C",
                           "n" = "N"))
# --- Compute residuals and identify outliers ---
soil_outliers <- soil_wide_loc %>%
  group_by(variable) %>%
  mutate(
    residual = SP - Composite,
    z = (residual - mean(residual, na.rm = TRUE)) / sd(residual, na.rm = TRUE),
    outlier = abs(z) > 2   # threshold
  )

# ✔ Check how many outliers per nutrient (optional)
soil_outliers %>% count(variable, outlier)

soil_no_outliers <- soil_outliers %>%
  filter(outlier == FALSE)

# --- Plot with 1:1 line and label only outliers ---
ggplot(soil_outliers, aes(x = Composite, y = SP)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  geom_text_repel(
    data = ~ dplyr::filter(.x, outlier == TRUE),
    aes(label = PointLabel),
    size = 2,
    fontface = "bold",
    max.overlaps = Inf,
    min.segment.length = 0
  ) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Composite vs Single Point with Outlier Labels",
    x = "Composite",
    y = "Single Point"
  ) +
  theme(strip.text = element_text(face = "bold"))
```

### Plot without outliers and R² values

```{r, warning=FALSE}

# --- Compute R² for each nutrient ---
r2_labels <- soil_no_outliers %>%
  filter(!is.na(SP), !is.na(Composite), !is.na(variable)) %>%
  group_by(variable) %>%
  summarise(
    r2 = cor(SP, Composite, use = "complete.obs")^2,
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 2)))

# --- Remove any NA variables before plotting ---
soil_no_outliers <- soil_no_outliers %>%
  filter(!is.na(variable), variable != "NA")

# --- Ensure custom facet order ---
var_order <- c("pH", "Ca", "K", "Mg", "Mn", "P", "Zn", "C", "N")
soil_no_outliers$variable <- factor(soil_no_outliers$variable, levels = var_order)
r2_labels$variable <- factor(r2_labels$variable, levels = var_order)

# --- Plot ---
ggplot(soil_no_outliers, aes(x = Composite, y = SP)) +
  geom_point(size = 2, alpha = 0.7, color = "#4C72B0") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "red", linewidth = 0.9) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.9) +
  geom_text(
    data = r2_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.1, vjust = 1.4,
    size = 3.3, fontface = "bold"
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Composite vs Single Point Nutrient Values",
    x = "Composite (ppm)",
    y = "Single Point (ppm)"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# ggsave("../outputs/figures/composite_vs_singlepoint_no_outliers_with_R2.png", width = 16, height = 9, units = "in", dpi = 300)

```

### Density Plots Comparing Distributions

```{r, warning=FALSE}
# ---- Prepare long-format dataset for density plots ----

# Select only the soil variables + dataset label
soil_SP_long <- soil_SP_full %>%
  mutate(dataset = "Single Point") %>%
  select(dataset, all_of(soil_vars)) %>%
  pivot_longer(cols = all_of(soil_vars), names_to = "variable", values_to = "value")

soil_C_long <- soil_C_full %>%
  mutate(dataset = "Composite") %>%
  select(dataset, all_of(soil_vars)) %>%
  pivot_longer(cols = all_of(soil_vars), names_to = "variable", values_to = "value")


# Optional: nicer variable names for plotting
soil_long_all <- soil_long_all %>%
  mutate(variable = recode(variable,
                           "lbc_1" = "LBC1",
                           "lb_ceq" = "LBCeq",
                           "p_h_2" = "pH",
                           "ca" = "Ca",
                           "k" = "K",
                           "mg" = "Mg",
                           "mn" = "Mn",
                           "p" = "P",
                           "zn" = "Zn",
                           "c" = "C",
                           "n" = "N"))
#plot all nutrients with SP on x axis and C on y axis

# Function to make each density plot with nutrient name as title
make_density <- function(var_name){
  ggplot(filter(soil_long_all, variable == var_name),
         aes(x = value, fill = dataset)) +
    geom_density(alpha = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    theme_minimal(base_size = 14) +
    labs(title = var_name, x = NULL, y = "Density", fill = "Sampling Method") +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      legend.position = "bottom"
    )
}

# Create each plot
p_pH <- make_density("pH")
p_Ca <- make_density("Ca")
p_K  <- make_density("K")
p_Mg <- make_density("Mg")
p_Mn <- make_density("Mn")
p_P  <- make_density("P")
p_Zn <- make_density("Zn")
p_C  <- make_density("C")
p_N  <- make_density("N")

# Arrange layout (Zn centered)
final_plot <- 
  (p_Ca | p_K | p_Mg) /
  (p_Mn | p_P | p_pH) /
  (p_C | p_Zn | p_N) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

# Add main title
# final_plot + 
#   plot_annotation(
#     title = "Soil Nutrient Distributions: Single Point vs Composite Sampling",
#     theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
#   )

#again but log scale all but pH

final_plot_log <-
  (
    (p_Ca + scale_x_log10() | p_K + scale_x_log10() | p_Mg + scale_x_log10()) /
    (p_Mn + scale_x_log10() | p_P + scale_x_log10() | p_pH) /
    (p_C + scale_x_log10()  | p_Zn + scale_x_log10() | p_N + scale_x_log10())
  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Soil Nutrient Distributions (Log Scale): Single Point vs Composite Sampling",
    theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  ) &
  theme(legend.position = "bottom")

final_plot_log

```
### Spatial Bootstrapping and Model Comparison

```{r, warning=FALSE}

# --- Spatial blocking function ---
generate_spatial_bootstrap_block <- function(sf_data,
                                             n_boot = 50,
                                             train_frac = 0.7,
                                             theRange = 50,     # in meters; adjust based on variogram range
                                             n_folds = 5,       # number of spatial blocks
                                             seed = 123) {
  set.seed(seed)

  # Run blockCV spatial blocking to assign folds based on variogram range
  sb <- spatialBlock(speciesData = sf_data,
                     theRange = theRange,       # block size = variogram range
                     k = n_folds,               # number of folds (blocks)
                     selection = "random",      # random assignment of folds
                     showBlocks = FALSE,
                     progress = FALSE)

  # Extract fold assignments
  sf_data$fold <- sb$foldID

  # --- Build list of bootstrapped train/test splits ---
  boot_list <- vector("list", n_boot)
  for (i in seq_len(n_boot)) {
    # randomly pick some folds for training (e.g., 70%)
    n_train_folds <- ceiling(train_frac * n_folds)
    train_folds <- sample(unique(sf_data$fold), n_train_folds)

    df <- sf_data %>%
      mutate(set = ifelse(fold %in% train_folds, "train", "test"))

    boot_list[[i]] <- df
  }

  return(boot_list)
}


# --- Safe R² function ---
safe_cor2 <- function(x, y){
  if(length(x)==0 || length(y)==0) return(NA)
  if(all(is.na(x)) || all(is.na(y))) return(NA)
  cor(x, y, use="complete.obs")^2
}
  #uses correlation with complete.obs to avoid NA issues... but I am not sure if this is the best way to handle it.

# --- Generate bootstraps ---
# Estimate an average variogram range first if you have it
# e.g., from your variogram fits:
# mean_range <- mean(fit_variograms$range, na.rm = TRUE)
# Otherwise start with 50 m as a field-scale approximation

boot_SP <- generate_spatial_bootstrap_block(soil_sf_SP,
                                            n_boot = 50,
                                            train_frac = 0.7,
                                            theRange = 50,
                                            n_folds = 5)

boot_C  <- generate_spatial_bootstrap_block(soil_sf_C,
                                            n_boot = 50,
                                            train_frac = 0.7,
                                            theRange = 50,
                                            n_folds = 5)


# --- Random Forest function ---
run_RF <- function(boot_list, covariates, soil_vars){
  results <- list()
  for(var in soil_vars){
    boot_results <- map(boot_list, function(df){
      train_df <- st_drop_geometry(df %>% filter(set=="train")) %>% select(all_of(c(covariates,var))) %>% na.omit()
      test_df  <- st_drop_geometry(df %>% filter(set=="test")) %>% select(all_of(c(covariates,var))) %>% na.omit()
      if(nrow(train_df) < 5 | nrow(test_df) < 5) return(list(RMSE=NA,R2=NA))
      rf_model <- randomForest(x = train_df[,covariates], y = train_df[[var]], ntree = 500)
      pred <- predict(rf_model, newdata = test_df[,covariates])
      list(model = rf_model, pred = pred,
           RMSE = sqrt(mean((pred - test_df[[var]])^2)),
           R2 = safe_cor2(pred, test_df[[var]]))
    })
    results[[var]] <- boot_results
  }
  return(results)
}

# --- Ordinary Kriging function with fallback ---
# --- Ordinary Kriging function with fallback (always returns R² even if only nugget) ---
run_OK_safe <- function(boot_list, soil_vars) {
  results <- list()

  for (var in soil_vars) {
    boot_results <- map(boot_list, function(df) {
      train_sf <- df %>% filter(set == "train")
      test_sf  <- df %>% filter(set == "test")

      train_vals <- train_sf[[var]]
      test_vals  <- test_sf[[var]]

      # --- skip if too few usable data points ---
      if (all(is.na(train_vals)) || length(unique(train_vals)) <= 1) {
        pred <- rep(mean(train_vals, na.rm = TRUE), nrow(test_sf))
        rmse <- sqrt(mean((pred - test_vals)^2, na.rm = TRUE))
        r2   <- ifelse(var(pred, na.rm = TRUE) > 0 && var(test_vals, na.rm = TRUE) > 0,
                       suppressWarnings(cor(pred, test_vals, use = "complete.obs")^2),
                       0)
        return(list(pred = pred, RMSE = rmse, R2 = r2))
      }

      # --- variogram and model fit (fallback to nugget if needed) ---
      vgm_model <- tryCatch({
        v_emp <- variogram(as.formula(paste(var, "~ 1")), train_sf)
        fit.variogram(v_emp, vgm(psill = var(train_vals, na.rm = TRUE),
                                 model = "Exp",
                                 nugget = 0.1,
                                 range = max(v_emp$dist, na.rm = TRUE)/3))
      }, error = function(e) NULL)

      if (is.null(vgm_model)) {
        vgm_model <- vgm(psill = var(train_vals, na.rm = TRUE),
                         model = "Nug", nugget = 0.1)
      }

      # --- try kriging prediction ---
      ok_pred <- tryCatch({
        krige(as.formula(paste(var, "~ 1")), train_sf, test_sf, model = vgm_model)
      }, error = function(e) NULL)

      # --- fallback if krige fails ---
      pred <- if (!is.null(ok_pred) && "var1.pred" %in% names(ok_pred)) {
        ok_pred$var1.pred
      } else {
        rep(mean(train_vals, na.rm = TRUE), nrow(test_sf))
      }

      # --- safe metrics (handles no complete pairs) ---
      rmse <- sqrt(mean((pred - test_vals)^2, na.rm = TRUE))
      r2 <- tryCatch({
        if (var(pred, na.rm = TRUE) == 0 || var(test_vals, na.rm = TRUE) == 0) 0
        else cor(pred, test_vals, use = "complete.obs")^2
      }, error = function(e) 0)

      list(pred = pred, RMSE = rmse, R2 = r2)
    })
    results[[var]] <- boot_results
  }
  return(results)
}


# --- Regression Kriging function with fallback ---
run_RK_safe <- function(boot_list, covariates, soil_vars){
  results <- list()
  for(var in soil_vars){
    boot_results <- map(boot_list, function(df){
      train_sf <- df %>% filter(set=="train")
      test_sf  <- df %>% filter(set=="test")
      
      # Remove rows with NA in predictors or target
      train_df <- st_drop_geometry(train_sf) %>% select(all_of(c(covariates,var))) %>% na.omit()
      test_df  <- st_drop_geometry(test_sf) %>% select(all_of(c(covariates,var))) %>% na.omit()
      
      # Skip if too few points
      if(nrow(train_df)<5 | nrow(test_df)<5) return(list(RMSE=NA,R2=NA))
      
      # Match sf objects to filtered rows
      train_sf <- train_sf[rownames(train_df), ]
      test_sf  <- test_sf[rownames(test_df), ]
      
      # Trend model (RF)
      rf_model <- randomForest(x=train_df[,covariates], y=train_df[[var]], ntree=500)
      trend_train <- predict(rf_model, newdata=train_df[,covariates])
      trend_test  <- predict(rf_model, newdata=test_df[,covariates])
      
      # Residuals
      residuals_train <- train_df[[var]] - trend_train
      
      # Make a new sf object with only the filtered residuals
      train_res_sf <- train_sf %>% mutate(residuals=residuals_train)
      
      # Variogram for residuals
      vgm_model <- tryCatch(
        variogram(residuals~1, train_res_sf) %>%
          fit.variogram(vgm(psill=var(residuals_train, na.rm=TRUE), model="Exp", nugget=0.1, range=100)),
        error=function(e) NULL
      )
      if(is.null(vgm_model)) vgm_model <- vgm(psill=var(residuals_train, na.rm=TRUE), model="Exp", nugget=0.1, range=100)
      
      # Krige residuals
      rk_pred <- tryCatch(
        krige(residuals~1, train_res_sf, test_sf, model=vgm_model),
        error=function(e) NULL
      )
      # fallback if krige fails or lengths mismatch
      if(is.null(rk_pred) || length(rk_pred$var1.pred)!=nrow(test_sf)){
        rk_pred <- data.frame(var1.pred=rep(0,nrow(test_sf)))
      }
      
      # Combine trend + residuals
      final_pred <- trend_test + rk_pred$var1.pred
      rmse <- sqrt(mean((final_pred - test_df[[var]])^2, na.rm=TRUE))
      r2   <- safe_cor2(final_pred, test_df[[var]])
      
      list(RMSE=rmse, R2=r2)
    })
    results[[var]] <- boot_results
  }
  return(results)
}

# --- Run all methods ---
RF_results_SP <- run_RF(boot_SP, covariates, soil_vars)
RF_results_C  <- run_RF(boot_C, covariates, soil_vars)
OK_results_SP <- run_OK_safe(boot_SP, soil_vars)
OK_results_C  <- run_OK_safe(boot_C, soil_vars)
RK_results_SP <- run_RK_safe(boot_SP, covariates, soil_vars)
RK_results_C  <- run_RK_safe(boot_C, covariates, soil_vars)

# --- Function to extract metrics ---
extract_metrics <- function(results_list, dataset_name){
  map_dfr(names(results_list), function(var){
    res <- results_list[[var]]
    tibble(variable=var, RMSE=map_dbl(res,"RMSE"), R2=map_dbl(res,"R2"))
  }) %>% mutate(dataset=dataset_name)
}

rf_metrics_all <- bind_rows(extract_metrics(RF_results_SP,"Single Point"),
                            extract_metrics(RF_results_C,"Composite"))
ok_metrics_all <- bind_rows(extract_metrics(OK_results_SP,"Single Point"),
                            extract_metrics(OK_results_C,"Composite"))
rk_metrics_all <- bind_rows(extract_metrics(RK_results_SP,"Single Point"),
                            extract_metrics(RK_results_C,"Composite"))

# --- Combine all for plotting ---
combined_metrics <- bind_rows(rf_metrics_all %>% mutate(method="RF"),
                              ok_metrics_all %>% mutate(method="OK"),
                              rk_metrics_all %>% mutate(method="RK")) %>%
  pivot_longer(cols=c(RMSE,R2), names_to="metric", values_to="value")

# --- Plot ---

ggplot(combined_metrics, aes(x=variable, y=value, fill=dataset)) +
  geom_violin(trim=FALSE, alpha=0.6) +
  geom_boxplot(width=0.1, position=position_dodge(width=0.9), outlier.shape=NA) +
  theme_minimal(base_size=13) +
  labs(title="Distribution of RMSE and R² across nutrients - All Methods",
       x="Nutrient", y="Value") +
  scale_fill_brewer(palette="Set2") +
  facet_grid(metric~method, scales="free_y")

```

```{r, warning=FALSE}
#better plot: Just R2 for OK RF and RK. Remove LBC1 and LBCeq
#rename variables for better readability
filtered_metrics <- combined_metrics %>%
  filter(metric=="R2", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(variable = recode(variable,
                           "p_h_2" = "pH",
                           "ca" = "Ca",
                           "k" = "K",
                           "mg" = "Mg",
                           "mn" = "Mn",
                           "p" = "P",
                           "zn" = "Zn",
                           "c" = "C",
                           "n" = "N"))

# Separate RMSE and R2 for plotting
filtered_metrics_RMSE <- combined_metrics %>%
  filter(metric == "RMSE", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(
    variable = recode(variable, "p_h_2"="pH","ca"="Ca","k"="K","mg"="Mg","mn"="Mn","p"="P","zn"="Zn","c"="C","n"="N")
  )

filtered_metrics_R2 <- combined_metrics %>%
  filter(metric == "R2", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(
    variable = recode(variable, "p_h_2"="pH","ca"="Ca","k"="K","mg"="Mg","mn"="Mn","p"="P","zn"="Zn","c"="C","n"="N")
  )
# Consistent colors for methods
method_colors <- c(RF = "#1b9e77", OK = "#d95f02", RK = "#7570b3")

# Factor levels to control ordering
filtered_metrics_RMSE <- filtered_metrics_RMSE %>%
  mutate(
    method  = factor(method, levels = c("RF","OK","RK")),
    dataset = factor(dataset, levels = c("Single Point","Composite"))
  )

# If you don't have `filtered_metrics_R2` yet, build it from your combined_metrics
filtered_metrics_R2 <- combined_metrics %>%
  filter(metric == "R2", !variable %in% c("lbc_1","lb_ceq")) %>%
  mutate(
    variable = recode(variable, "p_h_2"="pH","ca"="Ca","k"="K","mg"="Mg","mn"="Mn","p"="P","zn"="Zn","c"="C","n"="N"),
    method   = factor(method, levels = c("RF","OK","RK")),
    dataset  = factor(dataset, levels = c("Single Point","Composite"))
  )

# RMSE plot
p_rmse <- ggplot(filtered_metrics_RMSE, aes(x = dataset, y = value)) +
  geom_violin(aes(fill = method),
              position = position_dodge(width = 0.85),
              width = 0.8, trim = FALSE, alpha = 0.6) +
  # Box outlines fully colored by method (fill = NA so outlines/whiskers show)
  geom_boxplot(aes(color = method, group = interaction(dataset, method)),
               position = position_dodge(width = 0.85),
               width = 0.14, outlier.shape = NA, fill = NA, size = 0.7) +
  # Mean marker
  stat_summary(aes(color = method),
               fun = mean, geom = "point",
               position = position_dodge(width = 0.85),
               size = 2.5, shape = 18) +
  facet_wrap(~ variable, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = method_colors, name = "Method") +
  scale_color_manual(values = method_colors, name = "Method") +
  labs(title = "RMSE by nutrient, method, and dataset",
       x = "Dataset", y = "RMSE") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
print(p_rmse)

# R2 plot
p_r2 <- ggplot(filtered_metrics_R2, aes(x = dataset, y = value)) +
  geom_violin(aes(fill = method),
              position = position_dodge(width = 0.85),
              width = 0.8, trim = FALSE, alpha = 0.6) +
  geom_boxplot(aes(color = method, group = interaction(dataset, method)),
               position = position_dodge(width = 0.85),
               width = 0.14, outlier.shape = NA, fill = NA, size = 0.7) +
  stat_summary(aes(color = method),
               fun = mean, geom = "point",
               position = position_dodge(width = 0.85),
               size = 2.5, shape = 18) +
  facet_wrap(~ variable, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = method_colors, name = "Method") +
  scale_color_manual(values = method_colors, name = "Method") +
  labs(title = "R² by nutrient, method, and dataset",
       x = "Dataset", y = "R²") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))



make_panel <- function(df_one_metric, nut_label, ylab = "RMSE/R²") {
  ggplot(df_one_metric, aes(x = dataset, y = value)) +
    geom_violin(aes(fill = method),
                position = position_dodge(width = 0.85),
                width = 0.8, trim = FALSE, alpha = 0.6) +
    geom_boxplot(aes(color = method, group = interaction(dataset, method)),
                 position = position_dodge(width = 0.85),
                 width = 0.14, outlier.shape = NA, fill = NA, size = 0.7) +
    stat_summary(aes(color = method),
                 fun = mean, geom = "point",
                 position = position_dodge(width = 0.85),
                 size = 2.5, shape = 18) +
    scale_fill_manual(values = method_colors, name = "Method") +
    scale_color_manual(values = method_colors, name = "Method") +
    labs(title = nut_label, x = "Dataset", y = ylab) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
}

# Build per-nutrient panels for RMSE
vars_rmse <- levels(factor(filtered_metrics_RMSE$variable))
panels_rmse <- lapply(vars_rmse, function(v) {
  dfv <- filtered_metrics_RMSE %>% filter(variable == v)
  make_panel(dfv, v, ylab = "RMSE")
})

# If last row has a single panel (e.g., total facets %% 3 == 1), center it by adding spacers
ncol <- 3
rem <- length(panels_rmse) %% ncol
if (rem == 1) {
  # add one spacer before and one after the last panel to center it
  panels_rmse <- append(panels_rmse, list(plot_spacer()), after = length(panels_rmse) - 1)
  panels_rmse <- append(panels_rmse, list(plot_spacer()))
} else if (rem == 2) {
  # add one spacer after to visually center two panels (optional)
  panels_rmse <- append(panels_rmse, list(plot_spacer()))
}

p_rmse_centered <- wrap_plots(panels_rmse, ncol = ncol, guides = "collect") &
  theme(legend.position = "top")

# If you want a shared legend, add it manually from one base plot; for brevity we keep legends off here.

# Build per-nutrient panels for R2
vars_r2 <- levels(factor(filtered_metrics_R2$variable))
panels_r2 <- lapply(vars_r2, function(v) {
  dfv <- filtered_metrics_R2 %>% filter(variable == v)
  make_panel(dfv, v, ylab = "R²")
})
rem2 <- length(panels_r2) %% ncol
if (rem2 == 1) {
  panels_r2 <- append(panels_r2, list(plot_spacer()), after = length(panels_r2) - 1)
  panels_r2 <- append(panels_r2, list(plot_spacer()))
} else if (rem2 == 2) {
  panels_r2 <- append(panels_r2, list(plot_spacer()))
}
p_r2_centered <- wrap_plots(panels_r2, ncol = ncol, guides = "collect") &
  theme(legend.position = "top")
p_r2_centered

```

### Metric Extraction Helpers

```{r, warning=FALSE}

safe_cor2 <- function(x,y){
  if(length(x)==0 || length(y)==0) return(NA_real_)
  if(all(is.na(x)) || all(is.na(y))) return(NA_real_)
  cc <- suppressWarnings(cor(x,y,use="complete.obs"))
  if(is.na(cc)) NA_real_ else cc^2
}

# helper to rebuild test observed vector in same order as how predictions were produced in your functions
# df_boot: an sf tibble that has column 'set' with "train"/"test" (like your generate_spatial_bootstrap output)
# var: target variable name
# returns: test_df (data.frame) after the na.omit used in your run functions, plus row_ids to map back to original rows
build_test_df_from_boot <- function(df_boot, var, covariates){
  # df_boot is sf with geometry; use st_drop_geometry in same manner as your run functions
  df <- sf::st_drop_geometry(df_boot)
  test_df_full <- df %>% filter(set == "test") %>% mutate(.orig_row = row_number())
  # select covariates and var and remove rows with NA similarly to run_RF/run_RK
  sel_cols <- c(covariates, var)
  test_sel <- test_df_full %>% select(all_of(sel_cols), .orig_row)
  # na.omit - keep track of original row indices
  complete_idx <- stats::complete.cases(test_sel)
  test_sel2 <- test_sel[complete_idx, , drop = FALSE]
  # test_sel2 has covariates, var and .orig_row
  return(list(test_df = test_sel2, omitted = !complete_idx))
}

# generic function to compute metrics from prediction and observed vectors
compute_metrics_vecs <- function(obs, pred){
  ok <- !is.na(obs) & !is.na(pred)
  n <- sum(ok)
  if(n==0) return(tibble(n = 0, RMSE = NA_real_, MAE = NA_real_, Bias = NA_real_, R2_cor = NA_real_))
  err <- pred[ok] - obs[ok]
  tibble(
    n = n,
    RMSE = sqrt(mean(err^2)),
    MAE = mean(abs(err)),
    Bias = mean(err),
    R2_cor = safe_cor2(obs[ok], pred[ok])
  )
}

# main extractor that returns a list: predictions_long (tibble), per_boot_metrics (tibble)
extract_metrics_for_method <- function(results_list, boot_list, method_name, covariates, soil_vars){
  preds_long <- list()
  per_boot_rows <- list()
  n_nut <- length(soil_vars)
  for(nut in soil_vars){
    res_boots <- results_list[[nut]]
    # iterate boots; results_list[[nut]] should be a list of length n_boot
    for(bi in seq_along(res_boots)){
      res <- res_boots[[bi]]
      boot_sf <- boot_list[[bi]]
      # build test_df in same logic as your functions
      test_build <- build_test_df_from_boot(boot_sf, nut, covariates)
      test_df <- test_build$test_df
      # try to extract pred vector from res (narrow cases)
      pred_vec <- NULL
      obs_vec <- NULL
      # If res already stored pred and maybe observed, prefer them
      if(!is.null(res$pred) && !is.null(res$observed)){
        pred_vec <- res$pred
        obs_vec  <- res$observed
      } else if(!is.null(res$pred) && is.data.frame(test_df) && nrow(test_df) == length(res$pred)){
        # assume ordering matches the test_df after na.omit (this is how your run_RF created pred)
        pred_vec <- res$pred
        obs_vec  <- test_df[[nut]]
      } else if(!is.null(res$model) && inherits(res$model, "randomForest") && is.data.frame(test_df)){
        # RF stored model object; predict fresh on test covariates (should match original pred)
        # ensure columns in test_df align
        predictor_matrix <- test_df[, covariates, drop = FALSE]
        pred_vec <- tryCatch(predict(res$model, newdata = predictor_matrix), error = function(e) NULL)
        obs_vec  <- test_df[[nut]]
      } else {
        # As a fallback: attempt to recompute prediction using simple approach depending on method
        if(method_name == "RF"){
          # try to fit a quick RF on training partition of this boot and predict
          train_df <- sf::st_drop_geometry(boot_sf) %>% filter(set=="train") %>% select(all_of(c(covariates, nut))) %>% na.omit()
          test_df2  <- test_df
          if(nrow(train_df) >= 3 && nrow(test_df2) >= 1){
            rf_mod <- tryCatch(randomForest::randomForest(x = train_df[, covariates, drop = FALSE], y = train_df[[nut]], ntree = 200), error = function(e) NULL)
            if(!is.null(rf_mod)){
              pred_vec <- tryCatch(predict(rf_mod, newdata = test_df2[, covariates, drop = FALSE]), error = function(e) NULL)
              obs_vec  <- test_df2[[nut]]
            }
          }
        } else if(method_name == "OK"){
          # run ordinary kriging on this boot
          train_sf <- boot_sf %>% filter(set=="train")
          test_sf  <- boot_sf %>% filter(set=="test")
          # keep only rows with non-NA target
          if(sum(!is.na(train_sf[[nut]])) >= 3 && nrow(test_sf) >= 1){
            # ensure train_sf/test_sf are sf objects with projected CRS (your boot list used sf transformed earlier)
            v_emp <- tryCatch(variogram(as.formula(paste(nut,"~1")), train_sf), error = function(e) NULL)
            vgm_model <- tryCatch(fit.variogram(v_emp, model = vgm(psill = var(train_sf[[nut]], na.rm=TRUE), model = "Exp", nugget = 0.1, range = max(v_emp$dist, na.rm=TRUE)/3)), error = function(e) NULL)
            if(is.null(vgm_model)) vgm_model <- tryCatch(vgm(psill = var(train_sf[[nut]], na.rm=TRUE), model="Exp", nugget=0.1, range=100), error = function(e) NULL)
            ok_pred <- tryCatch(krige(as.formula(paste(nut,"~1")), train_sf, test_sf, model = vgm_model), error = function(e) NULL)
            if(!is.null(ok_pred) && "var1.pred" %in% names(ok_pred)){
              pred_vec <- ok_pred$var1.pred
              # align obs: remove NA rows similarly to how you computed predictions
              # build test_df2 as earlier
              obs_build <- build_test_df_from_boot(boot_sf, nut, covariates)
              obs_vec <- obs_build$test_df[[nut]]
            }
          }
        } else if(method_name == "RK"){
          # recompute RK: RF trend on train + krige residuals
          train_df <- sf::st_drop_geometry(boot_sf) %>% filter(set=="train") %>% select(all_of(c(covariates, nut))) %>% na.omit()
          test_df2  <- test_df
          if(nrow(train_df) >= 5 && nrow(test_df2) >= 1){
            train_sf_filtered <- boot_sf %>% filter(set=="train")
            # align train_sf_filtered to non-NA rows in train_df
            # We assume original row ordering remains consistent enough; best to join by coordinates if needed
            # Fit RF trend
            rf_mod <- tryCatch(randomForest::randomForest(x = train_df[, covariates, drop = FALSE], y = train_df[[nut]], ntree = 200), error = function(e) NULL)
            if(!is.null(rf_mod)){
              trend_train <- predict(rf_mod, newdata = train_df[, covariates, drop = FALSE])
              resid_train <- train_df[[nut]] - trend_train
              # attach residuals to train_sf_filtered: subset to rows matching train_df by coordinates
              # This is a heuristic mapping: select rows in train_sf_filtered with values matching train_df rows
              # Safer approach: if train_df had rownames carrying original row IDs, use that. If not, we proceed best-effort.
              train_sf_filtered2 <- train_sf_filtered
              # assign residuals in order of non-NA rows
              non_na_idx <- which(!is.na(train_sf_filtered2[[nut]]))
              assign_len <- min(length(resid_train), length(non_na_idx))
              if(assign_len > 0) train_sf_filtered2$resid <- NA_real_
              train_sf_filtered2$resid[non_na_idx[1:assign_len]] <- resid_train[1:assign_len]
              # fit variogram and krige residuals
              v_emp_r <- tryCatch(variogram(resid~1, train_sf_filtered2), error = function(e) NULL)
              vgm_model <- tryCatch(fit.variogram(v_emp_r, model = vgm(psill = var(resid_train, na.rm=TRUE), model="Exp", nugget=0.1, range = max(v_emp_r$dist, na.rm=TRUE)/3)), error = function(e) NULL)
              if(is.null(vgm_model)) vgm_model <- tryCatch(vgm(psill = var(resid_train, na.rm=TRUE), model="Exp", nugget=0.1, range=100), error = function(e) NULL)
              test_sf <- boot_sf %>% filter(set=="test")
              kr_res <- tryCatch(krige(resid~1, train_sf_filtered2, test_sf, model = vgm_model), error = function(e) NULL)
              if(!is.null(kr_res) && "var1.pred" %in% names(kr_res)){
                trend_test <- predict(rf_mod, newdata = test_df2[, covariates, drop = FALSE])
                pred_vec <- trend_test + kr_res$var1.pred
                obs_vec <- test_df2[[nut]]
              }
            }
          }
        }
      } # end fallback block

      # Now compute metrics for this boot if pred_vec and obs_vec exist; align lengths
      if(!is.null(pred_vec) && !is.null(obs_vec) && length(pred_vec) == length(obs_vec)){
        metrics_row <- compute_metrics_vecs(obs_vec, pred_vec)
        per_boot_rows[[length(per_boot_rows) + 1]] <- tibble(
          nutrient = nut, method = method_name, boot = bi,
          n_test = metrics_row$n, RMSE = metrics_row$RMSE, MAE = metrics_row$MAE, Bias = metrics_row$Bias, R2_cor = metrics_row$R2_cor
        )
        # store predictions with coordinates by mapping test_df .orig_row back to original geometry from boot_sf
        # test_df came from build_test_df_from_boot and has .orig_row -> position within test partition
        # We will include coordinates from boot_sf for the test rows used
        # Rebuild a matching sf test subset (after na.omit): 
        # Use the same logic as in build_test_df_from_boot to identify which test rows used
        test_sf_full <- boot_sf %>% filter(set == "test")
        # find rows in test_sf_full that correspond to the kept ones: use complete.cases on selected columns
        test_sel_full <- sf::st_drop_geometry(test_sf_full)[, c(covariates, nut), drop = FALSE]
        keep_mask <- stats::complete.cases(test_sel_full)
        if(sum(keep_mask) > 0){
          test_sf_kept <- test_sf_full[keep_mask, ]
          # Now ensure lengths match pred_vec
          if(nrow(test_sf_kept) == length(pred_vec)){
            preds_long[[length(preds_long) + 1]] <- tibble(
              nutrient = nut, method = method_name, boot = bi,
              x = as.numeric(sf::st_coordinates(test_sf_kept)[,1]),
              y = as.numeric(sf::st_coordinates(test_sf_kept)[,2]),
              observed = obs_vec,
              predicted = pred_vec
            )
          } else {
            # fallback: store without coords
            preds_long[[length(preds_long) + 1]] <- tibble(
              nutrient = nut, method = method_name, boot = bi,
              x = NA_real_, y = NA_real_,
              observed = obs_vec,
              predicted = pred_vec
            )
          }
        } else {
          preds_long[[length(preds_long) + 1]] <- tibble(
            nutrient = nut, method = method_name, boot = bi,
            x = NA_real_, y = NA_real_,
            observed = obs_vec,
            predicted = pred_vec
          )
        }
      } else {
        # record NA metrics if we couldn't get predictions
        per_boot_rows[[length(per_boot_rows) + 1]] <- tibble(
          nutrient = nut, method = method_name, boot = bi, n_test = 0,
          RMSE = NA_real_, MAE = NA_real_, Bias = NA_real_, R2_cor = NA_real_
        )
      }

    } # end boots loop
  } # end nutrients loop

  per_boot_metrics <- bind_rows(per_boot_rows)
  predictions_long <- bind_rows(preds_long)
  list(predictions_long = predictions_long, per_boot_metrics = per_boot_metrics)
}
#SP:
# Run extraction for Single Point boots (boot_SP) and Composite boots (boot_C)
message("Extracting metrics for Single Point (SP) boots...")
rf_SP_out <- extract_metrics_for_method(RF_results_SP, boot_SP, "RF", covariates, soil_vars)
ok_SP_out <- extract_metrics_for_method(OK_results_SP, boot_SP, "OK", covariates, soil_vars)
rk_SP_out <- extract_metrics_for_method(RK_results_SP, boot_SP, "RK", covariates, soil_vars)

# Combine SP predictions and metrics
predictions_long_SP <- bind_rows(rf_SP_out$predictions_long, ok_SP_out$predictions_long, rk_SP_out$predictions_long)
per_boot_metrics_SP <- bind_rows(rf_SP_out$per_boot_metrics, ok_SP_out$per_boot_metrics, rk_SP_out$per_boot_metrics)

# compute pooled metrics per nutrient + method from predictions_long_SP
pooled_metrics_SP <- predictions_long_SP %>%
  group_by(nutrient, method) %>%
  summarize(
    n_total = sum(!is.na(observed) & !is.na(predicted)),
    pooled_RMSE = ifelse(n_total==0, NA_real_, sqrt(mean((observed - predicted)^2, na.rm=TRUE))),
    pooled_MAE  = ifelse(n_total==0, NA_real_, mean(abs(observed - predicted), na.rm=TRUE)),
    pooled_Bias = ifelse(n_total==0, NA_real_, mean(predicted - observed, na.rm=TRUE)),
    pooled_R2_cor = ifelse(n_total <= 1, NA_real_, {cc <- suppressWarnings(cor(observed, predicted, use="complete.obs")); ifelse(is.na(cc), NA_real_, cc^2)}),
    .groups = "drop"
  )

# Save results
dir.create("../ouputs/spatial_cv_results", showWarnings = FALSE)
saveRDS(predictions_long_SP, file = "spatial_cv_results/predictions_long_SP.rds")
saveRDS(per_boot_metrics_SP, file = "spatial_cv_results/per_boot_metrics_SP.rds")
saveRDS(pooled_metrics_SP, file = "spatial_cv_results/pooled_metrics_SP.rds")

# Print short summaries
message("Per-boot metrics (first rows):")
print(head(per_boot_metrics_SP))
message("Pooled metrics (first rows):")
print(head(pooled_metrics_SP))

# Return invisibly
invisible(list(
  predictions_long_SP = predictions_long_SP,
  per_boot_metrics_SP = per_boot_metrics_SP,
  pooled_metrics_SP = pooled_metrics_SP
))

#C:
message("Extracting metrics for Composite (C) boots...")
rf_C_out <- extract_metrics_for_method(RF_results_C, boot_C, "RF", covariates, soil_vars)
ok_C_out <- extract_metrics_for_method(OK_results_C, boot_C, "OK", covariates, soil_vars)
rk_C_out <- extract_metrics_for_method(RK_results_C, boot_C, "RK", covariates, soil_vars)
# Combine C predictions and metrics
predictions_long_C <- bind_rows(rf_C_out$predictions_long, ok_C_out$predictions_long, rk_C_out$predictions_long)   
per_boot_metrics_C <- bind_rows(rf_C_out$per_boot_metrics, ok_C_out$per_boot_metrics, rk_C_out$per_boot_metrics)
# compute pooled metrics per nutrient + method from predictions_long_C
pooled_metrics_C <- predictions_long_C %>%
  group_by(nutrient, method) %>%
  summarize(
    n_total = sum(!is.na(observed) & !is.na(predicted)),
    pooled_RMSE = ifelse(n_total==0, NA_real_, sqrt(mean((observed - predicted)^2, na.rm=TRUE))),
    pooled_MAE  = ifelse(n_total==0, NA_real_, mean(abs(observed - predicted), na.rm=TRUE)),
    pooled_Bias = ifelse(n_total==0, NA_real_, mean(predicted - observed, na.rm=TRUE)),
    pooled_R2_cor = ifelse(n_total <= 1, NA_real_, {cc <- suppressWarnings(cor(observed, predicted, use="complete.obs")); ifelse(is.na(cc), NA_real_, cc^2)}),
    .groups = "drop"
  )
# Save results                                    

saveRDS(predictions_long_C, file = "spatial_cv_results/predictions_long_C.rds")
saveRDS(per_boot_metrics_C, file = "spatial_cv_results/per_boot_metrics_C.rds")
saveRDS(pooled_metrics_C, file = "spatial_cv_results/pooled_metrics_C.rds")
# Print short summaries
message("Per-boot metrics (first rows):")
print(head(per_boot_metrics_C))                
message("Pooled metrics (first rows):")
print(head(pooled_metrics_C))
# Return invisibly          
invisible(list(
  predictions_long_C = predictions_long_C,
  per_boot_metrics_C = per_boot_metrics_C,
  pooled_metrics_C = pooled_metrics_C
))           
                                       
```

### Statistical Testing Between Methods

```{r, warning=FALSE}
prepare_paired_data <- function(results_list, dataset_name, method_name){
  map_dfr(names(results_list), function(var){
    res <- results_list[[var]]
    tibble(variable=var, RMSE=map_dbl(res,"RMSE"), R2=map_dbl(res,"R2")) %>%
      mutate(dataset=dataset_name, method=method_name)
  })
}
rf_paired <- bind_rows(prepare_paired_data(RF_results_SP,"Single Point","RF"),
                       prepare_paired_data(RF_results_C,"Composite","RF"))
ok_paired <- bind_rows(prepare_paired_data(OK_results_SP,"Single Point","OK"),
                       prepare_paired_data(OK_results_C,"Composite","OK"))
rk_paired <- bind_rows(prepare_paired_data(RK_results_SP,"Single Point","RK"),
                       prepare_paired_data(RK_results_C,"Composite","RK"))

all_paired <- bind_rows(rf_paired, ok_paired, rk_paired)

# Paired tests between methods for each variable and dataset
test_results <- map_dfr(unique(all_paired$variable), function(var){
  map_dfr(unique(all_paired$dataset), function(ds){
    subset_data <- all_paired %>% filter(variable==var, dataset==ds)
    rf_data <- subset_data %>% filter(method=="RF")
    ok_data <- subset_data %>% filter(method=="OK")
    rk_data <- subset_data %>% filter(method=="RK")
    # RF vs OK
    rf_ok_test <- wilcox.test(rf_data$R2, ok_data$R2, paired=TRUE, alternative="two.sided")
    # RF vs RK
    rf_rk_test <- wilcox.test(rf_data$R2, rk_data$R2, paired=TRUE, alternative="two.sided")
    # OK vs RK
    ok_rk_test <- wilcox.test(ok_data$R2, rk_data$R2, paired=TRUE, alternative="two.sided")
    tibble(variable=var, dataset=ds,
           RF_vs_OK_p=rf_ok_test$p.value,
           RF_vs_RK_p=rf_rk_test$p.value,
           OK_vs_RK_p=ok_rk_test$p.value)
  })
})
#create formatted results with significance <0.05 marked with *
formatted_results <- test_results %>%
  #sort by lowest p-value in RF_vs_OK
  arrange(RF_vs_OK_p) %>%
  mutate(RF_vs_OK = ifelse(RF_vs_OK_p < 0.05, paste0(round(RF_vs_OK_p,4), "*"), round(RF_vs_OK_p,4)),
         RF_vs_RK = ifelse(RF_vs_RK_p < 0.05, paste0(round(RF_vs_RK_p,4), "*"), round(RF_vs_RK_p,4)),
         OK_vs_RK = ifelse(OK_vs_RK_p < 0.05, paste0(round(OK_vs_RK_p,4), "*"), round(OK_vs_RK_p,4))) %>%
  select(variable, dataset, RF_vs_OK, RF_vs_RK, OK_vs_RK)
formatted_results
```
### Covariate Name Matching

```{r}


# ------------------------------------------------------------
# 1) Define soil vars we do NOT want as predictors
# ------------------------------------------------------------
soil_vars_all <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")

# ------------------------------------------------------------
# 2) Build normalized-name matching between points and raster
# ------------------------------------------------------------
norm_name <- function(x){
  x %>%
    tolower() %>%
    str_replace_all("[^a-z0-9]+", "_") %>%  # collapse punctuation/spaces to _
    str_replace_all("_+", "_") %>%          # remove repeats
    str_replace_all("^_|_$", "")            # trim ends
}

stack_names <- names(covariate_stack)
sp_names <- names(st_drop_geometry(soil_sf_SP))

# candidates on points side = everything except soil vars + geometry-ish columns
sp_candidates <- setdiff(sp_names, c(soil_vars_all, "geometry", "set", "fold"))

# make lookup tables
stack_tbl <- tibble(stack = stack_names, stack_norm = norm_name(stack_names))
sp_tbl    <- tibble(sp = sp_candidates, sp_norm = norm_name(sp_candidates))

# join by normalized name
map_tbl <- sp_tbl %>%
  left_join(stack_tbl, by = c("sp_norm" = "stack_norm"))

# which point cols did not match any raster layer?
unmatched_sp <- map_tbl %>% filter(is.na(stack)) %>% pull(sp)

# which raster layers were not matched by any point col?
matched_stack <- na.omit(map_tbl$stack) %>% unique()
unmatched_stack <- setdiff(stack_names, matched_stack)

cat("Point covariate cols:", length(sp_candidates), "\n")
cat("Raster layers:", length(stack_names), "\n")
cat("Matched:", length(matched_stack), "\n")
cat("UNmatched point cols:", length(unmatched_sp), "\n")
cat("UNmatched raster layers:", length(unmatched_stack), "\n")

if (length(unmatched_sp) > 0) {
  cat("\n--- Unmatched point columns (first 30) ---\n")
  print(head(unmatched_sp, 30))
}
if (length(unmatched_stack) > 0) {
  cat("\n--- Unmatched raster layers (first 30) ---\n")
  print(head(unmatched_stack, 30))
}

# ------------------------------------------------------------
# 3) Actually rename point columns to match raster stack names
#    Only for those with a match, and avoid collisions.
# ------------------------------------------------------------
rename_vec <- map_tbl %>%
  filter(!is.na(stack)) %>%
  distinct(sp, stack) %>%
  # if multiple sp map to same stack (rare), keep first
  group_by(stack) %>% slice(1) %>% ungroup() %>%
  { setNames(.$stack, .$sp) }   # old -> new mapping (dplyr::rename expects new=old, so we’ll do base rename)

# base R rename for data.frames: names(df)[match(old, names(df))] <- new
apply_rename <- function(sf_obj, rename_vec){
  nm <- names(sf_obj)
  olds <- names(rename_vec)
  idx <- match(olds, nm)
  hit <- !is.na(idx)
  nm[idx[hit]] <- rename_vec[hit]
  names(sf_obj) <- nm
  sf_obj
}

soil_sf_SP <- apply_rename(soil_sf_SP, rename_vec)
soil_sf_C  <- apply_rename(soil_sf_C,  rename_vec)

# ------------------------------------------------------------
# 4) Now you can use ALL raster layers as covariates
#    (except soil variables if they exist as layers)
# ------------------------------------------------------------
covariates <- setdiff(names(covariate_stack), soil_vars_all)

# Final sanity
missing_now_SP <- setdiff(covariates, names(st_drop_geometry(soil_sf_SP)))
missing_now_C  <- setdiff(covariates, names(st_drop_geometry(soil_sf_C)))

cat("\nAfter renaming:\n")
cat("Missing in SP:", length(missing_now_SP), "\n")
cat("Missing in C :", length(missing_now_C), "\n")

if (length(missing_now_SP) > 0) {
  cat("\n--- Still missing in SP (first 30) ---\n")
  print(head(missing_now_SP, 30))
}

# Optional: reorder stack to covariates
covariate_stack <- covariate_stack[[covariates]]

```
### Final Map Prediction Using Best Boot Models

```{r}

# -----------------------------
# 0) Choose which soil vars to map
# -----------------------------
soil_vars_to_map <- c("ca","k","mg","mn","p","zn","c","n","p_h_2")  # edit if you want
# If you want to exclude pH from log stuff later, fine—here it maps same as others.

# -----------------------------
# 1) Combine per-boot metrics and pick "best boot" per nutrient/method/dataset
#    (high R2, low RMSE)
# -----------------------------
per_boot_metrics_all <- bind_rows(
  per_boot_metrics_SP %>% mutate(dataset = "SP"),
  per_boot_metrics_C  %>% mutate(dataset = "C")
) %>%
  rename(R2 = R2_cor) %>%
  filter(is.finite(RMSE), is.finite(R2))

# score within each nutrient+method+dataset so scale differences don't dominate
best_boot_tbl <- per_boot_metrics_all %>%
  filter(nutrient %in% soil_vars_to_map, method %in% c("RF","RK","OK")) %>%
  group_by(dataset, method, nutrient) %>%
  mutate(
    zR2   = as.numeric(scale(R2)),
    zRMSE = as.numeric(scale(RMSE)),
    score = zR2 - zRMSE
  ) %>%
  slice_max(order_by = score, n = 1, with_ties = FALSE) %>%
  ungroup()

print(best_boot_tbl)

# -----------------------------
# 2) Helpers: fit+predict FINAL maps (train on all points)
#    NOTE: covariates must match raster layer names.
# -----------------------------

# Build covariates from raster stack names (most robust)
soil_vars_all <- c("lbc_1","lb_ceq","p_h_2","ca","k","mg","mn","p","zn","c","n")
covariates <- setdiff(names(covariate_stack), soil_vars_all)

# Ensure point tables have these covariates
stopifnot(all(covariates %in% names(st_drop_geometry(soil_sf_SP))))
stopifnot(all(covariates %in% names(st_drop_geometry(soil_sf_C))))

# RF final map
predict_RF_final <- function(soil_sf, covariate_stack, covariates, nutrient, ntree = 500) {
  df <- st_drop_geometry(soil_sf) %>%
    select(all_of(c(covariates, nutrient))) %>%
    na.omit()

  if (nrow(df) < 10) stop("Too few training rows for RF: ", nutrient)

  rf_mod <- randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nutrient]],
    ntree = ntree
  )

  # Safe predict (no terra::predict): use as.data.frame + complete.cases
  pred_df <- terra::as.data.frame(covariate_stack[[covariates]], cells = TRUE, na.rm = FALSE)
  X <- pred_df[, covariates, drop = FALSE]
  keep <- complete.cases(X)

  pred <- rep(NA_real_, nrow(pred_df))
  pred[keep] <- as.numeric(predict(rf_mod, newdata = as.data.frame(X[keep, , drop = FALSE])))

  out <- covariate_stack[[1]]
  vals <- rep(NA_real_, terra::ncell(out))
  vals[pred_df$cell] <- pred
  terra::values(out) <- vals
  out
}

# OK final map
predict_OK_final <- function(soil_sf, template_raster, nutrient) {
  train_sf <- soil_sf
  vals <- train_sf[[nutrient]]
  if (sum(!is.na(vals)) < 10 || length(unique(na.omit(vals))) < 2) stop("Too few points for OK: ", nutrient)

  v_emp <- variogram(as.formula(paste(nutrient, "~ 1")), train_sf)
  v_fit <- tryCatch(
    fit.variogram(v_emp, vgm(
      psill  = var(vals, na.rm = TRUE),
      model  = "Exp",
      nugget = 0.1,
      range  = max(v_emp$dist, na.rm = TRUE)/3
    )),
    error = function(e) vgm(psill = var(vals, na.rm = TRUE), model = "Nug", nugget = 0.1)
  )

  xy <- terra::crds(template_raster, df = TRUE)
  grd_sf <- st_as_sf(xy, coords = c("x","y"), crs = st_crs(train_sf))

  ok <- krige(as.formula(paste(nutrient, "~ 1")), train_sf, grd_sf, model = v_fit)

  out <- template_raster
  values(out) <- ok$var1.pred
  out
}

# RK final map: RF trend + OK residuals
predict_RK_final <- function(soil_sf, covariate_stack, covariates, nutrient, ntree = 500) {
  # Fit RF trend on points
  df <- st_drop_geometry(soil_sf) %>%
    select(all_of(c(covariates, nutrient))) %>%
    na.omit()
  if (nrow(df) < 10) stop("Too few training rows for RK: ", nutrient)

  rf_mod <- randomForest(
    x = df[, covariates, drop = FALSE],
    y = df[[nutrient]],
    ntree = ntree
  )

  trend_pts <- predict(rf_mod, newdata = df[, covariates, drop = FALSE])
  resid <- df[[nutrient]] - trend_pts

  # attach residuals to a matching sf (subset soil_sf rows to df rownames)
  soil_sf2 <- soil_sf[rownames(df), ]
  train_res_sf <- soil_sf2 %>% mutate(resid = resid)

  # variogram for residuals
  v_emp <- variogram(resid ~ 1, train_res_sf)
  v_fit <- tryCatch(
    fit.variogram(v_emp, vgm(
      psill  = var(resid, na.rm = TRUE),
      model  = "Exp",
      nugget = 0.1,
      range  = max(v_emp$dist, na.rm = TRUE)/3
    )),
    error = function(e) vgm(psill = var(resid, na.rm = TRUE), model = "Nug", nugget = 0.1)
  )

  # trend raster (safe predict)
  pred_df <- terra::as.data.frame(covariate_stack[[covariates]], cells = TRUE, na.rm = FALSE)
  X <- pred_df[, covariates, drop = FALSE]
  keep <- complete.cases(X)

  trend_pred <- rep(NA_real_, nrow(pred_df))
  trend_pred[keep] <- as.numeric(predict(rf_mod, newdata = as.data.frame(X[keep, , drop = FALSE])))

  trend_r <- covariate_stack[[1]]
  vals <- rep(NA_real_, terra::ncell(trend_r))
  vals[pred_df$cell] <- trend_pred
  values(trend_r) <- vals

  # residual kriging to raster grid
  xy <- terra::crds(trend_r, df = TRUE)
  grd_sf <- st_as_sf(xy, coords = c("x","y"), crs = st_crs(train_res_sf))
  kr_res <- krige(resid ~ 1, train_res_sf, grd_sf, model = v_fit)

  res_r <- trend_r
  values(res_r) <- kr_res$var1.pred

  trend_r + res_r
}

# -----------------------------
# 3) Run: for each dataset + nutrient, pick the best method(s) and map them
# -----------------------------
out_dir <- "best_boot_maps"
dir.create(out_dir, showWarnings = FALSE)

template_r <- covariate_stack[[1]]

for (ds in c("SP","C")) {

  soil_sf_use <- if (ds == "SP") soil_sf_SP else soil_sf_C

  for (nut in soil_vars_to_map) {

    # best boot rows (one per method)
    bb <- best_boot_tbl %>% filter(dataset == ds, nutrient == nut)

    if (nrow(bb) == 0) next

    # Option: map ALL three methods; or only the top one.
    # Here: map all methods that exist in bb
    for (meth in bb$method) {

      cat("Mapping:", ds, nut, meth, "\n")

      r <- tryCatch({
        if (meth == "RF") predict_RF_final(soil_sf_use, covariate_stack, covariates, nut)
        else if (meth == "RK") predict_RK_final(soil_sf_use, covariate_stack, covariates, nut)
        else predict_OK_final(soil_sf_use, template_r, nut)
      }, error = function(e) {
        message("FAILED: ", ds, " ", nut, " ", meth, " -> ", e$message)
        NULL
      })

      if (!is.null(r)) {
        fn <- file.path(out_dir, sprintf("%s_%s_%s_bestboot.tif", ds, nut, meth))
        terra::writeRaster(r, fn, overwrite = TRUE)
      }
    }
  }
}

cat("Done. Outputs in:", out_dir, "\n")
```
### Example Leaflet Plot of a Best Boot Map

```{r}
#plot

ca_rf <- rast("best_boot_maps/SP_ca_RF_bestboot.tif")

# project to lat/long for leaflet
ca_rf_proj <- project(ca_rf, "EPSG:4326")

pal <- colorNumeric("viridis", values(ca_rf_proj), na.color = "transparent")

leaflet() %>%
  addTiles() %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite")  %>%
  addRasterImage(ca_rf_proj, colors = pal, opacity = 0.8) %>%
  addLegend(pal = pal, values = values(ca_rf_proj), title = "Ca (ppm)", opacity = 1)
  #basemap: satellite bottom layer

```








