---
title: "All_workflow_05_11_2026"
format: html
---

---
title: "Perry_MZ_umap_04_20_2026"
format: html
---

# ============================================================
# PHASE 0 — Libraries
# ============================================================

```{r libraries, message=FALSE, warning=FALSE}
library(terra)
library(tidyverse)
library(janitor)
library(sf)
library(sp)
library(gstat)
library(randomForest)
library(quantregForest)   # QRF prediction intervals
library(CAST)
library(caret)
library(umap)
library(dbscan)
library(beepr)
library(broom)
library(FNN)
library(uwot)
library(patchwork)
library(ggtext)
library(gt)
library(gtExtras)
library(ggrepel)
library(cowplot)
library(cluster)
library(scales)
library(tidytext)
```

# ============================================================
# PHASE 1 — Data Loading & Preparation
# ============================================================

```{r data-load}
# ── Raw datasets ───────────────────────────────────────────
soil_points_nutrients <- read.csv(
  "../LOCAL DATA DO NOT PUSH/soil/perry_soil_points_surface_03_2026.csv"
) %>% clean_names()

ap_val <- read_csv(
  "../LOCAL DATA DO NOT PUSH/Ap_hz_data/hz_0_15_wide_03_30_2026.csv"
)

covariate_stack <- rast(
  "../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m_all_with_SR_indices.tif"
)

# ── Agronomic soil variable subset ────────────────────────

soil_vars <- c("lbc_1", "lb_ceq", "p_h_2", "ca", "k", "mg", "mn", "p", "zn")
cov_names <- names(covariate_stack)

# ── Helper functions ───────────────────────────────────────
add_xy <- function(dat) {
  coords <- st_coordinates(dat)
  dat %>% mutate(X = coords[, 1], Y = coords[, 2])
}

extract_covariates <- function(dat_sf, stack) {
  vect_pts <- terra::vect(dat_sf)
  cov_vals <- terra::extract(stack, vect_pts)
  bind_cols(
    dat_sf %>% st_drop_geometry(),
    cov_vals %>% dplyr::select(-ID)
  ) %>%
    bind_cols(geometry = st_geometry(dat_sf)) %>%
    st_as_sf()
}

# ── SP and COMP modelling datasets ────────────────────────
soil_prop_SP <- soil_points_nutrients %>%
  filter(type == "SPID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

soil_prop_COMP <- soil_points_nutrients %>%
  filter(type == "CID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Validation dataset — HELD OUT, never used in CV ───────
ap_val_sf <- ap_val %>%
  st_as_sf(coords = c("POINT_X", "POINT_Y"), crs = 3857) %>%
  st_transform(26917) %>%
  transmute(
    point_label = paste0("VAL_", Point),
    type        = "VAL",
    lb_ceq = LBCeq_0_15,
    p_h_2  = pH.2_0_15,
    ca     = Ca_0_15,
    k      = K_0_15,
    p      = P_0_15,
     mg     = Mg_0_15,
    mn     = Mn_0_15,
    zn     = Zn_0_15,
    lbc_1  = LBC.1_0_15,
    geometry = geometry
  ) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Spatial distribution check ────────────────────────────
ggplot() +
  geom_sf(data = soil_prop_SP,   color = "steelblue",   size = 2) +
  geom_sf(data = soil_prop_COMP, color = "tomato",      size = 2, shape = 17) +
  geom_sf(data = ap_val_sf,      color = "forestgreen", size = 2, shape = 15) +
  theme_minimal() +
  labs(title = "Spatial distribution: SP (blue), COMP (red), VAL (green)")
```

# ============================================================
# PHASE 2 — Spatial Cross-Validation Setup
# ============================================================

```{r spatial-cv-setup}
make_spatial_folds <- function(sf_data, k = 5, seed = 20260401) {
  coords <- st_coordinates(sf_data)
  set.seed(seed)
  km <- kmeans(coords, centers = k, nstart = 25)
  sf_data$spatial_block <- as.character(km$cluster)

  folds <- CAST::CreateSpacetimeFolds(
    x        = sf_data %>% st_drop_geometry(),
    spacevar = "spatial_block",
    k        = k,
    seed     = seed
  )

  purrr::map(seq_along(folds$index), function(i) {
    list(
      iteration = i,
      seed      = seed,
      train_ids = folds$index[[i]],
      test_ids  = folds$indexOut[[i]]
    )
  })
}

resamples_SP   <- make_spatial_folds(soil_prop_SP,   k = 5)
resamples_COMP <- make_spatial_folds(soil_prop_COMP, k = 5)

make_cutoff <- function(sf_data) max(dist(st_coordinates(sf_data))) / 2
width_lag   <- 75

cutoff_SP   <- make_cutoff(soil_prop_SP)
cutoff_COMP <- make_cutoff(soil_prop_COMP)

# Visualise spatial blocks
soil_prop_SP$spatial_block <- as.character(
  kmeans(st_coordinates(soil_prop_SP), centers = 5, nstart = 25)$cluster
)
ggplot() +
  geom_sf(data = soil_prop_SP, aes(color = spatial_block), size = 3) +
  theme_minimal() +
  labs(title = "Spatial CV blocks — SP dataset", color = "Block")
```

# ============================================================
# PHASE 3 — Model Fitting Helpers (RF, OK, RK)
# ============================================================

```{r model-helpers}

calc_ioa <- function(obs, pred) {
  keep <- is.finite(obs) & is.finite(pred)
  obs <- obs[keep]
  pred <- pred[keep]
  
  if (length(obs) < 2) return(NA_real_)
  
  obs_mean <- mean(obs)
  num <- sum((pred - obs)^2)
  den <- sum((abs(pred - obs_mean) + abs(obs - obs_mean))^2)
  
  if (den == 0) return(NA_real_)
  1 - (num / den)
}

calc_metrics <- function(obs, pred) {
  tibble(
    n          = length(obs),
    RMSE       = sqrt(mean((obs - pred)^2, na.rm = TRUE)),
    MAE        = mean(abs(obs - pred), na.rm = TRUE),
    R2         = cor(obs, pred, use = "complete.obs")^2,
    IOA        = calc_ioa(obs, pred),
    mean_error = mean(obs - pred, na.rm = TRUE)
  )
}

tune_rf_oob <- function(train_df, response, predictors, ntree = 500) {
  x_train   <- train_df %>% select(all_of(predictors))
  y_train   <- train_df[[response]]
  p         <- length(predictors)
  mtry_grid <- unique(pmax(1, c(
    floor(sqrt(p)), floor(p / 4), floor(p / 3), floor(p / 2)
  )))

  fits <- purrr::map(mtry_grid, function(m) {
    set.seed(sample.int(1e6, 1))
    rf <- randomForest(x = x_train, y = y_train,
                       ntree = ntree, mtry = m, importance = TRUE)
    tibble(mtry = m, oob_mse = tail(rf$mse, 1))
  }) %>% bind_rows()

  list(
    best_mtry    = fits %>% slice_min(oob_mse, n = 1) %>% pull(mtry),
    tuning_table = fits
  )
}

fit_variogram_models <- function(vg) {
  vg_df       <- as.data.frame(vg)
  init_nugget <- max(0, min(vg_df$gamma, na.rm = TRUE) * 0.5)
  init_sill   <- max(vg_df$gamma, na.rm = TRUE)
  init_psill  <- max(init_sill - init_nugget, 1e-6)
  init_range  <- max(vg_df$dist, na.rm = TRUE) / 3

  fits <- purrr::map(c("Sph", "Exp"), function(m) {
    tryCatch({
      fit <- fit.variogram(
        vg,
        model = vgm(psill = init_psill, model = m,
                    range = init_range,  nugget = init_nugget)
      )
      if (any(!is.finite(fit$psill)))               return(NULL)
      if (any(!is.finite(fit$range)))               return(NULL)
      if (any(fit$range[fit$model != "Nug"] <= 0))  return(NULL)
      fit
    }, error = function(e) NULL)
  })

  fits <- fits[!sapply(fits, is.null)]
  if (length(fits) == 0) return(NULL)

  sse      <- sapply(fits, function(f) attr(f, "SSErr"))
  best_fit <- fits[[which.min(sse)]]

  # Degeneracy guard — pure nugget = constant surface
  str_row <- best_fit %>% filter(model != "Nug")
  if (nrow(str_row) > 0 && str_row$psill < 1e-6) {
    warning("Degenerate variogram fit: psill ≈ 0. Returning NULL.")
    return(NULL)
  }
  best_fit
}

# ── RF ─────────────────────────────────────────────────────
run_rf_model <- function(train_sf, test_sf, response,
                          predictors, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  tuning  <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  preds <- predict(rf_fit, newdata = test_df %>% select(all_of(predictors)))
  list(
    model       = rf_fit,
    predictions = tibble(obs = test_df[[response]], pred = preds),
    metrics     = calc_metrics(test_df[[response]], preds),
    best_mtry   = tuning$best_mtry
  )
}

# ── OK ─────────────────────────────────────────────────────
run_ok_model <- function(train_sf, test_sf, response, cutoff, width) {
  train_sp <- as(train_sf, "Spatial")
  test_sp  <- as(test_sf,  "Spatial")
  f        <- as.formula(paste(response, "~ 1"))
  vg       <- variogram(f, data = train_sp, cutoff = cutoff, width = width)
  fit      <- fit_variogram_models(vg)
  if (is.null(fit)) return(NULL)

  kr   <- krige(f, locations = train_sp, newdata = test_sp, model = fit)
  pred <- kr$var1.pred
  obs  <- test_sf %>% st_drop_geometry() %>% pull(.data[[response]])
  list(
    variogram   = vg,
    fit         = fit,
    predictions = tibble(obs = obs, pred = pred),
    metrics     = calc_metrics(obs, pred)
  )
}

# ── RK ─────────────────────────────────────────────────────
run_rk_model <- function(train_sf, test_sf, response, predictors,
                          cutoff, width, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  train_sf2 <- train_sf %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))
  test_sf2  <- test_sf  %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))

  tuning <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )

  trend_train        <- predict(rf_fit, newdata = train_df %>%
                                  select(all_of(predictors)))
  trend_test         <- predict(rf_fit, newdata = test_df  %>%
                                  select(all_of(predictors)))
  train_sf2$residual <- train_df[[response]] - trend_train

  train_sp  <- as(train_sf2, "Spatial")
  test_sp   <- as(test_sf2,  "Spatial")
  vg_resid  <- variogram(residual ~ 1, data = train_sp,
                          cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  if (is.null(fit_resid)) {
    return(list(
      metrics               = calc_metrics(test_df[[response]], trend_test),
      best_mtry             = tuning$best_mtry,
      used_residual_kriging = FALSE
    ))
  }

  kr_resid <- krige(residual ~ 1, locations = train_sp,
                    newdata = test_sp, model = fit_resid)
  pred     <- trend_test + kr_resid$var1.pred
  obs      <- test_df[[response]]

  list(
    metrics               = calc_metrics(obs, pred),
    best_mtry             = tuning$best_mtry,
    used_residual_kriging = TRUE
  )
}

# ── Combined iteration runner ──────────────────────────────
run_one_iteration <- function(sf_data, split, response, predictors,
                               dataset_name, cutoff, width, ntree = 500) {
  train_sf <- sf_data[split$train_ids, ]
  test_sf  <- sf_data[split$test_ids,  ]

  rf_res <- run_rf_model(train_sf, test_sf, response, predictors, ntree)
  ok_res <- run_ok_model(train_sf, test_sf, response, cutoff, width)
  rk_res <- run_rk_model(train_sf, test_sf, response, predictors,
                          cutoff, width, ntree)

  bind_rows(
    rf_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RF",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rf_res$best_mtry),
    if (!is.null(ok_res))
      ok_res$metrics %>%
        mutate(dataset = dataset_name, variable = response, method = "OK",
               iteration = split$iteration, seed = split$seed,
               best_mtry = NA_real_),
    rk_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RK",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rk_res$best_mtry,
             used_residual_kriging = rk_res$used_residual_kriging)
  )
}
```

```{r run-cv, warning=FALSE}
results_SP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_SP, function(split_obj) {
    run_one_iteration(soil_prop_SP, split_obj, v, cov_names,
                      "SP", cutoff_SP, width_lag)
  })
})

results_COMP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_COMP, function(split_obj) {
    run_one_iteration(soil_prop_COMP, split_obj, v, cov_names,
                      "COMP", cutoff_COMP, width_lag)
  })
})

results_all <- bind_rows(results_SP, results_COMP)
beep("wilhelm")

# ── CV summary ────────────────────────────────────────────
cv_summary <- results_all %>%
  group_by(dataset, variable, method) %>%
  summarise(
    n_folds     = n(),
    median_R2   = median(R2,   na.rm = TRUE),
    q25_R2      = quantile(R2,   0.25, na.rm = TRUE),
    q75_R2      = quantile(R2,   0.75, na.rm = TRUE),
    median_RMSE = median(RMSE, na.rm = TRUE),
    median_MAE  = median(MAE,  na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  arrange(variable, dataset, desc(median_R2))

print(cv_summary)
write_csv(cv_summary, "../outputs/cv_summary_spatial_folds.csv")

# ── CV violin plots ───────────────────────────────────────
results_all %>%
  mutate(dataset = factor(dataset, c("SP", "COMP")),
         method  = factor(method,  c("OK", "RF", "RK"))) %>%
  ggplot(aes(x = method, y = R2, fill = method)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  facet_grid(variable ~ dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "Spatial CV: R² by dataset and method",
       x = "Method", y = expression(R^2)) +
  theme(legend.position = "none")
```

```{r}
cv_summary <- results_all %>%
  group_by(dataset, variable, method) %>%
  summarise(
    n_folds     = n(),
    median_R2   = median(R2, na.rm = TRUE),
    q25_R2      = quantile(R2, 0.25, na.rm = TRUE),
    q75_R2      = quantile(R2, 0.75, na.rm = TRUE),
    median_RMSE = median(RMSE, na.rm = TRUE),
    cv_RMSE     = sd(RMSE, na.rm = TRUE) / mean(RMSE, na.rm = TRUE),
    median_MAE  = median(MAE, na.rm = TRUE),
    median_IOA  = median(IOA, na.rm = TRUE),
    .groups     = "drop"
  )
```


# ============================================================
# PHASE 4 — Final Full-Dataset Models + Held-Out Validation
# ============================================================

```{r final-models}
fit_final_rf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)
  randomForest(
    x          = df %>% select(all_of(predictors)),
    y          = df[[response]],
    ntree      = ntree,
    mtry       = tuning$best_mtry,
    importance = TRUE
  )
}

fit_final_ok <- function(sf_data, response, cutoff, width) {
  sp_data <- as(sf_data, "Spatial")
  f       <- as.formula(paste(response, "~ 1"))
  vg      <- variogram(f, data = sp_data, cutoff = cutoff, width = width)
  fit     <- fit_variogram_models(vg)
  list(variogram = vg, fit = fit, formula = f, sp_data = sp_data)
}

fit_final_rk <- function(sf_data, response, predictors,
                          cutoff, width, ntree = 500) {
  df      <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  sf_data2 <- sf_data %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))

  tuning  <- tune_rf_oob(df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = df %>% select(all_of(predictors)),
    y = df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  trend             <- predict(rf_fit, newdata = df %>%
                                 select(all_of(predictors)))
  sf_data2$residual <- df[[response]] - trend

  sp_data  <- as(sf_data2, "Spatial")
  vg_resid <- variogram(residual ~ 1, data = sp_data,
                         cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  list(rf_fit = rf_fit, fit_resid = fit_resid, sp_data = sp_data)
}

final_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_SP, v, cov_names),
    ok = fit_final_ok(soil_prop_SP, v, cutoff_SP, width_lag),
    rk = fit_final_rk(soil_prop_SP, v, cov_names, cutoff_SP, width_lag)
  )
)

final_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_COMP, v, cov_names),
    ok = fit_final_ok(soil_prop_COMP, v, cutoff_COMP, width_lag),
    rk = fit_final_rk(soil_prop_COMP, v, cov_names, cutoff_COMP, width_lag)
  )
)

beep("fanfare")
```

```{r variogram-summary}
# ── Phase 4b: Variogram model summary ─────────────────────
extract_variogram_summary <- function(final_models, soil_vars, label) {
  purrr::map_dfr(soil_vars, function(v) {
    fit <- final_models[[v]]$ok$fit
    if (is.null(fit)) {
      return(tibble(variable = v, dataset = label,
                    model = NA, nugget = NA, psill = NA,
                    total_sill = NA, range_m = NA,
                    nugget_sill_ratio = NA))
    }
    nug_row  <- fit %>% filter(model == "Nug")
    str_row  <- fit %>% filter(model != "Nug")
    nugget   <- if (nrow(nug_row) > 0) nug_row$psill else 0
    psill    <- if (nrow(str_row) > 0) str_row$psill else NA
    rng      <- if (nrow(str_row) > 0) str_row$range else NA
    mod_type <- if (nrow(str_row) > 0) as.character(str_row$model) else NA
    tibble(variable = v, dataset = label, model = mod_type,
           nugget   = round(nugget, 4), psill = round(psill, 4),
           total_sill = round(nugget + psill, 4),
           range_m    = round(rng, 1),
           nugget_sill_ratio = round(nugget / (nugget + psill), 3))
  })
}

vg_summary_all <- bind_rows(
  extract_variogram_summary(final_models_SP,   soil_vars, "SP"),
  extract_variogram_summary(final_models_COMP, soil_vars, "COMP")
)

vg_summary_all %>%
  mutate(spatial_dependence = case_when(
    nugget_sill_ratio < 0.25 ~ "Strong",
    nugget_sill_ratio < 0.75 ~ "Moderate",
    TRUE                     ~ "Weak — kriging limited"
  )) %>%
  arrange(dataset, variable) %>%
  print(n = 20)

write_csv(vg_summary_all, "../outputs/variogram_summary_table.csv")
```

```{r held-out-validation}
# ── Predict to held-out validation set ────────────────────
predict_to_points <- function(models_list, response, new_sf, predictors) {
  
  # keep one clean validation table for all methods
  new_sf_use <- new_sf %>%
    filter(
      complete.cases(
        st_drop_geometry(select(., any_of(c(response, predictors))))
      )
    )
  
  new_df <- new_sf_use %>%
    st_drop_geometry() %>%
    select(any_of(c(response, predictors)))
  
  obs <- new_df[[response]]
  
  # RF
  rf_pred <- predict(
    models_list$rf,
    newdata = new_df %>% select(all_of(predictors))
  )
  
  # OK
  ok_pred <- tryCatch({
    kr <- krige(
      models_list$ok$formula,
      locations = models_list$ok$sp_data,
      newdata   = as(new_sf_use, "Spatial"),
      model     = models_list$ok$fit
    )
    kr$var1.pred
  }, error = function(e) rep(NA_real_, nrow(new_df)))
  
  # RK
  rk_trend <- predict(
    models_list$rk$rf_fit,
    newdata = new_df %>% select(all_of(predictors))
  )
  
  rk_resid <- tryCatch({
    kr2 <- krige(
      residual ~ 1,
      locations = models_list$rk$sp_data,
      newdata   = as(new_sf_use, "Spatial"),
      model     = models_list$rk$fit_resid
    )
    kr2$var1.pred
  }, error = function(e) rep(0, nrow(new_df)))
  
  rk_pred <- rk_trend + rk_resid
  
  bind_rows(
    calc_metrics(obs, rf_pred) %>%
      mutate(variable = response, method = "RF"),
    calc_metrics(obs, ok_pred) %>%
      mutate(variable = response, method = "OK"),
    calc_metrics(obs, rk_pred) %>%
      mutate(variable = response, method = "RK")
  )
}

val_results_all <- bind_rows(
  purrr::map_dfr(soil_vars, function(v) {
    predict_to_points(final_models_SP[[v]], v, ap_val_sf,
                      cov_names, cutoff_SP, width_lag) %>%
      mutate(dataset = "SP")
  }),
  purrr::map_dfr(soil_vars, function(v) {
    predict_to_points(final_models_COMP[[v]], v, ap_val_sf,
                      cov_names, cutoff_COMP, width_lag) %>%
      mutate(dataset = "COMP")
  })
)

write_csv(val_results_all, "../outputs/validation_results_held_out.csv")
val_results_all %>% arrange(variable, dataset, desc(R2))
```

# ============================================================
# PHASE 4c — RF Variable Importance Analysis
# ============================================================

```{r rf-importance}
# ── Extract importance from final RF models ────────────────
# Uses %IncMSE (permutation importance) — more reliable
# than node impurity (%IncNodePurity) for correlated predictors.
# Top-20 covariates shown per variable × dataset.

extract_importance <- function(rf_model, variable, dataset) {
  imp <- importance(rf_model, type = 1) # type=1 = %IncMSE
  tibble(
    covariate = rownames(imp),
    pct_inc_mse = imp[, 1],
    variable  = variable,
    dataset   = dataset
  ) %>%
    arrange(desc(pct_inc_mse))
}

importance_SP <- purrr::map_dfr(soil_vars, function(v) {
  extract_importance(final_models_SP[[v]]$rf, v, "SP")
})

importance_COMP <- purrr::map_dfr(soil_vars, function(v) {
  extract_importance(final_models_COMP[[v]]$rf, v, "COMP")
})

importance_all <- bind_rows(importance_SP, importance_COMP)

# ── Per-variable importance plots (top 20 covariates) ─────
plot_importance_per_var <- function(imp_df, dataset_label) {
  imp_df %>%
    filter(dataset == dataset_label) %>%
    group_by(variable) %>%
    slice_max(order_by = pct_inc_mse, n = 20) %>%
    ungroup() %>%
    mutate(covariate = tidytext::reorder_within(
      covariate, pct_inc_mse, variable)) %>%
    ggplot(aes(x = covariate, y = pct_inc_mse,
               fill = pct_inc_mse)) +
    geom_col(show.legend = FALSE) +
    tidytext::scale_x_reordered() +
    scale_fill_viridis_c(option = "plasma", direction = -1) +
    coord_flip() +
    facet_wrap(~ variable, scales = "free") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("RF Variable Importance — ",
                        dataset_label, " dataset"),
      subtitle = "Top 20 covariates by %IncMSE (permutation importance)",
      x        = NULL,
      y        = "% Increase in MSE"
    )
}

# Note: tidytext is used for reorder_within().
# If not installed: install.packages("tidytext")
# Alternative without tidytext: use fct_reorder within facets.

print(plot_importance_per_var(importance_all, "SP"))
print(plot_importance_per_var(importance_all, "COMP"))

# ── Aggregate importance: mean %IncMSE across all soil vars ─
# Identifies covariates that are broadly useful across
# the full agronomic variable set, not just one variable.

agg_importance <- importance_all %>%
  group_by(covariate, dataset) %>%
  summarise(
    mean_pct_inc_mse   = mean(pct_inc_mse,   na.rm = TRUE),
    median_pct_inc_mse = median(pct_inc_mse, na.rm = TRUE),
    n_vars_top20 = sum(rank(desc(pct_inc_mse)) <= 20,
                       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(dataset, desc(mean_pct_inc_mse))

# Top-30 aggregate importance
p_agg_imp <- agg_importance %>%
  group_by(dataset) %>%
  slice_max(mean_pct_inc_mse, n = 30) %>%
  ungroup() %>%
  mutate(
    covariate = tidytext::reorder_within(
      covariate, mean_pct_inc_mse, dataset)
  ) %>%
  ggplot(aes(x = covariate, y = mean_pct_inc_mse,
             fill = dataset)) +
  geom_col(show.legend = FALSE) +
  tidytext::scale_x_reordered() +
  scale_fill_manual(values = c("SP"   = "#2980b9",
                                "COMP" = "#e74c3c")) +
  coord_flip() +
  facet_wrap(~ dataset, scales = "free") +
  theme_bw(base_size = 9) +
  labs(
    title    = "Aggregate RF Variable Importance",
    subtitle = paste0("Mean %IncMSE across all ",
                      length(soil_vars),
                      " agronomic soil variables"),
    x        = NULL,
    y        = "Mean % Increase in MSE"
  )

print(p_agg_imp)

# ── Covariate overlap: top-15 per variable heatmap ────────
top15_per_var <- importance_all %>%
  group_by(dataset, variable) %>%
  slice_max(pct_inc_mse, n = 15) %>%
  ungroup() %>%
  mutate(in_top15 = TRUE)

# Which covariates appear in top-15 for multiple variables?
overlap_counts <- top15_per_var %>%
  group_by(covariate, dataset) %>%
  summarise(n_vars_in_top15 = n(), .groups = "drop") %>%
  filter(n_vars_in_top15 >= 2)

p_overlap <- top15_per_var %>%
  filter(covariate %in% overlap_counts$covariate) %>%
  ggplot(aes(x = variable, y = covariate,
             fill = pct_inc_mse)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                        name = "%IncMSE") +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title    = "Covariate importance overlap across soil variables",
    subtitle = "Covariates appearing in top-15 for ≥2 variables",
    x = "Soil variable", y = "Covariate"
  )

print(p_overlap)
single_bands <- c("coastal_blue", "blue", "green_i", "green",
                  "yellow", "red", "rededge", "nir")

# Soil variable display names
soil_var_labels <- c(
  "lbc_1"  = "Lime Buffer Capacity",
  "lb_ceq" = "LBC Equilibrium",
  "p_h_2"  = "pH",
  "ca"     = "Calcium",
  "k"      = "Potassium",
  "mg"     = "Magnesium",
  "mn"     = "Manganese",
  "p"      = "Phosphorus",
  "zn"     = "Zinc"
)

# ── Covariate overlap: top-15 per variable heatmap ────────
top15_per_var <- importance_all %>%
  filter(!covariate %in% single_bands) %>%
  group_by(dataset, variable) %>%
  slice_max(pct_inc_mse, n = 15) %>%
  ungroup() %>%
  mutate(in_top15 = TRUE)

overlap_counts <- top15_per_var %>%
  group_by(covariate, dataset) %>%
  summarise(n_vars_in_top15 = n(), .groups = "drop") %>%
  filter(n_vars_in_top15 >= 2)

p_overlap2 <- top15_per_var %>%
  filter(covariate %in% overlap_counts$covariate) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable,
                      levels = soil_var_labels)
  ) %>%
  ggplot(aes(x = variable, y = covariate,
             fill = pct_inc_mse)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                        name = "%IncMSE") +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title    = "Covariate importance overlap across soil variables",
    subtitle = "Covariates appearing in top-15 for ≥2 variables (single bands excluded)",
    x = "Soil variable", y = "Covariate"
  )

print(p_overlap2)
#save:

ggsave("../outputs/fig_rf_variable_importance_overlap.png",
       p_overlap2,
       width = 10, height = 6, dpi = 300)


write_csv(importance_all,
          "../outputs/rf_importance_per_variable.csv")
write_csv(agg_importance,
          "../outputs/rf_importance_aggregate.csv")

soil_var_labels <- c(
  "lbc_1"  = "Lime Buffer Capacity",
  "lb_ceq" = "Lime Buffer Capacity Eq.",
  "p_h_2"  = "pH",
  "ca"     = "Calcium (Ca)",
  "k"      = "Potassium (K)",
  "mg"     = "Magnesium (Mg)",
  "mn"     = "Manganese (Mn)",
  "p"      = "Phosphorus (P)",
  "zn"     = "Zinc (Zn)"
)

# ── Soil variable × covariate correlation heatmap ─────────
# Pearson correlation between each soil variable and each
# covariate at SP and COMP sampling locations combined.
# Helps interpret which environmental drivers are most
# associated with each soil property.

cor_data <- bind_rows(
  soil_prop_SP   %>% st_drop_geometry() %>% mutate(dataset = "SP"),
  soil_prop_COMP %>% st_drop_geometry() %>% mutate(dataset = "COMP")
) %>%
  select(all_of(soil_vars), all_of(cov_names)) %>%
  na.omit()

cor_mat <- cor(
  cor_data %>% select(all_of(soil_vars)),
  cor_data %>% select(all_of(cov_names)),
  use = "complete.obs"
) %>%
  as.data.frame() %>%
  rownames_to_column("soil_var") %>%
  pivot_longer(-soil_var,
               names_to  = "covariate",
               values_to = "r") %>%
    mutate(
        soil_var  = factor(recode(soil_var, !!!soil_var_labels),
                           levels = soil_var_labels),
        covariate = factor(covariate, levels = cov_names)
      ) %>%
   filter(!covariate %in% single_bands)

p_cor_heatmap <- cor_mat %>%
  ggplot(aes(x = soil_var, y = covariate, fill = r)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = ifelse(abs(r) >= 0.10,
                                sprintf("%.2f", r), "")),
            size = 2.0, color = "grey10") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9850",
    midpoint = 0,
    limits   = c(-1, 1),
    oob      = scales::squish,
    name     = "Pearson r"
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1,
                                     size = 8, face = "bold"),
    axis.text.y      = element_text(size = 7),
    panel.grid       = element_blank(),
    legend.position  = "right"
  ) +
  labs(
    title    = "Soil property × covariate correlations",
    subtitle = "SP and COMP sampling locations combined (n = 272). Values shown where |r| ≥ 0.10",
    x        = "Soil variable",
    y        = "Covariate"
  )

print(p_cor_heatmap)
ggsave("../outputs/fig_soil_covariate_correlations.png",
       p_cor_heatmap,
       width  = 10,
       height = 14,
       dpi    = 300)
write_csv(cor_mat, "../outputs/soil_covariate_correlations.csv")

#pH is the most predictable variable (top values 10–11%) — highest importance signal of any variable, consistent with it being the dominant UMAP1 driver
# Zn is the least predictable (top values 3–4%) — consistent with its pure nugget variogram and near-zero spatial structure
# P SP has unusually high mean importance (2.79%) relative to its top value — suggesting many covariates contribute modestly rather than one dominating
# COMP generally has higher top values than SP — composite sampling smooths local noise, making covariate-soil relationships cleaner and easier for RF to detect
```

# ============================================================
# PHASE 5 — Full-Grid Spatial Prediction
# ============================================================

```{r grid-prediction}
grid_df <- as.data.frame(covariate_stack, xy = TRUE, na.rm = TRUE)
grid_sf <- st_as_sf(grid_df, coords = c("x", "y"), crs = 26917)
grid_sp <- as(grid_sf, "Spatial")

predict_grid_rf <- function(rf_model, grid_df, predictors) {
  predict(rf_model, newdata = grid_df %>% select(all_of(predictors)))
}

predict_grid_ok <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    kr$var1.pred
  }, error = function(e) {
    warning("OK prediction failed; returning NA")
    rep(NA_real_, nrow(grid_sp@data))
  })
}

predict_grid_rk <- function(rk_obj, grid_df, grid_sp, predictors) {
  trend <- predict(rk_obj$rf_fit,
                   newdata = grid_df %>% select(all_of(predictors)))
  resid <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    kr2$var1.pred
  }, error = function(e) rep(0, length(trend)))
  trend + resid
}

build_prediction_stack <- function(final_models, grid_df, grid_sf,
                                    grid_sp, soil_vars, predictors,
                                    label) {
  preds <- purrr::map(setNames(soil_vars, soil_vars), function(v) {
    m <- final_models[[v]]
    tibble(
      !!paste0(v, "_RF") := predict_grid_rf(m$rf, grid_df, predictors),
      !!paste0(v, "_RK") := predict_grid_rk(m$rk, grid_df, grid_sp, predictors),
      !!paste0(v, "_OK") := predict_grid_ok(m$ok, grid_sp)
    )
  }) %>% bind_cols()

  bind_cols(grid_df %>% select(x, y), preds, dataset = label)
}

pred_stack_SP   <- build_prediction_stack(
  final_models_SP,   grid_df, grid_sf, grid_sp,
  soil_vars, cov_names, "SP")
pred_stack_COMP <- build_prediction_stack(
  final_models_COMP, grid_df, grid_sf, grid_sp,
  soil_vars, cov_names, "COMP")

beep("mario")
save.image("../outputs/workspace_phase4b.RData")
beep("coin")
```

# ============================================================
# PHASE 5b — Spatial Lag Features
# ============================================================

```{r spatial-lags}
add_spatial_lags_focal <- function(pred_df, soil_vars, methods,
                                    crs = "EPSG:26917") {
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_df)]

  cat("Computing spatial lags via terra::focal for",
      length(col_names), "layers...\n")

  pred_rast <- pred_df %>%
    select(x, y, all_of(col_names)) %>%
    rast(type = "xyz", crs = crs)

  w       <- matrix(1, 3, 3)
  w[2, 2] <- 0
  lag_rast <- terra::focal(pred_rast, w = w,
                            fun = "mean", na.rm = TRUE)
  names(lag_rast) <- paste0(col_names, "_lag")

  pts    <- terra::vect(pred_df %>% select(x, y),
                        geom = c("x", "y"), crs = crs)
  lag_df <- terra::extract(lag_rast, pts, ID = FALSE)

  cat("pred_df rows:", nrow(pred_df),
      "| lag rows:", nrow(lag_df), "\n")
  bind_cols(pred_df, lag_df)
}

pred_stack_SP_lag   <- add_spatial_lags_focal(
  pred_stack_SP,   soil_vars, c("RF", "RK"))
pred_stack_COMP_lag <- add_spatial_lags_focal(
  pred_stack_COMP, soil_vars, c("RF", "RK"))

cat("SP lag stack columns:",   ncol(pred_stack_SP_lag),   "\n")
cat("COMP lag stack columns:", ncol(pred_stack_COMP_lag), "\n")
```

# ============================================================
# PHASE 5c — Uncertainty Analysis
# ============================================================

```{r uncertainty-analysis}
# ── 1. Quantile Regression Forest (RF Prediction Intervals) ─
# QRF extends RF to estimate conditional quantiles.
# We compute the 10th and 90th percentiles to produce
# an 80% prediction interval for each grid pixel.
# Interval width = spatial map of RF prediction uncertainty.

fit_qrf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)

  set.seed(20260401)
  quantregForest(
    x    = as.matrix(df %>% select(all_of(predictors))),
    y    = df[[response]],
    ntree = ntree,
    mtry  = tuning$best_mtry
  )
}

cat("Fitting QRF models for SP dataset...\n")
qrf_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  ~ fit_qrf(soil_prop_SP, .x, cov_names)
)

cat("Fitting QRF models for COMP dataset...\n")
qrf_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  ~ fit_qrf(soil_prop_COMP, .x, cov_names)
)

# Predict 80% PI to grid
grid_mat <- as.matrix(grid_df %>% select(all_of(cov_names)))

predict_qrf_grid <- function(qrf_model, grid_mat,
                              quantiles = c(0.10, 0.90)) {
  preds <- predict(qrf_model, newdata = grid_mat,
                   what = quantiles)
  tibble(
    pi_lower = preds[, 1],
    pi_upper = preds[, 2],
    pi_width = preds[, 2] - preds[, 1]
  )
}

qrf_uncertainty_SP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    p <- predict_qrf_grid(qrf_models_SP[[v]], grid_mat)
    setNames(p, paste0(v, "_", names(p)))
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

qrf_uncertainty_COMP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    p <- predict_qrf_grid(qrf_models_COMP[[v]], grid_mat)
    setNames(p, paste0(v, "_", names(p)))
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

# ── QRF uncertainty maps — one panel per variable ─────────
plot_qrf_width <- function(qrf_df, dataset_label) {
  width_cols <- paste0(soil_vars, "_pi_width")
  width_cols <- width_cols[width_cols %in% names(qrf_df)]

  qrf_df %>%
    select(x, y, all_of(width_cols)) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "pi_width") %>%
    mutate(variable = str_remove(variable, "_pi_width")) %>%
    ggplot(aes(x = x, y = y, fill = pi_width)) +
    geom_raster() +
    scale_fill_viridis_c(option = "inferno", direction = -1,
                          name = "PI width\n(80%)") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("QRF 80% Prediction Interval Width — ",
                        dataset_label),
      subtitle = "Wider = higher RF prediction uncertainty",
      x = "Easting", y = "Northing"
    )
}

print(plot_qrf_width(qrf_uncertainty_SP,   "SP"))
print(plot_qrf_width(qrf_uncertainty_COMP, "COMP"))

write_csv(qrf_uncertainty_SP,
          "../outputs/qrf_uncertainty_SP.csv")
write_csv(qrf_uncertainty_COMP,
          "../outputs/qrf_uncertainty_COMP.csv")

# ── 2. Kriging Variance Surfaces (OK and RK) ──────────────
# krige() returns var1.var — the kriging variance at each
# prediction location. This is the geostatistical uncertainty
# estimate; sqrt(var1.var) = standard error of the kriging
# prediction. For RK, this is the residual kriging variance.

predict_grid_ok_with_var <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    tibble(
      pred = kr$var1.pred,
      krig_var = kr$var1.var
    )
  }, error = function(e) {
    tibble(pred    = rep(NA_real_, nrow(grid_sp@data)),
           krig_var = rep(NA_real_, nrow(grid_sp@data)))
  })
}

predict_grid_rk_with_var <- function(rk_obj, grid_df,
                                      grid_sp, predictors) {
  trend <- predict(rk_obj$rf_fit,
                   newdata = grid_df %>% select(all_of(predictors)))
  resid_pred <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    tibble(resid = kr2$var1.pred,
           krig_var = kr2$var1.var)
  }, error = function(e) {
    tibble(resid    = rep(0, length(trend)),
           krig_var = rep(NA_real_, length(trend)))
  })
  tibble(
    pred     = trend + resid_pred$resid,
    krig_var = resid_pred$krig_var
  )
}

krig_var_SP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    m <- final_models_SP[[v]]
    ok <- predict_grid_ok_with_var(m$ok, grid_sp)
    rk <- predict_grid_rk_with_var(m$rk, grid_df, grid_sp, cov_names)
    tibble(
      !!paste0(v, "_ok_krig_var") := ok$krig_var,
      !!paste0(v, "_rk_krig_var") := rk$krig_var
    )
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

krig_var_COMP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    m <- final_models_COMP[[v]]
    ok <- predict_grid_ok_with_var(m$ok, grid_sp)
    rk <- predict_grid_rk_with_var(m$rk, grid_df, grid_sp, cov_names)
    tibble(
      !!paste0(v, "_ok_krig_var") := ok$krig_var,
      !!paste0(v, "_rk_krig_var") := rk$krig_var
    )
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

# ── Kriging SE maps ───────────────────────────────────────
plot_krig_se <- function(kv_df, dataset_label, method_suffix) {
  se_cols <- paste0(soil_vars, method_suffix)
  se_cols <- se_cols[se_cols %in% names(kv_df)]

  kv_df %>%
    select(x, y, all_of(se_cols)) %>%
    mutate(across(all_of(se_cols), sqrt)) %>%  # var → SE
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "krig_se") %>%
    mutate(variable = str_remove(
      variable, str_escape(method_suffix))) %>%
    ggplot(aes(x = x, y = y, fill = krig_se)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1,
                          name = "Kriging SE",
                          na.value = "grey80") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("Kriging Standard Error — ",
                        dataset_label, " (",
                        str_remove(method_suffix,
                                   "_krig_var"), ")"),
      subtitle = "sqrt(kriging variance) at each grid pixel",
      x = "Easting", y = "Northing"
    )
}

print(plot_krig_se(krig_var_SP,   "SP",   "_ok_krig_var"))
print(plot_krig_se(krig_var_SP,   "SP",   "_rk_krig_var"))
print(plot_krig_se(krig_var_COMP, "COMP", "_ok_krig_var"))
print(plot_krig_se(krig_var_COMP, "COMP", "_rk_krig_var"))

write_csv(krig_var_SP,   "../outputs/kriging_variance_SP.csv")
write_csv(krig_var_COMP, "../outputs/kriging_variance_COMP.csv")

# ── 3. Bootstrap Uncertainty for RF Predictions ───────────
# Bootstrap resampling of training data (B=100 replicates)
# to estimate prediction SD at each grid pixel.
# SD captures the sensitivity of RF predictions to the
# specific set of training observations used.

bootstrap_rf_uncertainty <- function(sf_data, soil_vars,
                                      predictors, grid_df,
                                      B = 100, ntree = 500,
                                      seed = 20260401) {
  cat("Bootstrap RF uncertainty (B =", B, "replicates)...\n")

  df_full <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(soil_vars, predictors))) %>% na.omit()

  set.seed(seed)
  boot_preds <- purrr::map(seq_len(B), function(b) {
    if (b %% 10 == 0) cat(" Bootstrap rep:", b, "/", B, "\n")
    boot_idx <- sample(nrow(df_full), replace = TRUE)
    boot_df  <- df_full[boot_idx, ]

    purrr::map_dfc(soil_vars, function(v) {
      rf_b <- tryCatch({
        randomForest(
          x = boot_df %>% select(all_of(predictors)),
          y = boot_df[[v]],
          ntree = ntree, mtry = floor(sqrt(length(predictors)))
        )
      }, error = function(e) NULL)

      if (is.null(rf_b)) {
        return(tibble(!!v := rep(NA_real_, nrow(grid_df))))
      }

      tibble(
        !!v := predict(rf_b,
                       newdata = grid_df %>%
                         select(all_of(predictors)))
      )
    })
  })

  # Compute SD across bootstrap replicates per pixel
  purrr::map_dfc(soil_vars, function(v) {
    boot_mat <- do.call(cbind, purrr::map(boot_preds, ~ .x[[v]]))
    tibble(
      !!paste0(v, "_boot_sd")   := apply(boot_mat, 1,
                                          sd, na.rm = TRUE),
      !!paste0(v, "_boot_mean") := apply(boot_mat, 1,
                                          mean, na.rm = TRUE)
    )
  }) %>%
    bind_cols(grid_df %>% select(x, y), .)
}

# B=100 replicates — increase to 500 for publication
cat("Running bootstrap for SP dataset...\n")
boot_uncertainty_SP   <- bootstrap_rf_uncertainty(
  soil_prop_SP,   soil_vars, cov_names, grid_df, B = 100)

cat("Running bootstrap for COMP dataset...\n")
boot_uncertainty_COMP <- bootstrap_rf_uncertainty(
  soil_prop_COMP, soil_vars, cov_names, grid_df, B = 100)

# ── Bootstrap SD maps ─────────────────────────────────────
plot_boot_sd <- function(boot_df, dataset_label) {
  sd_cols <- paste0(soil_vars, "_boot_sd")
  sd_cols <- sd_cols[sd_cols %in% names(boot_df)]

  boot_df %>%
    select(x, y, all_of(sd_cols)) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd")) %>%
    ggplot(aes(x = x, y = y, fill = boot_sd)) +
    geom_raster() +
    scale_fill_viridis_c(option = "rocket", direction = -1,
                          name = "Bootstrap SD") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("RF Bootstrap Uncertainty (B=100) — ",
                        dataset_label),
      subtitle = "SD of predictions across 100 bootstrap replicates",
      x = "Easting", y = "Northing"
    )
}

print(plot_boot_sd(boot_uncertainty_SP,   "SP"))
print(plot_boot_sd(boot_uncertainty_COMP, "COMP"))

write_csv(boot_uncertainty_SP,
          "../outputs/bootstrap_uncertainty_SP.csv")
write_csv(boot_uncertainty_COMP,
          "../outputs/bootstrap_uncertainty_COMP.csv")

# ── 4. Zone Assignment Uncertainty (membership probability) ─
# Already computed within HDBSCAN — handled in Phase 6b.
# Membership probability ∈ [0,1]: 1 = core cluster member,
# 0 = peripherally assigned (borderline pixel).
# Noise pixels assigned by NN have membership_prob = 0.
# Reported spatially and summarised per zone below.
# See plot_confidence() in Phase 6b for spatial maps.

# ── Uncertainty comparison summary ────────────────────────
# Side-by-side comparison of uncertainty magnitude by method
uncertainty_summary <- bind_rows(
  # QRF mean PI width
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable  = str_remove(variable, "_pi_width"),
           method    = "QRF 80% PI width",
           dataset   = "SP"),

  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           method   = "QRF 80% PI width",
           dataset  = "COMP"),

  # Bootstrap SD
  boot_uncertainty_SP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           method   = "Bootstrap SD",
           dataset  = "SP"),

  boot_uncertainty_COMP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           method   = "Bootstrap SD",
           dataset  = "COMP")
) %>%
  group_by(variable, method, dataset) %>%
  summarise(
    mean_unc   = round(mean(value,   na.rm = TRUE), 3),
    median_unc = round(median(value, na.rm = TRUE), 3),
    .groups    = "drop"
  )

print(uncertainty_summary, n = 40)
write_csv(uncertainty_summary,
          "../outputs/uncertainty_summary.csv")

beep("wilhelm")
save.image("../outputs/workspace_uncertainty.RData")
#ERROR:
#Per-variable interpretation:

  # pH — bootstrap SD ~0.07–0.10, PI width ~0.9–1.1. Very tight, makes sense — pH varies narrowly at field scale and RF predicts it well
  # P — bootstrap SD ~2–3, PI width ~20–23. Low uncertainty relative to the other nutrients, consistent with P often being spatially structured by management history
  # K — moderate uncertainty, COMP tighter than SP which is expected given composite sampling smooths local variability
  # Ca and lb_ceq — largest absolute uncertainty, which makes sense as these are highly variable and correlated with soil texture and organic matter gradients
#The SP > COMP pattern for most variables is also exactly what I expect w/ composite samples average over a larger area so the RF trained on them sees less local noise, producing tighter predictions.

#P flips the pattern (COMP SD slightly higher than SP). That's worth a sentence in my methods ...it may reflect that composite sampling obscures real P spatial structure that single-point sampling captures, since P is often highly localized around application history.
# Kriging SE summary — OK
bind_rows(
  krig_var_SP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "SP", method = "OK kriging SE"),

  krig_var_COMP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "COMP", method = "OK kriging SE"),

  krig_var_SP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "SP", method = "RK kriging SE"),

  krig_var_COMP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "COMP", method = "RK kriging SE")
) %>%
  mutate(
    variable = str_extract(name, paste(soil_vars, collapse = "|")),
    stat     = if_else(str_detect(name, "_mean"), "mean", "median")
  ) %>%
  select(variable, stat, dataset, method, value) %>%
  arrange(variable, method, dataset, stat) 

# FINAL UNCERTAINTY SUMMARY:
# SP > COMP for most variables (SP higher uncertainty), except P (COMP slightly higher)
# OK > RK > RF in uncertainty magnitude
# HOWEVER — these are not apples-to-apples comparisons:
#   - OK SE is purely geometry-driven (data configuration)
#   - RK SE is residual kriging component only (incomplete — missing RF trend uncertainty)
#   - RF bootstrap SD captures training data sensitivity only
# RK total uncertainty = RF trend uncertainty + residual kriging variance (neither alone is complete)
# All three are complementary, not directly comparable on the same scale


#RF is not highly-sensitive to particular variables, and is producing stable predictions (based on QRF ratios) RK has high uncertainty due to it being hybrid. 
#OK predicted constantly across each map for SE, except for SP Mg, which had a great variogram fit and strong autocorrelation. 
```

# ============================================================
# PHASE 6 — UMAP + HDBSCAN Management Zone Delineation
# ============================================================

```{r umap-hdbscan}
run_umap_hdbscan <- function(pred_df,
                              methods          = c("RF", "RK"),
                              soil_vars,
                              include_lags     = FALSE,
                              n_components     = 2,
                              min_dist         = 0.001,
                              n_neighbors      = 30,
                              minPts           = 50,
                              cluster_eps      = 0,
                              label            = "dataset") {

  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  if (include_lags) {
    lag_cols  <- paste0(col_names, "_lag")
    col_names <- c(col_names, lag_cols[lag_cols %in% names(pred_df)])
  }
  col_names <- col_names[col_names %in% names(pred_df)]

  mat           <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  mat_clean     <- mat[complete_rows, ]
  xy_clean      <- pred_df %>% select(x, y) %>% slice(which(complete_rows))

  cat(label, "— pixels:", nrow(mat_clean),
      "| cols:", ncol(mat_clean),
      "| lags:", include_lags, "\n")

  # Zero-variance guard
  col_ranges <- apply(mat_clean, 2, function(x)
                      diff(range(x, na.rm = TRUE)))
  zero_var   <- col_ranges < 1e-10

  if (any(zero_var)) {
    cat("Warning: dropping", sum(zero_var), "constant column(s):",
        paste(colnames(mat_clean)[zero_var], collapse = ", "), "\n")
    mat_clean <- mat_clean[, !zero_var, drop = FALSE]
  }

  if (ncol(mat_clean) < 2) {
    cat("ERROR: fewer than 2 valid columns for", label, "\n")
    return(NULL)
  }

  mat_scaled <- scale(mat_clean)
  mat_scaled[!is.finite(mat_scaled)] <- 0
  stopifnot("NA values remain" = !anyNA(mat_scaled))

  cat("Fitting UMAP...\n")
  umap_fit <- uwot::umap(
    mat_scaled,
    n_components = n_components,
    min_dist     = min_dist,
    n_neighbors  = n_neighbors,
    ret_model    = TRUE,
    seed         = 20260401,
    verbose      = TRUE
  )

  emb_df        <- as.data.frame(umap_fit$embedding)
  names(emb_df) <- paste0("UMAP", seq_len(n_components))

  hdbscan_sample_size <- min(nrow(emb_df), 10000)
  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df), hdbscan_sample_size))
  emb_sample <- emb_df[sample_idx, ]

  cat("Running HDBSCAN on subsample of", hdbscan_sample_size,
      "points...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts,
    cluster_selection_epsilon = cluster_eps
  )

  cat("Clusters on subsample:", max(hdb_fit$cluster),
      "| noise:", round(mean(hdb_fit$cluster == 0) * 100, 1), "%\n")

  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  cat("Full grid — zones:", max(full_clusters),
      "| noise:", round(mean(full_clusters == 0) * 100, 1), "%\n")

  bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard    = full_clusters,
           membership_prob = full_memb)
  ) %>%
    mutate(
      dataset      = label,
      spatial_lags = include_lags,
      zone         = if_else(cluster_hard == 0L, NA_integer_,
                             as.integer(cluster_hard))
    )
}

wrap_mz_result <- function(df) {
  if (is.data.frame(df)) {
    zones <- df$zone[!is.na(df$zone)]
    return(list(result_df   = df,
                n_zones     = length(unique(zones)),
                pct_noise   = mean(is.na(df$zone)) * 100,
                umap_fit    = NULL,
                hdbscan_fit = NULL))
  }
  zones <- df$result_df$zone[!is.na(df$result_df$zone)]
  df$n_zones   <- length(unique(zones))
  df$pct_noise <- mean(is.na(df$result_df$zone)) * 100
  df
}

# ── 10 UMAP+HDBSCAN runs ──────────────────────────────────
mz_SP_RF_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RF_nolags")

mz_SP_RF_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RF_lags")

mz_SP_RK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RK_nolags")

mz_SP_RK_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RK_lags")

mz_SP_OK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_OK_nolags")

mz_COMP_RF_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RF_nolags")

mz_COMP_RF_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RF_lags")

mz_COMP_RK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RK_nolags")

mz_COMP_RK_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RK_lags")

mz_COMP_OK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_OK_nolags")

save.image("../outputs/workspace_phase6_complete.RData")
beep("mario")

mz_named_tuned <- list(
  SP_RF_nolags   = wrap_mz_result(mz_SP_RF_nolags),
  SP_RF_lags     = wrap_mz_result(mz_SP_RF_lags),
  SP_RK_nolags   = wrap_mz_result(mz_SP_RK_nolags),
  SP_RK_lags     = wrap_mz_result(mz_SP_RK_lags),
  SP_OK_nolags   = wrap_mz_result(mz_SP_OK_nolags),
  COMP_RF_nolags = wrap_mz_result(mz_COMP_RF_nolags),
  COMP_RF_lags   = wrap_mz_result(mz_COMP_RF_lags),
  COMP_RK_nolags = wrap_mz_result(mz_COMP_RK_nolags),
  COMP_RK_lags   = wrap_mz_result(mz_COMP_RK_lags),
  COMP_OK_nolags = wrap_mz_result(mz_COMP_OK_nolags)
)

purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})
```

# ============================================================
# PHASE 6c — HDBSCAN Retuning
# ============================================================

```{r hdbscan-retune}
retune_hdbscan <- function(mz_obj, minPts_new,
                            cluster_eps = 0,
                            hdbscan_n   = 10000,
                            label       = NULL) {
  emb_df       <- mz_obj$result_df %>% select(UMAP1, UMAP2)
  xy_clean     <- mz_obj$result_df %>% select(x, y)
  spatial_lags <- mz_obj$result_df$spatial_lags[1]
  lbl          <- if (!is.null(label)) label else
                    mz_obj$result_df$dataset[1]
  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df),
                             min(hdbscan_n, nrow(emb_df))))
  emb_sample <- emb_df[sample_idx, ]
  cat("Retuning:", lbl, "| minPts =", minPts_new, "...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts_new,
    cluster_selection_epsilon = cluster_eps
  )
  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]
  result_df <- bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard = full_clusters,
           membership_prob = full_memb)
  ) %>%
    mutate(dataset      = lbl,
           spatial_lags = spatial_lags,
           zone         = if_else(cluster_hard == 0L,
                                  NA_integer_,
                                  as.integer(cluster_hard)))
  list(result_df   = result_df,
       hdbscan_fit = hdb_fit,
       n_zones     = length(unique(full_clusters[full_clusters != 0])),
       pct_noise   = mean(full_clusters == 0) * 100)
}

# ── Snapshot the source list before any tuning ─────────────
mz_named_source <- mz_named_tuned

# ── Map each configuration to its pred_df + method ────────
config_pred_map <- list(
  SP_RF_nolags   = list(pred_df = pred_stack_SP,        method = "RF"),
  SP_RF_lags     = list(pred_df = pred_stack_SP_lag,    method = "RF"),
  SP_RK_nolags   = list(pred_df = pred_stack_SP,        method = "RK"),
  SP_RK_lags     = list(pred_df = pred_stack_SP_lag,    method = "RK"),
  SP_OK_nolags   = list(pred_df = pred_stack_SP,        method = "OK"),
  COMP_RF_nolags = list(pred_df = pred_stack_COMP,      method = "RF"),
  COMP_RF_lags   = list(pred_df = pred_stack_COMP_lag,  method = "RF"),
  COMP_RK_nolags = list(pred_df = pred_stack_COMP,      method = "RK"),
  COMP_RK_lags   = list(pred_df = pred_stack_COMP_lag,  method = "RK"),
  COMP_OK_nolags = list(pred_df = pred_stack_COMP,      method = "OK")
)

minpts_candidates <- c(50, 100, 200, 400, 600, 800, 1000)

# ── Sensitivity scan with VR_pred per configuration ───────
# For each (config, minPts) pair, retune HDBSCAN and compute
# the resulting variance reduction on the prediction grid.
cat("\n=== Sensitivity scan: computing n_zones, noise %, and VR per minPts ===\n\n")

sensitivity_results <- purrr::map_dfr(minpts_candidates, function(mp) {
  purrr::imap_dfr(config_pred_map, function(cfg, nm) {
    r <- retune_hdbscan(mz_named_source[[nm]], mp, label = nm)
    vr <- tryCatch(
      calc_variance_reduction(cfg$pred_df, r$result_df$zone,
                              soil_vars, cfg$method),
      error = function(e) NA_real_
    )
    tibble(label     = nm,
           minPts    = mp,
           n_zones   = r$n_zones,
           pct_noise = round(r$pct_noise, 1),
           VR        = round(vr, 3))
  })
})

# ── Display the cluster count and VR by config ────────────
cat("\n=== Cluster counts by minPts and configuration ===\n")
sensitivity_results %>%
  select(label, minPts, n_zones) %>%
  pivot_wider(names_from = label, values_from = n_zones) %>%
  arrange(minPts) %>%
  print(width = Inf)

cat("\n=== VR by minPts and configuration ===\n")
sensitivity_results %>%
  select(label, minPts, VR) %>%
  pivot_wider(names_from = label, values_from = VR) %>%
  arrange(minPts) %>%
  print(width = Inf)

write_csv(sensitivity_results,
          "../outputs/hdbscan_minpts_tuning.csv")

# ── Per-configuration selection rule ───────────────────────
# 1. Prefer minPts producing 3–8 zones (agronomically interpretable)
# 2. Among those, maximize VR
# 3. If no candidate falls in 3–8 zones, fall back to the option
#    with the most zones > 1 (avoid pure-noise collapse)
select_best_minpts <- function(config_tuning,
                                min_zones = 3,
                                max_zones = 8,
                                hard_max_zones = 12) {
  
  # Tier 1: prefer minPts producing the agronomic sweet spot (3–8 zones)
  valid <- config_tuning %>%
    filter(n_zones >= min_zones, n_zones <= max_zones,
           !is.na(VR))
  
  if (nrow(valid) > 0) {
    return(valid %>% slice_max(VR, n = 1, with_ties = FALSE) %>%
                     pull(minPts))
  }
  
  # Tier 2: accept slightly wider range (up to hard_max_zones)
  # but cap at 12 — anything above that is unusable for management
  loose <- config_tuning %>%
    filter(n_zones >= 2, n_zones <= hard_max_zones,
           !is.na(VR))
  
  if (nrow(loose) > 0) {
    return(loose %>% slice_max(VR, n = 1, with_ties = FALSE) %>%
                     pull(minPts))
  }
  
  # Tier 3: fallback — best VR among configs with 2+ zones,
  # but explicitly NOT picking solutions with > hard_max_zones
  fallback <- config_tuning %>%
    filter(n_zones >= 2, n_zones <= hard_max_zones)
  
  if (nrow(fallback) > 0) {
    return(fallback %>% slice_max(n_zones, n = 1, with_ties = FALSE) %>%
                       pull(minPts))
  }
  
  200  # safe default
}

minpts_final <- sensitivity_results %>%
  split(.$label) %>%
  purrr::map(~ select_best_minpts(.x))

cat("\n=== Selected minPts per configuration ===\n")
purrr::iwalk(minpts_final, function(mp, nm) {
  selected_row <- sensitivity_results %>%
    filter(label == nm, minPts == mp)
  cat(sprintf("  %-20s minPts = %4d  →  zones = %d, noise = %.1f%%, VR = %.3f\n",
              nm, mp, selected_row$n_zones,
              selected_row$pct_noise, selected_row$VR))
})

# ── Apply selected minPts to retune all configurations ────
mz_named_tuned <- purrr::imap(minpts_final, function(mp, nm) {
  retune_hdbscan(mz_named_source[[nm]], mp, label = nm)
})

cat("\n=== Final tuned configurations ===\n")
purrr::iwalk(mz_named_tuned, function(x, nm) {
  cat(sprintf("  %-20s | zones: %d | noise: %.1f%% | minPts: %d\n",
              nm, x$n_zones, x$pct_noise, minpts_final[[nm]]))
})

save.image("../outputs/workspace_phase6c_retuned.RData")
beep("coin")
```

# ============================================================
# PHASE 6b — Post-processing + Zone Uncertainty Summary
# ============================================================

```{r post-processing}
post_process_zones <- function(result_df, zone_rast_crs = "EPSG:26917") {
  n_total <- nrow(result_df)
  n_noise <- sum(is.na(result_df$zone))
  cat("Noise pixels before fill:", n_noise,
      "(", round(100 * n_noise / n_total, 1), "% of field)\n")

  if (n_noise == n_total) {
    cat("WARNING: all pixels are noise — no zones found.\n")
    return(list(result_df_filled = result_df,
                zone_rast_raw    = NULL,
                zone_rast_smooth = NULL))
  }

  assigned  <- result_df %>% filter(!is.na(zone))
  noise_pts <- result_df %>% filter( is.na(zone))

  if (nrow(noise_pts) > 0) {
    nn_idx <- FNN::get.knnx(
      data  = assigned  %>% select(UMAP1, UMAP2) %>% as.matrix(),
      query = noise_pts %>% select(UMAP1, UMAP2) %>% as.matrix(),
      k     = 1
    )$nn.index
    noise_pts$zone            <- assigned$zone[nn_idx]
    noise_pts$membership_prob <- 0
    result_df <- bind_rows(assigned, noise_pts)
  }

  cat("Noise pixels after fill: 0\n")

  zone_rast        <- result_df %>%
    select(x, y, zone) %>%
    rast(type = "xyz", crs = zone_rast_crs)
  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)

  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("Pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "% of field)\n")

  list(result_df_filled = result_df,
       zone_rast_raw    = zone_rast,
       zone_rast_smooth = zone_rast_smooth)
}

pp_named <- purrr::imap(mz_named_tuned, function(x, nm) {
  cat("\n---", nm, "---\n")
  post_process_zones(x$result_df)
})

write_zone_raster <- function(pp_obj, path) {
  if (is.null(pp_obj$zone_rast_smooth)) {
    cat("Skipping", path, "— no zones\n")
    return(invisible(NULL))
  }
  writeRaster(pp_obj$zone_rast_smooth, path, overwrite = TRUE)
  cat("Saved:", path, "\n")
}

purrr::iwalk(pp_named, function(pp, nm) {
  write_zone_raster(pp, paste0("../outputs/zones_", nm, "_smooth.tif"))
})

# ── UMAP / spatial zone / confidence plots ────────────────
plot_umap_zones <- function(res, title) {
  res$result_df %>%
    ggplot(aes(UMAP1, UMAP2, color = factor(zone),
               alpha = membership_prob)) +
    geom_point(size = 0.4) +
    scale_alpha_continuous(range = c(0.2, 1)) +
    theme_minimal() +
    labs(title = title, color = "Zone",
         alpha = "Membership\nconfidence") +
    guides(color = guide_legend(
      override.aes = list(size = 3, alpha = 1)))
}

plot_spatial_zones <- function(res, title) {
  res$result_df %>%
    filter(!is.na(zone)) %>%
    ggplot(aes(x, y, fill = factor(zone))) +
    geom_raster() + coord_equal() + theme_minimal() +
    labs(title = title, fill = "Zone",
         x = "Easting", y = "Northing")
}

plot_confidence <- function(res, title) {
  res$result_df %>%
    ggplot(aes(x, y, fill = membership_prob)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1) +
    coord_equal() + theme_minimal() +
    labs(title = title, fill = "Confidence",
         x = "Easting", y = "Northing")
}

purrr::iwalk(mz_named_tuned, function(x, nm) {
  print(plot_umap_zones(x,    paste("UMAP —", nm)))
  print(plot_spatial_zones(x, paste("Zones —", nm)))
  print(plot_confidence(x,    paste("Confidence —", nm)))
})

# ── Zone assignment uncertainty summary per config ────────
# Summarises membership probability distribution per zone
# to flag zones with high proportions of low-confidence
# (noise-filled) pixels.

zone_membership_summary <- purrr::imap_dfr(mz_named_tuned,
  function(x, nm) {
    x$result_df %>%
      mutate(zone = if_else(is.na(zone),
                            as.integer(-1L), zone)) %>%
      group_by(zone) %>%
      summarise(
        n_pixels          = n(),
        mean_memb_prob    = round(mean(membership_prob,
                                       na.rm = TRUE), 3),
        pct_low_conf      = round(
          100 * mean(membership_prob < 0.3,
                     na.rm = TRUE), 1),
        .groups = "drop"
      ) %>%
      mutate(configuration = nm)
  }
)

print(zone_membership_summary, n = 40)
write_csv(zone_membership_summary,
          "../outputs/zone_membership_uncertainty.csv")
```
```{r}
#plot all final post-processed zones side by side
purrr::iwalk(pp_named, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat("Skipping", nm, "— no zones to plot\n")
    return(invisible(NULL))
  }
  df_plot <- as.data.frame(pp$zone_rast_smooth, xy = TRUE)

  names(df_plot) <- c("x", "y", "zone")
  df_plot <- df_plot %>% filter(!is.na(zone))
  
  p <- ggplot(df_plot, aes(x = x, y = y, fill = factor(zone))) +
    geom_raster() + coord_equal() + theme_minimal() +
    labs(title = paste("Final Zones —", nm), fill = "Zone",
         x = "Easting", y = "Northing")
  
  print(p)
})
```

# ============================================================
# PHASE 7 — Baseline Comparison: PCA + k-means
# ============================================================

```{r pca-kmeans-baseline}
run_pca_kmeans <- function(pred_df, methods = c("RF", "RK"),
                           soil_vars, include_lags = FALSE,
                           n_pcs = 2, k_range = 2:8,
                           label = "dataset") {
  
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  if (include_lags) {
    lag_cols  <- paste0(col_names, "_lag")
    col_names <- c(col_names, lag_cols[lag_cols %in% names(pred_df)])
  }
  col_names <- col_names[col_names %in% names(pred_df)]
  
  mat <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  mat_clean <- mat[complete_rows, , drop = FALSE]
  xy_clean <- pred_df %>% select(x, y) %>% slice(which(complete_rows))
  
  col_vars <- apply(mat_clean, 2, var, na.rm = TRUE)
  mat_clean <- mat_clean[, col_vars > 1e-10, drop = FALSE]
  
  mat_scaled <- scale(mat_clean)
  pca_out <- prcomp(mat_scaled, center = FALSE, scale. = FALSE)
  pca_scores <- pca_out$x[, seq_len(min(n_pcs, ncol(pca_out$x))), drop = FALSE]
  
  # elbow profile
  wss <- purrr::map_dbl(k_range, function(k) {
    set.seed(20260401)
    km <- kmeans(pca_scores, centers = k, nstart = 25, iter.max = 100)
    km$tot.withinss
  })
  
  # choose elbow by largest second derivative
  # (simple automatic elbow rule)
  d1 <- diff(wss)
  d2 <- diff(d1)
  
  if (length(d2) > 0) {
    best_k <- k_range[which.max(-d2) + 1]
  } else {
    best_k <- k_range[1]
  }
  
  cat(label, "— best k (elbow):", best_k, "\n")
  
  set.seed(20260401)
  km_best <- kmeans(pca_scores, centers = best_k, nstart = 25, iter.max = 100)
  
  result_df <- bind_cols(
    xy_clean,
    tibble(zone_kmeans = km_best$cluster)
  ) %>%
    mutate(dataset = label, method = "PCA_kmeans")
  
  list(
    result_df   = result_df,
    pca         = pca_out,
    km          = km_best,
    elbow_profile = tibble(k = k_range, wss = wss),
    best_k      = best_k
  )
}

baseline_SP_RF        <- run_pca_kmeans(pred_stack_SP,   "RF", soil_vars, label = "SP_RF")
baseline_SP_RK        <- run_pca_kmeans(pred_stack_SP,   "RK", soil_vars, label = "SP_RK")
baseline_SP_OK        <- run_pca_kmeans(pred_stack_SP,   "OK", soil_vars, label = "SP_OK")
baseline_COMP_RF      <- run_pca_kmeans(pred_stack_COMP, "RF", soil_vars, label = "COMP_RF")
baseline_COMP_RK      <- run_pca_kmeans(pred_stack_COMP, "RK", soil_vars, label = "COMP_RK")
baseline_COMP_OK      <- run_pca_kmeans(pred_stack_COMP, "OK", soil_vars, label = "COMP_OK")
baseline_SP_RF_lags   <- run_pca_kmeans(pred_stack_SP_lag,   "RF", soil_vars,
                                        include_lags = TRUE, label = "SP_RF_lags_kmeans")
baseline_SP_RK_lags   <- run_pca_kmeans(pred_stack_SP_lag,   "RK", soil_vars,
                                        include_lags = TRUE, label = "SP_RK_lags_kmeans")
baseline_COMP_RF_lags <- run_pca_kmeans(pred_stack_COMP_lag, "RF", soil_vars,
                                        include_lags = TRUE, label = "COMP_RF_lags_kmeans")
baseline_COMP_RK_lags <- run_pca_kmeans(pred_stack_COMP_lag, "RK", soil_vars,
                                        include_lags = TRUE, label = "COMP_RK_lags_kmeans")

# Silhouette profiles
# ── PCA+k-means elbow profiles with best-k indicated in legend ──
elbow_combined <- bind_rows(
  baseline_SP_RF$elbow_profile        %>% mutate(config = "SP_RF",
                                                 best_k = baseline_SP_RF$best_k),
  baseline_SP_RK$elbow_profile        %>% mutate(config = "SP_RK",
                                                 best_k = baseline_SP_RK$best_k),
  baseline_SP_OK$elbow_profile        %>% mutate(config = "SP_OK",
                                                 best_k = baseline_SP_OK$best_k),
  baseline_COMP_RF$elbow_profile      %>% mutate(config = "COMP_RF",
                                                 best_k = baseline_COMP_RF$best_k),
  baseline_COMP_RK$elbow_profile      %>% mutate(config = "COMP_RK",
                                                 best_k = baseline_COMP_RK$best_k),
  baseline_COMP_OK$elbow_profile      %>% mutate(config = "COMP_OK",
                                                 best_k = baseline_COMP_OK$best_k),
  baseline_SP_RF_lags$elbow_profile   %>% mutate(config = "SP_RF_lags",
                                                 best_k = baseline_SP_RF_lags$best_k),
  baseline_SP_RK_lags$elbow_profile   %>% mutate(config = "SP_RK_lags",
                                                 best_k = baseline_SP_RK_lags$best_k),
  baseline_COMP_RF_lags$elbow_profile %>% mutate(config = "COMP_RF_lags",
                                                 best_k = baseline_COMP_RF_lags$best_k),
  baseline_COMP_RK_lags$elbow_profile %>% mutate(config = "COMP_RK_lags",
                                                 best_k = baseline_COMP_RK_lags$best_k)
) %>%
  # Build a display label that shows the selected k alongside the config name
  mutate(config_label = paste0(config, " (k = ", best_k, ")"))

# Filter to just the best-k rows for the marker overlay
best_k_points <- elbow_combined %>%
  filter(k == best_k)

p_elbow <- elbow_combined %>%
  ggplot(aes(x = k, y = wss, color = config_label, group = config_label)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.8) +

  scale_x_continuous(breaks = 2:8) +
  theme_minimal(base_size = 10) +
  theme(legend.position  = "right",
        legend.text      = element_text(size = 9),
        panel.grid.minor = element_blank()) +
  labs(title    = "PCA+k-means elbow profiles",
       subtitle = "Black stars mark the selected k (second-derivative elbow rule)",
       x        = "Number of clusters (k)",
       y        = "Total within-cluster sum of squares",
       color    = "Configuration (selected k)")

print(p_elbow)

ggsave("../outputs/fig_pca_elbow_profiles.png",
       p_elbow, width = 9.5, height = 6, dpi = 300)
```

```{r pca-kmeans-postprocess}
post_process_kmeans <- function(result_df,
                                 zone_rast_crs = "EPSG:26917") {
  n_total   <- nrow(result_df)
  zone_rast <- result_df %>%
    select(x, y, zone = zone_kmeans) %>%
    rast(type = "xyz", crs = zone_rast_crs)

  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)
  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("PCA+kmeans: pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "%)\n")

  list(result_df_filled = result_df %>% mutate(zone = zone_kmeans),
       zone_rast_raw    = zone_rast,
       zone_rast_smooth = zone_rast_smooth)
}

pp_baseline <- list(
  SP_RF_kmeans        = post_process_kmeans(baseline_SP_RF$result_df),
  SP_RK_kmeans        = post_process_kmeans(baseline_SP_RK$result_df),
  SP_OK_kmeans        = post_process_kmeans(baseline_SP_OK$result_df),
  COMP_RF_kmeans      = post_process_kmeans(baseline_COMP_RF$result_df),
  COMP_RK_kmeans      = post_process_kmeans(baseline_COMP_RK$result_df),
  COMP_OK_kmeans      = post_process_kmeans(baseline_COMP_OK$result_df),
  SP_RF_lags_kmeans   = post_process_kmeans(baseline_SP_RF_lags$result_df),
  SP_RK_lags_kmeans   = post_process_kmeans(baseline_SP_RK_lags$result_df),
  COMP_RF_lags_kmeans = post_process_kmeans(baseline_COMP_RF_lags$result_df),
  COMP_RK_lags_kmeans = post_process_kmeans(baseline_COMP_RK_lags$result_df)
)

purrr::iwalk(pp_baseline, function(pp, nm) {
  writeRaster(pp$zone_rast_smooth,
              paste0("../outputs/zones_", nm, "_smooth.tif"),
              overwrite = TRUE)
  cat("Saved:", nm, "\n")
})

#plot PCA+kmeans zones
# plot all PCA+kmeans smooth zone rasters from pp_baseline
plot_all_pca_zone_rasters <- function(pp_list, ncol_plot = 3) {
  n_maps <- length(pp_list)
  nrow_plot <- ceiling(n_maps / ncol_plot)
  
  op <- par(mfrow = c(nrow_plot, ncol_plot), mar = c(2, 2, 3, 5))
  
  for (nm in names(pp_list)) {
    r <- pp_list[[nm]]$zone_rast_smooth
    
    if (!is.null(r)) {
      plot(r, main = nm, type = "classes")
    } else {
      plot.new()
      title(main = paste(nm, "\nNo raster"))
    }
  }
  
  par(op)
}

plot_all_pca_zone_rasters(pp_baseline)

pca_zone_files <- list.files(
  "../outputs",
  pattern = "^zones_.*_kmeans_smooth\\.tif$",
  full.names = TRUE
)

pca_zone_rasters <- setNames(
  lapply(pca_zone_files, terra::rast),
  tools::file_path_sans_ext(basename(pca_zone_files))
)

plot_all_pca_zone_rasters_disk <- function(rast_list, ncol_plot = 3) {
  n_maps <- length(rast_list)
  nrow_plot <- ceiling(n_maps / ncol_plot)
  
  op <- par(mfrow = c(nrow_plot, ncol_plot), mar = c(2, 2, 3, 5))
  
  for (nm in names(rast_list)) {
    plot(rast_list[[nm]], main = nm, type = "classes")
  }
  
  par(op)
}

plot_all_pca_zone_rasters_disk(pca_zone_rasters)



```

# ============================================================
# PHASE 8 — Zone Validation Metrics
# ============================================================

```{r zone-validation}
calc_variance_reduction <- function(pred_df, zone_vec,
                                     soil_vars, methods) {
  col_names  <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names  <- col_names[col_names %in% names(pred_df)]
  mat        <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  col_ranges <- apply(mat, 2, function(x)
                       diff(range(x, na.rm = TRUE)))
  mat        <- mat[, col_ranges > 1e-10, drop = FALSE]

  keep      <- !is.na(zone_vec) & complete.cases(mat)
  mat_k     <- mat[keep, ]
  zones_k   <- zone_vec[keep]
  total_var <- mean(apply(mat_k, 2, var, na.rm = TRUE))
  within_var <- mean(sapply(unique(zones_k), function(z) {
    mean(apply(mat_k[zones_k == z, , drop = FALSE],
               2, var, na.rm = TRUE))
  }))
  1 - (within_var / total_var)
}

calc_silhouette_sub <- function(emb_df, zone_vec, n_sub = 5000) {
  keep  <- !is.na(zone_vec)
  emb_k <- emb_df[keep, ]
  z_k   <- zone_vec[keep]
  set.seed(20260401)
  idx   <- sample(nrow(emb_k), min(n_sub, nrow(emb_k)))
  sil   <- cluster::silhouette(z_k[idx], dist(emb_k[idx, ]))
  mean(sil[, 3])
}

# UMAP VR + silhouette
zone_validation_umap <- purrr::imap_dfr(mz_named_tuned,
  function(x, nm) {
    parts  <- strsplit(nm, "_")[[1]]
    samp   <- parts[1]; method <- parts[2]
    lag_yn <- length(parts) >= 3 && parts[3] == "lags"

    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = pred_stack_SP,   SP_RK = pred_stack_SP,
      SP_OK   = pred_stack_SP,   COMP_RF = pred_stack_COMP,
      COMP_RK = pred_stack_COMP, COMP_OK = pred_stack_COMP)

    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = lag_yn,
      approach      = "UMAP+HDBSCAN",
      VR = calc_variance_reduction(
        pred_df, x$result_df$zone, soil_vars, method),
      silhouette = calc_silhouette_sub(
        x$result_df %>% select(UMAP1, UMAP2),
        x$result_df$zone)
    )
  })

# PCA+kmeans VR + silhouette
baseline_list <- list(
  SP_RF_kmeans        = baseline_SP_RF,
  SP_RK_kmeans        = baseline_SP_RK,
  SP_OK_kmeans        = baseline_SP_OK,
  COMP_RF_kmeans      = baseline_COMP_RF,
  COMP_RK_kmeans      = baseline_COMP_RK,
  COMP_OK_kmeans      = baseline_COMP_OK,
  SP_RF_lags_kmeans   = baseline_SP_RF_lags,
  SP_RK_lags_kmeans   = baseline_SP_RK_lags,
  COMP_RF_lags_kmeans = baseline_COMP_RF_lags,
  COMP_RK_lags_kmeans = baseline_COMP_RK_lags
)

zone_validation_pca <- purrr::imap_dfr(baseline_list,
  function(b, nm) {
    clean   <- str_remove(nm, "_kmeans")
    has_lag <- str_detect(clean, "_lags")
    clean   <- str_remove(clean, "_lags")
    parts   <- strsplit(clean, "_")[[1]]
    samp    <- parts[1]; method <- parts[2]

    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_OK   = pred_stack_SP,
      COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_OK = pred_stack_COMP)

    ok_cols <- pred_df %>%
      select(any_of(paste0(soil_vars, "_", method))) %>% na.omit()
    col_vars <- apply(ok_cols, 2, var, na.rm = TRUE)
    ok_cols  <- ok_cols[, col_vars > 1e-10, drop = FALSE]

    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = has_lag,
      approach      = "PCA+kmeans",
      VR = calc_variance_reduction(
        pred_df, b$result_df$zone_kmeans, soil_vars, method),
      silhouette = calc_silhouette_sub(
        prcomp(scale(ok_cols), center = FALSE)$x[, 1:2],
        b$result_df$zone_kmeans)
    )
  })

zone_validation_all <- bind_rows(zone_validation_umap,
                                   zone_validation_pca) %>%
  arrange(sampling, pred_method, lags, approach)

print(zone_validation_all)
write_csv(zone_validation_all,
          "../outputs/zone_validation_metrics.csv")
```

# ============================================================
# PHASE 8b — Agronomic Validation (held-out ap_val_sf)
# ============================================================

```{r agronomic-validation}
# Load validate_zones_with_holdout from Chapter 1 workspace
validate_zones_with_holdout <- function(zone_rast_smooth, val_sf,
                                        pred_stack, soil_vars,
                                        methods = c("RF", "RK"),
                                        label = "dataset") {

  # ── 1. Extract zone for each validation point ─────────────
  val_vect    <- terra::vect(val_sf)
  val_zone    <- terra::extract(zone_rast_smooth, val_vect)
  val_sf$zone <- val_zone[, 2]

  n_assigned  <- sum(!is.na(val_sf$zone))
  cat(label, "— validation points with zone assignment:",
      n_assigned, "of", nrow(val_sf), "\n")

# Replace step 2 inside validate_zones_with_holdout

  # ── 2. Per-zone summary of PREDICTED values ───────────────
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_stack)]

  # Extract zone from smoothed raster at pred_stack coordinates
  # rather than joining by x/y — avoids floating point mismatch
  pred_pts  <- terra::vect(
    pred_stack %>% select(x, y),
    geom = c("x", "y"), crs = "EPSG:26917"
  )
  pred_zone_extract <- terra::extract(zone_rast_smooth, pred_pts, ID = FALSE)
  pred_stack_zoned  <- pred_stack %>%
    mutate(zone = pred_zone_extract[[1]])

  zone_profiles <- pred_stack_zoned %>%
    filter(!is.na(zone)) %>%
    group_by(zone) %>%
    summarise(across(all_of(col_names), mean, na.rm = TRUE),
              .groups = "drop")

  # ── 3. Compare observed val points to zone predicted means ─
  val_long <- val_sf %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    pivot_longer(-zone, names_to = "variable", values_to = "obs")

  # FIX: use the actual method suffix, not hardcoded "_RF"
  method_suffix <- paste0("_", methods[1])   # "_RF" or "_RK"

  pred_long <- zone_profiles %>%
    select(zone, ends_with(method_suffix)) %>%
    pivot_longer(-zone, names_to = "col", values_to = "pred_zone_mean") %>%
    mutate(variable = str_remove(col, method_suffix)) %>%
    select(zone, variable, pred_zone_mean)

  comparison <- val_long %>%
    left_join(pred_long, by = c("zone", "variable")) %>%
    group_by(zone, variable) %>%
    summarise(
      n_val_pts      = n(),
      mean_obs       = mean(obs,                        na.rm = TRUE),
      pred_zone_mean = first(pred_zone_mean),
      bias           = mean(obs - pred_zone_mean,       na.rm = TRUE),
      rmse_vs_zone   = sqrt(mean((obs - pred_zone_mean)^2, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(dataset = label)

  # ── 4. Kruskal-Wallis: do zones differ in observed values? ─
  kw_results <- purrr::map_dfr(soil_vars, function(v) {
    df <- val_sf %>%
      st_drop_geometry() %>%
      filter(!is.na(zone), !is.na(.data[[v]])) %>%
      select(zone, value = all_of(v))

    if (length(unique(df$zone)) < 2 || nrow(df) < 5) {
      return(tibble(variable = v, kw_p = NA_real_, dataset = label))
    }

    kt <- kruskal.test(value ~ factor(zone), data = df)
    tibble(variable = v, kw_p = kt$p.value, dataset = label)
  })

  list(
    comparison     = comparison,
    kw_results     = kw_results,
    val_with_zones = val_sf
  )
}
get_pred_stack <- function(nm) {
  clean   <- nm %>%
    str_remove("_kmeans") %>%
    str_remove("_nolags") %>%
    str_remove("_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]; method <- parts[2]
  has_lag <- str_detect(nm, "_lags")

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_OK = pred_stack_COMP)

  list(pred_df = pred_df, method = method,
       samp = samp, has_lag = has_lag)
}

# validate_zones_with_holdout() — must already be defined
# in workspace or sourced separately.
# Rehydrate UMAP+HDBSCAN smoothed zone rasters from disk
pp_named <- purrr::imap(pp_named, function(pp, nm) {
  tif_path <- paste0("../outputs/zones_", nm, "_smooth.tif")
  if (file.exists(tif_path)) {
    pp$zone_rast_smooth <- terra::rast(tif_path)
    pp$zone_rast_raw    <- terra::rast(tif_path)  # if needed downstream
  }
  pp
})

# Rehydrate PCA+kmeans smoothed zone rasters from disk
pp_baseline <- purrr::imap(pp_baseline, function(pp, nm) {
  # Note: after your rename, the file path matches the new key directly
  tif_path <- paste0("../outputs/zones_", nm, "_smooth.tif")
  if (file.exists(tif_path)) {
    pp$zone_rast_smooth <- terra::rast(tif_path)
  }
  pp
})

# Also rehydrate the validation data and pred stacks if they have rasters
ap_val_sf  # this is sf, should be fine
val_umap <- purrr::imap(pp_named, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- UMAP+HDBSCAN:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method, label = nm)
})
val_umap <- val_umap[!sapply(val_umap, is.null)]

val_kmeans <- purrr::imap(pp_baseline, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- PCA+kmeans:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method,
    label   = nm)
})
val_kmeans <- val_kmeans[!sapply(val_kmeans, is.null)]

# Add OK configs missed in original loop
# val_ok_missing <- list(
#   SP_OK_nolags   = validate_zones_with_holdout(
#     pp_named[["SP_OK_nolags"]]$zone_rast_smooth,
#     ap_val_sf, pred_stack_SP, soil_vars,
#     methods = "OK", label = "SP_OK_nolags"),
#   COMP_OK_nolags = validate_zones_with_holdout(
#     pp_named[["COMP_OK_nolags"]]$zone_rast_smooth,
#     ap_val_sf, pred_stack_COMP, soil_vars,
#     methods = "OK", label = "COMP_OK_nolags")
# )

val_named <- val_umap  # all 10 configs including OK
all_val   <- c(val_named, val_kmeans)

all_kw <- purrr::map_dfr(all_val, ~ .x$kw_results) %>%
  mutate(
    significant = kw_p < 0.05,
    approach    = if_else(str_detect(dataset, "_kmeans"),
                          "PCA+kmeans", "UMAP+HDBSCAN"),
    clean       = dataset %>%
                    str_remove("_kmeans") %>%
                    str_remove("_nolags") %>%
                    str_remove("_lags"),
    sampling    = str_extract(clean, "^(SP|COMP)"),
    pred_method = str_extract(clean, "(RF|RK|OK)"),
    lags        = str_detect(dataset, "_lags")
  ) %>%
  select(-clean)

all_comparisons      <- purrr::map_dfr(val_umap,   ~ .x$comparison)
all_comparisons_pca  <- purrr::map_dfr(val_kmeans, ~ .x$comparison)
all_comparisons_full <- bind_rows(all_comparisons, all_comparisons_pca)

kw_summary <- all_kw %>%
  group_by(dataset, approach, sampling, pred_method, lags) %>%
  summarise(
    n_significant   = sum(significant, na.rm = TRUE),
    n_total         = n(),
    pct_significant = round(100 * mean(significant, na.rm = TRUE)),
    sig_vars        = paste(variable[significant], collapse = ", "),
    .groups         = "drop"
  ) %>%
  arrange(approach, sampling, pred_method, lags)

print(kw_summary, n = 30)
write_csv(all_kw,      "../outputs/zone_kruskal_wallis_all.csv")
write_csv(kw_summary,  "../outputs/kw_summary_all.csv")
```

# ============================================================
# PHASE 8c — Per-Variable Observed VR + Effect Sizes
# ============================================================

```{r observed-vr-and-effect-sizes}
calc_zone_consistency <- function(val_with_zones,
                                   soil_vars, label) {
  df <- val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    na.omit()

  if (nrow(df) < 5 || length(unique(df$zone)) < 2) {
    return(tibble(configuration = label,
                  observed_VR_mean = NA_real_,
                  observed_VR_med  = NA_real_,
                  n_pos_VR         = NA_integer_))
  }

  total_var_vec <- df %>%
    select(all_of(soil_vars)) %>%
    summarise(across(everything(),
                     ~ var(.x, na.rm = TRUE))) %>%
    unlist()

  per_var_vr <- purrr::map_dbl(soil_vars, function(v) {
    within_var <- df %>%
      group_by(zone) %>%
      summarise(wv = var(.data[[v]], na.rm = TRUE),
                .groups = "drop") %>%
      pull(wv) %>%
      mean(na.rm = TRUE)
    1 - (within_var / total_var_vec[[v]])
  })
  names(per_var_vr) <- soil_vars

  tibble(configuration    = label,
         observed_VR_mean = round(mean(per_var_vr,   na.rm = TRUE), 3),
         observed_VR_med  = round(median(per_var_vr, na.rm = TRUE), 3),
         n_pos_VR         = sum(per_var_vr > 0, na.rm = TRUE))
}

obs_vr_all <- bind_rows(
  purrr::imap_dfr(val_named,  ~ calc_zone_consistency(
    .x$val_with_zones, soil_vars, .y)) %>%
    rename(config_join = configuration),
  purrr::imap_dfr(val_kmeans, ~ calc_zone_consistency(
    .x$val_with_zones, soil_vars, .y)) %>%
    rename(config_join = configuration)
)

# Zone prediction RMSE
zone_pred_rmse <- all_comparisons_full %>%
  filter(!is.na(rmse_vs_zone)) %>%
  group_by(dataset) %>%
  summarise(
    mean_zone_rmse   = round(mean(rmse_vs_zone,   na.rm = TRUE), 3),
    median_zone_rmse = round(median(rmse_vs_zone, na.rm = TRUE), 3),
    .groups          = "drop"
  ) %>%
  rename(config_join = dataset)

# R²
#old:
zone_r2 <- all_comparisons_full %>%
  filter(!is.na(pred_zone_mean), !is.na(mean_obs)) %>%
  group_by(dataset) %>%
  summarise(
    r2_obs_pred = round(
      cor(mean_obs, pred_zone_mean, use = "complete.obs")^2, 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)
# R²
#new:
# R² — coefficient of determination at the POINT level
# Each validation point's observed value vs the zone-mean predicted
# value for that point's assigned zone. Computed per variable, then
# aggregated by median across variables.
zone_r2 <- point_pairs_all %>%
  group_by(configuration, variable) %>%
  summarise(
    n_pts  = n(),
    sse    = sum((obs - pred_zone_mean)^2, na.rm = TRUE),
    sst    = sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE),
    r2_var = ifelse(sst > 0 & n_pts >= 5, 1 - sse / sst, NA_real_),
    .groups = "drop"
  ) %>%
  group_by(configuration) %>%
  summarise(
    r2_obs_pred = round(median(r2_var, na.rm = TRUE), 3),
    n_vars_r2   = sum(!is.na(r2_var)),
    .groups     = "drop"
  ) %>%
  rename(config_join = configuration)

cat("R² aggregation check:\n")
print(zone_r2)

# Diagnostic — confirm all configs have R² computed from enough variables
cat("R² aggregation check:\n")
print(zone_r2)
# Per-variable observed VR for heatmap
obs_vr_by_var_all <- bind_rows(
  purrr::imap_dfr(val_named, function(x, nm) {
    df <- x$val_with_zones %>% st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>% na.omit()
    if (nrow(df) < 5 || length(unique(df$zone)) < 2) return(NULL)
    tv <- df %>% select(all_of(soil_vars)) %>%
      summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
      unlist()
    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>% group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
        pull(wv) %>% mean(na.rm = TRUE)
      tibble(configuration = nm, variable = v,
             obs_VR = round(1 - wv / tv[[v]], 3))
    })
  }),
  purrr::imap_dfr(val_kmeans, function(x, nm) {
    df <- x$val_with_zones %>% st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>% na.omit()
    if (nrow(df) < 5 || length(unique(df$zone)) < 2) return(NULL)
    tv <- df %>% select(all_of(soil_vars)) %>%
      summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
      unlist()
    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>% group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
        pull(wv) %>% mean(na.rm = TRUE)
      tibble(configuration = paste0(nm), variable = v,
             obs_VR = round(1 - wv / tv[[v]], 3))
    })
  })
)

# Supplementary heatmap
p_obs_vr <- obs_vr_by_var_all %>%
  left_join(
    bind_rows(
      zone_validation_umap %>%
        select(configuration, clustering = approach),
      zone_validation_pca  %>%
        mutate(configuration = if_else(
          !str_detect(configuration, "_kmeans"),
          paste0(configuration, "_kmeans"),
          configuration)) %>%
        select(configuration, clustering = approach)
    ),
    by = "configuration"
  ) %>%
  mutate(
    config_label = str_replace_all(
      configuration,
      c("_nolags" = "", "_lags" = "+lags",
            "_kmeans" = "",  "COMP_" = "COMP/", "SP_" = "SP/")),
              clustering = if_else(
                is.na(clustering),
                if_else(str_detect(configuration, "_kmeans"),
                        "PCA+kmeans",
                        "UMAP+HDBSCAN"), clustering),
    approach   = factor(clustering,
                        levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = config_label, y = variable, fill = obs_VR)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", obs_VR)),
            size = 2.2) +
  facet_wrap(~ approach, scales = "free_x") +
  scale_fill_gradient2(
    low = "#d73027", mid = "#ffffbf", high = "#1a9850",
    midpoint = 0, limits = c(-1, 1),
    oob  = scales::squish, name = "Observed VR"
  ) +
  theme_bw(base_size = 9) +
  theme(axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
        strip.background = element_rect(fill = "grey92"),
        strip.text       = element_text(face = "bold"),
        panel.grid       = element_blank()) +
  labs(
    title    = "Supplementary Figure S1. Per-variable observed variance reduction",
    subtitle = "Held-out validation points (n = 59), fully independent of prediction surfaces",
    x = NULL, y = "Soil variable"
  )

print(p_obs_vr)
ggsave("../outputs/figS1_observed_vr_heatmap.png",
       p_obs_vr, width = 12, height = 4, dpi = 300)
write_csv(obs_vr_by_var_all,
          "../outputs/observed_vr_per_variable.csv")
```

# ============================================================
# Final Summary + Results Table
# ============================================================

```{r final-summary-and-table}
# ============================================================
# Final Summary + Results Table — focused metric set
# ============================================================
# Metrics retained:
#   Configuration block:  Sampling, Method, Lags, k
#   Zone structure:       VR_pred
#   Agronomic validation: N_sig, VR_obs, N_pos, Coverage
#   Overall:              Rank
#
# Metrics computed but excluded (reported in discussion):
#   Silhouette  — UMAP and PCA spaces not directly comparable
#   R²          — point-level metric, inappropriate for zone validation
#   RMSE_zone   — penalizes within-zone variation (a soil feature, not a zone failure)
#   IOA         — narrow range, redundant with VR_obs / Coverage
#   CV-RMSE     — narrow range, discussion-only finding (all values > 1.0)

# ── Build summary tables ───────────────────────────────────
summary_umap <- purrr::imap_dfr(mz_named_tuned, function(x, nm) {
  parts  <- strsplit(nm, "_")[[1]]
  samp   <- parts[1]; method <- parts[2]
  lag_yn <- length(parts) >= 3 && parts[3] == "lags"
  tibble(configuration   = nm,
         clustering      = "UMAP+HDBSCAN",
         sampling_design = ifelse(samp == "SP",
                                  "Single-point", "Composite"),
         pred_method     = method,
         spatial_lags    = lag_yn,
         n_zones         = x$n_zones)
}) %>%
  left_join(
    all_kw %>% filter(approach == "UMAP+HDBSCAN") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                sig_vars   = paste(variable[significant], collapse = ", "),
                .groups    = "drop") %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(n_sig_vars = replace_na(n_sig_vars, 0L),
         sig_vars   = ifelse(is.na(sig_vars) | sig_vars == "",
                             "none", sig_vars)) %>%
  left_join(
    zone_validation_all %>% filter(approach == "UMAP+HDBSCAN") %>%
      select(configuration, VR),
    by = "configuration"
  )

summary_pca <- purrr::imap_dfr(baseline_list, function(b, nm) {
  clean   <- str_remove(nm, "_kmeans")
  has_lag <- str_detect(clean, "_lags")
  clean   <- str_remove(clean, "_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]; method <- parts[2]
  tibble(configuration   = nm,
         clustering      = "PCA+kmeans",
         sampling_design = ifelse(samp == "SP",
                                  "Single-point", "Composite"),
         pred_method     = method,
         spatial_lags    = has_lag,
         n_zones         = b$best_k)
}) %>%
  left_join(
    all_kw %>% filter(approach == "PCA+kmeans") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                sig_vars   = paste(variable[significant], collapse = ", "),
                .groups    = "drop") %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(n_sig_vars = replace_na(n_sig_vars, 0L),
         sig_vars   = ifelse(is.na(sig_vars) | sig_vars == "",
                             "none", sig_vars)) %>%
  left_join(
    zone_validation_all %>% filter(approach == "PCA+kmeans") %>%
      select(configuration, VR),
    by = "configuration"
  )

# ── Join all metrics ───────────────────────────────────────
final_summary <- bind_rows(summary_umap, summary_pca) %>%
  mutate(
    config_join = case_when(
      clustering == "PCA+kmeans" &
        !str_detect(configuration, "_kmeans") ~
        paste0(configuration, "_kmeans"),
      TRUE ~ configuration
    )
  ) %>%
  left_join(obs_vr_all,    by = "config_join") %>%
  left_join(zone_coverage, by = "config_join") %>%
  select(-config_join)

# ── Composite Rank — 5 metrics, all in the final table ─────
final_ranked <- final_summary %>%
  mutate(
    vr_rank     = rank(-VR,                na.last = "keep"),
    kw_rank     = rank(-n_sig_vars,        na.last = "keep"),
    obs_vr_rank = rank(-observed_VR_med,   na.last = "keep"),
    npos_rank   = rank(-n_pos_VR,          na.last = "keep"),
    cov_rank    = rank(-pct_in_full_range, na.last = "keep"),
    Rank = round(vr_rank + kw_rank + obs_vr_rank +
                 npos_rank + cov_rank, 1)
  ) %>%
  select(configuration, Rank)

cat("\n=== Composite ranking (5-metric) ===\n")
final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  arrange(Rank) %>%
  select(configuration, clustering, n_zones, VR, n_sig_vars,
         observed_VR_med, n_pos_VR, pct_in_full_range, Rank) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  print(n = 20)

write_csv(final_summary,
          "../outputs/final_results_summary_complete.csv")

# ── gt Results Table ───────────────────────────────────────
table_data <- final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  mutate(
    Approach = clustering,
    Sampling = ifelse(sampling_design == "Single-point", "SP", "COMP"),
    Method   = pred_method,
    Lags     = ifelse(spatial_lags, "Yes", "No"),
    Zones    = n_zones,
    VR_pred  = round(VR,                3),
    N_sig    = n_sig_vars,
    VR_obs   = round(observed_VR_med,   3),
    N_pos    = n_pos_VR,
    Coverage = round(pct_in_full_range, 1)
  ) %>%
  select(Approach, Sampling, Method, Lags, Zones,
         VR_pred, N_sig, VR_obs, N_pos, Coverage, Rank) %>%
  arrange(Approach, Sampling, Method, Lags)

tbl <- table_data %>%
  gt(groupname_col = "Approach") %>%
  cols_label(
    Sampling = "Sampling",
    Method   = "Pred. method",
    Lags     = "Lags",
    Zones    = md("*k*"),
    VR_pred  = md("VR~pred~"),
    N_sig    = md("*N*~sig~"),
    VR_obs   = md("VR~obs~ (median)"),
    N_pos    = md("*N*~pos~"),
    Coverage = "Coverage (%)",
    Rank     = "Rank"
  ) %>%
  tab_spanner(label   = md("**Configuration**"),
              columns = c(Sampling, Method, Lags, Zones)) %>%
  tab_spanner(label   = md("**Zone structure** *(all grid pixels)*"),
              columns = VR_pred) %>%
  tab_spanner(label   = md("**Agronomic validation** *(held-out, n = 59)*"),
              columns = c(N_sig, VR_obs, N_pos, Coverage)) %>%
  tab_spanner(label   = md("**Overall**"),
              columns = Rank) %>%
  tab_style(
    style     = list(cell_fill(color = "#d6eaf8"),
                     cell_text(weight = "bold")),
    locations = cells_body(
      rows = Sampling == "COMP" & Method == "RF" &
             Lags == "Yes" & Approach == "UMAP+HDBSCAN")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fde8e8"),
    locations = cells_body(columns = VR_obs,
                           rows = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  tab_style(
    style     = cell_text(color = "#c0392b"),
    locations = cells_body(columns = VR_obs,
                           rows = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—") %>%
  fmt_number(columns = c(VR_pred, VR_obs),  decimals = 3) %>%
  fmt_number(columns = Coverage,            decimals = 1) %>%
  fmt_number(columns = Rank,                decimals = 1) %>%
  fmt_integer(columns = c(Zones, N_sig, N_pos)) %>%
  tab_caption(
    md("**Table 1.** Zone validation metrics for all 20 management zone configurations. SP = single-point; COMP = composite; RF = random forest; RK = regression kriging; OK = ordinary kriging. *k* = zones after post-processing. VR~pred~ = variance reduction on the predicted soil surface (higher = more internally homogeneous zones). *N*~sig~ = number of soil variables (of 9) showing significant between-zone differences in held-out observations (Kruskal-Wallis, p < 0.05). VR~obs~ = median variance reduction across variables computed on held-out observations. *N*~pos~ = number of variables with positive VR~obs~. Coverage = % of held-out points whose observed values fall within their assigned zone's predicted value range. Rank = composite ordering across the five metrics (lower = better). Blue row = best overall configuration. Red = negative VR~obs~.")
  ) %>%
  tab_options(
    table.font.size                   = px(11),
    column_labels.font.size           = px(11),
    table.width                       = pct(100),
    data_row.padding                  = px(4),
    table.border.top.style            = "solid",
    table.border.top.width            = px(2),
    table.border.top.color            = "#2c3e50",
    column_labels.border.bottom.style = "solid",
    column_labels.border.bottom.width = px(1),
    column_labels.border.bottom.color = "#7f8c8d"
  )

tbl
gt::gtsave(tbl, "../outputs/table1_zone_metrics.html")
cat("Table saved.\n")

save.image("../outputs/workspace_final.RData")

#UMAP is more sensitive to variable reduction than PCA+kmeans — PCA+kmeans performance is relatively stable across both variable sets, while UMAP configurations shift substantially. This makes sense because UMAP's manifold learning is more dependent on the full multivariate structure of the input space
# Lag augmentation loses its benefit — COMP_RF_lags was your Chapter 1 winner but drops to 0 sig vars here. The spatial lag features may have been capturing structure in the dropped variables (lbc_1, mn, zn) more than in the agronomic subset
# SP configurations become more competitive — SP_RF_nolags and SP_RK_nolags now match or beat COMP configs on n_sig, suggesting composite sampling's advantage diminishes when the variable set is reduced
# The best agronomic config is SP_RF_nolags or SP_RK_nolags with 3 significant variables each
```



# ============================================================
# PHASE 9 — Three-panel thesis figure
# ============================================================

```{r thesis-figure}
fig_data <- final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  mutate(
    x_group = factor(
      paste0(pred_method, ifelse(spatial_lags, "+lags", "")),
      levels = c("RF", "RF+lags", "RK", "RK+lags", "OK")
    ),
    sampling_f   = factor(sampling_design,
                          levels = c("Single-point", "Composite")),
    clustering_f = factor(clustering,
                          levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  )

clust_colors <- c("PCA+kmeans"   = "#4E9AF1",
                  "UMAP+HDBSCAN" = "#E05C3A")
lag_alpha    <- c("No lags" = 0.5, "With lags" = 1.0)

lag_aes <- aes(
  x     = x_group,
  fill  = clustering_f,
  alpha = factor(spatial_lags,
                 levels = c(FALSE, TRUE),
                 labels = c("No lags", "With lags"))
)

base_theme <- theme_bw(base_size = 9) +
  theme(
    legend.position    = "none",
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(size = 8, face = "bold"),
    axis.text.x        = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.major.x = element_blank()
  )

pA <- fig_data %>%
  ggplot(lag_aes) + aes(y = VR) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.40, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha, name = "Spatial lags") +
  scale_y_continuous(limits = c(0, 0.85), breaks = seq(0, 0.8, 0.2)) +
  labs(title = "A).  Variance Reduction",
       subtitle = "Higher = zones more internally homogeneous",
       x = NULL, y = "VR") +
  base_theme

pB <- fig_data %>%
  ggplot(lag_aes) + aes(y = silhouette) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.50, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha, name = "Spatial lags") +
  scale_y_continuous(limits = c(0, 0.65), breaks = seq(0, 0.6, 0.2)) +
  labs(title = "B).  Silhouette Coefficient",
       subtitle = "Higher = zones more spatially distinct",
       x = NULL, y = "Silhouette") +
  base_theme

pC <- fig_data %>%
  ggplot(lag_aes) + aes(y = n_sig_vars) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_text(aes(label = ifelse(n_sig_vars > 0,
                                as.character(n_sig_vars), "")),
            position = position_dodge(0.8),
            vjust = -0.4, size = 2.5, color = "grey30") +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha, name = "Spatial lags") +
  scale_y_continuous(breaks = 0:5,
                     limits = c(0, length(soil_vars) + 0.5)) +
  labs(title = paste0("C). Significant Soil Properties (of ",
                       length(soil_vars), ") (KW, p < 0.05)"),
       subtitle = "Validation points: n = 59",
       x = "Prediction method",
       y = "# of properties significant") +
  base_theme

# Shared legend
legend_plot <- fig_data %>%
  ggplot(lag_aes) + aes(y = VR) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha,   name = "Spatial lags") +
  theme_bw(base_size = 9) +
  theme(legend.position  = "bottom",
        legend.box       = "horizontal",
        legend.title     = element_text(size = 8, face = "bold"),
        legend.text      = element_text(size = 8),
        legend.key.size  = unit(0.4, "cm"))

shared_legend <- cowplot::get_legend(legend_plot)

fig_combined <- cowplot::plot_grid(
  pA / pB / pC,
  shared_legend,
  ncol    = 1,
  rel_heights = c(10, 1)
)

print(fig_combined)
ggsave("../outputs/fig_zone_quality_metrics.png",
       fig_combined, width = 9, height = 11, dpi = 300)
cat("Figure saved.\n")
```

```{r}
# For a given config, correlate each soil var prediction with UMAP coords
umap_contributions <- purrr::map_dfr(soil_vars, function(v) {
  pred_col <- paste0(v, "_RF")  # or _RK, _OK
  umap_df  <- mz_named_tuned[["COMP_RF_lags"]]$result_df
  
  pred_vals <- pred_stack_COMP_lag %>%
    filter(complete.cases(.)) %>%
    pull(pred_col)
  
  tibble(
    variable = v,
    cor_UMAP1 = cor(pred_vals, umap_df$UMAP1, use = "complete.obs"),
    cor_UMAP2 = cor(pred_vals, umap_df$UMAP2, use = "complete.obs")
  )
})
print(umap_contributions)

# Drop one variable at a time, rerun UMAP, record VR
loo_results <- purrr::map_dfr(soil_vars, function(drop_var) {
  vars_subset <- soil_vars[soil_vars != drop_var]
  
  mz <- run_umap_hdbscan(
    pred_stack_COMP_lag,
    methods      = "RF",
    soil_vars    = vars_subset,
    include_lags = TRUE,
    label        = paste0("drop_", drop_var)
  )
  
  pp <- post_process_zones(mz)
  vr <- calc_variance_reduction(
    pred_stack_COMP_lag, mz$zone, vars_subset, "RF")
  
  tibble(dropped = drop_var, n_vars = length(vars_subset), VR = vr)
})
print(loo_results)

pred_cols <- paste0(soil_vars, "_RF")

pca_contributions <- pred_stack_COMP %>%
  select(all_of(pred_cols)) %>%
  na.omit() %>%
  scale() %>%
  prcomp()

# Loadings on first 2 PCs
loadings <- as.data.frame(pca_contributions$rotation[, 1:2]) %>%
  rownames_to_column("variable") %>%
  mutate(variable = str_remove(variable, "_RF"),
         loading_magnitude = sqrt(PC1^2 + PC2^2)) %>%
  arrange(desc(loading_magnitude))

print(umap_contributions)
print(loo_results)
print(loadings)
# pH and K are the primary drivers of zone structure (UMAP1), while the broad fertility variables (Ca, Mg, P, lb_ceq) collectively define UMAP2. Zn and Ca may be candidates for exclusion in future work as they appear to add noise rather than structure — but removing the full agronomic subset simultaneously collapsed both axes, explaining the Chapter 2 performance drop.
```

```{r}
#final figure plot
# ============================================================
# Final zone map figures — two separate panels (UMAP, PCA)
# Layout: rows = sampling (SP, COMP), columns = method (OK, RF, RK)
# Lag variants distinguished within RF and RK columns
# ============================================================

# ── Helper: load and convert a zone raster from disk ──────
make_zone_df_from_disk <- function(file_path, nm, approach) {
  if (!file.exists(file_path)) {
    cat("Skipping", nm, "— file not found\n")
    return(NULL)
  }
  r <- terra::rast(file_path)
  df <- as.data.frame(r, xy = TRUE)
  names(df) <- c("x", "y", "zone")
  df %>%
    filter(!is.na(zone)) %>%
    mutate(configuration = nm, approach = approach)
}

# ── Build file path lookups ───────────────────────────────
umap_configs <- names(pp_named)
pca_configs  <- names(pp_baseline)

umap_files <- setNames(
  paste0("../outputs/zones_", umap_configs, "_smooth.tif"),
  umap_configs)

pca_files <- setNames(
  paste0("../outputs/zones_", pca_configs, "_kmeans_smooth.tif"),
  pca_configs)

# ── Pull all rasters into tidy data frames ────────────────
zone_maps_umap <- purrr::imap_dfr(umap_files, function(fp, nm) {
  make_zone_df_from_disk(fp, nm, "UMAP+HDBSCAN")
})

zone_maps_pca <- purrr::imap_dfr(pca_files, function(fp, nm) {
  make_zone_df_from_disk(fp, nm, "PCA+kmeans")
})

# ── Parse configuration into Sampling / Method / Lags ─────
parse_config <- function(df) {
  df %>%
    mutate(
      clean    = configuration %>% 
                 str_remove("_kmeans$") %>%
                 str_remove("_nolags$"),
      has_lags = str_detect(clean, "_lags$"),
      clean    = str_remove(clean, "_lags$"),
      sampling = str_extract(clean, "^(SP|COMP)"),
      method   = str_extract(clean, "(OK|RF|RK)"),
      # Column header — combine method with lag variant
      method_label = case_when(
        method == "OK"          ~ "OK",
        method == "RF" & !has_lags ~ "RF (no lags)",
        method == "RF" &  has_lags ~ "RF (+ lags)",
        method == "RK" & !has_lags ~ "RK (no lags)",
        method == "RK" &  has_lags ~ "RK (+ lags)"
      ),
      method_label = factor(method_label, levels = c(
        "OK", "RF (no lags)", "RF (+ lags)", "RK (no lags)", "RK (+ lags)"
      )),
      sampling = factor(sampling, levels = c("SP", "COMP"))
    ) %>%
    select(-clean)
}

zone_maps_umap <- parse_config(zone_maps_umap)
zone_maps_pca  <- parse_config(zone_maps_pca)

# ── Colorblind-friendly palette (Okabe-Ito) ───────────────
okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#999999",
               "#000000")

all_zones <- sort(unique(c(
  as.character(zone_maps_umap$zone),
  as.character(zone_maps_pca$zone)
)))

zone_palette <- setNames(okabe_ito[seq_along(all_zones)], all_zones)

# ── Plotting function — shared layout for both figures ────
make_zone_figure <- function(df, title_text) {
  df %>%
    mutate(zone = factor(zone, levels = all_zones)) %>%
    ggplot(aes(x = x, y = y, fill = zone)) +
    geom_raster() +
    # Use fixed coordinate ratio without coord_equal so free scales work
    facet_grid(sampling ~ method_label, scales = "free", switch = "y") +
    scale_fill_manual(values  = zone_palette,
                      name    = "Zone",
                      na.value = "grey90",
                      drop    = FALSE) +
    theme_minimal(base_size = 9) +
    theme(
      strip.background.x = element_rect(fill = "grey92", color = NA),
      strip.background.y = element_rect(fill = "grey85", color = NA),
      strip.text.x       = element_text(face = "bold", size = 9),
      strip.text.y       = element_text(face = "bold", size = 10),
      strip.placement    = "outside",
      axis.text          = element_text(size = 6),
      axis.title         = element_text(size = 9),
      panel.spacing      = unit(0.7, "lines"),
      legend.position    = "bottom",
      legend.key.size    = unit(0.4, "cm"),
      panel.grid.minor   = element_blank(),
      plot.title         = element_text(face = "bold", size = 11)
    ) +
    labs(
      title = title_text,
      x = "Easting (m)", y = "Northing (m)"
    )
}

# ── UMAP figure ───────────────────────────────────────────
p_umap <- make_zone_figure(
  zone_maps_umap,
  "UMAP+HDBSCAN management zones — all configurations"
)

print(p_umap)
ggsave("../outputs/fig_zone_maps_umap.png",
       p_umap, width = 12, height = 6, dpi = 300)

# ── PCA figure ────────────────────────────────────────────
p_pca <- make_zone_figure(
  zone_maps_pca,
  "PCA+kmeans management zones — all configurations"
)

print(p_pca)
ggsave("../outputs/fig_zone_maps_pca.png",
       p_pca, width = 12, height = 6, dpi = 300)

cat("Both figures saved.\n")

# ============================================================
# UMAP feature-space figure — all configurations
# Same grid layout as the zone maps
# ============================================================

# ── Pull UMAP embedding + zone assignment for all configs ─
# mz_named_tuned contains result_df with UMAP1, UMAP2, zone
make_umap_df <- function(mz_obj, nm) {
  if (is.null(mz_obj$result_df)) return(NULL)
  mz_obj$result_df %>%
    select(UMAP1, UMAP2, zone, membership_prob) %>%
    mutate(configuration = nm)
}

umap_space_df <- purrr::imap_dfr(mz_named_tuned, make_umap_df) %>%
  parse_config()  # reuses the parse_config function from the zone map chunk

# ── UMAP feature-space figure ─────────────────────────────
p_umap_space <- umap_space_df %>%
  mutate(zone = factor(zone, levels = all_zones)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  # Noise/unassigned points first, so they sit behind the colored ones
  geom_point(data = . %>% filter(is.na(zone)),
             color = "grey75",
             size  = 0.25,#more translucent
             alpha = 0.25) +
  # Assigned-zone points on top
  geom_point(data = . %>% filter(!is.na(zone)),
             aes(color = zone, alpha = membership_prob),
             size = 0.3) +
  facet_grid(sampling ~ method_label, scales = "free", switch = "y") +
  scale_color_manual(values  = zone_palette,
                     name    = "Zone",
                     na.value = "grey75",
                     drop    = FALSE) +
  scale_alpha_continuous(range = c(0.2, 1),
                         name  = "Membership\nconfidence",
                         guide = guide_legend(override.aes = list(size = 2))) +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_minimal(base_size = 9) +
  theme(
    strip.background.x = element_rect(fill = "grey92", color = NA),
    strip.background.y = element_rect(fill = "grey85", color = NA),
    strip.text.x       = element_text(face = "bold", size = 9),
    strip.text.y       = element_text(face = "bold", size = 10),
    strip.placement    = "outside",
    axis.text          = element_text(size = 6),
    axis.title         = element_text(size = 9),
    panel.spacing      = unit(0.7, "lines"),
    legend.position    = "bottom",
    legend.box         = "horizontal",
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 11)
  ) +
  labs(
    title    = "UMAP feature space — all configurations",
    subtitle = "Colored points = HDBSCAN cluster assignment (zone). Grey = noise (unclustered). Transparency = membership confidence.",
    x = "UMAP1", y = "UMAP2"
  )

print(p_umap_space)
ggsave("../outputs/fig_umap_feature_space2.png",
       p_umap_space, width = 12, height = 6, dpi = 300)
cat("Saved UMAP feature space figure.\n")
# ============================================================
# PCA projection figure with eigenvectors — all configurations
# Same grid layout, colored by zone assignment
# ============================================================

# ── Helper: build PCA projection + loadings for one config ─
make_pca_projection_df <- function(baseline_obj, nm) {
  if (is.null(baseline_obj$pca) || is.null(baseline_obj$result_df)) {
    return(list(scores = NULL, loadings = NULL))
  }
  
  # Scores (PC coordinates per pixel)
  scores <- as.data.frame(baseline_obj$pca$x[, 1:2])
  names(scores) <- c("PC1", "PC2")
  scores$zone <- baseline_obj$result_df$zone_kmeans
  scores$configuration <- nm
  
  # Loadings (variable contributions to PC1, PC2)
  loadings <- as.data.frame(baseline_obj$pca$rotation[, 1:2]) %>%
    rownames_to_column("variable") %>%
    mutate(
      configuration = nm,
      # Clean variable names — strip method suffix and lag suffix
      variable = variable %>%
                 str_remove("_RF$|_RK$|_OK$") %>%
                 str_replace("_lag$", " (lag)"),
      # Magnitude for arrow filtering
      magnitude = sqrt(PC1^2 + PC2^2)
    )
  
  list(scores = scores, loadings = loadings)
}

pca_projections <- purrr::imap(baseline_list, make_pca_projection_df)

# Combine scores
pca_scores_df <- purrr::map_dfr(pca_projections, ~ .x$scores) %>%
  parse_config()

# Combine loadings
pca_loadings_df <- purrr::map_dfr(pca_projections, ~ .x$loadings) %>%
  parse_config()

# ── Scale loadings to fit on the score plot ───────────────
# Use ~80% of the score range as the arrow scale, per facet
loading_scale <- pca_scores_df %>%
  group_by(configuration) %>%
  summarise(
    pc1_range = max(PC1, na.rm = TRUE) - min(PC1, na.rm = TRUE),
    pc2_range = max(PC2, na.rm = TRUE) - min(PC2, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(arrow_scale = 0.4 * pmin(pc1_range, pc2_range))

# Per-config loading max-magnitude for normalizing arrow lengths
loading_max <- pca_loadings_df %>%
  group_by(configuration) %>%
  summarise(max_mag = max(magnitude, na.rm = TRUE), .groups = "drop")

pca_loadings_scaled <- pca_loadings_df %>%
  left_join(loading_scale, by = "configuration") %>%
  left_join(loading_max,   by = "configuration") %>%
  mutate(
    PC1_scaled = (PC1 / max_mag) * arrow_scale,
    PC2_scaled = (PC2 / max_mag) * arrow_scale
  ) %>%
  # Keep only the top 5 loadings per configuration to avoid clutter
  group_by(configuration) %>%
  slice_max(magnitude, n = 5) %>%
  ungroup()

# ── PCA projection figure ─────────────────────────────────
p_pca_space <- pca_scores_df %>%
  mutate(zone = factor(zone, levels = all_zones)) %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(color = zone),
             size = 0.3, alpha = 0.7) +
  # Eigenvector arrows
  geom_segment(data = pca_loadings_scaled,
               aes(x = 0, y = 0,
                   xend = PC1_scaled, yend = PC2_scaled),
               arrow    = arrow(length = unit(0.15, "cm"),
                                type = "closed"),
               color    = "black",
               linewidth = 0.4) +
  # Variable labels at arrow tips
  geom_text_repel(data = pca_loadings_scaled,
                  aes(x = PC1_scaled, y = PC2_scaled,
                      label = variable),
                  size = 2.2,
                  color = "black",
                  fontface = "bold",
                  segment.color = "grey60",
                  segment.size  = 0.2,
                  min.segment.length = 0.1,
                  max.overlaps  = 20,
                  box.padding   = 0.3) +
  facet_grid(sampling ~ method_label, scales = "free", switch = "y") +
  scale_color_manual(values  = zone_palette,
                     name    = "Zone",
                     na.value = "grey80",
                     drop    = FALSE) +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_minimal(base_size = 9) +
  theme(
    strip.background.x = element_rect(fill = "grey92", color = NA),
    strip.background.y = element_rect(fill = "grey85", color = NA),
    strip.text.x       = element_text(face = "bold", size = 9),
    strip.text.y       = element_text(face = "bold", size = 10),
    strip.placement    = "outside",
    axis.text          = element_text(size = 6),
    axis.title         = element_text(size = 9),
    panel.spacing      = unit(0.7, "lines"),
    legend.position    = "bottom",
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 11)
  ) +
  labs(
    title    = "PCA projection space with variable loadings — all configurations",
    subtitle = "Points colored by zone assignment; arrows = top-5 variable eigenvectors",
    x = "PC1", y = "PC2"
  )

print(p_pca_space)
ggsave("../outputs/fig_pca_projection_space.png",
       p_pca_space, width = 14, height = 7, dpi = 300)
cat("Saved PCA projection figure.\n")
```

