---
title: "Perry_analysis_updated_workflow"
format: html
---

# ============================================================
# PHASE 0 — Libraries
# ============================================================

```{r libraries, message=FALSE, warning=FALSE}
library(terra)
library(tidyverse)
library(janitor)
library(sf)
library(sp)
library(gstat)
library(randomForest)
library(CAST)        # spatial CV folds
library(caret)       # needed by CAST
library(umap)        # UMAP dimensionality reduction
library(dbscan)      # HDBSCAN
library(beepr)
library(broom)
library(FNN)
library(uwot)
library(patchwork)
library(ggtext)
library(gt)
library(gtExtras)
```

# ============================================================
# PHASE 1 — Data Loading & Preparation
# ============================================================

```{r data-load}
# ── raw datasets ──────────────────────────────────────────
soil_points_nutrients <- read.csv(
  "../LOCAL DATA DO NOT PUSH/soil/perry_soil_points_surface_03_2026.csv"
) %>% clean_names()

ap_val <- read_csv(
  "../LOCAL DATA DO NOT PUSH/Ap_hz_data/hz_0_15_wide_03_30_2026.csv"
)

covariate_stack <- rast(
  "../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m_all_with_SR_indices.tif"
)

soil_vars <- c("lbc_1", "lb_ceq", "p_h_2", "ca", "k", "mg", "mn", "p", "zn")
cov_names <- names(covariate_stack)

# ── helper functions ──────────────────────────────────────
add_xy <- function(dat) {
  coords <- st_coordinates(dat)
  dat %>% mutate(X = coords[, 1], Y = coords[, 2])
}

extract_covariates <- function(dat_sf, stack) {
  vect_pts <- terra::vect(dat_sf)
  cov_vals <- terra::extract(stack, vect_pts)
  bind_cols(
    dat_sf %>% st_drop_geometry(),
    cov_vals %>% dplyr::select(-ID)
  ) %>%
    bind_cols(geometry = st_geometry(dat_sf)) %>%
    st_as_sf()
}

# ── SP and COMP modelling datasets ────────────────────────
soil_prop_SP <- soil_points_nutrients %>%
  filter(type == "SPID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

soil_prop_COMP <- soil_points_nutrients %>%
  filter(type == "CID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Validation dataset — HELD OUT, never used in CV ───────
# NOTE: ap_val is NOT pooled with modelling data.
# It is only used in Phase 4 for final held-out evaluation.
ap_val_sf <- ap_val %>%
  st_as_sf(coords = c("POINT_X", "POINT_Y"), crs = 3857) %>%
  st_transform(26917) %>%
  transmute(
    point_label = paste0("VAL_", Point),
    type        = "VAL",
    lbc_1  = LBC.1_0_15,
    lb_ceq = LBCeq_0_15,
    p_h_2  = pH.2_0_15,
    ca     = Ca_0_15,
    k      = K_0_15,
    mg     = Mg_0_15,
    mn     = Mn_0_15,
    p      = P_0_15,
    zn     = Zn_0_15,
    geometry = geometry
  ) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Quick spatial distribution check ──────────────────────
ggplot() +
  geom_sf(data = soil_prop_SP,   color = "steelblue", size = 2) +
  geom_sf(data = soil_prop_COMP, color = "tomato",    size = 2, shape = 17) +
  geom_sf(data = ap_val_sf,      color = "forestgreen", size = 2, shape = 15) +
  theme_minimal() +
  labs(title = "Spatial distribution: SP (blue), COMP (red), VAL (green)")
```


```{r data-summary}
# ============================================================
# PHASE 2 — Spatial Cross-Validation Setup
# ============================================================

# CreateSpacetimeFolds requires pre-defined spatial groups.
# We create them by k-means clustering on coordinates,
# then pass the cluster ID as spacevar.
# This is the standard pattern for field-scale soil data
# and avoids the deprecated knncv function.

make_spatial_folds <- function(sf_data, k = 5, seed = 20260401) {
  coords <- st_coordinates(sf_data)          # matrix of X, Y

  set.seed(seed)
  km <- kmeans(coords, centers = k, nstart = 25)

  # Attach cluster ID as a column — required by CreateSpacetimeFolds
  sf_data$spatial_block <- as.character(km$cluster)

  folds <- CAST::CreateSpacetimeFolds(
    x        = sf_data %>% st_drop_geometry(),
    spacevar = "spatial_block",
    k        = k,
    seed     = seed
  )

  # Convert CAST output (index / indexOut lists) to the
  # train_ids / test_ids format used by run_one_iteration()
  purrr::map(seq_along(folds$index), function(i) {
    list(
      iteration = i,
      seed      = seed,
      train_ids = folds$index[[i]],
      test_ids  = folds$indexOut[[i]]
    )
  })
}

folds_SP   <- make_spatial_folds(soil_prop_SP,   k = 5, seed = 20260401)
folds_COMP <- make_spatial_folds(soil_prop_COMP, k = 5, seed = 20260401)

# These are already in the right format — rename to match
# the rest of the script
resamples_SP   <- folds_SP
resamples_COMP <- folds_COMP

# Variogram settings
make_cutoff <- function(sf_data) max(dist(st_coordinates(sf_data))) / 2
width_lag   <- 75

cutoff_SP   <- make_cutoff(soil_prop_SP)
cutoff_COMP <- make_cutoff(soil_prop_COMP)

# Optional: visualise the spatial blocks to confirm separation
soil_prop_SP$spatial_block <- as.character(
  kmeans(st_coordinates(soil_prop_SP), centers = 5, nstart = 25)$cluster
)
ggplot() +
  geom_sf(data = soil_prop_SP, aes(color = spatial_block), size = 3) +
  theme_minimal() +
  labs(title = "Spatial CV blocks — SP dataset", color = "Block")


```

# ============================================================
# PHASE 3 — Model Fitting (RF, OK, RK)
# ============================================================

```{r model-helpers}
calc_metrics <- function(obs, pred) {
  tibble(
    n          = length(obs),
    RMSE       = sqrt(mean((obs - pred)^2, na.rm = TRUE)),
    MAE        = mean(abs(obs - pred),      na.rm = TRUE),
    R2         = cor(obs, pred, use = "complete.obs")^2,
    mean_error = mean(obs - pred,           na.rm = TRUE)
  )
}

tune_rf_oob <- function(train_df, response, predictors, ntree = 500) {
  x_train <- train_df %>% select(all_of(predictors))
  y_train <- train_df[[response]]
  p       <- length(predictors)
  mtry_grid <- unique(pmax(1, c(
    floor(sqrt(p)), floor(p / 4), floor(p / 3), floor(p / 2)
  )))

  fits <- purrr::map(mtry_grid, function(m) {
    set.seed(sample.int(1e6, 1))
    rf <- randomForest(x = x_train, y = y_train,
                       ntree = ntree, mtry = m, importance = TRUE)
    tibble(mtry = m, oob_mse = tail(rf$mse, 1))
  }) %>% bind_rows()

  list(
    best_mtry    = fits %>% slice_min(oob_mse, n = 1) %>% pull(mtry),
    tuning_table = fits
  )
}

fit_variogram_models <- function(vg) {
  vg_df        <- as.data.frame(vg)
  init_nugget  <- max(0, min(vg_df$gamma, na.rm = TRUE) * 0.5)
  init_sill    <- max(vg_df$gamma, na.rm = TRUE)
  init_psill   <- max(init_sill - init_nugget, 1e-6)
  init_range   <- max(vg_df$dist, na.rm = TRUE) / 3

  fits <- purrr::map(c("Sph", "Exp"), function(m) {
    tryCatch({
      fit <- fit.variogram(
        vg,
        model = vgm(psill = init_psill, model = m,
                    range = init_range,  nugget = init_nugget)
      )
      if (any(!is.finite(fit$psill)))               return(NULL)
      if (any(!is.finite(fit$range)))               return(NULL)
      if (any(fit$range[fit$model != "Nug"] <= 0))  return(NULL)
      fit
    }, error = function(e) NULL)
  })

  fits <- fits[!sapply(fits, is.null)]
  if (length(fits) == 0) return(NULL)

  sse      <- sapply(fits, function(f) attr(f, "SSErr"))
  best_fit <- fits[[which.min(sse)]]

  # ── Degeneracy guard ──────────────────────────────────────
  # A psill of zero means the optimizer found no spatial
  # structure — kriging will return the global mean for every
  # pixel (a constant surface). Treat as a failed fit so
  # downstream code handles it via NA rather than silently
  # producing a meaningless constant prediction.
  str_row <- best_fit %>% filter(model != "Nug")
  if (nrow(str_row) > 0 && str_row$psill < 1e-6) {
    warning("Degenerate variogram fit: psill ≈ 0. ",
            "Returning NULL — variable will be excluded from OK predictions.")
    return(NULL)
  }

  best_fit
}

# ── RF ────────────────────────────────────────────────────
run_rf_model <- function(train_sf, test_sf, response, predictors, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  tuning  <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  preds <- predict(rf_fit, newdata = test_df %>% select(all_of(predictors)))
  list(
    model       = rf_fit,
    predictions = tibble(obs = test_df[[response]], pred = preds),
    metrics     = calc_metrics(test_df[[response]], preds),
    best_mtry   = tuning$best_mtry
  )
}

# ── OK ────────────────────────────────────────────────────
run_ok_model <- function(train_sf, test_sf, response, cutoff, width) {
  train_sp <- as(train_sf, "Spatial")
  test_sp  <- as(test_sf,  "Spatial")
  f        <- as.formula(paste(response, "~ 1"))
  vg       <- variogram(f, data = train_sp, cutoff = cutoff, width = width)
  fit      <- fit_variogram_models(vg)
  if (is.null(fit)) return(NULL)

  kr   <- krige(f, locations = train_sp, newdata = test_sp, model = fit)
  pred <- kr$var1.pred
  obs  <- test_sf %>% st_drop_geometry() %>% pull(.data[[response]])
  list(
    variogram   = vg,
    fit         = fit,
    predictions = tibble(obs = obs, pred = pred),
    metrics     = calc_metrics(obs, pred)
  )
}

# ── RK (RF trend + kriged residuals) ──────────────────────
run_rk_model <- function(train_sf, test_sf, response, predictors,
                         cutoff, width, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  train_sf2 <- train_sf %>%
    filter(complete.cases(st_drop_geometry(select(., all_of(c(response, predictors))))))
  test_sf2  <- test_sf  %>%
    filter(complete.cases(st_drop_geometry(select(., all_of(c(response, predictors))))))

  tuning <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )

  trend_train         <- predict(rf_fit, newdata = train_df %>% select(all_of(predictors)))
  trend_test          <- predict(rf_fit, newdata = test_df  %>% select(all_of(predictors)))
  train_sf2$residual  <- train_df[[response]] - trend_train

  train_sp  <- as(train_sf2, "Spatial")
  test_sp   <- as(test_sf2,  "Spatial")
  vg_resid  <- variogram(residual ~ 1, data = train_sp, cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  # Graceful fallback if residual variogram fails
  if (is.null(fit_resid)) {
    return(list(
      metrics              = calc_metrics(test_df[[response]], trend_test),
      best_mtry            = tuning$best_mtry,
      used_residual_kriging = FALSE
    ))
  }

  kr_resid <- krige(residual ~ 1,
                    locations = train_sp, newdata = test_sp, model = fit_resid)
  pred     <- trend_test + kr_resid$var1.pred
  obs      <- test_df[[response]]

  list(
    metrics              = calc_metrics(obs, pred),
    best_mtry            = tuning$best_mtry,
    used_residual_kriging = TRUE
  )
}

# ── Combined iteration runner ──────────────────────────────
run_one_iteration <- function(sf_data, split, response, predictors,
                              dataset_name, cutoff, width, ntree = 500) {
  train_sf <- sf_data[split$train_ids, ]
  test_sf  <- sf_data[split$test_ids,  ]

  rf_res <- run_rf_model(train_sf, test_sf, response, predictors, ntree)
  ok_res <- run_ok_model(train_sf, test_sf, response, cutoff, width)
  rk_res <- run_rk_model(train_sf, test_sf, response, predictors, cutoff, width, ntree)

  bind_rows(
    rf_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RF",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rf_res$best_mtry),

    if (!is.null(ok_res))
      ok_res$metrics %>%
        mutate(dataset = dataset_name, variable = response, method = "OK",
               iteration = split$iteration, seed = split$seed,
               best_mtry = NA_real_),

    rk_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RK",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rk_res$best_mtry,
             used_residual_kriging = rk_res$used_residual_kriging)
  )
}
```

```{r run-cv, warning=FALSE}
# ── Run spatial CV for SP dataset ─────────────────────────
results_SP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_SP, function(split_obj) {
    run_one_iteration(
      sf_data      = soil_prop_SP,
      split        = split_obj,
      response     = v,
      predictors   = cov_names,
      dataset_name = "SP",
      cutoff       = cutoff_SP,
      width        = width_lag,
      ntree        = 500
    )
  })
})

# ── Run spatial CV for COMP dataset ───────────────────────
results_COMP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_COMP, function(split_obj) {
    run_one_iteration(
      sf_data      = soil_prop_COMP,
      split        = split_obj,
      response     = v,
      predictors   = cov_names,
      dataset_name = "COMP",
      cutoff       = cutoff_COMP,
      width        = width_lag,
      ntree        = 500
    )
  })
})

results_all <- bind_rows(results_SP, results_COMP)
beep("wilhelm")
```

```{r cv-summary-plots}
# ── Violin + boxplot: R² ──────────────────────────────────
results_all %>%
  mutate(
    dataset = factor(dataset, levels = c("SP", "COMP")),
    method  = factor(method,  levels = c("OK", "RF", "RK"))
  ) %>%
  ggplot(aes(x = method, y = R2, fill = method)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  facet_grid(variable ~ dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "Spatial CV: R² by dataset and method",
       x = "Method", y = expression(R^2)) +
  theme(legend.position = "none")

# ── Violin + boxplot: RMSE ────────────────────────────────
results_all %>%
  mutate(
    dataset = factor(dataset, levels = c("SP", "COMP")),
    method  = factor(method,  levels = c("OK", "RF", "RK"))
  ) %>%
  ggplot(aes(x = method, y = RMSE, fill = method)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  facet_grid(variable ~ dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "Spatial CV: RMSE by dataset and method",
       x = "Method", y = "RMSE") +
  theme(legend.position = "none")

# ── Summary table ─────────────────────────────────────────
cv_summary <- results_all %>%
  group_by(dataset, variable, method) %>%
  summarise(
    n_folds    = n(),
    median_R2   = median(R2,   na.rm = TRUE),
    q25_R2      = quantile(R2,   0.25, na.rm = TRUE),
    q75_R2      = quantile(R2,   0.75, na.rm = TRUE),
    median_RMSE = median(RMSE, na.rm = TRUE),
    median_MAE  = median(MAE,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(variable, dataset, desc(median_R2))

cv_summary
write_csv(cv_summary, "../outputs/cv_summary_spatial_folds.csv")
```

# ============================================================
# PHASE 4 — Final Full-Dataset Models (for spatial prediction)
# ============================================================
  # Train on ALL modelling points (SP or COMP) — no hold-out.
  # These models are used to predict to the full raster grid.
  # The validation dataset (ap_val_sf) is evaluated here.

```{r final-models}
fit_final_rf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)
  randomForest(
    x         = df %>% select(all_of(predictors)),
    y         = df[[response]],
    ntree     = ntree,
    mtry      = tuning$best_mtry,
    importance = TRUE
  )
}

fit_final_ok <- function(sf_data, response, cutoff, width) {
  sp_data <- as(sf_data, "Spatial")
  f       <- as.formula(paste(response, "~ 1"))
  vg      <- variogram(f, data = sp_data, cutoff = cutoff, width = width)
  fit     <- fit_variogram_models(vg)
  list(variogram = vg, fit = fit, formula = f, sp_data = sp_data)
}

fit_final_rk <- function(sf_data, response, predictors, cutoff, width, ntree = 500) {
  df      <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  sf_data2 <- sf_data %>%
    filter(complete.cases(st_drop_geometry(select(., all_of(c(response, predictors))))))

  tuning  <- tune_rf_oob(df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = df %>% select(all_of(predictors)),
    y = df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  trend            <- predict(rf_fit, newdata = df %>% select(all_of(predictors)))
  sf_data2$residual <- df[[response]] - trend

  sp_data  <- as(sf_data2, "Spatial")
  vg_resid <- variogram(residual ~ 1, data = sp_data, cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  list(rf_fit = rf_fit, fit_resid = fit_resid, sp_data = sp_data)
}

# Fit final models for both datasets
final_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_SP, v, cov_names),
    ok = fit_final_ok(soil_prop_SP, v, cutoff_SP, width_lag),
    rk = fit_final_rk(soil_prop_SP, v, cov_names, cutoff_SP, width_lag)
  )
)

final_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_COMP, v, cov_names),
    ok = fit_final_ok(soil_prop_COMP, v, cutoff_COMP, width_lag),
    rk = fit_final_rk(soil_prop_COMP, v, cov_names, cutoff_COMP, width_lag)
  )
)

beep("fanfare")
```
# ============================================================
# PHASE 4b — Variogram model summary table
# ============================================================
```{r variogram-summary}

# Extracts nugget, partial sill, total sill, range, and
# nugget-to-sill ratio for every variable × dataset.
# High nugget:sill (> 0.75) flags variables where kriging
# adds little — important to discuss in methods.

extract_variogram_summary <- function(final_models, soil_vars, label) {
  purrr::map_dfr(soil_vars, function(v) {
    fit <- final_models[[v]]$ok$fit
    if (is.null(fit)) {
      return(tibble(variable = v, dataset = label,
                    model = NA, nugget = NA, psill = NA,
                    total_sill = NA, range_m = NA, nugget_sill_ratio = NA))
    }

    nug_row  <- fit %>% filter(model == "Nug")
    str_row  <- fit %>% filter(model != "Nug")

    nugget   <- if (nrow(nug_row) > 0) nug_row$psill else 0
    psill    <- if (nrow(str_row) > 0) str_row$psill else NA
    rng      <- if (nrow(str_row) > 0) str_row$range else NA
    mod_type <- if (nrow(str_row) > 0) as.character(str_row$model) else NA

    tibble(
      variable         = v,
      dataset          = label,
      model            = mod_type,
      nugget           = round(nugget, 4),
      psill            = round(psill,  4),
      total_sill       = round(nugget + psill, 4),
      range_m          = round(rng, 1),
      nugget_sill_ratio = round(nugget / (nugget + psill), 3)
    )
  })
}

vg_summary_SP   <- extract_variogram_summary(final_models_SP,   soil_vars, "SP")
vg_summary_COMP <- extract_variogram_summary(final_models_COMP, soil_vars, "COMP")

vg_summary_all <- bind_rows(vg_summary_SP, vg_summary_COMP)

# Flag variables with weak spatial structure
vg_summary_all %>%
  mutate(
    spatial_dependence = case_when(
      nugget_sill_ratio < 0.25 ~ "Strong",
      nugget_sill_ratio < 0.75 ~ "Moderate",
      TRUE                     ~ "Weak — kriging limited"
    )
  ) %>%
  arrange(dataset, variable) %>%
  print(n = 30)

write_csv(vg_summary_all, "../outputs/variogram_summary_table.csv")

```


```{r validation-evaluation}
# ── Evaluate final models on HELD-OUT validation set ──────
# This is the only place ap_val_sf is used for performance
# metrics — it was never seen during CV or model fitting.

predict_to_points <- function(models_list, response, new_sf, predictors, cutoff, width) {
  new_df <- new_sf %>% st_drop_geometry() %>%
    select(any_of(c(response, predictors))) %>% na.omit()
  obs    <- new_df[[response]]

  # RF
  rf_pred <- predict(models_list$rf, newdata = new_df %>% select(all_of(predictors)))

  # OK
  ok_pred <- tryCatch({
    kr <- krige(
      models_list$ok$formula,
      locations = models_list$ok$sp_data,
      newdata   = as(new_sf, "Spatial"),
      model     = models_list$ok$fit
    )
    kr$var1.pred
  }, error = function(e) rep(NA_real_, nrow(new_df)))

  # RK
  rk_trend <- predict(models_list$rk$rf_fit,
                      newdata = new_df %>% select(all_of(predictors)))
  rk_resid <- tryCatch({
    new_sf2 <- new_sf %>%
      filter(complete.cases(st_drop_geometry(select(., any_of(predictors)))))
    kr2 <- krige(
      residual ~ 1,
      locations = models_list$rk$sp_data,
      newdata   = as(new_sf2, "Spatial"),
      model     = models_list$rk$fit_resid
    )
    kr2$var1.pred
  }, error = function(e) rep(0, length(rk_trend)))
  rk_pred <- rk_trend + rk_resid

  bind_rows(
    calc_metrics(obs, rf_pred) %>% mutate(variable = response, method = "RF"),
    calc_metrics(obs, ok_pred) %>% mutate(variable = response, method = "OK"),
    calc_metrics(obs, rk_pred) %>% mutate(variable = response, method = "RK")
  )
}

val_results_SP <- purrr::map_dfr(soil_vars, function(v) {
  predict_to_points(final_models_SP[[v]], v, ap_val_sf, cov_names,
                    cutoff_SP, width_lag) %>% mutate(dataset = "SP")
})

val_results_COMP <- purrr::map_dfr(soil_vars, function(v) {
  predict_to_points(final_models_COMP[[v]], v, ap_val_sf, cov_names,
                    cutoff_COMP, width_lag) %>% mutate(dataset = "COMP")
})

val_results_all <- bind_rows(val_results_SP, val_results_COMP)
write_csv(val_results_all, "../outputs/validation_results_held_out.csv")

val_results_all %>%
  arrange(variable, dataset, desc(R2))

```

# ============================================================
# PHASE 5 — Full-Grid Spatial Prediction (raster outputs)
# ============================================================

```{r grid-prediction}

# Extract covariate values for every non-NA pixel in the raster
grid_df <- as.data.frame(covariate_stack, xy = TRUE, na.rm = TRUE)

# Convert to sf for kriging steps
grid_sf <- st_as_sf(grid_df, coords = c("x", "y"), crs = 26917)
grid_sp <- as(grid_sf, "Spatial")

# ── Predict all variables to grid for each dataset/method ─
predict_grid_rf <- function(rf_model, grid_df, predictors) {
  predict(rf_model, newdata = grid_df %>% select(all_of(predictors)))
}

predict_grid_ok <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    kr$var1.pred
  }, error = function(e) {
    warning("OK prediction to grid failed; returning NA")
    rep(NA_real_, nrow(grid_sp@data))
  })
}

predict_grid_rk <- function(rk_obj, grid_df, grid_sp, predictors) {
  trend  <- predict(rk_obj$rf_fit,
                    newdata = grid_df %>% select(all_of(predictors)))
  resid  <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    kr2$var1.pred
  }, error = function(e) rep(0, length(trend)))
  trend + resid
}

# Build prediction data frames — one column per variable per method
build_prediction_stack <- function(final_models, grid_df, grid_sf, grid_sp,
                                   soil_vars, predictors, label) {
  preds <- purrr::map(setNames(soil_vars, soil_vars), function(v) {
    m <- final_models[[v]]
    tibble(
      !!paste0(v, "_RF") := predict_grid_rf(m$rf,  grid_df, predictors),
      !!paste0(v, "_RK") := predict_grid_rk(m$rk,  grid_df, grid_sp, predictors),
      !!paste0(v, "_OK") := predict_grid_ok(m$ok,  grid_sp)
    )
  }) %>% bind_cols()

  bind_cols(
    grid_df %>% select(x, y),
    preds,
    dataset = label
  )
}

pred_stack_SP   <- build_prediction_stack(
  final_models_SP,   grid_df, grid_sf, grid_sp, soil_vars, cov_names, "SP"
)
pred_stack_COMP <- build_prediction_stack(
  final_models_COMP, grid_df, grid_sf, grid_sp, soil_vars, cov_names, "COMP"
)

beep("mario")
save.image("../outputs/workspace_phase4b.RData")
beep("coin")
```
# ============================================================
# PHASE 5b — Optional: add spatial lag features to pred stack
# ============================================================
  # For each predicted property column, compute the mean of
  # the k nearest geographic neighbours as an additional feature.
  # This encodes "what does the local neighbourhood look like"
  # without using raw coordinates.
```{r spatial-lags}
#load("../outputs/workspace_phase4b.RData")
# ============================================================
# PHASE 5b — Spatial lags via terra::focal (memory safe)
# ============================================================
# focal() computes lags entirely in terra's C++ backend.
# Never loads a distance matrix — safe at any raster size.
# 3x3 window at 3m resolution = 9m neighbourhood mean,
# which is agronomically meaningful (approx. boom section).


add_spatial_lags_focal <- function(pred_df, soil_vars, methods,
                                   crs = "EPSG:26917") {

  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_df)]

  cat("Computing spatial lags via terra::focal for",
      length(col_names), "layers...\n")

  # Convert prediction stack to SpatRaster
  pred_rast <- pred_df %>%
    select(x, y, all_of(col_names)) %>%
    rast(type = "xyz", crs = crs)

  # 3x3 window, neighbours only (centre excluded = pure spatial lag)
  w       <- matrix(1, 3, 3)
  w[2, 2] <- 0
  lag_rast <- terra::focal(pred_rast, w = w, fun = "mean", na.rm = TRUE)
  names(lag_rast) <- paste0(col_names, "_lag")

  # ── KEY FIX: extract at original coordinates ──────────────
  # as.data.frame() on the lag raster changes row count because
  # focal fills edge NAs. Instead, extract at the exact xy
  # locations from pred_df to guarantee row alignment.
  pts     <- terra::vect(pred_df %>% select(x, y),
                         geom = c("x", "y"), crs = crs)
  lag_df  <- terra::extract(lag_rast, pts, ID = FALSE)

  cat("pred_df rows:", nrow(pred_df),
      "| lag rows:", nrow(lag_df), "\n")  # should match

  bind_cols(pred_df, lag_df)
}

pred_stack_SP_lag   <- add_spatial_lags_focal(pred_stack_SP,
                                               soil_vars, c("RF", "RK"))
pred_stack_COMP_lag <- add_spatial_lags_focal(pred_stack_COMP,
                                               soil_vars, c("RF", "RK"))

cat("SP lag stack columns:",   ncol(pred_stack_SP_lag),   "\n")
cat("COMP lag stack columns:", ncol(pred_stack_COMP_lag), "\n")
```


# ============================================================
# PHASE 6 — UMAP + HDBSCAN Management Zone Delineation
# ============================================================
  # NOTE: UMAP is applied AFTER spatial prediction.
  # Scaling happens HERE, not before RF/OK/RK inputs.
  # Applied to RF and RK prediction stacks separately.
  # OK is kept as a standalone geostatistical benchmark only.

```{r umap-hdbscan}
run_umap_hdbscan <- function(pred_df,
                              methods          = c("RF", "RK"),
                              soil_vars,
                              include_lags     = FALSE,
                              n_components     = 2,
                              min_dist         = 0.001,
                              n_neighbors      = 30,
                              minPts           = 50,
                              cluster_eps      = 0,
                              label            = "dataset") {

  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  if (include_lags) {
    lag_cols  <- paste0(col_names, "_lag")
    col_names <- c(col_names, lag_cols[lag_cols %in% names(pred_df)])
  }
  col_names <- col_names[col_names %in% names(pred_df)]

  mat           <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  mat_clean     <- mat[complete_rows, ]
  xy_clean      <- pred_df %>% select(x, y) %>% slice(which(complete_rows))

  cat(label, "— pixels:", nrow(mat_clean),
      "| cols:", ncol(mat_clean),
      "| lags:", include_lags, "\n")

  # ── 2. Scale — with zero-variance and NA guard ────────────
  col_ranges <- apply(mat_clean, 2, function(x)
                      diff(range(x, na.rm = TRUE)))
  zero_var   <- col_ranges < 1e-10

  if (any(zero_var)) {
    cat("Warning: dropping", sum(zero_var),
        "constant column(s):",
        paste(colnames(mat_clean)[zero_var], collapse = ", "), "\n")
    mat_clean <- mat_clean[, !zero_var, drop = FALSE]
  }

  if (ncol(mat_clean) < 2) {
    cat("ERROR: fewer than 2 valid columns after variance filtering.",
        "Cannot run UMAP for", label, "\n")
    return(NULL)
  }

  mat_scaled <- scale(mat_clean)
  mat_scaled[!is.finite(mat_scaled)] <- 0
  cat("Columns entering UMAP:", ncol(mat_scaled), "\n")
  stopifnot("NA values remain" = !anyNA(mat_scaled))

  # ── 3. UMAP ───────────────────────────────────────────────
  cat("Fitting UMAP...\n")
  umap_fit <- uwot::umap(
    mat_scaled,
    n_components = n_components,
    min_dist     = min_dist,
    n_neighbors  = n_neighbors,
    ret_model    = TRUE,
    seed         = 20260401,
    verbose      = TRUE
  )

  emb_df        <- as.data.frame(umap_fit$embedding)
  names(emb_df) <- paste0("UMAP", seq_len(n_components))

  # ── 4. HDBSCAN on subsample ───────────────────────────────
  hdbscan_sample_size <- min(nrow(emb_df), 10000)
  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df), hdbscan_sample_size))
  emb_sample <- emb_df[sample_idx, ]

  cat("Running HDBSCAN on subsample of", hdbscan_sample_size, "points...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts,
    cluster_selection_epsilon = cluster_eps
  )

  cat("Clusters on subsample:", max(hdb_fit$cluster),
      "| noise:", round(mean(hdb_fit$cluster == 0) * 100, 1), "%\n")

  cat("Assigning full grid via NN...\n")
  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  cat("Full grid — zones:", max(full_clusters),
      "| noise:", round(mean(full_clusters == 0) * 100, 1), "%\n")

  # ── 5. Assemble ───────────────────────────────────────────
  bind_cols(
    xy_clean,
    emb_df,
    tibble(
      cluster_hard    = full_clusters,
      membership_prob = full_memb
    )
  ) %>%
    mutate(
      dataset      = label,
      spatial_lags = include_lags,
      zone         = if_else(cluster_hard == 0L, NA_integer_,
                             as.integer(cluster_hard))
    )
}

# ── wrap_mz_result helper ─────────────────────────────────
wrap_mz_result <- function(df) {
  if (is.data.frame(df)) {
    zones <- df$zone[!is.na(df$zone)]
    return(list(
      result_df   = df,
      n_zones     = length(unique(zones)),
      pct_noise   = mean(is.na(df$zone)) * 100,
      umap_fit    = NULL,
      hdbscan_fit = NULL
    ))
  }
  # already wrapped — recompute metadata from result_df
  zones <- df$result_df$zone[!is.na(df$result_df$zone)]
  df$n_zones   <- length(unique(zones))
  df$pct_noise <- mean(is.na(df$result_df$zone)) * 100
  df
}

# ── 10 runs: RF, RK, OK × SP, COMP ───────────────────────
mz_SP_RF_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RF_nolags")

mz_SP_RF_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RF_lags")

mz_SP_RK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RK_nolags")

mz_SP_RK_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RK_lags")

mz_SP_OK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_OK_nolags")

mz_COMP_RF_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RF_nolags")

mz_COMP_RF_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RF_lags")

mz_COMP_RK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RK_nolags")

mz_COMP_RK_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RK_lags")

mz_COMP_OK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_OK_nolags")

save.image("../outputs/workspace_phase6_complete.RData")
beep("mario")

# ── Wrap all 10 and build named list ─────────────────────
mz_named_tuned <- list(
  SP_RF_nolags   = wrap_mz_result(mz_SP_RF_nolags),
  SP_RF_lags     = wrap_mz_result(mz_SP_RF_lags),
  SP_RK_nolags   = wrap_mz_result(mz_SP_RK_nolags),
  SP_RK_lags     = wrap_mz_result(mz_SP_RK_lags),
  SP_OK_nolags   = wrap_mz_result(mz_SP_OK_nolags),
  COMP_RF_nolags = wrap_mz_result(mz_COMP_RF_nolags),
  COMP_RF_lags   = wrap_mz_result(mz_COMP_RF_lags),
  COMP_RK_nolags = wrap_mz_result(mz_COMP_RK_nolags),
  COMP_RK_lags   = wrap_mz_result(mz_COMP_RK_lags),
  COMP_OK_nolags = wrap_mz_result(mz_COMP_OK_nolags)
)

purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})
```
# ============================================================
# PHASE 6c — HDBSCAN retuning for agronomic zone counts
# ============================================================
```{r hdbscan-retune}

retune_hdbscan <- function(mz_obj, minPts_new,
                            cluster_eps  = 0,
                            hdbscan_n    = 10000,
                            label        = NULL) {

  emb_df       <- mz_obj$result_df %>% select(UMAP1, UMAP2)
  xy_clean     <- mz_obj$result_df %>% select(x, y)
  spatial_lags <- mz_obj$result_df$spatial_lags[1]
  lbl          <- if (!is.null(label)) label else
                    mz_obj$result_df$dataset[1]

  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df), min(hdbscan_n, nrow(emb_df))))
  emb_sample <- emb_df[sample_idx, ]

  cat("Retuning:", lbl, "| minPts =", minPts_new, "...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts_new,
    cluster_selection_epsilon = cluster_eps
  )

  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  result_df <- bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard = full_clusters, membership_prob = full_memb)
  ) %>%
    mutate(
      dataset      = lbl,
      spatial_lags = spatial_lags,
      zone         = if_else(cluster_hard == 0L, NA_integer_,
                             as.integer(cluster_hard))
    )

  list(
    result_df   = result_df,
    hdbscan_fit = hdb_fit,
    n_zones     = length(unique(full_clusters[full_clusters != 0])),
    pct_noise   = mean(full_clusters == 0) * 100
  )
}

# ── Sensitivity scan ──────────────────────────────────────
minpts_candidates <- c(50, 100, 200, 400, 600, 800, 1000)

sensitivity_results <- purrr::map_dfr(minpts_candidates, function(mp) {
  purrr::map_dfr(names(mz_named_tuned), function(nm) {
    r <- retune_hdbscan(mz_named_tuned[[nm]], mp, label = nm)
    tibble(label = nm, minPts = mp,
           n_zones = r$n_zones, pct_noise = round(r$pct_noise, 1))
  })
})

sensitivity_results %>%
  select(label, minPts, n_zones) %>%
  pivot_wider(names_from = label, values_from = n_zones) %>%
  arrange(minPts) %>%
  print(width = Inf)

sensitivity_results %>%
  select(label, minPts, pct_noise) %>%
  pivot_wider(names_from = label, values_from = pct_noise,
              names_prefix = "noise_") %>%
  arrange(minPts) %>%
  print(width = Inf)

# ── Set per-config minPts after reviewing table ───────────
# Adjust values below based on sensitivity scan output
minpts_final <- list(
  SP_RF_nolags   = 200,
  SP_RF_lags     = 200,
  SP_RK_nolags   = 200,
  SP_RK_lags     = 200,
  SP_OK_nolags   = 200,   # adjust if OK collapses at 200
  COMP_RF_nolags = 200,
  COMP_RF_lags   = 200,
  COMP_RK_nolags = 200,
  COMP_RK_lags   = 200,
  COMP_OK_nolags = 200    # adjust if OK collapses at 200
)

# Keep original as source, write to new object
mz_named_source <- mz_named_tuned  # snapshot before overwriting

mz_named_tuned <- purrr::imap(minpts_final, function(mp, nm) {
  retune_hdbscan(mz_named_source[[nm]], mp, label = nm)
})

# Verify
purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})

save.image("../outputs/workspace_phase6c_retuned.RData")
beep("coin")

#SP_OK_nolags at 79.6% noise is worth flagging in methods but not necessarily changing — lowering to minPts=100 gives 12 zones which is too many for 59 validation points. The 3-zone/79.6% noise result is itself a finding about OK's limitations. Keeping minPts=200 across the board and document the noise percentages explicitly in methods table.
```

```{r umap-hdbscan-plots}
plot_umap_zones <- function(res, title) {
  res$result_df %>%
    ggplot(aes(x = UMAP1, y = UMAP2,
               color = factor(zone),
               alpha = membership_prob)) +
    geom_point(size = 0.4) +
    scale_alpha_continuous(range = c(0.2, 1)) +
    theme_minimal() +
    labs(title = title, color = "Zone",
         alpha = "Membership\nconfidence") +
    guides(color = guide_legend(
      override.aes = list(size = 3, alpha = 1)))
}

plot_spatial_zones <- function(res, title) {
  res$result_df %>%
    filter(!is.na(zone)) %>%
    ggplot(aes(x = x, y = y, fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    theme_minimal() +
    labs(title = title, fill = "Zone",
         x = "Easting", y = "Northing")
}

plot_confidence <- function(res, title) {
  res$result_df %>%
    ggplot(aes(x = x, y = y, fill = membership_prob)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1) +
    coord_equal() +
    theme_minimal() +
    labs(title = title, fill = "Confidence",
         x = "Easting", y = "Northing")
}

# All 8 UMAP embeddings
purrr::iwalk(mz_named_tuned, function(x, nm) {
  print(plot_umap_zones(x, paste("UMAP —", nm)))
})

# All 8 spatial zone maps
purrr::iwalk(mz_named_tuned, function(x, nm) {
  print(plot_spatial_zones(x, paste("Zones —", nm)))
})

# All 8 confidence surfaces
purrr::iwalk(mz_named_tuned, function(x, nm) {
  print(plot_confidence(x, paste("Confidence —", nm)))
})
```

# ============================================================
# PHASE 6b — Post-processing: noise fill + contiguity filter
# ============================================================

```{r post-processing}
post_process_zones <- function(result_df, zone_rast_crs = "EPSG:26917") {

  n_total <- nrow(result_df)
  n_noise <- sum(is.na(result_df$zone))
  cat("Noise pixels before fill:", n_noise,
      "(", round(100 * n_noise / n_total, 1), "% of field)\n")

  # ── Guard: if ALL pixels are noise, skip processing ───────
  if (n_noise == n_total) {
    cat("WARNING: all pixels are noise — no zones found at this minPts.\n",
        "Consider lowering minPts for this configuration.\n")
    return(list(
      result_df_filled = result_df,
      zone_rast_raw    = NULL,
      zone_rast_smooth = NULL
    ))
  }

  assigned  <- result_df %>% filter(!is.na(zone))
  noise_pts <- result_df %>% filter( is.na(zone))

  if (nrow(noise_pts) > 0) {
    nn_idx <- FNN::get.knnx(
      data  = assigned  %>% select(UMAP1, UMAP2) %>% as.matrix(),
      query = noise_pts %>% select(UMAP1, UMAP2) %>% as.matrix(),
      k     = 1
    )$nn.index
    noise_pts$zone            <- assigned$zone[nn_idx]
    noise_pts$membership_prob <- 0
    result_df <- bind_rows(assigned, noise_pts)
  }

  cat("Noise pixels after fill: 0\n")

  zone_rast        <- result_df %>%
    select(x, y, zone) %>%
    rast(type = "xyz", crs = zone_rast_crs)

  zone_rast_smooth <- focal(zone_rast, w = 5, fun = "modal", na.rm = TRUE)

  changed <- sum(values(zone_rast) != values(zone_rast_smooth), na.rm = TRUE)
  cat("Pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "% of field)\n")

  list(
    result_df_filled = result_df,
    zone_rast_raw    = zone_rast,
    zone_rast_smooth = zone_rast_smooth
  )
}

pp_named <- purrr::imap(mz_named_tuned, function(x, nm) {
  cat("\n---", nm, "---\n")
  post_process_zones(x$result_df)
})

# Safe raster writer
write_zone_raster <- function(pp_obj, path) {
  if (is.null(pp_obj$zone_rast_smooth)) {
    cat("Skipping", path, "— no zones (all noise)\n")
    return(invisible(NULL))
  }
  writeRaster(pp_obj$zone_rast_smooth, path, overwrite = TRUE)
  cat("Saved:", path, "\n")
}

purrr::iwalk(pp_named, function(pp, nm) {
  write_zone_raster(pp,
    paste0("../outputs/zones_", nm, "_smooth.tif"))
})


```



# ============================================================
# PHASE 7 — Baseline Comparison: PCA + k-means
# ============================================================

```{r pca-kmeans-baseline}
run_pca_kmeans <- function(pred_df, methods = c("RF", "RK"),
                           soil_vars, k_range = 2:8,
                           label = "dataset") {
  col_names     <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names     <- col_names[col_names %in% names(pred_df)]
  mat           <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  xy_clean      <- pred_df %>% select(x, y) %>% slice(which(complete_rows))

  mat_scaled <- scale(mat[complete_rows, ])

  # Drop constant columns — needed for OK which may have zero variance
  col_ranges  <- apply(mat_scaled, 2, function(x)
                       diff(range(x, na.rm = TRUE)))
  const_cols  <- !is.finite(mat_scaled[1, ]) | col_ranges < 1e-10
  if (any(const_cols)) {
    cat("Dropping", sum(const_cols), "constant column(s):",
        paste(colnames(mat_scaled)[const_cols], collapse = ", "), "\n")
    mat_scaled <- mat_scaled[, !const_cols, drop = FALSE]
  }
  mat_scaled[!is.finite(mat_scaled)] <- 0

  pca_out    <- prcomp(mat_scaled, center = FALSE, scale. = FALSE)
  cum_var    <- cumsum(pca_out$sdev^2) / sum(pca_out$sdev^2)
  n_comp     <- which(cum_var >= 0.95)[1]
  pca_scores <- pca_out$x[, seq_len(n_comp), drop = FALSE]

  set.seed(20260401)
  sil_idx <- sample(nrow(pca_scores), min(5000, nrow(pca_scores)))
  pca_sub <- pca_scores[sil_idx, ]

  sil_scores <- purrr::map_dbl(k_range, function(k) {
    set.seed(20260401)
    km  <- kmeans(pca_sub, centers = k, nstart = 25, iter.max = 100)
    sil <- cluster::silhouette(km$cluster, dist(pca_sub))
    mean(sil[, 3])
  })
  best_k <- k_range[which.max(sil_scores)]
  cat(label, "— best k:", best_k, "\n")

  set.seed(20260401)
  km_best <- kmeans(pca_scores, centers = best_k,
                    nstart = 25, iter.max = 100)

  result_df <- bind_cols(
    xy_clean,
    tibble(zone_kmeans = km_best$cluster)
  ) %>% mutate(dataset = label, method = "PCA_kmeans")

  list(
    result_df   = result_df,
    pca         = pca_out,
    km          = km_best,
    sil_profile = tibble(k = k_range, silhouette = sil_scores),
    best_k      = best_k
  )
}

# ── No lags ───────────────────────────────────────────────
baseline_SP_RF   <- run_pca_kmeans(pred_stack_SP,
                     methods = "RF", soil_vars = soil_vars, label = "SP_RF")
baseline_SP_RK   <- run_pca_kmeans(pred_stack_SP,
                     methods = "RK", soil_vars = soil_vars, label = "SP_RK")
baseline_SP_OK   <- run_pca_kmeans(pred_stack_SP,
                     methods = "OK", soil_vars = soil_vars, label = "SP_OK")
baseline_COMP_RF <- run_pca_kmeans(pred_stack_COMP,
                     methods = "RF", soil_vars = soil_vars, label = "COMP_RF")
baseline_COMP_RK <- run_pca_kmeans(pred_stack_COMP,
                     methods = "RK", soil_vars = soil_vars, label = "COMP_RK")
baseline_COMP_OK <- run_pca_kmeans(pred_stack_COMP,
                     methods = "OK", soil_vars = soil_vars, label = "COMP_OK")

# ── With lags ─────────────────────────────────────────────
baseline_SP_RF_lags   <- run_pca_kmeans(pred_stack_SP_lag,
                     methods = "RF", soil_vars = soil_vars,
                     label = "SP_RF_lags_kmeans")
baseline_SP_RK_lags   <- run_pca_kmeans(pred_stack_SP_lag,
                     methods = "RK", soil_vars = soil_vars,
                     label = "SP_RK_lags_kmeans")
baseline_COMP_RF_lags <- run_pca_kmeans(pred_stack_COMP_lag,
                     methods = "RF", soil_vars = soil_vars,
                     label = "COMP_RF_lags_kmeans")
baseline_COMP_RK_lags <- run_pca_kmeans(pred_stack_COMP_lag,
                     methods = "RK", soil_vars = soil_vars,
                     label = "COMP_RK_lags_kmeans")

# ── Silhouette profiles ───────────────────────────────────
bind_rows(
  baseline_SP_RF$sil_profile        %>% mutate(config = "SP_RF"),
  baseline_SP_RK$sil_profile        %>% mutate(config = "SP_RK"),
  baseline_SP_OK$sil_profile        %>% mutate(config = "SP_OK"),
  baseline_COMP_RF$sil_profile      %>% mutate(config = "COMP_RF"),
  baseline_COMP_RK$sil_profile      %>% mutate(config = "COMP_RK"),
  baseline_COMP_OK$sil_profile      %>% mutate(config = "COMP_OK"),
  baseline_SP_RF_lags$sil_profile   %>% mutate(config = "SP_RF_lags"),
  baseline_SP_RK_lags$sil_profile   %>% mutate(config = "SP_RK_lags"),
  baseline_COMP_RF_lags$sil_profile %>% mutate(config = "COMP_RF_lags"),
  baseline_COMP_RK_lags$sil_profile %>% mutate(config = "COMP_RK_lags")
) %>%
  ggplot(aes(x = k, y = silhouette, color = config)) +
  geom_line() + geom_point() +
  theme_minimal() +
  labs(title = "PCA+k-means silhouette profiles",
       x = "k", y = "Mean silhouette")
```

# ============================================================
# PHASE 7b — Post-processing PCA+kmeans zones
# ============================================================

```{r pca-kmeans-postprocess}
# PCA+kmeans zones need the same noise fill + majority filter
# treatment as UMAP+HDBSCAN zones for a fair comparison.
# Note: PCA+kmeans has no noise by definition (every pixel
# gets a cluster assignment) so the fill step is skipped.
# We still apply the majority filter for spatial coherence.
post_process_kmeans <- function(result_df,
                                 zone_rast_crs = "EPSG:26917") {
  n_total <- nrow(result_df)
  cat("PCA+kmeans: no noise pixels (all assigned)\n")

  zone_rast <- result_df %>%
    select(x, y, zone = zone_kmeans) %>%
    rast(type = "xyz", crs = zone_rast_crs)

  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)

  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("Pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "% of field)\n")

  list(
    result_df_filled = result_df %>% mutate(zone = zone_kmeans),
    zone_rast_raw    = zone_rast,
    zone_rast_smooth = zone_rast_smooth
  )
}

pp_baseline <- list(
  SP_RF          = post_process_kmeans(baseline_SP_RF$result_df),
  SP_RK          = post_process_kmeans(baseline_SP_RK$result_df),
  SP_OK          = post_process_kmeans(baseline_SP_OK$result_df),
  COMP_RF        = post_process_kmeans(baseline_COMP_RF$result_df),
  COMP_RK        = post_process_kmeans(baseline_COMP_RK$result_df),
  COMP_OK        = post_process_kmeans(baseline_COMP_OK$result_df),
  SP_RF_lags     = post_process_kmeans(baseline_SP_RF_lags$result_df),
  SP_RK_lags     = post_process_kmeans(baseline_SP_RK_lags$result_df),
  COMP_RF_lags   = post_process_kmeans(baseline_COMP_RF_lags$result_df),
  COMP_RK_lags   = post_process_kmeans(baseline_COMP_RK_lags$result_df)
)

purrr::iwalk(pp_baseline, function(pp, nm) {
  writeRaster(pp$zone_rast_smooth,
              paste0("../outputs/zones_", nm, "_kmeans_smooth.tif"),
              overwrite = TRUE)
  cat("Saved:", nm, "\n")
})
```

# ============================================================
# PHASE 7c — Agronomic validation of PCA+kmeans zones
# ============================================================

```{r pca-kmeans-validation}
baseline_configs <- list(
  SP_RF        = list(pp = pp_baseline[["SP_RF"]],
                      pred = pred_stack_SP,       method = "RF"),
  SP_RK        = list(pp = pp_baseline[["SP_RK"]],
                      pred = pred_stack_SP,       method = "RK"),
  SP_OK        = list(pp = pp_baseline[["SP_OK"]],
                      pred = pred_stack_SP,       method = "OK"),
  COMP_RF      = list(pp = pp_baseline[["COMP_RF"]],
                      pred = pred_stack_COMP,     method = "RF"),
  COMP_RK      = list(pp = pp_baseline[["COMP_RK"]],
                      pred = pred_stack_COMP,     method = "RK"),
  COMP_OK      = list(pp = pp_baseline[["COMP_OK"]],
                      pred = pred_stack_COMP,     method = "OK"),
  SP_RF_lags   = list(pp = pp_baseline[["SP_RF_lags"]],
                      pred = pred_stack_SP_lag,   method = "RF"),
  SP_RK_lags   = list(pp = pp_baseline[["SP_RK_lags"]],
                      pred = pred_stack_SP_lag,   method = "RK"),
  COMP_RF_lags = list(pp = pp_baseline[["COMP_RF_lags"]],
                      pred = pred_stack_COMP_lag, method = "RF"),
  COMP_RK_lags = list(pp = pp_baseline[["COMP_RK_lags"]],
                      pred = pred_stack_COMP_lag, method = "RK")
)

val_baseline <- purrr::imap(baseline_configs, function(cfg, nm) {
  cat("\n---", nm, "PCA+kmeans ---\n")
  validate_zones_with_holdout(
    zone_rast_smooth = cfg$pp$zone_rast_smooth,
    val_sf           = ap_val_sf,
    pred_stack       = cfg$pred,
    soil_vars        = soil_vars,
    methods          = cfg$method,
    label            = paste0(nm, "_kmeans")
  )
})

kw_baseline <- purrr::map_dfr(val_baseline, ~ .x$kw_results) %>%
  mutate(significant = kw_p < 0.05)

kw_baseline_summary <- kw_baseline %>%
  group_by(dataset) %>%
  summarise(
    n_significant   = sum(significant, na.rm = TRUE),
    n_total         = n(),
    pct_significant = round(100 * mean(significant, na.rm = TRUE)),
    sig_vars        = paste(variable[significant], collapse = ", "),
    .groups         = "drop"
  ) %>%
  arrange(desc(n_significant))

print(kw_baseline_summary)
write_csv(kw_baseline,         "../outputs/kw_baseline_kmeans.csv")
write_csv(kw_baseline_summary, "../outputs/kw_baseline_summary.csv")
```

# ============================================================
# PHASE 7d — Head-to-head comparison: UMAP+HDBSCAN vs PCA+kmeans
# ============================================================

```{r method-comparison}
# Combine KW results from both approaches for direct comparison

kw_all_methods <- bind_rows(
  all_kw %>% mutate(approach = "UMAP+HDBSCAN"),
  kw_baseline %>% mutate(
    approach = "PCA+kmeans",
    # Harmonise label — strip "_kmeans" suffix for joining
    dataset_clean = str_remove(dataset, "_kmeans")
  )
)

# Summary comparison table
comparison_summary <- bind_rows(

  # UMAP+HDBSCAN
  all_kw %>%
    mutate(approach = "UMAP+HDBSCAN") %>%
    group_by(dataset, approach) %>%
    summarise(
      n_sig    = sum(significant, na.rm = TRUE),
      sig_vars = paste(variable[significant], collapse = ", "),
      .groups  = "drop"
    ) %>%
    mutate(
      parts      = strsplit(dataset, "_"),
      sampling   = map_chr(parts, 1),
      method     = map_chr(parts, 2),
      lags       = map_chr(parts, 3)
    ) %>%
    select(-parts),

  # PCA+kmeans
  kw_baseline %>%
    mutate(approach = "PCA+kmeans") %>%
    group_by(dataset, approach) %>%
    summarise(
      n_sig    = sum(significant, na.rm = TRUE),
      sig_vars = paste(variable[significant], collapse = ", "),
      .groups  = "drop"
    ) %>%
    mutate(
      clean    = str_remove(dataset, "_kmeans"),
      parts    = strsplit(clean, "_"),
      sampling = map_chr(parts, 1),
      method   = map_chr(parts, 2),
      lags     = "nolags"
    ) %>%
    select(-parts, -clean)

) %>%
  arrange(sampling, method, lags, approach)

cat("\n=== Head-to-head: UMAP+HDBSCAN vs PCA+kmeans ===\n")
comparison_summary %>%
  select(sampling, method, lags, approach, n_sig, sig_vars) %>%
  print(n = 40)

write_csv(comparison_summary, "../outputs/method_comparison_summary.csv")

# Visual: bar chart of significant variables per config × approach
comparison_summary %>%
  mutate(
    config = paste(sampling, method, lags, sep = "_"),
    approach = factor(approach,
                      levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = config, y = n_sig, fill = approach)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Significant soil variables per zone configuration",
    subtitle = "Kruskal-Wallis p < 0.05 on held-out validation points",
    x = "Configuration",
    y = "N significant variables (of 9)",
    fill = "Clustering approach"
  )
```

# ============================================================
# PHASE 8 — Zone Validation Metrics
# ============================================================

```{r zone-validation}
calc_variance_reduction <- function(pred_df, zone_vec,
                                     soil_vars, methods) {
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_df)]
  mat       <- pred_df %>% select(all_of(col_names)) %>% as.matrix()

  # Drop constant columns — prevents artificial VR inflation
  # for OK configurations where some variables have zero variance
  col_ranges <- apply(mat, 2, function(x)
                      diff(range(x, na.rm = TRUE)))
  mat        <- mat[, col_ranges > 1e-10, drop = FALSE]

  keep       <- !is.na(zone_vec) & complete.cases(mat)
  mat_k      <- mat[keep, ]
  zones_k    <- zone_vec[keep]
  total_var  <- mean(apply(mat_k, 2, var, na.rm = TRUE))
  within_var <- mean(sapply(unique(zones_k), function(z) {
    mean(apply(mat_k[zones_k == z, , drop = FALSE],
               2, var, na.rm = TRUE))
  }))
  1 - (within_var / total_var)
}

calc_silhouette_sub <- function(emb_df, zone_vec, n_sub = 5000) {
  keep  <- !is.na(zone_vec)
  emb_k <- emb_df[keep, ]
  z_k   <- zone_vec[keep]
  set.seed(20260401)
  idx   <- sample(nrow(emb_k), min(n_sub, nrow(emb_k)))
  sil   <- cluster::silhouette(z_k[idx], dist(emb_k[idx, ]))
  mean(sil[, 3])
}

# ── UMAP+HDBSCAN validation metrics ──────────────────────
zone_validation_umap <- purrr::imap_dfr(mz_named_tuned, function(x, nm) {
  parts  <- strsplit(nm, "_")[[1]]
  samp   <- parts[1]
  method <- parts[2]
  lag_yn <- length(parts) >= 3 && parts[3] == "lags"

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = pred_stack_SP,
    SP_RK   = pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = pred_stack_COMP,
    COMP_RK = pred_stack_COMP,
    COMP_OK = pred_stack_COMP
  )

  tibble(
    configuration = nm,
    sampling      = samp,
    pred_method   = method,
    lags          = lag_yn,
    approach      = "UMAP+HDBSCAN",
    VR = calc_variance_reduction(
      pred_df, x$result_df$zone, soil_vars, method),
    silhouette = calc_silhouette_sub(
      x$result_df %>% select(UMAP1, UMAP2),
      x$result_df$zone)
  )
})

# ── PCA+kmeans validation metrics ─────────────────────────
baseline_list <- list(
  SP_RF               = baseline_SP_RF,
  SP_RK               = baseline_SP_RK,
  SP_OK               = baseline_SP_OK,
  COMP_RF             = baseline_COMP_RF,
  COMP_RK             = baseline_COMP_RK,
  COMP_OK             = baseline_COMP_OK,
  SP_RF_lags_kmeans   = baseline_SP_RF_lags,
  SP_RK_lags_kmeans   = baseline_SP_RK_lags,
  COMP_RF_lags_kmeans = baseline_COMP_RF_lags,
  COMP_RK_lags_kmeans = baseline_COMP_RK_lags
)

zone_validation_pca <- purrr::imap_dfr(baseline_list, function(b, nm) {
  clean   <- str_remove(nm, "_kmeans")
  has_lag <- str_detect(clean, "_lags")
  clean   <- str_remove(clean, "_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]
  method  <- parts[2]

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_OK = pred_stack_COMP
  )

  # Guard for OK constant columns in prcomp
  ok_cols <- pred_df %>%
    select(any_of(paste0(soil_vars, "_", method))) %>%
    na.omit()
  col_vars <- apply(ok_cols, 2, var, na.rm = TRUE)
  ok_cols  <- ok_cols[, col_vars > 1e-10, drop = FALSE]

  tibble(
    configuration = nm,
    sampling      = samp,
    pred_method   = method,
    lags          = has_lag,
    approach      = "PCA+kmeans",
    VR = calc_variance_reduction(
      pred_df, b$result_df$zone_kmeans, soil_vars, method),
    silhouette = calc_silhouette_sub(
      prcomp(scale(ok_cols), center = FALSE)$x[, 1:2],
      b$result_df$zone_kmeans)
  )
})

zone_validation_all <- bind_rows(zone_validation_umap,
                                  zone_validation_pca) %>%
  arrange(sampling, pred_method, lags, approach)

zone_validation_all
write_csv(zone_validation_all, "../outputs/zone_validation_metrics.csv")

# ── Plot ──────────────────────────────────────────────────
zone_validation_all %>%
  pivot_longer(cols = c(VR, silhouette),
               names_to = "metric", values_to = "value") %>%
  mutate(
    config_label = paste(sampling, pred_method, sep = "\n"),
    approach     = factor(approach,
                          levels = c("PCA+kmeans", "UMAP+HDBSCAN")),
    lags_label   = ifelse(lags, "With lags", "No lags"),
    metric       = factor(metric,
                          levels = c("VR", "silhouette"),
                          labels = c("Variance Reduction",
                                     "Silhouette Coefficient"))
  ) %>%
  ggplot(aes(x = config_label, y = value,
             fill = approach, alpha = lags_label)) +
  geom_col(position = "dodge") +
  scale_alpha_manual(values = c("No lags" = 0.6,
                                "With lags" = 1.0)) +
  facet_wrap(~ metric, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 9)) +
  labs(
    title    = "Zone quality: UMAP+HDBSCAN vs PCA+k-means",
    x        = "Sampling × Method",
    y        = "Metric value",
    fill     = "Clustering approach",
    alpha    = "Spatial lags"
  )
```

# ============================================================
# PHASE 8b — Agronomic validation using held-out ap_val_sf
# ============================================================


```{r agronomic-validation}

# validate_zones_with_holdout() stays exactly as written above
# ── helper to get correct pred_stack per config ───────────
# ── Helper: get correct pred stack and method per config ──
get_pred_stack <- function(nm) {
  clean  <- nm %>%
    str_remove("_kmeans") %>%
    str_remove("_nolags") %>%
    str_remove("_lags")
  parts  <- strsplit(clean, "_")[[1]]
  samp   <- parts[1]
  method <- parts[2]
  has_lag <- str_detect(nm, "_lags")

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_OK = pred_stack_COMP
  )

  list(pred_df = pred_df, method = method,
       samp = samp, has_lag = has_lag)
}

# ── UMAP+HDBSCAN validation ───────────────────────────────
val_umap <- purrr::imap(pp_named, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- UMAP+HDBSCAN:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method, label = nm
  )
})
val_umap <- val_umap[!sapply(val_umap, is.null)]

# ── PCA+kmeans validation ─────────────────────────────────
val_kmeans <- purrr::imap(pp_baseline, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- PCA+kmeans:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method,
    label   = paste0(nm, "_kmeans")
  )
})
val_kmeans <- val_kmeans[!sapply(val_kmeans, is.null)]

# ── Combine ───────────────────────────────────────────────
all_val <- c(val_umap, val_kmeans)

all_kw <- purrr::map_dfr(all_val, ~ .x$kw_results) %>%
  mutate(
    significant = kw_p < 0.05,
    approach    = if_else(str_detect(dataset, "_kmeans"),
                          "PCA+kmeans", "UMAP+HDBSCAN"),
    clean       = dataset %>%
                    str_remove("_kmeans") %>%
                    str_remove("_nolags") %>%
                    str_remove("_lags"),
    sampling    = str_extract(clean, "^(SP|COMP)"),
    pred_method = str_extract(clean, "(RF|RK|OK)"),  # ← OK included
    lags        = str_detect(dataset, "_lags")
  ) %>%
  select(-clean)

all_comparisons <- purrr::map_dfr(all_val, ~ .x$comparison)

# ── KW summary ────────────────────────────────────────────
kw_summary <- all_kw %>%
  group_by(dataset, approach, sampling, pred_method, lags) %>%
  summarise(
    n_significant   = sum(significant, na.rm = TRUE),
    n_total         = n(),
    pct_significant = round(100 * mean(significant, na.rm = TRUE)),
    sig_vars        = paste(variable[significant], collapse = ", "),
    .groups         = "drop"
  ) %>%
  arrange(approach, sampling, pred_method, lags)

kw_summary %>%
  select(approach, sampling, pred_method, lags,
         n_significant, sig_vars) %>%
  print(n = 30)

# ── Plot ──────────────────────────────────────────────────
kw_summary %>%
  mutate(
    config     = paste(sampling, pred_method, sep = "_"),
    lags_label = ifelse(lags, "With lags", "No lags"),
    approach   = factor(approach,
                        levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = config, y = n_significant,
             fill = approach, alpha = lags_label)) +
  geom_col(position = "dodge") +
  scale_alpha_manual(values = c("No lags" = 0.6,
                                "With lags" = 1.0)) +
  scale_fill_manual(values = c("PCA+kmeans"   = "#4E9AF1",
                               "UMAP+HDBSCAN" = "#E05C3A")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title    = "Agronomic zone validation: significant soil variables",
    subtitle = "Kruskal-Wallis p < 0.05, held-out validation (n=59)",
    x        = "Sampling × Method",
    y        = "N significant variables (of 9)",
    fill     = "Clustering approach",
    alpha    = "Spatial lags"
  )

write_csv(all_kw,          "../outputs/zone_kruskal_wallis_all.csv")
write_csv(all_comparisons, "../outputs/zone_agronomic_validation_all.csv")
write_csv(kw_summary,      "../outputs/kw_summary_all.csv")

# ── Pts per zone diagnostic ───────────────────────────────
purrr::walk(all_val, function(x) {
  pts <- x$val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    count(zone) %>%
    summarise(n_zones    = n(),
              median_pts = median(n),
              min_pts    = min(n),
              zones_lt3  = sum(n < 3))
  cat(x$comparison$dataset[1], "\n")
  print(pts); cat("\n")
})

# ── Effect sizes (eta²) ───────────────────────────────────
calc_kw_effect_size <- function(zone_rast_smooth, val_sf,
                                 soil_vars, label) {
  val_vect    <- terra::vect(val_sf)
  val_zone    <- terra::extract(zone_rast_smooth, val_vect)
  val_sf$zone <- val_zone[, 2]

  purrr::map_dfr(soil_vars, function(v) {
    df <- val_sf %>%
      st_drop_geometry() %>%
      filter(!is.na(zone), !is.na(.data[[v]])) %>%
      select(zone, value = all_of(v))

    if (length(unique(df$zone)) < 2 || nrow(df) < 5) {
      return(tibble(variable = v, H = NA, k = NA,
                    n = NA, eta2 = NA, dataset = label))
    }
    kt   <- kruskal.test(value ~ factor(zone), data = df)
    k    <- length(unique(df$zone))
    n    <- nrow(df)
    H    <- kt$statistic
    eta2 <- (H - k + 1) / (n - k)
    tibble(variable = v, H = H, k = k, n = n,
           eta2 = round(eta2, 4), dataset = label)
  })
}

effect_sizes_umap <- purrr::imap_dfr(pp_named, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) return(NULL)
  calc_kw_effect_size(pp$zone_rast_smooth, ap_val_sf,
                       soil_vars, label = nm)
})

effect_sizes_kmeans <- purrr::imap_dfr(pp_baseline, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) return(NULL)
  calc_kw_effect_size(pp$zone_rast_smooth, ap_val_sf,
                       soil_vars, label = paste0(nm, "_kmeans"))
})

effect_sizes_all <- bind_rows(effect_sizes_umap,
                               effect_sizes_kmeans) %>%
  mutate(
    approach    = if_else(str_detect(dataset, "_kmeans"),
                          "PCA+kmeans", "UMAP+HDBSCAN"),
    clean       = dataset %>%
                    str_remove("_kmeans") %>%
                    str_remove("_nolags") %>%
                    str_remove("_lags"),
    sampling    = str_extract(clean, "^(SP|COMP)"),
    pred_method = str_extract(clean, "(RF|RK|OK)"),  # ← OK included
    lags        = str_detect(dataset, "_lags") &
                    !str_detect(dataset, "_kmeans") |
                    str_detect(dataset, "lags_kmeans")
  ) %>%
  select(-clean)

eta2_summary <- effect_sizes_all %>%
  group_by(dataset, approach, sampling, pred_method, lags) %>%
  summarise(
    mean_eta2   = round(mean(eta2,   na.rm = TRUE), 4),
    median_eta2 = round(median(eta2, na.rm = TRUE), 4),
    n_zones     = first(k),
    .groups     = "drop"
  ) %>%
  arrange(approach, desc(mean_eta2))

print(eta2_summary, n = 30)
write_csv(effect_sizes_all, "../outputs/kw_effect_sizes.csv")
write_csv(eta2_summary,     "../outputs/kw_eta2_summary.csv")

# ── eta² scatter ──────────────────────────────────────────
eta2_summary %>%
  mutate(
    config     = paste(sampling, pred_method, sep = "_"),
    lags_label = ifelse(lags, "With lags", "No lags"),
    approach   = factor(approach,
                        levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = n_zones, y = mean_eta2,
             color = approach, shape = lags_label, label = config)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3) +
  scale_color_manual(values = c("PCA+kmeans"   = "#4E9AF1",
                                "UMAP+HDBSCAN" = "#E05C3A")) +
  theme_minimal() +
  labs(title    = "Zone differentiation strength vs number of zones",
       subtitle = "eta² controls for pts/zone differences",
       x = "Number of zones", y = "Mean eta²",
       color = "Clustering approach", shape = "Spatial lags")

# ── eta² bar chart ────────────────────────────────────────
eta2_summary %>%
  mutate(
    config     = paste(sampling, pred_method, sep = "_"),
    lags_label = ifelse(lags, "With lags", "No lags"),
    approach   = factor(approach,
                        levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = reorder(paste(config, lags_label), mean_eta2),
             y = mean_eta2, fill = approach, alpha = lags_label)) +
  geom_col() +
  geom_text(aes(label = paste0("k=", n_zones)),
            hjust = -0.2, size = 3) +
  scale_fill_manual(values = c("PCA+kmeans"   = "#4E9AF1",
                               "UMAP+HDBSCAN" = "#E05C3A")) +
  scale_alpha_manual(values = c("No lags" = 0.6, "With lags" = 1.0)) +
  coord_flip() +
  expand_limits(y = 0.17) +
  theme_minimal() +
  labs(title    = "Zone differentiation effect size (eta²)",
       subtitle = "k = number of zones",
       x = NULL, y = "Mean eta²",
       fill = "Clustering approach", alpha = "Spatial lags")

save.image("../outputs/workspace_final.RData")
beep("complete")
```
# ============================================================
# Final consolidated results summary
# ============================================================

```{r final-summary}
# ── UMAP+HDBSCAN summary ──────────────────────────────────
summary_umap <- purrr::imap_dfr(mz_named_tuned, function(x, nm) {
  parts  <- strsplit(nm, "_")[[1]]
  samp   <- parts[1]
  method <- parts[2]
  lag_yn <- length(parts) >= 3 && parts[3] == "lags"

  tibble(
    configuration     = nm,
    clustering        = "UMAP+HDBSCAN",
    sampling_design   = ifelse(samp == "SP", "Single-point", "Composite"),
    pred_method       = method,
    spatial_lags      = lag_yn,
    minPts_used       = minpts_final[[nm]],
    n_zones           = x$n_zones,
    pct_noise_prefill = round(x$pct_noise, 1)
  )
}) %>%
  left_join(
    all_kw %>%
      filter(approach == "UMAP+HDBSCAN") %>%
      group_by(dataset) %>%
      summarise(
        n_sig_vars = sum(significant, na.rm = TRUE),
        sig_vars   = paste(variable[significant], collapse = ", "),
        .groups    = "drop"
      ) %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(
    n_sig_vars = replace_na(n_sig_vars, 0L),
    sig_vars   = replace_na(sig_vars,   "none"),
    sig_vars = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(
    zone_validation_all %>%
      filter(approach == "UMAP+HDBSCAN") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

# ── PCA+kmeans summary ────────────────────────────────────
summary_pca <- purrr::imap_dfr(baseline_list, function(b, nm) {
  clean   <- str_remove(nm, "_kmeans")
  has_lag <- str_detect(clean, "_lags")
  clean   <- str_remove(clean, "_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]
  method  <- parts[2]

  tibble(
    configuration     = nm,
    clustering        = "PCA+kmeans",
    sampling_design   = ifelse(samp == "SP", "Single-point", "Composite"),
    pred_method       = method,
    spatial_lags      = has_lag,
    minPts_used       = NA_real_,   # not applicable for kmeans
    n_zones           = b$best_k,
    pct_noise_prefill = 0           # kmeans has no noise
  )
}) %>%
  left_join(
    all_kw %>%
      filter(approach == "PCA+kmeans") %>%
      group_by(dataset) %>%
      summarise(
        n_sig_vars = sum(significant, na.rm = TRUE),
        sig_vars   = paste(variable[significant], collapse = ", "),
        .groups    = "drop"
      ) %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(
    n_sig_vars = replace_na(n_sig_vars, 0L),
    sig_vars   = replace_na(sig_vars,   "none"),
    sig_vars = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(
    zone_validation_all %>%
      filter(approach == "PCA+kmeans") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

# ── Combined final summary ────────────────────────────────
final_summary %>%
  select(clustering, sampling_design, pred_method,
         spatial_lags, n_zones, pct_noise_prefill,
         n_sig_vars, sig_vars, VR, silhouette) %>%
  mutate(
    VR         = round(VR,         3),
    silhouette = round(silhouette, 3)
  ) %>%
  arrange(clustering, sampling_design, pred_method, spatial_lags) %>%
  print(n = 30)
final_summary %>%
  mutate(
    VR             = round(VR,             3),
    silhouette     = round(silhouette,     3),
    pct_noise_prefill = round(pct_noise_prefill, 1),
    sig_vars       = ifelse(sig_vars == "" | is.na(sig_vars),
                            "none", sig_vars)
  ) %>%
  arrange(clustering, sampling_design, pred_method, spatial_lags) %>%
  write_csv("../outputs/final_results_summary_clean.csv")


save.image("../outputs/workspace_final.RData")
beep("complete")
#Among the ten UMAP+HDBSCAN configurations evaluated, composite sampling with RF predictions and spatial lag augmentation (COMP_RF_lags) produced the most agronomically interpretable management zones, with four significantly differentiated soil variables (lbc_1, lb_ceq, ca, p), a variance reduction of 0.463, and a silhouette coefficient of 0.529. Single-point RF without lags (SP_RF_nolags) achieved the highest variance reduction overall (0.756) but with only three significant variables and high noise (43.7%), suggesting overfitting to local sampling variability. Ordinary kriging configurations consistently underperformed RF and regression kriging — SP_OK_nolags produced no significantly differentiated variables and the highest noise prefill percentage (79.6%), while COMP_OK_nolags found 25 zones but only one significant variable, reflecting the limited spatial information content of kriging-only prediction surfaces for zone delineation. Regression kriging configurations occupied an intermediate position, generally outperforming OK but underperforming RF, consistent with the spatial CV results in Phase 3.

#best method: # Rank by composite score
final_summary %>%
  filter(clustering == "UMAP+HDBSCAN") %>%
  mutate(
    vr_rank  = rank(-VR),
    sil_rank = rank(-silhouette),
    kw_rank  = rank(-n_sig_vars),
    composite_rank = vr_rank + sil_rank + kw_rank
  ) %>%
  arrange(composite_rank) %>%
  select(configuration, VR, silhouette, n_sig_vars, composite_rank)
```
```{r validation-metrics}
# ============================================================
# Extended validation metrics — all configurations
# ============================================================
# Combines UMAP+HDBSCAN (val_named) and PCA+kmeans (val_baseline)
# into one unified comparison object first, then computes
# zone_pred_rmse, observed_VR, and R² for everything.
all_comparisons_pca  <- purrr::map_dfr(val_baseline,
                                        ~ .x$comparison)
all_comparisons_full <- bind_rows(all_comparisons,
                                   all_comparisons_pca)

# ── 2. Zone prediction RMSE ───────────────────────────────
zone_pred_rmse <- all_comparisons_full %>%
  filter(!is.na(rmse_vs_zone)) %>%
  group_by(dataset) %>%
  summarise(
    mean_zone_rmse   = round(mean(rmse_vs_zone,   na.rm = TRUE), 3),
    median_zone_rmse = round(median(rmse_vs_zone, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

# ── 3. Observed VR — per variable, then summarise ─────────
# Averaging VR across variables with very different spatial
# distributions is misleading. Instead:
#   - compute VR per variable per configuration
#   - report median (robust to outliers like Zn)
#   - report n_pos_VR (how many variables show positive VR)

calc_zone_consistency <- function(val_with_zones,
                                   soil_vars, label) {
  df <- val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    na.omit()

  if (nrow(df) < 5 || length(unique(df$zone)) < 2) {
    return(tibble(
      configuration    = label,
      observed_VR_mean = NA_real_,
      observed_VR_med  = NA_real_,
      n_pos_VR         = NA_integer_
    ))
  }

  # Per-variable total variance
  total_var_vec <- df %>%
    select(all_of(soil_vars)) %>%
    summarise(across(everything(),
                     ~ var(.x, na.rm = TRUE))) %>%
    unlist()

  # Per-variable within-zone variance (mean across zones)
  per_var_vr <- purrr::map_dbl(soil_vars, function(v) {
    within_var <- df %>%
      group_by(zone) %>%
      summarise(wv = var(.data[[v]], na.rm = TRUE),
                .groups = "drop") %>%
      pull(wv) %>%
      mean(na.rm = TRUE)

    1 - (within_var / total_var_vec[[v]])
  })
  names(per_var_vr) <- soil_vars

  tibble(
    configuration    = label,
    observed_VR_mean = round(mean(per_var_vr,   na.rm = TRUE), 3),
    observed_VR_med  = round(median(per_var_vr, na.rm = TRUE), 3),
    n_pos_VR         = sum(per_var_vr > 0,      na.rm = TRUE)
  )
}

# UMAP configs
obs_vr_umap <- purrr::imap_dfr(val_named, function(x, nm) {
  calc_zone_consistency(x$val_with_zones, soil_vars,
                        label = nm)
}) %>% rename(config_join = configuration)

# PCA+kmeans configs
obs_vr_pca <- purrr::imap_dfr(val_baseline, function(x, nm) {
  calc_zone_consistency(x$val_with_zones, soil_vars,
                        label = paste0(nm, "_kmeans"))
}) %>% rename(config_join = configuration)

obs_vr_all <- bind_rows(obs_vr_umap, obs_vr_pca)

# Print per-var breakdown for key configs to inspect Zn effect
cat("\n=== Per-variable observed VR — key configs ===\n")
list(
  COMP_RF_lags = val_named[["COMP_RF_lags"]],
  SP_RF_nolags = val_named[["SP_RF_nolags"]]
) %>%
  purrr::imap_dfr(function(x, nm) {
    df <- x$val_with_zones %>%
      st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>%
      na.omit()

    total_var_vec <- df %>%
      select(all_of(soil_vars)) %>%
      summarise(across(everything(),
                       ~ var(.x, na.rm = TRUE))) %>%
      unlist()

    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>%
        group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE),
                  .groups = "drop") %>%
        pull(wv) %>%
        mean(na.rm = TRUE)
      tibble(config = nm, variable = v,
             obs_VR = round(1 - wv / total_var_vec[[v]], 3))
    })
  }) %>%
  pivot_wider(names_from = config, values_from = obs_VR) %>%
  arrange(COMP_RF_lags) %>%
  print()

# ── 4. R² ─────────────────────────────────────────────────
zone_r2 <- all_comparisons_full %>%
  filter(!is.na(pred_zone_mean), !is.na(mean_obs)) %>%
  group_by(dataset) %>%
  summarise(
    r2_obs_pred = round(
      cor(mean_obs, pred_zone_mean,
          use = "complete.obs")^2, 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

# ── 5. Rebuild final_summary ──────────────────────────────
final_summary <- bind_rows(summary_umap, summary_pca) %>%
  mutate(
    config_join = case_when(
      clustering == "PCA+kmeans" &
        !str_detect(configuration, "_kmeans") ~
        paste0(configuration, "_kmeans"),
      TRUE ~ configuration
    ),
    sig_vars = ifelse(sig_vars == "" | is.na(sig_vars),
                      "none", sig_vars)
  ) %>%
  left_join(zone_pred_rmse,         by = "config_join") %>%
  left_join(obs_vr_all,             by = "config_join") %>%
  left_join(zone_r2 %>%
              select(config_join, r2_obs_pred),
            by = "config_join") %>%
  select(-config_join)

# ── 6. Print ──────────────────────────────────────────────
cat("\n=== Full validation metrics ===\n")
final_summary %>%
  select(clustering, sampling_design, pred_method,
         spatial_lags, n_zones, VR, silhouette,
         n_sig_vars, median_zone_rmse,
         observed_VR_med, n_pos_VR, r2_obs_pred) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  arrange(clustering, sampling_design,
          pred_method, spatial_lags) %>%
  print(n = 30)

# ── 7. Composite ranking ──────────────────────────────────
cat("\n=== Composite ranking — all configurations ===\n")
final_summary %>%
  mutate(
    vr_rank      = rank(-VR,                na.last = "keep"),
    sil_rank     = rank(-silhouette,        na.last = "keep"),
    kw_rank      = rank(-n_sig_vars,        na.last = "keep"),
    rmse_rank    = rank(median_zone_rmse,   na.last = "keep"),
    obs_vr_rank  = rank(-observed_VR_med,   na.last = "keep"),
    npos_rank    = rank(-n_pos_VR,          na.last = "keep"),
    r2_rank      = rank(-r2_obs_pred,       na.last = "keep"),
    composite_rank = vr_rank + sil_rank + kw_rank +
                     rmse_rank + obs_vr_rank +
                     npos_rank + r2_rank
  ) %>%
  arrange(composite_rank) %>%
  select(configuration, clustering, VR, silhouette,
         n_sig_vars, median_zone_rmse,
         observed_VR_med, n_pos_VR,
         r2_obs_pred, composite_rank) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  print(n = 20)

# ── 8. Save ───────────────────────────────────────────────
write_csv(final_summary,
          "../outputs/final_results_summary_complete.csv")
save.image("../outputs/workspace_final.RData")
beep("complete")
```
```{r}
# ── Fix 1: Run OK validation that was missed ──────────────
val_ok_missing <- list(
  SP_OK_nolags   = validate_zones_with_holdout(
    pp_named[["SP_OK_nolags"]]$zone_rast_smooth,
    ap_val_sf, pred_stack_SP, soil_vars,
    methods = "OK", label = "SP_OK_nolags"),
  COMP_OK_nolags = validate_zones_with_holdout(
    pp_named[["COMP_OK_nolags"]]$zone_rast_smooth,
    ap_val_sf, pred_stack_COMP, soil_vars,
    methods = "OK", label = "COMP_OK_nolags")
)

# Add to val_named
val_named <- c(val_named, val_ok_missing)

# ── Fix 2: Harmonise PCA labels ───────────────────────────
# val_baseline comparison datasets use _kmeans suffix
# but final_summary PCA no-lags configs are stored as
# SP_RF, COMP_RF etc. — add _kmeans to match

# Check what final_summary PCA configurations are called
final_summary %>%
  filter(clustering == "PCA+kmeans") %>%
  pull(configuration)
# Should show: SP_RF, SP_RK, SP_OK, COMP_RF, COMP_RK, COMP_OK,
#              SP_RF_lags_kmeans, SP_RK_lags_kmeans, etc.

# The no-lags PCA configs need _kmeans appended to match
# val_baseline dataset labels
final_summary <- final_summary %>%
  mutate(
    # Standardise PCA+kmeans config names to always have _kmeans
    config_join = case_when(
      clustering == "PCA+kmeans" &
        !str_detect(configuration, "_kmeans") ~
        paste0(configuration, "_kmeans"),
      TRUE ~ configuration
    )
  )

# ── Fix 3: Rebuild all metrics using config_join key ──────
all_comparisons_pca <- purrr::map_dfr(val_baseline,
                                       ~ .x$comparison)

all_comparisons_full <- bind_rows(
  all_comparisons,       # UMAP — labels like SP_RF_nolags
  all_comparisons_pca    # PCA  — labels like SP_RF_kmeans
)

zone_pred_rmse <- all_comparisons_full %>%
  filter(!is.na(rmse_vs_zone)) %>%
  group_by(dataset) %>%
  summarise(
    mean_zone_rmse   = round(mean(rmse_vs_zone,   na.rm = TRUE), 3),
    median_zone_rmse = round(median(rmse_vs_zone, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

zone_r2 <- all_comparisons_full %>%
  filter(!is.na(pred_zone_mean), !is.na(mean_obs)) %>%
  group_by(dataset) %>%
  summarise(
    r2_obs_pred = round(
      cor(mean_obs, pred_zone_mean, use = "complete.obs")^2, 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

# Observed VR — UMAP
obs_vr_umap <- purrr::imap_dfr(val_named, function(x, nm) {
  calc_zone_consistency(x$val_with_zones, soil_vars, label = nm)
}) %>% rename(config_join = configuration)

# Observed VR — PCA (labels already have _kmeans)
obs_vr_pca <- purrr::imap_dfr(val_baseline, function(x, nm) {
  calc_zone_consistency(
    x$val_with_zones, soil_vars,
    label = paste0(nm, "_kmeans")
  )
}) %>% rename(config_join = configuration)

obs_vr_all <- bind_rows(obs_vr_umap, obs_vr_pca)

# ── Join everything via config_join ───────────────────────
final_summary <- bind_rows(summary_umap, summary_pca) %>%
  mutate(
    config_join = case_when(
      clustering == "PCA+kmeans" &
        !str_detect(configuration, "_kmeans") ~
        paste0(configuration, "_kmeans"),
      TRUE ~ configuration
    ),
    sig_vars = ifelse(sig_vars == "" | is.na(sig_vars),
                      "none", sig_vars)
  ) %>%
  left_join(zone_pred_rmse, by = "config_join") %>%
  left_join(obs_vr_all %>%
              select(config_join, observed_VR),
            by = "config_join") %>%
  left_join(zone_r2 %>%
              select(config_join, r2_obs_pred),
            by = "config_join") %>%
  select(-config_join)   # drop helper column

# ── Print and save ────────────────────────────────────────
cat("\n=== Full validation metrics — all configurations ===\n")
final_summary %>%
  select(clustering, sampling_design, pred_method,
         spatial_lags, n_zones,
         VR, silhouette, n_sig_vars,
         mean_zone_rmse, observed_VR, r2_obs_pred) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  arrange(clustering, sampling_design, pred_method,
          spatial_lags) %>%
  print(n = 30)

cat("\n=== UMAP+HDBSCAN composite ranking (6 metrics) ===\n")
final_summary %>%
  filter(clustering == "UMAP+HDBSCAN") %>%
  mutate(
    vr_rank     = rank(-VR),
    sil_rank    = rank(-silhouette),
    kw_rank     = rank(-n_sig_vars),
    rmse_rank   = rank(mean_zone_rmse),
    obs_vr_rank = rank(-observed_VR,  na.last = "keep"),
    r2_rank     = rank(-r2_obs_pred,  na.last = "keep"),
    composite_rank = vr_rank + sil_rank + kw_rank +
                     rmse_rank + obs_vr_rank + r2_rank
  ) %>%
  arrange(composite_rank) %>%
  select(configuration, VR, silhouette, n_sig_vars,
         mean_zone_rmse, observed_VR, r2_obs_pred,
         composite_rank) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  print(n = 10)

write_csv(final_summary,
          "../outputs/final_results_summary_complete.csv")
save.image("../outputs/workspace_final.RData")
beep("complete")
```


```{r final figure comparison}
# ============================================================
# THESIS FIGURE — Zone quality metrics across all configurations
# ============================================================



# ── Prep data ─────────────────────────────────────────────
fig_data <- final_summary %>%
  mutate(
    x_group = factor(
      paste0(pred_method, ifelse(spatial_lags, "+lags", "")),
      levels = c("RF", "RF+lags", "RK", "RK+lags", "OK")
    ),
    sampling_f   = factor(sampling_design,
                          levels = c("Single-point", "Composite")),
    clustering_f = factor(clustering,
                          levels = c("PCA+kmeans", "UMAP+HDBSCAN")),
    # Clean sig_vars for labelling — shorten names
    sig_label = case_when(
      n_sig_vars == 0 ~ "",
      n_sig_vars <= 2 ~ sig_vars,
      TRUE            ~ paste0(n_sig_vars, " vars")
    )
  )
fig_data <- fig_data %>%
  mutate(
    is_best = configuration == "COMP_RF_lags",
    # Add star to label
    x_group_label = ifelse(
      is_best,
      paste0(as.character(x_group), " ★"),
      as.character(x_group)
    )
  )
clust_colors <- c("PCA+kmeans"   = "#4E9AF1",
                  "UMAP+HDBSCAN" = "#E05C3A")
lag_alpha    <- c("No lags" = 0.5, "With lags" = 1.0)
lag_aes      <- aes(x     = x_group,
                    fill  = clustering_f,
                    alpha = factor(spatial_lags,
                                   levels = c(FALSE, TRUE),
                                   labels = c("No lags",
                                              "With lags")))

base_theme <- theme_bw(base_size = 9) +
  theme(
    legend.position  = "none",
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(size = 8, face = "bold"),
    axis.text.x      = element_text(size = 7, angle = 30,
                                     hjust = 1),
    panel.grid.major.x = element_blank()
  )

# ── Panel A — VR ──────────────────────────────────────────
pA <- fig_data %>%
  ggplot(lag_aes) +
  aes(y = VR) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.40, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors,
                    name   = "Clustering") +
  scale_alpha_manual(values = lag_alpha,
                     name   = "Spatial lags") +
  scale_y_continuous(limits = c(0, 0.85),
                     breaks = seq(0, 0.8, 0.2)) +
  labs(title = "A).  Variance Reduction",
       subtitle = "Higher = zones more internally homogeneous, max is 1.0",
       x = NULL, y = "VR") +
  base_theme

# ── Panel B — Silhouette ──────────────────────────────────
pB <- fig_data %>%
  ggplot(lag_aes) +
  aes(y = silhouette) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.50, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors,
                    name   = "Clustering") +
  scale_alpha_manual(values = lag_alpha,
                     name   = "Spatial lags") +
  scale_y_continuous(limits = c(0, 0.65),
                     breaks = seq(0, 0.6, 0.2)) +
  labs(title = "B).  Silhouette Coefficient",
       subtitle = "Higher = zones more spatially distinct, max is 1.0",
       x = NULL, y = "Silhouette") +
  base_theme

# ── Panel C — KW sig vars (no text labels, just numbers) ──
pC <- fig_data %>%
  ggplot(lag_aes) +
  aes(y = n_sig_vars) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_text(aes(label = ifelse(n_sig_vars > 0,
                                as.character(n_sig_vars), "")),
            position  = position_dodge(0.8),
            vjust     = -0.4,
            size      = 2.5,
            color     = "grey30") +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors,
                    name   = "Clustering") +
  scale_alpha_manual(values = lag_alpha,
                     name   = "Spatial lags") +
  scale_y_continuous(breaks = 0:6, limits = c(0, 6.5)) +
  labs(title    = "C). Significant Soil Properties (Out of 9) (KW, p < 0.05)",
       subtitle = "Validation points: n = 59",
       x        = "Prediction method",
       y        = "# of properties significant") +
  base_theme

# ── Shared legend panel ───────────────────────────────────
legend_plot <- fig_data %>%
  ggplot(lag_aes) +
  aes(y = VR) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha,   name = "Spatial lags") +
  theme_bw(base_size = 9) +
  theme(legend.position  = "bottom",
        legend.box       = "horizontal",
        legend.title     = element_text(size = 8, face = "bold"),
        legend.text      = element_text(size = 8),
        legend.key.size  = unit(0.4, "cm"))

shared_legend <- cowplot::get_legend(legend_plot)
# install.packages("cowplot") if needed

# ── Combine ───────────────────────────────────────────────
fig_combined <- (pA / pB / pC) +
  plot_layout(heights = c(1, 1, 1.2))

# Print interactively first
print(fig_combined)
```
```{r final table}
# ============================================================
# THESIS RESULTS TABLE — All configurations, all metrics
# ============================================================

# ── 1. Composite rank ─────────────────────────────────────
all_ranked <- final_summary %>%
  mutate(
    vr_rank     = rank(-VR,              na.last = "keep"),
    sil_rank    = rank(-silhouette,      na.last = "keep"),
    kw_rank     = rank(-n_sig_vars,      na.last = "keep"),
    rmse_rank   = rank(median_zone_rmse, na.last = "keep"),
    obs_vr_rank = rank(-observed_VR_med, na.last = "keep"),
    npos_rank   = rank(-n_pos_VR,        na.last = "keep"),
    r2_rank     = rank(-r2_obs_pred,     na.last = "keep"),
    Rank        = round(vr_rank + sil_rank + kw_rank +
                          rmse_rank + obs_vr_rank +
                          npos_rank + r2_rank, 1)
  ) %>%
  select(configuration, Rank)

# ── 2. Table data ─────────────────────────────────────────
table_data <- final_summary %>%
  left_join(all_ranked, by = "configuration") %>%
  mutate(
    Approach  = clustering,
    Sampling  = ifelse(sampling_design == "Single-point",
                       "SP", "COMP"),
    Method    = pred_method,
    Lags      = ifelse(spatial_lags, "Yes", "No"),
    Zones     = n_zones,
    VR_pred   = round(VR,               3),
    Sil       = round(silhouette,       3),
    N_sig     = n_sig_vars,
    RMSE_zone = round(median_zone_rmse, 1),
    VR_obs    = round(observed_VR_med,  3),   # median
    N_pos     = n_pos_VR,                      # new column
    R2        = round(r2_obs_pred,      3)
  ) %>%
  select(Approach, Sampling, Method, Lags, Zones,
         VR_pred, Sil, N_sig, RMSE_zone,
         VR_obs, N_pos, R2, Rank) %>%
  arrange(Approach, Sampling, Method, Lags)

# ── 3. Best values ────────────────────────────────────────
best <- list(
  VR_pred   = max(table_data$VR_pred,   na.rm = TRUE),
  Sil       = max(table_data$Sil,       na.rm = TRUE),
  N_sig     = max(table_data$N_sig,     na.rm = TRUE),
  RMSE_zone = min(table_data$RMSE_zone, na.rm = TRUE),
  VR_obs    = max(table_data$VR_obs,    na.rm = TRUE),
  N_pos     = max(table_data$N_pos,     na.rm = TRUE),
  R2        = max(table_data$R2,        na.rm = TRUE),
  Rank      = min(table_data$Rank,      na.rm = TRUE)
)

# ── 4. Build gt table ─────────────────────────────────────
tbl <- table_data %>%
  gt(groupname_col = "Approach") %>%

  cols_label(
    Sampling  = "Sampling",
    Method    = "Pred. method",
    Lags      = "Lags",
    Zones     = md("*k*"),
    VR_pred   = md("VR~pred~"),
    Sil       = "Silhouette",
    N_sig     = md("*N*~sig~"),
    RMSE_zone = md("RMSE~zone~"),
    VR_obs    = md("VR~obs~ (median)"),
    N_pos     = md("*N*~pos~"),
    R2        = md("*R*²"),
    Rank      = "Rank"
  ) %>%

  tab_spanner(
    label   = md("**Configuration**"),
    columns = c(Sampling, Method, Lags, Zones)
  ) %>%
  tab_spanner(
    label   = md("**Zone structure** *(all 117,677 grid pixels)*"),
    columns = c(VR_pred, Sil)
  ) %>%
  tab_spanner(
    label   = md("**Agronomic validation** *(held-out, n = 59)*"),
    columns = c(N_sig, RMSE_zone, VR_obs, N_pos, R2)
  ) %>%
  tab_spanner(
    label   = md("**Overall**"),
    columns = Rank
  ) %>%

  # Row group styling
  tab_style(
    style     = cell_fill(color = "#f0f4f8"),
    locations = cells_row_groups()
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", size = px(11)),
    locations = cells_row_groups()
  ) %>%

  # Bold best per column
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = VR_pred,
                           rows    = VR_pred == best$VR_pred)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = Sil,
                           rows    = Sil == best$Sil)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = N_sig,
                           rows    = N_sig == best$N_sig)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = RMSE_zone,
                           rows    = !is.na(RMSE_zone) &
                                       RMSE_zone == best$RMSE_zone)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = VR_obs,
                           rows    = !is.na(VR_obs) &
                                       VR_obs == best$VR_obs)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = N_pos,
                           rows    = !is.na(N_pos) &
                                       N_pos == best$N_pos)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = R2,
                           rows    = !is.na(R2) &
                                       R2 == best$R2)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(columns = Rank,
                           rows    = !is.na(Rank) &
                                       Rank == best$Rank)
  ) %>%

  # Blue highlight — best overall row
  tab_style(
    style = list(
      cell_fill(color = "#d6eaf8"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Sampling == "COMP" & Method == "RF" &
               Lags == "Yes" & Approach == "UMAP+HDBSCAN"
    )
  ) %>%

  # Red shading for negative VR_obs
  tab_style(
    style     = cell_fill(color = "#fde8e8"),
    locations = cells_body(columns = VR_obs,
                           rows    = !is.na(VR_obs) &
                                       VR_obs < 0)
  ) %>%
  tab_style(
    style     = cell_text(color = "#c0392b"),
    locations = cells_body(columns = VR_obs,
                           rows    = !is.na(VR_obs) &
                                       VR_obs < 0)
  ) %>%

  # Grey redundant PCA no-lag rows
  tab_style(
    style     = cell_text(color = "grey55"),
    locations = cells_body(
      columns = c(VR_pred, Sil, RMSE_zone,
                  VR_obs, N_pos, R2, Rank),
      rows    = Approach == "PCA+kmeans" &
                  Method %in% c("RF", "RK") &
                  Lags == "No"
    )
  ) %>%

  # Footnotes
  tab_footnote(
    footnote  = md("PCA+k-means zone assignments are insensitive to spatial lag augmentation for RF and RK — all zone structure and validation metrics are identical between lag and no-lag rows. Only *N*~sig~ differs. Greyed values indicate redundancy."),
    locations = cells_column_spanners(
      spanners = "**Configuration**"
    )
  ) %>%
  tab_footnote(
    footnote  = md("COMP OK UMAP+HDBSCAN: degenerate variogram fits (psill = 0) for lbc_1 and zn under composite sampling resulted in only 2 zones with 56.9% noise pre-fill. Zone boundaries are spatially arbitrary, yielding negative VR~obs~ and *N*~pos~ = 2 of 9. See variogram summary table."),
    locations = cells_body(
      columns = VR_obs,
      rows    = Sampling == "COMP" & Method == "OK" &
                  Approach == "UMAP+HDBSCAN"
    )
  ) %>%

  fmt_missing(columns = everything(), missing_text = "—") %>%

  fmt_number(columns = c(VR_pred, Sil, VR_obs, R2),
             decimals = 3) %>%
  fmt_number(columns = RMSE_zone, decimals = 1) %>%
  fmt_number(columns = Rank,      decimals = 1) %>%
  fmt_integer(columns = c(Zones, N_sig, N_pos)) %>%

  tab_caption(
    md("**Table 1.** Zone quality metrics for all 20 management zone configurations. SP = single-point; COMP = composite; RF = random forest; RK = regression kriging; OK = ordinary kriging. *k* = zones after post-processing.")
  ) %>%

  tab_source_note(md("**Metric definitions and data sources:**")) %>%
  tab_source_note(
    md("**VR~pred~** *(all grid pixels)*: proportion of variance in predicted soil property surfaces explained by zone membership. Higher = more internally homogeneous zones. Values ≥ 0.40 considered acceptable (Yuan et al. 2022).")
  ) %>%
  tab_source_note(
    md("**Silhouette** *(all grid pixels)*: mean silhouette coefficient in UMAP embedding space (UMAP+HDBSCAN) or first two PCA scores (PCA+k-means), 5,000-point subsample. Range −1 to 1; higher = more distinct zones.")
  ) %>%
  tab_source_note(
    md("***N*~sig~** *(held-out, n = 59)*: number of soil variables (of 9) with significantly different observed values across zones (Kruskal–Wallis, p < 0.05). Higher = more agronomically differentiated zones.")
  ) %>%
  tab_source_note(
    md("**RMSE~zone~** *(held-out, n = 59)*: median RMSE between observed values at validation points and their zone's mean predicted value across 9 soil variables. Lower = zone profiles better represent independent observations.")
  ) %>%
  tab_source_note(
    md("**VR~obs~ (median)** *(held-out, n = 59)*: median per-variable observed variance reduction across 9 soil variables at validation points, fully independent of prediction surfaces. Median used in place of mean to reduce sensitivity to variables with extreme spatial distributions (notably Zn, which exhibited high nugget variance relative to field-scale gradients). Negative values (red) = majority of variables show no zone-aligned structure in independent observations.")
  ) %>%
  tab_source_note(
    md("***N*~pos~** *(held-out, n = 59)*: number of soil variables (of 9) with positive observed VR at validation points. Higher = more variables showing zone-consistent spatial structure in independent data. Complements VR~obs~ (median) by indicating breadth of zone validity across soil properties.")
  ) %>%
  tab_source_note(
    md("***R*²** *(held-out, n = 59)*: squared Pearson correlation between zone mean predicted and observed values across all soil variables. Measures rank-order agreement between predicted and observed zone profiles. Higher = better.")
  ) %>%
  tab_source_note(
    md("**Rank**: composite rank = sum of per-metric ranks across seven metrics (VR~pred~, Silhouette, *N*~sig~, RMSE~zone~, VR~obs~, *N*~pos~, R²), where rank 1 = best for each metric. Lower = better overall performance.")
  ) %>%
  tab_source_note(
    md("Blue row = best overall configuration (COMP RF + lags, UMAP+HDBSCAN). **Bold** = best per column. Greyed = redundant PCA+k-means no-lag rows (see footnote). '—' = not applicable.")
  ) %>%

  tab_options(
    table.font.size                   = px(11),
    column_labels.font.size           = px(11),
    row_group.font.size               = px(11),
    source_notes.font.size            = px(9),
    footnotes.font.size               = px(9),
    table.width                       = pct(100),
    data_row.padding                  = px(4),
    column_labels.padding             = px(6),
    row_group.padding                 = px(5),
    table.border.top.style            = "solid",
    table.border.top.width            = px(2),
    table.border.top.color            = "#2c3e50",
    column_labels.border.bottom.style = "solid",
    column_labels.border.bottom.width = px(1),
    column_labels.border.bottom.color = "#7f8c8d",
    row_group.border.bottom.style     = "hidden",
    source_notes.border.bottom.style  = "hidden",
    footnotes.border.bottom.style     = "hidden"
  )

tbl
gt::gtsave(tbl, "../outputs/table1_zone_metrics.html")
cat("Table saved to ../outputs/\n")

#Negative median observed variance reduction was observed exclusively in configurations producing k=2 zones, consistent with the known sensitivity of binary field partitioning to boundary misalignment when validation points are sparse relative to field heterogeneity. The exception — COMP RF nolags UMAP+HDBSCAN (k=2, positive observed VR) — demonstrates that prediction surface quality modulates this sensitivity: covariate-informed RF predictions produced a 2-zone boundary sufficiently aligned with real property gradients to yield positive observed VR despite the small zone count. PCA+k-means configurations, which optimise boundaries on linear feature combinations of predicted surfaces rather than the nonlinear manifold structure captured by UMAP, showed the most severe negative observed VR values (median −0.22 to −0.67), suggesting systematic misalignment between PCA-derived zone boundaries and independent observations regardless of prediction method.
```


```{r}
# Per-variable observed VR heatmap — clean supplementary figure
# Already computed in calc_zone_consistency_by_var

obs_vr_by_var_all <- bind_rows(
  # UMAP configs
  purrr::imap_dfr(val_named, function(x, nm) {
    df <- x$val_with_zones %>%
      st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>%
      na.omit()

    if (nrow(df) < 5 || length(unique(df$zone)) < 2)
      return(NULL)

    total_var_vec <- df %>%
      select(all_of(soil_vars)) %>%
      summarise(across(everything(),
                       ~ var(.x, na.rm = TRUE))) %>%
      unlist()

    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>%
        group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE),
                  .groups = "drop") %>%
        pull(wv) %>%
        mean(na.rm = TRUE)

      tibble(
        configuration = nm,
        variable      = v,
        obs_VR        = round(1 - wv /
                                total_var_vec[[v]], 3)
      )
    })
  }),

  # PCA configs
  purrr::imap_dfr(val_baseline, function(x, nm) {
    df <- x$val_with_zones %>%
      st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>%
      na.omit()

    if (nrow(df) < 5 || length(unique(df$zone)) < 2)
      return(NULL)

    total_var_vec <- df %>%
      select(all_of(soil_vars)) %>%
      summarise(across(everything(),
                       ~ var(.x, na.rm = TRUE))) %>%
      unlist()

    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>%
        group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE),
                  .groups = "drop") %>%
        pull(wv) %>%
        mean(na.rm = TRUE)

      tibble(
        configuration = paste0(nm, "_kmeans"),
        variable      = v,
        obs_VR        = round(1 - wv /
                                total_var_vec[[v]], 3)
      )
    })
  })
) %>%
  left_join(
    final_summary %>%
      select(configuration, clustering,
             pred_method, sampling_design,
             spatial_lags, n_zones),
    by = "configuration"
  ) %>%
  # After the left_join to final_summary, add:
  mutate(
    clustering = case_when(
      is.na(clustering) & str_detect(configuration, "_kmeans") ~
        "PCA+kmeans",
      is.na(clustering) ~
        "UMAP+HDBSCAN",
      TRUE ~ clustering
    ),
    approach = factor(clustering,
                      levels = c("PCA+kmeans",
                                 "UMAP+HDBSCAN"))
  ) %>%
  mutate(
    config_label = str_replace_all(
      configuration,
      c("_nolags" = "", "_lags" = "+lags",
        "_kmeans" = "", "COMP_" = "COMP/",
        "SP_" = "SP/")
    ),
    approach = factor(clustering,
                      levels = c("PCA+kmeans",
                                 "UMAP+HDBSCAN")),
    # Clean variable names for display
    variable = factor(variable,
                      levels = c("lbc_1", "lb_ceq",
                                 "ca", "k", "mg",
                                 "mn", "p", "p_h_2",
                                 "zn"))
  )

# Heatmap
p_obs_vr <- obs_vr_by_var_all %>%
  ggplot(aes(x    = config_label,
             y    = variable,
             fill = obs_VR)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", obs_VR)),
            size = 2.2,
            color = ifelse(
              obs_vr_by_var_all$obs_VR > 0.3 |
                obs_vr_by_var_all$obs_VR < -0.3,
              "white", "grey20")
            ) +
  facet_wrap(~ approach, scales = "free_x") +
      scale_fill_gradient2(
      low      = "#d73027",
      mid      = "#ffffbf",
      high     = "#1a9850",
      midpoint = 0,
      limits   = c(-1, 1),
      oob      = scales::squish,   # ← clamps outliers to red/green
      name     = "Observed VR",
      labels   = scales::number_format(accuracy = 0.1)
    ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x      = element_text(angle = 45,
                                     hjust = 1,
                                     size  = 7),
    axis.text.y      = element_text(size  = 8),
    strip.background = element_rect(fill  = "grey92"),
    strip.text       = element_text(face  = "bold",
                                     size  = 9),
    legend.position  = "right",
    panel.grid       = element_blank()
  ) +
  labs(
    title    = "Supplementary Figure S1. Per-variable observed variance reduction",
    subtitle = paste0(
      "Observed VR computed on held-out validation points ",
      "(n = 59) independently of all prediction surfaces.\n",
      "Green = zones capture variance structure in ",
      "independent observations; ",
      "red = zones do not align with real gradients."
    ),
    x        = NULL,
    y        = "Soil variable",
    caption  = paste0(
      "Observed VR = 1 − (mean within-zone variance / ",
      "total variance) at validation points.\n",
      "Computed per variable independently — avoids ",
      "cross-variable averaging assumptions.\n",
      "Zn shows negative observed VR across most ",
      "configurations, consistent with high nugget:sill ",
      "ratio (SP = 0.937, COMP = 1.000)\n",
      "indicating absent field-scale spatial structure ",
      "for this variable regardless of zone method."
    )
  )

print(p_obs_vr)

ggsave("../outputs/figS1_observed_vr_heatmap.png",
       p_obs_vr,
       width  = 12,
       height = 5,
       dpi    = 300)

write_csv(obs_vr_by_var_all,
          "../outputs/observed_vr_per_variable.csv")

cat("Supplementary figure saved\n")
```
```{r}
calc_obs_effect_size <- function(val_with_zones,
                                  soil_vars, label,
                                  method = "epsilon") {
  df <- val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    na.omit() %>%
    mutate(zone = factor(zone))

  if (nrow(df) < 5 || length(unique(df$zone)) < 2) {
    return(tibble(
      configuration = label,
      variable      = soil_vars,
      effect_size   = NA_real_,
      metric        = method
    ))
  }

  purrr::map_dfr(soil_vars, function(v) {
    aov_fit <- tryCatch(
      aov(reformulate("zone", response = v), data = df),
      error = function(e) NULL
    )
    if (is.null(aov_fit)) {
      return(tibble(configuration = label,
                    variable = v,
                    effect_size = NA_real_,
                    metric = method))
    }

    ss     <- summary(aov_fit)[[1]]
    ss_bet <- ss["zone",      "Sum Sq"]
    ss_wit <- ss["Residuals", "Sum Sq"]
    ss_tot <- ss_bet + ss_wit
    ms_wit <- ss["Residuals", "Mean Sq"]
    k      <- length(unique(df$zone))

    es <- if (method == "omega") {
      # ω² = (SS_between - (k-1)*MS_within) /
      #       (SS_total + MS_within)
      (ss_bet - (k - 1) * ms_wit) / (ss_tot + ms_wit)
    } else {
      # ε² = (SS_between - (k-1)*MS_within) / SS_total
      # Analogous to adjusted R², least biased
      (ss_bet - (k - 1) * ms_wit) / ss_tot
    }

    tibble(
      configuration = label,
      variable      = v,
      effect_size   = round(es, 3),
      metric        = method
    )
  })
}

# Run for all UMAP configs
obs_es_umap <- purrr::imap_dfr(val_named, function(x, nm) {
  calc_obs_effect_size(x$val_with_zones, soil_vars,
                       label  = nm,
                       method = "epsilon")
})

# Run for all PCA+kmeans configs
obs_es_pca <- purrr::imap_dfr(val_baseline, function(x, nm) {
  calc_obs_effect_size(x$val_with_zones, soil_vars,
                       label  = paste0(nm, "_kmeans"),
                       method = "epsilon")
})

obs_es_all <- bind_rows(obs_es_umap, obs_es_pca)

# Summary per config
obs_es_summary <- obs_es_all %>%
  group_by(configuration) %>%
  summarise(
    median_es = round(median(effect_size, na.rm = TRUE), 3),
    mean_es   = round(mean(effect_size,   na.rm = TRUE), 3),
    n_pos_es  = sum(effect_size > 0,      na.rm = TRUE),
    .groups   = "drop"
  )

obs_es_summary %>% arrange(desc(median_es))


obs_es_summary %>%
  left_join(
    final_summary %>%
      # Add _kmeans to no-lags PCA configs to match
      mutate(
        config_join = case_when(
          clustering == "PCA+kmeans" &
            !str_detect(configuration, "_kmeans") ~
            paste0(configuration, "_kmeans"),
          TRUE ~ configuration
        )
      ) %>%
      select(config_join, n_zones, clustering,
             pred_method, sampling_design,
             spatial_lags),
    by = c("configuration" = "config_join")
  ) %>%
  mutate(
    k_group = case_when(
      n_zones == 2       ~ "k=2",
      n_zones == 3       ~ "k=3",
      n_zones %in% 4:5   ~ "k=4-5",
      n_zones >= 6       ~ "k=6+",
      TRUE               ~ NA_character_
    )
  ) %>%
  arrange(k_group, desc(median_es)) %>%
  select(configuration, clustering, n_zones,
         k_group, median_es, n_pos_es) %>%
  print(n = 25)

#UMAP RF configurations (0.022–0.046) outperform PCA OK (0.003–0.039) at comparable zone counts. The three UMAP RF configs are all positive and meaningfully above zero. SP_OK_kmeans at 0.003 is essentially zero — 6 zones from PCA on OK predictions explains almost nothing at validation points

#Using Cohen's benchmarks for eta-squared family effect sizes:
# 
# ε² ≥ 0.01 = small effect
# ε² ≥ 0.06 = medium effect
# ε² ≥ 0.14 = large effect

#epsilon^2 = (SS_between - (k-1)*MS_within) / SS_total
```


```{r picp-supplementary}
# ============================================================
# SUPPLEMENTARY — Prediction Interval Coverage Probability
# ============================================================
# PICP: proportion of observed validation point values that
# fall within their zone's predicted mean ± 1 SD.
# Expected coverage under normality = 68.3%.
# Configurations below ~50% have zones too wide or too
# poorly calibrated for reliable variable-rate prescription.
#
# Reported as supplementary calibration check only —
# NOT included in composite rank (partially redundant
# with RMSE_zone and would double-weight that dimension).
#
# Reference: Wadoux et al. (2021) Geoderma — PICP as
# spatial prediction calibration metric; Brus et al. (2011)
# European Journal of Soil Science — coverage probability
# for soil mapping validation.
# ============================================================

# ── 1. Compute zone SD from prediction stack ──────────────
# For each config, extract predicted values for all pixels
# in each zone, compute SD per zone per variable.
# Then check whether observed val point value falls within
# zone_mean ± 1 SD.

calc_picp <- function(val_result, pred_stack,
                       soil_vars, methods, label) {

  val_sf <- val_result$val_with_zones

  if (is.null(val_sf) || nrow(val_sf) == 0) {
    cat(label, "— skipped (no validation data)\n")
    return(NULL)
  }

  # Extract zone assignment for each validation point
  val_df <- val_sf %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars))

  if (nrow(val_df) < 5) return(NULL)

  # For each variable, get zone mean and SD from pred stack
  purrr::map_dfr(soil_vars, function(v) {
    col_nm <- paste0(v, "_", methods)
    if (!col_nm %in% names(pred_stack)) return(NULL)

    # Join zone assignments to pred stack pixels
    # Use val_result$comparison which already has
    # pred_zone_mean — we need SD additionally
    zone_sd <- pred_stack %>%
      select(x, y, pred_val = all_of(col_nm)) %>%
      # Get zone for each pixel via spatial join
      # Use the smoothed raster approach — extract zone
      # for all pred_stack pixels then summarise
      # Actually: derive from comparison object + full stack
      # Simpler: compute from all_comparisons_full directly
      # which has pred_zone_mean; we compute SD separately
      { . }  # placeholder — see below

    # Simpler approach: compute zone SD directly from
    # pred_stack pixels joined to zone raster
    # Get zone assignments for all grid pixels
    # This is available via pp_named or pp_baseline

    # Use comparison object for mean, compute SD from
    # the pred_stack filtered to each zone
    comp_sub <- all_comparisons_full %>%
      filter(dataset == label, variable == v) %>%
      select(zone, pred_zone_mean)

    if (nrow(comp_sub) == 0) return(NULL)

    # For SD: we need all predicted values per zone
    # Join pred_stack to zone assignments
    # Zone assignments are in val_result$val_with_zones
    # for validation points only — not full grid
    # Use a different approach: compute SD from the
    # observed values within each zone as a proxy,
    # OR use the pred_stack with zone raster

    # Most practical: use observed values spread within
    # each zone as the interval basis
    # This is the "empirical coverage" version
    obs_by_zone <- val_df %>%
      select(zone, obs = all_of(v)) %>%
      filter(!is.na(obs), !is.na(zone))

    if (nrow(obs_by_zone) < 3) return(NULL)

    # Join zone mean predictions
    covered <- obs_by_zone %>%
      left_join(comp_sub, by = "zone") %>%
      filter(!is.na(pred_zone_mean)) %>%
      # Get zone SD from pred stack
      left_join(
        pred_stack %>%
          select(pred_val = all_of(col_nm)) %>%
          # We need zone labels for pred stack pixels
          # Use the zone raster — extract zone for all pixels
          # This is expensive; use val_result zone summary
          # instead: SD of observed values per zone
          bind_cols(
            val_df %>%
              select(zone) %>%
              slice(rep(1:n(), length.out = nrow(pred_stack)))
          ) %>%
          group_by(zone) %>%
          summarise(pred_zone_sd = sd(pred_val, na.rm = TRUE),
                    .groups = "drop"),
        by = "zone"
      ) %>%
      mutate(
        lower    = pred_zone_mean - pred_zone_sd,
        upper    = pred_zone_mean + pred_zone_sd,
        covered  = obs >= lower & obs <= upper
      )

    tibble(
      dataset  = label,
      variable = v,
      n_pts    = nrow(covered),
      n_covered = sum(covered$covered, na.rm = TRUE),
      picp     = round(mean(covered$covered, na.rm = TRUE), 3)
    )
  })
}



#Actually that approach gets complicated fast because we need zone SD from the full prediction stack, not just the validation points. Here's the cleaner implementation that uses the prediction stack directly:


# ============================================================
# SUPPLEMENTARY — Prediction Interval Coverage Probability
# ============================================================

# ── 1. Build zone SD lookup from prediction stack ─────────
# For each configuration, compute SD of predicted values
# per zone per variable using the full grid pixels

get_zone_stats <- function(pred_stack, zone_rast,
                            soil_vars, methods, crs = "EPSG:26917") {
  col_names <- paste0(soil_vars, "_", methods)
  col_names <- col_names[col_names %in% names(pred_stack)]

  # Convert pred_stack to raster and extract zone per pixel
  pred_rast <- pred_stack %>%
    select(x, y, all_of(col_names)) %>%
    rast(type = "xyz", crs = crs)

  # Stack zone raster with predictions
  zone_vals <- as.data.frame(zone_rast, xy = TRUE) %>%
    rename(zone = 3) %>%
    filter(!is.na(zone))

  pred_vals <- as.data.frame(pred_rast, xy = TRUE,
                              na.rm = FALSE)

  combined <- zone_vals %>%
    left_join(pred_vals, by = c("x", "y"))

  # Zone mean and SD per variable
  purrr::map_dfr(col_names, function(col) {
    v <- str_remove(col, paste0("_", methods))
    combined %>%
      filter(!is.na(.data[[col]])) %>%
      group_by(zone) %>%
      summarise(
        pred_zone_mean = mean(.data[[col]], na.rm = TRUE),
        pred_zone_sd   = sd(.data[[col]],   na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(variable = v, method = methods)
  })
}

# ── 2. PICP per config ────────────────────────────────────
calc_picp_config <- function(val_result, zone_stats,
                              soil_vars, label) {
  val_df <- val_result$val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars))

  if (nrow(val_df) < 5) return(NULL)

  purrr::map_dfr(soil_vars, function(v) {
    zs <- zone_stats %>%
      filter(variable == v) %>%
      select(zone, pred_zone_mean, pred_zone_sd)

    obs_df <- val_df %>%
      select(zone, obs = all_of(v)) %>%
      filter(!is.na(obs)) %>%
      left_join(zs, by = "zone") %>%
      filter(!is.na(pred_zone_mean),
             !is.na(pred_zone_sd)) %>%
      mutate(
        lower   = pred_zone_mean - pred_zone_sd,
        upper   = pred_zone_mean + pred_zone_sd,
        covered = obs >= lower & obs <= upper
      )

    if (nrow(obs_df) < 3) return(NULL)

    tibble(
      configuration = label,
      variable      = v,
      n_pts         = nrow(obs_df),
      n_covered     = sum(obs_df$covered, na.rm = TRUE),
      picp          = round(mean(obs_df$covered,
                                  na.rm = TRUE), 3)
    )
  })
}

# ── 3. Run for all UMAP+HDBSCAN configs ──────────────────
picp_umap <- purrr::imap_dfr(val_named, function(x, nm) {
  cfg <- get_pred_stack(nm)

  # Get zone raster for this config
  zone_r <- pp_named[[nm]]$zone_rast_smooth
  if (is.null(zone_r)) return(NULL)

  cat("Computing zone SD for", nm, "...\n")
  zs <- get_zone_stats(cfg$pred_df, zone_r,
                        soil_vars, cfg$method)

  calc_picp_config(x, zs, soil_vars, label = nm)
})

# ── Run for PCA+kmeans configs ────────────────────────────
picp_pca <- purrr::imap_dfr(val_baseline, function(x, nm) {
  # Parse sampling and method directly from baseline name
  # nm is like "SP_RF", "COMP_RK_lags", "SP_OK" etc.
  clean   <- str_remove(nm, "_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]   # SP or COMP
  method  <- parts[2]   # RF, RK, or OK
  has_lag <- str_detect(nm, "_lags")

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_OK = pred_stack_COMP,
    { cat("No pred stack found for", nm, "\n"); return(NULL) }
  )

  zone_r <- pp_baseline[[nm]]$zone_rast_smooth
  if (is.null(zone_r)) {
    cat(nm, "— skipped (NULL zone raster)\n")
    return(NULL)
  }

  cat("Computing zone SD for PCA", nm, "...\n")
  zs <- get_zone_stats(pred_df, zone_r, soil_vars, method)

  calc_picp_config(x, zs, soil_vars,
                   label = paste0(nm, "_kmeans"))
})

picp_all <- bind_rows(picp_umap, picp_pca)

# ── 5. Summary per configuration ─────────────────────────
picp_summary <- picp_all %>%
  group_by(configuration) %>%
  summarise(
    mean_picp   = round(mean(picp,   na.rm = TRUE), 3),
    median_picp = round(median(picp, na.rm = TRUE), 3),
    n_above_50  = sum(picp >= 0.50,  na.rm = TRUE),
    n_above_68  = sum(picp >= 0.683, na.rm = TRUE),
    n_vars      = n(),
    .groups     = "drop"
  ) %>%
  left_join(
    final_summary %>%
      select(configuration, clustering,
             sampling_design, pred_method,
             spatial_lags, n_zones),
    by = "configuration"
  ) %>%
  arrange(desc(median_picp))

cat("\n=== PICP summary — all configurations ===\n")
print(picp_summary, n = 25)

# ── 6. Supplementary figure ──────────────────────────────
# Heatmap: variable × configuration, coloured by PICP
# Reference line at 68.3% (expected under normality)

picp_plot_data <- picp_all %>%
  left_join(
    final_summary %>%
      select(configuration, clustering,
             pred_method, sampling_design,
             spatial_lags),
    by = "configuration"
  ) %>%
  mutate(
    config_label = str_replace_all(
      configuration,
      c("_nolags" = "", "_lags" = "+lags",
        "_kmeans" = "", "COMP_" = "COMP/",
        "SP_" = "SP/")
    ),
    approach = factor(clustering,
                      levels = c("PCA+kmeans",
                                 "UMAP+HDBSCAN")),
    coverage_band = case_when(
      picp >= 0.683 ~ "≥68% (good)",
      picp >= 0.50  ~ "50–68% (marginal)",
      TRUE          ~ "<50% (poor)"
    ),
    coverage_band = factor(
      coverage_band,
      levels = c("≥68% (good)",
                 "50–68% (marginal)",
                 "<50% (poor)")
    )
  )

# Tile plot
p_picp <- picp_plot_data %>%
  ggplot(aes(x    = config_label,
             y    = variable,
             fill = picp)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = scales::percent(picp,
                                         accuracy = 1)),
            size = 2.2, color = "grey20") +
  facet_wrap(~ approach, scales = "free_x") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9850",
    midpoint = 0.683,
    limits   = c(0, 1),
    name     = "PICP",
    labels   = scales::percent
  ) +
  geom_hline(yintercept = 0,  # not meaningful here
             linetype = "blank") +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x      = element_text(angle = 45,
                                     hjust = 1,
                                     size  = 7),
    axis.text.y      = element_text(size = 8),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold",
                                     size = 9),
    legend.position  = "right",
    panel.grid       = element_blank()
  ) +
  labs(
    title    = "Supplementary Figure S1. Prediction interval coverage probability (PICP)",
    subtitle = paste0(
      "Proportion of held-out validation points (n = 59) ",
      "whose observed value falls within\n",
      "zone predicted mean ± 1 SD. ",
      "Green = ≥68.3% (expected under normality); ",
      "red = <50% (poor calibration)."
    ),
    x = NULL,
    y = "Soil variable",
    caption = paste0(
      "PICP computed per variable per configuration. ",
      "Zone SD derived from full prediction grid (117,677 pixels).\n",
      "Not included in composite ranking — reported as ",
      "supplementary calibration check only.\n",
      "Reference: Wadoux et al. (2021) Geoderma; ",
      "Brus et al. (2011) Eur. J. Soil Sci."
    )
  )

print(p_picp)

# ── 7. Summary bar chart ──────────────────────────────────
# In p_picp_bar, update reference lines and calibration bands
# since none clear 50%, show relative performance instead

p_picp_bar <- picp_summary %>%
  mutate(
    config_label = str_replace_all(
      configuration,
      c("_nolags" = "", "_lags" = "+lags",
        "_kmeans" = "", "COMP_" = "COMP/",
        "SP_" = "SP/")
    ),
    approach = factor(clustering,
                      levels = c("PCA+kmeans",
                                 "UMAP+HDBSCAN")),
    # Reframe bands given overall low coverage
    calibration = case_when(
      median_picp >= 0.40 ~ "Moderate (≥40%)",
      median_picp >= 0.25 ~ "Low (25–40%)",
      TRUE                ~ "Very low (<25%)"
    ),
    calibration = factor(
      calibration,
      levels = c("Moderate (≥40%)",
                 "Low (25–40%)",
                 "Very low (<25%)")
    )
  ) %>%
  ggplot(aes(x    = reorder(config_label, median_picp),
             y    = median_picp,
             fill = calibration)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0.683,
             linetype   = "dashed",
             color      = "grey40",
             linewidth  = 0.5) +
  geom_hline(yintercept = 0.40,
             linetype   = "dotted",
             color      = "grey60",
             linewidth  = 0.4) +
  annotate("text", x = 0.6, y = 0.70,
           label = "68.3% expected", size = 2.5,
           color = "grey40", hjust = 0) +
  annotate("text", x = 0.6, y = 0.42,
           label = "40% threshold", size = 2.5,
           color = "grey60", hjust = 0) +
  facet_wrap(~ approach, scales = "free_x") +
  scale_fill_manual(
    values = c("Moderate (≥40%)"  = "#74add1",
               "Low (25–40%)"     = "#fee08b",
               "Very low (<25%)"  = "#d73027"),
    name   = "Coverage"
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 0.75)
  ) +
  coord_flip() +
  theme_bw(base_size = 9) +
  theme(
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold",
                                       size = 9),
    axis.text.y        = element_text(size = 7),
    legend.position    = "bottom",
    panel.grid.major.y = element_blank()
  ) +
  labs(
    title    = "Supplementary Figure S2. Median PICP per configuration",
    subtitle = paste0(
      "Median across 9 soil variables. ",
      "Dashed = 68.3% theoretical expectation (normality); ",
      "dotted = 40% reference.\n",
      "All configurations below 50%, indicating systematic ",
      "underestimation of within-zone variability ",
      "by prediction surfaces."
    ),
    x        = NULL,
    y        = "Median PICP",
    caption  = paste0(
      "Low PICP across all configurations reflects ",
      "smooth prediction surfaces underestimating\n",
      "the nugget-scale variability present in ",
      "independent validation observations.\n",
      "PICP is reported as a supplementary calibration ",
      "check and is not included in the composite ranking."
    )
  )

print(p_picp_bar)
# ── 8. Save ───────────────────────────────────────────────
write_csv(picp_all,     "../outputs/picp_per_variable.csv")
write_csv(picp_summary, "../outputs/picp_summary.csv")

ggsave("../outputs/figS1_picp_heatmap.png",
       p_picp,     width = 12, height = 7,  dpi = 300)
ggsave("../outputs/figS2_picp_bar.png",
       p_picp_bar, width = 8,  height = 10, dpi = 300)

cat("PICP outputs saved to ../outputs/\n")
# <!-- The key design decisions worth noting — the midpoint of the colour scale in the heatmap is set to 0.683 (the theoretical expected coverage under normality) so green = at or above expectation and red = below. The bar chart uses three calibration bands with explicit thresholds so a reader can immediately see which configurations are prescription-ready versus marginal versus poor. -->
# 
# <!-- The `get_zone_stats` function extracts SD from the full prediction grid pixels per zone rather than from the validation points — this is important because computing SD from only ~30 validation points per zone would be very noisy. Using the full 117k pixels gives a stable estimate of the true predicted distribution within each zone. -->
# <!-- ``` -->

```







# ============================================================
# PHASE 9 — Save Outputs
# ============================================================

```{r save-outputs}
# Prediction stacks
write_csv(pred_stack_SP,       "../outputs/prediction_stack_SP.csv")
write_csv(pred_stack_COMP,     "../outputs/prediction_stack_COMP.csv")
write_csv(pred_stack_SP_lag,   "../outputs/prediction_stack_SP_lag.csv")
write_csv(pred_stack_COMP_lag, "../outputs/prediction_stack_COMP_lag.csv")

# Zone assignments — all 10 UMAP+HDBSCAN configurations
purrr::iwalk(mz_named_tuned, function(x, nm) {
  write_csv(x$result_df,
            paste0("../outputs/zones_", nm, ".csv"))
})

# PCA+kmeans baselines — all 10
write_csv(baseline_SP_RF$result_df,
          "../outputs/zones_SP_RF_PCA_kmeans.csv")
write_csv(baseline_SP_RK$result_df,
          "../outputs/zones_SP_RK_PCA_kmeans.csv")
write_csv(baseline_SP_OK$result_df,
          "../outputs/zones_SP_OK_PCA_kmeans.csv")
write_csv(baseline_COMP_RF$result_df,
          "../outputs/zones_COMP_RF_PCA_kmeans.csv")
write_csv(baseline_COMP_RK$result_df,
          "../outputs/zones_COMP_RK_PCA_kmeans.csv")
write_csv(baseline_COMP_OK$result_df,
          "../outputs/zones_COMP_OK_PCA_kmeans.csv")

# Final consolidated summary
final_summary <- kw_summary %>%
  left_join(
    zone_validation_all %>%
      select(configuration, sampling, pred_method,
             lags, approach, VR, silhouette),
    by = c("dataset" = "configuration")
  )

write_csv(final_summary, "../outputs/final_results_summary.csv")
save.image("../outputs/workspace_final.RData")
beep("complete")
```
