---
title: "Perry_MZ_umap_PCA_analysis_04_30_26"
format: html
---


# ============================================================
# PHASE 0 — Libraries
# ============================================================

```{r libraries, message=FALSE, warning=FALSE}
library(terra)
library(tidyverse)
library(janitor)
library(sf)
library(sp)
library(gstat)
library(randomForest)
library(quantregForest)   # QRF prediction intervals
library(CAST)
library(caret)
library(umap)
library(dbscan)
library(beepr)
library(broom)
library(FNN)
library(uwot)
library(patchwork)
library(ggtext)
library(gt)
library(gtExtras)
library(ggrepel)
library(cowplot)
library(cluster)
library(scales)
```

# ============================================================
# PHASE 1 — Data Loading & Preparation
# ============================================================

```{r data-load}
# ── Raw datasets ───────────────────────────────────────────
soil_points_nutrients <- read.csv(
  "../LOCAL DATA DO NOT PUSH/soil/perry_soil_points_surface_03_2026.csv"
) %>% clean_names()

ap_val <- read_csv(
  "../LOCAL DATA DO NOT PUSH/Ap_hz_data/hz_0_15_wide_03_30_2026.csv"
)

covariate_stack <- rast(
  "../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m_all_with_SR_indices.tif"
)

# ── Agronomic soil variable subset ────────────────────────

soil_vars <- c("lbc_1", "lb_ceq", "p_h_2", "ca", "k", "mg", "mn", "p", "zn")
cov_names <- names(covariate_stack)

# ── Helper functions ───────────────────────────────────────
add_xy <- function(dat) {
  coords <- st_coordinates(dat)
  dat %>% mutate(X = coords[, 1], Y = coords[, 2])
}

extract_covariates <- function(dat_sf, stack) {
  vect_pts <- terra::vect(dat_sf)
  cov_vals <- terra::extract(stack, vect_pts)
  bind_cols(
    dat_sf %>% st_drop_geometry(),
    cov_vals %>% dplyr::select(-ID)
  ) %>%
    bind_cols(geometry = st_geometry(dat_sf)) %>%
    st_as_sf()
}

# ── SP and COMP modelling datasets ────────────────────────
soil_prop_SP <- soil_points_nutrients %>%
  filter(type == "SPID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

soil_prop_COMP <- soil_points_nutrients %>%
  filter(type == "CID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Validation dataset — HELD OUT, never used in CV ───────
ap_val_sf <- ap_val %>%
  st_as_sf(coords = c("POINT_X", "POINT_Y"), crs = 3857) %>%
  st_transform(26917) %>%
  transmute(
    point_label = paste0("VAL_", Point),
    type        = "VAL",
    lb_ceq = LBCeq_0_15,
    p_h_2  = pH.2_0_15,
    ca     = Ca_0_15,
    k      = K_0_15,
    p      = P_0_15,
     mg     = Mg_0_15,
    mn     = Mn_0_15,
    zn     = Zn_0_15,
    lbc_1  = LBC.1_0_15,
    geometry = geometry
  ) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Spatial distribution check ────────────────────────────
ggplot() +
  geom_sf(data = soil_prop_SP,   color = "steelblue",   size = 2) +
  geom_sf(data = soil_prop_COMP, color = "tomato",      size = 2, shape = 17) +
  geom_sf(data = ap_val_sf,      color = "forestgreen", size = 2, shape = 15) +
  theme_minimal() +
  labs(title = "Spatial distribution: SP (blue), COMP (red), VAL (green)")
```

# ============================================================
# PHASE 2 — Spatial Cross-Validation Setup
# ============================================================

```{r spatial-cv-setup}
make_spatial_folds <- function(sf_data, k = 5, seed = 20260401) {
  coords <- st_coordinates(sf_data)
  set.seed(seed)
  km <- kmeans(coords, centers = k, nstart = 25)
  sf_data$spatial_block <- as.character(km$cluster)

  folds <- CAST::CreateSpacetimeFolds(
    x        = sf_data %>% st_drop_geometry(),
    spacevar = "spatial_block",
    k        = k,
    seed     = seed
  )

  purrr::map(seq_along(folds$index), function(i) {
    list(
      iteration = i,
      seed      = seed,
      train_ids = folds$index[[i]],
      test_ids  = folds$indexOut[[i]]
    )
  })
}

resamples_SP   <- make_spatial_folds(soil_prop_SP,   k = 5)
resamples_COMP <- make_spatial_folds(soil_prop_COMP, k = 5)

make_cutoff <- function(sf_data) max(dist(st_coordinates(sf_data))) / 2
width_lag   <- 75

cutoff_SP   <- make_cutoff(soil_prop_SP)
cutoff_COMP <- make_cutoff(soil_prop_COMP)

# Visualise spatial blocks
soil_prop_SP$spatial_block <- as.character(
  kmeans(st_coordinates(soil_prop_SP), centers = 5, nstart = 25)$cluster
)
ggplot() +
  geom_sf(data = soil_prop_SP, aes(color = spatial_block), size = 3) +
  theme_minimal() +
  labs(title = "Spatial CV blocks — SP dataset", color = "Block")
```

# ============================================================
# PHASE 3 — Model Fitting Helpers (RF, OK, RK)
# ============================================================

```{r model-helpers}
calc_metrics <- function(obs, pred) {
  tibble(
    n          = length(obs),
    RMSE       = sqrt(mean((obs - pred)^2, na.rm = TRUE)),
    MAE        = mean(abs(obs - pred),      na.rm = TRUE),
    R2         = cor(obs, pred, use = "complete.obs")^2,
    mean_error = mean(obs - pred,           na.rm = TRUE)
  )
}

tune_rf_oob <- function(train_df, response, predictors, ntree = 500) {
  x_train   <- train_df %>% select(all_of(predictors))
  y_train   <- train_df[[response]]
  p         <- length(predictors)
  mtry_grid <- unique(pmax(1, c(
    floor(sqrt(p)), floor(p / 4), floor(p / 3), floor(p / 2)
  )))

  fits <- purrr::map(mtry_grid, function(m) {
    set.seed(sample.int(1e6, 1))
    rf <- randomForest(x = x_train, y = y_train,
                       ntree = ntree, mtry = m, importance = TRUE)
    tibble(mtry = m, oob_mse = tail(rf$mse, 1))
  }) %>% bind_rows()

  list(
    best_mtry    = fits %>% slice_min(oob_mse, n = 1) %>% pull(mtry),
    tuning_table = fits
  )
}

fit_variogram_models <- function(vg) {
  vg_df       <- as.data.frame(vg)
  init_nugget <- max(0, min(vg_df$gamma, na.rm = TRUE) * 0.5)
  init_sill   <- max(vg_df$gamma, na.rm = TRUE)
  init_psill  <- max(init_sill - init_nugget, 1e-6)
  init_range  <- max(vg_df$dist, na.rm = TRUE) / 3

  fits <- purrr::map(c("Sph", "Exp"), function(m) {
    tryCatch({
      fit <- fit.variogram(
        vg,
        model = vgm(psill = init_psill, model = m,
                    range = init_range,  nugget = init_nugget)
      )
      if (any(!is.finite(fit$psill)))               return(NULL)
      if (any(!is.finite(fit$range)))               return(NULL)
      if (any(fit$range[fit$model != "Nug"] <= 0))  return(NULL)
      fit
    }, error = function(e) NULL)
  })

  fits <- fits[!sapply(fits, is.null)]
  if (length(fits) == 0) return(NULL)

  sse      <- sapply(fits, function(f) attr(f, "SSErr"))
  best_fit <- fits[[which.min(sse)]]

  # Degeneracy guard — pure nugget = constant surface
  str_row <- best_fit %>% filter(model != "Nug")
  if (nrow(str_row) > 0 && str_row$psill < 1e-6) {
    warning("Degenerate variogram fit: psill ≈ 0. Returning NULL.")
    return(NULL)
  }
  best_fit
}

# ── RF ─────────────────────────────────────────────────────
run_rf_model <- function(train_sf, test_sf, response,
                          predictors, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  tuning  <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  preds <- predict(rf_fit, newdata = test_df %>% select(all_of(predictors)))
  list(
    model       = rf_fit,
    predictions = tibble(obs = test_df[[response]], pred = preds),
    metrics     = calc_metrics(test_df[[response]], preds),
    best_mtry   = tuning$best_mtry
  )
}

# ── OK ─────────────────────────────────────────────────────
run_ok_model <- function(train_sf, test_sf, response, cutoff, width) {
  train_sp <- as(train_sf, "Spatial")
  test_sp  <- as(test_sf,  "Spatial")
  f        <- as.formula(paste(response, "~ 1"))
  vg       <- variogram(f, data = train_sp, cutoff = cutoff, width = width)
  fit      <- fit_variogram_models(vg)
  if (is.null(fit)) return(NULL)

  kr   <- krige(f, locations = train_sp, newdata = test_sp, model = fit)
  pred <- kr$var1.pred
  obs  <- test_sf %>% st_drop_geometry() %>% pull(.data[[response]])
  list(
    variogram   = vg,
    fit         = fit,
    predictions = tibble(obs = obs, pred = pred),
    metrics     = calc_metrics(obs, pred)
  )
}

# ── RK ─────────────────────────────────────────────────────
run_rk_model <- function(train_sf, test_sf, response, predictors,
                          cutoff, width, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  train_sf2 <- train_sf %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))
  test_sf2  <- test_sf  %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))

  tuning <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )

  trend_train        <- predict(rf_fit, newdata = train_df %>%
                                  select(all_of(predictors)))
  trend_test         <- predict(rf_fit, newdata = test_df  %>%
                                  select(all_of(predictors)))
  train_sf2$residual <- train_df[[response]] - trend_train

  train_sp  <- as(train_sf2, "Spatial")
  test_sp   <- as(test_sf2,  "Spatial")
  vg_resid  <- variogram(residual ~ 1, data = train_sp,
                          cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  if (is.null(fit_resid)) {
    return(list(
      metrics               = calc_metrics(test_df[[response]], trend_test),
      best_mtry             = tuning$best_mtry,
      used_residual_kriging = FALSE
    ))
  }

  kr_resid <- krige(residual ~ 1, locations = train_sp,
                    newdata = test_sp, model = fit_resid)
  pred     <- trend_test + kr_resid$var1.pred
  obs      <- test_df[[response]]

  list(
    metrics               = calc_metrics(obs, pred),
    best_mtry             = tuning$best_mtry,
    used_residual_kriging = TRUE
  )
}

# ── Combined iteration runner ──────────────────────────────
run_one_iteration <- function(sf_data, split, response, predictors,
                               dataset_name, cutoff, width, ntree = 500) {
  train_sf <- sf_data[split$train_ids, ]
  test_sf  <- sf_data[split$test_ids,  ]

  rf_res <- run_rf_model(train_sf, test_sf, response, predictors, ntree)
  ok_res <- run_ok_model(train_sf, test_sf, response, cutoff, width)
  rk_res <- run_rk_model(train_sf, test_sf, response, predictors,
                          cutoff, width, ntree)

  bind_rows(
    rf_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RF",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rf_res$best_mtry),
    if (!is.null(ok_res))
      ok_res$metrics %>%
        mutate(dataset = dataset_name, variable = response, method = "OK",
               iteration = split$iteration, seed = split$seed,
               best_mtry = NA_real_),
    rk_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RK",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rk_res$best_mtry,
             used_residual_kriging = rk_res$used_residual_kriging)
  )
}
```

```{r run-cv, warning=FALSE}
results_SP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_SP, function(split_obj) {
    run_one_iteration(soil_prop_SP, split_obj, v, cov_names,
                      "SP", cutoff_SP, width_lag)
  })
})

results_COMP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_COMP, function(split_obj) {
    run_one_iteration(soil_prop_COMP, split_obj, v, cov_names,
                      "COMP", cutoff_COMP, width_lag)
  })
})

results_all <- bind_rows(results_SP, results_COMP)
beep("wilhelm")

# ── CV summary ────────────────────────────────────────────
cv_summary <- results_all %>%
  group_by(dataset, variable, method) %>%
  summarise(
    n_folds     = n(),
    median_R2   = median(R2,   na.rm = TRUE),
    q25_R2      = quantile(R2,   0.25, na.rm = TRUE),
    q75_R2      = quantile(R2,   0.75, na.rm = TRUE),
    median_RMSE = median(RMSE, na.rm = TRUE),
    median_MAE  = median(MAE,  na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  arrange(variable, dataset, desc(median_R2))

print(cv_summary)
write_csv(cv_summary, "../outputs/cv_summary_spatial_folds.csv")

# ── CV violin plots ───────────────────────────────────────
results_all %>%
  mutate(dataset = factor(dataset, c("SP", "COMP")),
         method  = factor(method,  c("OK", "RF", "RK"))) %>%
  ggplot(aes(x = method, y = R2, fill = method)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  facet_grid(variable ~ dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "Spatial CV: R² by dataset and method",
       x = "Method", y = expression(R^2)) +
  theme(legend.position = "none")

library(rstatix)

# Reshape CV results for statistical comparison
cv_long <- cv_summary %>%
  mutate(method_design = paste(method, dataset, sep = "_"))

# Friedman test — does method × sampling design affect R²?
friedman_result <- cv_long %>%
  friedman_test(median_R2 ~ method_design | variable)

print(friedman_result)

# Post-hoc pairwise Wilcoxon signed-rank tests
posthoc_result <- cv_long %>%
  wilcox_test(median_R2 ~ method_design,
              paired = TRUE,
              p.adjust.method = "bonferroni") %>%
  add_significance()

print(posthoc_result, n = 30)
# SP vs COMP comparison within each method
sp_comp_comparison <- cv_summary %>%
  select(variable, method, dataset, median_R2) %>%
  pivot_wider(names_from = dataset,
              values_from = median_R2) %>%
  group_by(method) %>%
  summarise(
    wilcox_p = wilcox.test(SP, COMP,
                            paired = TRUE,
                            exact = FALSE)$p.value,
    median_diff = median(SP - COMP, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(significant = wilcox_p < 0.05)

print(sp_comp_comparison)
# Method comparison within SP and COMP
method_comparison <- cv_summary %>%
  select(variable, method, dataset, median_R2) %>%
  group_by(dataset) %>%
  summarise(
    # RF vs OK
    rf_ok_p = wilcox.test(
      median_R2[method == "RF"],
      median_R2[method == "OK"],
      paired = TRUE, exact = FALSE)$p.value,
    # RF vs RK
    rf_rk_p = wilcox.test(
      median_R2[method == "RF"],
      median_R2[method == "RK"],
      paired = TRUE, exact = FALSE)$p.value,
    # RK vs OK
    rk_ok_p = wilcox.test(
      median_R2[method == "RK"],
      median_R2[method == "OK"],
      paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  )

print(method_comparison)

# SP outperforms COMP for RF
#SP outperforms COMP for RK
#RF_SP vs RK_COMP (p = 0.055) — borderline

#RF: significant (p = 0.033, median diff = 0.098) — SP outperforms COMP by about 0.10 R² units for RF, statistically significant
# RK: significant (p = 0.044, median diff = 0.050) — SP outperforms COMP for RK, statistically significant
# OK: not significant (p = 0.477) — no meaningful SP vs COMP difference for OK
#COMP: RF vs RK significant (p = 0.044) — RK outperforms RF under COMP sampling
```

# ============================================================
# PHASE 4 — Final Full-Dataset Models + Held-Out Validation
# ============================================================

```{r final-models}
fit_final_rf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)
  randomForest(
    x          = df %>% select(all_of(predictors)),
    y          = df[[response]],
    ntree      = ntree,
    mtry       = tuning$best_mtry,
    importance = TRUE
  )
}

fit_final_ok <- function(sf_data, response, cutoff, width) {
  sp_data <- as(sf_data, "Spatial")
  f       <- as.formula(paste(response, "~ 1"))
  vg      <- variogram(f, data = sp_data, cutoff = cutoff, width = width)
  fit     <- fit_variogram_models(vg)
  list(variogram = vg, fit = fit, formula = f, sp_data = sp_data)
}

fit_final_rk <- function(sf_data, response, predictors,
                          cutoff, width, ntree = 500) {
  df      <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  sf_data2 <- sf_data %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))

  tuning  <- tune_rf_oob(df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = df %>% select(all_of(predictors)),
    y = df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  trend             <- predict(rf_fit, newdata = df %>%
                                 select(all_of(predictors)))
  sf_data2$residual <- df[[response]] - trend

  sp_data  <- as(sf_data2, "Spatial")
  vg_resid <- variogram(residual ~ 1, data = sp_data,
                         cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  list(rf_fit = rf_fit, fit_resid = fit_resid, sp_data = sp_data)
}

final_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_SP, v, cov_names),
    ok = fit_final_ok(soil_prop_SP, v, cutoff_SP, width_lag),
    rk = fit_final_rk(soil_prop_SP, v, cov_names, cutoff_SP, width_lag)
  )
)

final_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_COMP, v, cov_names),
    ok = fit_final_ok(soil_prop_COMP, v, cutoff_COMP, width_lag),
    rk = fit_final_rk(soil_prop_COMP, v, cov_names, cutoff_COMP, width_lag)
  )
)

beep("fanfare")
```

```{r variogram-summary}
# ── Phase 4b: Variogram model summary ─────────────────────
extract_variogram_summary <- function(final_models, soil_vars, label) {
  purrr::map_dfr(soil_vars, function(v) {
    fit <- final_models[[v]]$ok$fit
    if (is.null(fit)) {
      return(tibble(variable = v, dataset = label,
                    model = NA, nugget = NA, psill = NA,
                    total_sill = NA, range_m = NA,
                    nugget_sill_ratio = NA))
    }
    nug_row  <- fit %>% filter(model == "Nug")
    str_row  <- fit %>% filter(model != "Nug")
    nugget   <- if (nrow(nug_row) > 0) nug_row$psill else 0
    psill    <- if (nrow(str_row) > 0) str_row$psill else NA
    rng      <- if (nrow(str_row) > 0) str_row$range else NA
    mod_type <- if (nrow(str_row) > 0) as.character(str_row$model) else NA
    tibble(variable = v, dataset = label, model = mod_type,
           nugget   = round(nugget, 4), psill = round(psill, 4),
           total_sill = round(nugget + psill, 4),
           range_m    = round(rng, 1),
           nugget_sill_ratio = round(nugget / (nugget + psill), 3))
  })
}

vg_summary_all <- bind_rows(
  extract_variogram_summary(final_models_SP,   soil_vars, "SP"),
  extract_variogram_summary(final_models_COMP, soil_vars, "COMP")
)

vg_summary_all %>%
  mutate(spatial_dependence = case_when(
    nugget_sill_ratio < 0.25 ~ "Strong",
    nugget_sill_ratio < 0.75 ~ "Moderate",
    TRUE                     ~ "Weak — kriging limited"
  )) %>%
  arrange(dataset, variable) %>%
  print(n = 20)

write_csv(vg_summary_all, "../outputs/variogram_summary_table.csv")
```

```{r held-out-validation}
# ── Predict to held-out validation set ────────────────────
predict_to_points <- function(models_list, response, new_sf,
                               predictors, cutoff, width) {
  new_df <- new_sf %>% st_drop_geometry() %>%
    select(any_of(c(response, predictors))) %>% na.omit()
  obs    <- new_df[[response]]

  rf_pred <- predict(models_list$rf,
                     newdata = new_df %>% select(all_of(predictors)))

  ok_pred <- tryCatch({
    kr <- krige(models_list$ok$formula,
                locations = models_list$ok$sp_data,
                newdata   = as(new_sf, "Spatial"),
                model     = models_list$ok$fit)
    kr$var1.pred
  }, error = function(e) rep(NA_real_, nrow(new_df)))

  rk_trend <- predict(models_list$rk$rf_fit,
                      newdata = new_df %>% select(all_of(predictors)))
  rk_resid <- tryCatch({
    new_sf2 <- new_sf %>%
      filter(complete.cases(st_drop_geometry(
        select(., any_of(predictors)))))
    kr2 <- krige(residual ~ 1,
                 locations = models_list$rk$sp_data,
                 newdata   = as(new_sf2, "Spatial"),
                 model     = models_list$rk$fit_resid)
    kr2$var1.pred
  }, error = function(e) rep(0, length(rk_trend)))
  rk_pred <- rk_trend + rk_resid

  bind_rows(
    calc_metrics(obs, rf_pred) %>%
      mutate(variable = response, method = "RF"),
    calc_metrics(obs, ok_pred) %>%
      mutate(variable = response, method = "OK"),
    calc_metrics(obs, rk_pred) %>%
      mutate(variable = response, method = "RK")
  )
}

val_results_all <- bind_rows(
  purrr::map_dfr(soil_vars, function(v) {
    predict_to_points(final_models_SP[[v]], v, ap_val_sf,
                      cov_names, cutoff_SP, width_lag) %>%
      mutate(dataset = "SP")
  }),
  purrr::map_dfr(soil_vars, function(v) {
    predict_to_points(final_models_COMP[[v]], v, ap_val_sf,
                      cov_names, cutoff_COMP, width_lag) %>%
      mutate(dataset = "COMP")
  })
)

write_csv(val_results_all, "../outputs/validation_results_held_out.csv")
val_results_all %>% arrange(variable, dataset, desc(R2))

#Unlike internal CV where OK occasionally outperformed RF/RK (e.g. K COMP, pH), the held-out validation tells a cleaner story — RF and RK almost universally produce higher R² than OK across all variables and both sampling designs. This is the more reliable finding since it comes from truly independent data.
```

# ============================================================
# PHASE 4c — RF Variable Importance Analysis
# ============================================================

```{r rf-importance}
# ── Extract importance from final RF models ────────────────
# Uses %IncMSE (permutation importance) — more reliable
# than node impurity (%IncNodePurity) for correlated predictors.
# Top-20 covariates shown per variable × dataset.

extract_importance <- function(rf_model, variable, dataset) {
  imp <- importance(rf_model, type = 1) # type=1 = %IncMSE
  tibble(
    covariate = rownames(imp),
    pct_inc_mse = imp[, 1],
    variable  = variable,
    dataset   = dataset
  ) %>%
    arrange(desc(pct_inc_mse))
}

importance_SP <- purrr::map_dfr(soil_vars, function(v) {
  extract_importance(final_models_SP[[v]]$rf, v, "SP")
})

importance_COMP <- purrr::map_dfr(soil_vars, function(v) {
  extract_importance(final_models_COMP[[v]]$rf, v, "COMP")
})

importance_all <- bind_rows(importance_SP, importance_COMP)

# ── Per-variable importance plots (top 20 covariates) ─────
plot_importance_per_var <- function(imp_df, dataset_label) {
  imp_df %>%
    filter(dataset == dataset_label) %>%
    group_by(variable) %>%
    slice_max(order_by = pct_inc_mse, n = 20) %>%
    ungroup() %>%
    mutate(covariate = tidytext::reorder_within(
      covariate, pct_inc_mse, variable)) %>%
    ggplot(aes(x = covariate, y = pct_inc_mse,
               fill = pct_inc_mse)) +
    geom_col(show.legend = FALSE) +
    tidytext::scale_x_reordered() +
    scale_fill_viridis_c(option = "plasma", direction = -1) +
    coord_flip() +
    facet_wrap(~ variable, scales = "free") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("RF Variable Importance — ",
                        dataset_label, " dataset"),
      subtitle = "Top 20 covariates by %IncMSE (permutation importance)",
      x        = NULL,
      y        = "% Increase in MSE"
    )
}

# Note: tidytext is used for reorder_within().
# If not installed: install.packages("tidytext")
# Alternative without tidytext: use fct_reorder within facets.

print(plot_importance_per_var(importance_all, "SP"))
print(plot_importance_per_var(importance_all, "COMP"))

# ── Aggregate importance: mean %IncMSE across all soil vars ─
# Identifies covariates that are broadly useful across
# the full agronomic variable set, not just one variable.

agg_importance <- importance_all %>%
  group_by(covariate, dataset) %>%
  summarise(
    mean_pct_inc_mse   = mean(pct_inc_mse,   na.rm = TRUE),
    median_pct_inc_mse = median(pct_inc_mse, na.rm = TRUE),
    n_vars_top20 = sum(rank(desc(pct_inc_mse)) <= 20,
                       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(dataset, desc(mean_pct_inc_mse))

# Top-30 aggregate importance
p_agg_imp <- agg_importance %>%
  group_by(dataset) %>%
  slice_max(mean_pct_inc_mse, n = 30) %>%
  ungroup() %>%
  mutate(
    covariate = tidytext::reorder_within(
      covariate, mean_pct_inc_mse, dataset)
  ) %>%
  ggplot(aes(x = covariate, y = mean_pct_inc_mse,
             fill = dataset)) +
  geom_col(show.legend = FALSE) +
  tidytext::scale_x_reordered() +
  scale_fill_manual(values = c("SP"   = "#2980b9",
                                "COMP" = "#e74c3c")) +
  coord_flip() +
  facet_wrap(~ dataset, scales = "free") +
  theme_bw(base_size = 9) +
  labs(
    title    = "Aggregate RF Variable Importance",
    subtitle = paste0("Mean %IncMSE across all ",
                      length(soil_vars),
                      " agronomic soil variables"),
    x        = NULL,
    y        = "Mean % Increase in MSE"
  )

print(p_agg_imp)

# ── Covariate overlap: top-15 per variable heatmap ────────
top15_per_var <- importance_all %>%
  group_by(dataset, variable) %>%
  slice_max(pct_inc_mse, n = 15) %>%
  ungroup() %>%
  mutate(in_top15 = TRUE)

# Which covariates appear in top-15 for multiple variables?
overlap_counts <- top15_per_var %>%
  group_by(covariate, dataset) %>%
  summarise(n_vars_in_top15 = n(), .groups = "drop") %>%
  filter(n_vars_in_top15 >= 2)

p_overlap <- top15_per_var %>%
  filter(covariate %in% overlap_counts$covariate) %>%
  ggplot(aes(x = variable, y = covariate,
             fill = pct_inc_mse)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                        name = "%IncMSE") +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title    = "Covariate importance overlap across soil variables",
    subtitle = "Covariates appearing in top-15 for ≥2 variables",
    x = "Soil variable", y = "Covariate"
  )

print(p_overlap)
single_bands <- c("coastal_blue", "blue", "green_i", "green",
                  "yellow", "red", "rededge", "nir")

# Soil variable display names
soil_var_labels <- c(
  "lbc_1"  = "Lime Buffer Capacity",
  "lb_ceq" = "LBC Equivalent",
  "p_h_2"  = "pH",
  "ca"     = "Calcium",
  "k"      = "Potassium",
  "mg"     = "Magnesium",
  "mn"     = "Manganese",
  "p"      = "Phosphorus",
  "zn"     = "Zinc"
)

# ── Covariate overlap: top-15 per variable heatmap ────────
top15_per_var <- importance_all %>%
  filter(!covariate %in% single_bands) %>%
  group_by(dataset, variable) %>%
  slice_max(pct_inc_mse, n = 15) %>%
  ungroup() %>%
  mutate(in_top15 = TRUE)

overlap_counts <- top15_per_var %>%
  group_by(covariate, dataset) %>%
  summarise(n_vars_in_top15 = n(), .groups = "drop") %>%
  filter(n_vars_in_top15 >= 2)

p_overlap2 <- top15_per_var %>%
  filter(covariate %in% overlap_counts$covariate) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable,
                      levels = soil_var_labels)
  ) %>%
  ggplot(aes(x = variable, y = covariate,
             fill = pct_inc_mse)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                        name = "%IncMSE") +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title    = "Covariate importance overlap across soil variables",
    subtitle = "Covariates appearing in top-15 for ≥2 variables (single bands excluded)",
    x = "Soil variable", y = "Covariate"
  )

print(p_overlap2)
write_csv(importance_all,
          "../outputs/rf_importance_per_variable.csv")
write_csv(agg_importance,
          "../outputs/rf_importance_aggregate.csv")

soil_var_labels <- c(
  "lbc_1"  = "Lime Buffer Capacity",
  "lb_ceq" = "Lime Buffer Capacity Eq.",
  "p_h_2"  = "pH",
  "ca"     = "Calcium (Ca)",
  "k"      = "Potassium (K)",
  "mg"     = "Magnesium (Mg)",
  "mn"     = "Manganese (Mn)",
  "p"      = "Phosphorus (P)",
  "zn"     = "Zinc (Zn)"
)

# ── Soil variable × covariate correlation heatmap ─────────
# Pearson correlation between each soil variable and each
# covariate at SP and COMP sampling locations combined.
# Helps interpret which environmental drivers are most
# associated with each soil property.

cor_data <- bind_rows(
  soil_prop_SP   %>% st_drop_geometry() %>% mutate(dataset = "SP"),
  soil_prop_COMP %>% st_drop_geometry() %>% mutate(dataset = "COMP")
) %>%
  select(all_of(soil_vars), all_of(cov_names)) %>%
  na.omit()

cor_mat <- cor(
  cor_data %>% select(all_of(soil_vars)),
  cor_data %>% select(all_of(cov_names)),
  use = "complete.obs"
) %>%
  as.data.frame() %>%
  rownames_to_column("soil_var") %>%
  pivot_longer(-soil_var,
               names_to  = "covariate",
               values_to = "r") %>%
    mutate(
        soil_var  = factor(recode(soil_var, !!!soil_var_labels),
                           levels = soil_var_labels),
        covariate = factor(covariate, levels = cov_names)
      )

p_cor_heatmap <- cor_mat %>%
  ggplot(aes(x = soil_var, y = covariate, fill = r)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = ifelse(abs(r) >= 0.30,
                                sprintf("%.2f", r), "")),
            size = 2.0, color = "grey10") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9850",
    midpoint = 0,
    limits   = c(-1, 1),
    oob      = scales::squish,
    name     = "Pearson r"
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1,
                                     size = 8, face = "bold"),
    axis.text.y      = element_text(size = 7),
    panel.grid       = element_blank(),
    legend.position  = "right"
  ) +
  labs(
    title    = "Soil property × covariate correlations",
    subtitle = "SP and COMP sampling locations combined (n = 272). Values shown where |r| ≥ 0.30",
    x        = "Soil variable",
    y        = "Covariate"
  )

print(p_cor_heatmap)
ggsave("../outputs/fig_soil_covariate_correlations.png",
       p_cor_heatmap,
       width  = 10,
       height = 14,
       dpi    = 300)
write_csv(cor_mat, "../outputs/soil_covariate_correlations.csv")
```

# ============================================================
# PHASE 5 — Full-Grid Spatial Prediction
# ============================================================

```{r grid-prediction}
grid_df <- as.data.frame(covariate_stack, xy = TRUE, na.rm = TRUE)
grid_sf <- st_as_sf(grid_df, coords = c("x", "y"), crs = 26917)
grid_sp <- as(grid_sf, "Spatial")

predict_grid_rf <- function(rf_model, grid_df, predictors) {
  predict(rf_model, newdata = grid_df %>% select(all_of(predictors)))
}

predict_grid_ok <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    kr$var1.pred
  }, error = function(e) {
    warning("OK prediction failed; returning NA")
    rep(NA_real_, nrow(grid_sp@data))
  })
}

predict_grid_rk <- function(rk_obj, grid_df, grid_sp, predictors) {
  trend <- predict(rk_obj$rf_fit,
                   newdata = grid_df %>% select(all_of(predictors)))
  resid <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    kr2$var1.pred
  }, error = function(e) rep(0, length(trend)))
  trend + resid
}

build_prediction_stack <- function(final_models, grid_df, grid_sf,
                                    grid_sp, soil_vars, predictors,
                                    label) {
  preds <- purrr::map(setNames(soil_vars, soil_vars), function(v) {
    m <- final_models[[v]]
    tibble(
      !!paste0(v, "_RF") := predict_grid_rf(m$rf, grid_df, predictors),
      !!paste0(v, "_RK") := predict_grid_rk(m$rk, grid_df, grid_sp, predictors),
      !!paste0(v, "_OK") := predict_grid_ok(m$ok, grid_sp)
    )
  }) %>% bind_cols()

  bind_cols(grid_df %>% select(x, y), preds, dataset = label)
}

pred_stack_SP   <- build_prediction_stack(
  final_models_SP,   grid_df, grid_sf, grid_sp,
  soil_vars, cov_names, "SP")
pred_stack_COMP <- build_prediction_stack(
  final_models_COMP, grid_df, grid_sf, grid_sp,
  soil_vars, cov_names, "COMP")

beep("mario")
save.image("../outputs/workspace_phase4b.RData")
beep("coin")
```

# ============================================================
# PHASE 5b — Spatial Lag Features
# ============================================================

```{r spatial-lags}
add_spatial_lags_focal <- function(pred_df, soil_vars, methods,
                                    crs = "EPSG:26917") {
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_df)]

  cat("Computing spatial lags via terra::focal for",
      length(col_names), "layers...\n")

  pred_rast <- pred_df %>%
    select(x, y, all_of(col_names)) %>%
    rast(type = "xyz", crs = crs)

  w       <- matrix(1, 3, 3)
  w[2, 2] <- 0
  lag_rast <- terra::focal(pred_rast, w = w,
                            fun = "mean", na.rm = TRUE)
  names(lag_rast) <- paste0(col_names, "_lag")

  pts    <- terra::vect(pred_df %>% select(x, y),
                        geom = c("x", "y"), crs = crs)
  lag_df <- terra::extract(lag_rast, pts, ID = FALSE)

  cat("pred_df rows:", nrow(pred_df),
      "| lag rows:", nrow(lag_df), "\n")
  bind_cols(pred_df, lag_df)
}

pred_stack_SP_lag   <- add_spatial_lags_focal(
  pred_stack_SP,   soil_vars, c("RF", "RK"))
pred_stack_COMP_lag <- add_spatial_lags_focal(
  pred_stack_COMP, soil_vars, c("RF", "RK"))

cat("SP lag stack columns:",   ncol(pred_stack_SP_lag),   "\n")
cat("COMP lag stack columns:", ncol(pred_stack_COMP_lag), "\n")
```

# ============================================================
# PHASE 5c — Uncertainty Analysis
# ============================================================

```{r uncertainty-analysis}
# ── 1. Quantile Regression Forest (RF Prediction Intervals) ─
# QRF extends RF to estimate conditional quantiles.
# We compute the 10th and 90th percentiles to produce
# an 80% prediction interval for each grid pixel.
# Interval width = spatial map of RF prediction uncertainty.

fit_qrf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)

  set.seed(20260401)
  quantregForest(
    x    = as.matrix(df %>% select(all_of(predictors))),
    y    = df[[response]],
    ntree = ntree,
    mtry  = tuning$best_mtry
  )
}

cat("Fitting QRF models for SP dataset...\n")
qrf_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  ~ fit_qrf(soil_prop_SP, .x, cov_names)
)

cat("Fitting QRF models for COMP dataset...\n")
qrf_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  ~ fit_qrf(soil_prop_COMP, .x, cov_names)
)

# Predict 80% PI to grid
grid_mat <- as.matrix(grid_df %>% select(all_of(cov_names)))

predict_qrf_grid <- function(qrf_model, grid_mat,
                              quantiles = c(0.10, 0.90)) {
  preds <- predict(qrf_model, newdata = grid_mat,
                   what = quantiles)
  tibble(
    pi_lower = preds[, 1],
    pi_upper = preds[, 2],
    pi_width = preds[, 2] - preds[, 1]
  )
}

qrf_uncertainty_SP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    p <- predict_qrf_grid(qrf_models_SP[[v]], grid_mat)
    setNames(p, paste0(v, "_", names(p)))
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

qrf_uncertainty_COMP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    p <- predict_qrf_grid(qrf_models_COMP[[v]], grid_mat)
    setNames(p, paste0(v, "_", names(p)))
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

# ── QRF uncertainty maps — one panel per variable ─────────
plot_qrf_width <- function(qrf_df, dataset_label) {
  width_cols <- paste0(soil_vars, "_pi_width")
  width_cols <- width_cols[width_cols %in% names(qrf_df)]

  qrf_df %>%
    select(x, y, all_of(width_cols)) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "pi_width") %>%
    mutate(variable = str_remove(variable, "_pi_width")) %>%
    ggplot(aes(x = x, y = y, fill = pi_width)) +
    geom_raster() +
    scale_fill_viridis_c(option = "inferno", direction = -1,
                          name = "PI width\n(80%)") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("QRF 80% Prediction Interval Width — ",
                        dataset_label),
      subtitle = "Wider = higher RF prediction uncertainty",
      x = "Easting", y = "Northing"
    )
}

print(plot_qrf_width(qrf_uncertainty_SP,   "SP"))
print(plot_qrf_width(qrf_uncertainty_COMP, "COMP"))

write_csv(qrf_uncertainty_SP,
          "../outputs/qrf_uncertainty_SP.csv")
write_csv(qrf_uncertainty_COMP,
          "../outputs/qrf_uncertainty_COMP.csv")

# ── 2. Kriging Variance Surfaces (OK and RK) ──────────────
# krige() returns var1.var — the kriging variance at each
# prediction location. This is the geostatistical uncertainty
# estimate; sqrt(var1.var) = standard error of the kriging
# prediction. For RK, this is the residual kriging variance.

predict_grid_ok_with_var <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    tibble(
      pred = kr$var1.pred,
      krig_var = kr$var1.var
    )
  }, error = function(e) {
    tibble(pred    = rep(NA_real_, nrow(grid_sp@data)),
           krig_var = rep(NA_real_, nrow(grid_sp@data)))
  })
}

predict_grid_rk_with_var <- function(rk_obj, grid_df,
                                      grid_sp, predictors) {
  trend <- predict(rk_obj$rf_fit,
                   newdata = grid_df %>% select(all_of(predictors)))
  resid_pred <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    tibble(resid = kr2$var1.pred,
           krig_var = kr2$var1.var)
  }, error = function(e) {
    tibble(resid    = rep(0, length(trend)),
           krig_var = rep(NA_real_, length(trend)))
  })
  tibble(
    pred     = trend + resid_pred$resid,
    krig_var = resid_pred$krig_var
  )
}

krig_var_SP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    m <- final_models_SP[[v]]
    ok <- predict_grid_ok_with_var(m$ok, grid_sp)
    rk <- predict_grid_rk_with_var(m$rk, grid_df, grid_sp, cov_names)
    tibble(
      !!paste0(v, "_ok_krig_var") := ok$krig_var,
      !!paste0(v, "_rk_krig_var") := rk$krig_var
    )
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

krig_var_COMP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    m <- final_models_COMP[[v]]
    ok <- predict_grid_ok_with_var(m$ok, grid_sp)
    rk <- predict_grid_rk_with_var(m$rk, grid_df, grid_sp, cov_names)
    tibble(
      !!paste0(v, "_ok_krig_var") := ok$krig_var,
      !!paste0(v, "_rk_krig_var") := rk$krig_var
    )
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

# ── Kriging SE maps ───────────────────────────────────────
plot_krig_se <- function(kv_df, dataset_label, method_suffix) {
  se_cols <- paste0(soil_vars, method_suffix)
  se_cols <- se_cols[se_cols %in% names(kv_df)]

  kv_df %>%
    select(x, y, all_of(se_cols)) %>%
    mutate(across(all_of(se_cols), sqrt)) %>%  # var → SE
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "krig_se") %>%
    mutate(variable = str_remove(
      variable, str_escape(method_suffix))) %>%
    ggplot(aes(x = x, y = y, fill = krig_se)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1,
                          name = "Kriging SE",
                          na.value = "grey80") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("Kriging Standard Error — ",
                        dataset_label, " (",
                        str_remove(method_suffix,
                                   "_krig_var"), ")"),
      subtitle = "sqrt(kriging variance) at each grid pixel",
      x = "Easting", y = "Northing"
    )
}

print(plot_krig_se(krig_var_SP,   "SP",   "_ok_krig_var"))
print(plot_krig_se(krig_var_SP,   "SP",   "_rk_krig_var"))
print(plot_krig_se(krig_var_COMP, "COMP", "_ok_krig_var"))
print(plot_krig_se(krig_var_COMP, "COMP", "_rk_krig_var"))

write_csv(krig_var_SP,   "../outputs/kriging_variance_SP.csv")
write_csv(krig_var_COMP, "../outputs/kriging_variance_COMP.csv")

# ── 3. Bootstrap Uncertainty for RF Predictions ───────────
# Bootstrap resampling of training data (B=100 replicates)
# to estimate prediction SD at each grid pixel.
# SD captures the sensitivity of RF predictions to the
# specific set of training observations used.

bootstrap_rf_uncertainty <- function(sf_data, soil_vars,
                                      predictors, grid_df,
                                      B = 100, ntree = 500,
                                      seed = 20260401) {
  cat("Bootstrap RF uncertainty (B =", B, "replicates)...\n")

  df_full <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(soil_vars, predictors))) %>% na.omit()

  set.seed(seed)
  boot_preds <- purrr::map(seq_len(B), function(b) {
    if (b %% 10 == 0) cat(" Bootstrap rep:", b, "/", B, "\n")
    boot_idx <- sample(nrow(df_full), replace = TRUE)
    boot_df  <- df_full[boot_idx, ]

    purrr::map_dfc(soil_vars, function(v) {
      rf_b <- tryCatch({
        randomForest(
          x = boot_df %>% select(all_of(predictors)),
          y = boot_df[[v]],
          ntree = ntree, mtry = floor(sqrt(length(predictors)))
        )
      }, error = function(e) NULL)

      if (is.null(rf_b)) {
        return(tibble(!!v := rep(NA_real_, nrow(grid_df))))
      }

      tibble(
        !!v := predict(rf_b,
                       newdata = grid_df %>%
                         select(all_of(predictors)))
      )
    })
  })

  # Compute SD across bootstrap replicates per pixel
  purrr::map_dfc(soil_vars, function(v) {
    boot_mat <- do.call(cbind, purrr::map(boot_preds, ~ .x[[v]]))
    tibble(
      !!paste0(v, "_boot_sd")   := apply(boot_mat, 1,
                                          sd, na.rm = TRUE),
      !!paste0(v, "_boot_mean") := apply(boot_mat, 1,
                                          mean, na.rm = TRUE)
    )
  }) %>%
    bind_cols(grid_df %>% select(x, y), .)
}

# B=100 replicates — increase to 500 for publication
cat("Running bootstrap for SP dataset...\n")
boot_uncertainty_SP   <- bootstrap_rf_uncertainty(
  soil_prop_SP,   soil_vars, cov_names, grid_df, B = 100)

cat("Running bootstrap for COMP dataset...\n")
boot_uncertainty_COMP <- bootstrap_rf_uncertainty(
  soil_prop_COMP, soil_vars, cov_names, grid_df, B = 100)

# ── Bootstrap SD maps ─────────────────────────────────────
plot_boot_sd <- function(boot_df, dataset_label) {
  sd_cols <- paste0(soil_vars, "_boot_sd")
  sd_cols <- sd_cols[sd_cols %in% names(boot_df)]

  boot_df %>%
    select(x, y, all_of(sd_cols)) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd")) %>%
    ggplot(aes(x = x, y = y, fill = boot_sd)) +
    geom_raster() +
    scale_fill_viridis_c(option = "rocket", direction = -1,
                          name = "Bootstrap SD") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("RF Bootstrap Uncertainty (B=100) — ",
                        dataset_label),
      subtitle = "SD of predictions across 100 bootstrap replicates",
      x = "Easting", y = "Northing"
    )
}

print(plot_boot_sd(boot_uncertainty_SP,   "SP"))
print(plot_boot_sd(boot_uncertainty_COMP, "COMP"))

write_csv(boot_uncertainty_SP,
          "../outputs/bootstrap_uncertainty_SP.csv")
write_csv(boot_uncertainty_COMP,
          "../outputs/bootstrap_uncertainty_COMP.csv")

# ── 4. Zone Assignment Uncertainty (membership probability) ─
# Already computed within HDBSCAN — handled in Phase 6b.
# Membership probability ∈ [0,1]: 1 = core cluster member,
# 0 = peripherally assigned (borderline pixel).
# Noise pixels assigned by NN have membership_prob = 0.
# Reported spatially and summarised per zone below.
# See plot_confidence() in Phase 6b for spatial maps.

# ── Uncertainty comparison summary ────────────────────────
# Side-by-side comparison of uncertainty magnitude by method
uncertainty_summary <- bind_rows(
  # QRF mean PI width
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable  = str_remove(variable, "_pi_width"),
           method    = "QRF 80% PI width",
           dataset   = "SP"),

  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           method   = "QRF 80% PI width",
           dataset  = "COMP"),

  # Bootstrap SD
  boot_uncertainty_SP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           method   = "Bootstrap SD",
           dataset  = "SP"),

  boot_uncertainty_COMP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           method   = "Bootstrap SD",
           dataset  = "COMP")
) %>%
  group_by(variable, method, dataset) %>%
  summarise(
    mean_unc   = round(mean(value,   na.rm = TRUE), 3),
    median_unc = round(median(value, na.rm = TRUE), 3),
    .groups    = "drop"
  )

print(uncertainty_summary, n = 40)
write_csv(uncertainty_summary,
          "../outputs/uncertainty_summary.csv")

beep("wilhelm")
save.image("../outputs/workspace_uncertainty.RData")
#ERROR:
#Per-variable interpretation:

  # pH — bootstrap SD ~0.07–0.10, PI width ~0.9–1.1. Very tight, makes sense — pH varies narrowly at field scale and RF predicts it well
  # P — bootstrap SD ~2–3, PI width ~20–23. Low uncertainty relative to the other nutrients, consistent with P often being spatially structured by management history
  # K — moderate uncertainty, COMP tighter than SP which is expected given composite sampling smooths local variability
  # Ca and lb_ceq — largest absolute uncertainty, which makes sense as these are highly variable and correlated with soil texture and organic matter gradients
#The SP > COMP pattern for most variables is also exactly what I expect w/ composite samples average over a larger area so the RF trained on them sees less local noise, producing tighter predictions.

#P flips the pattern (COMP SD slightly higher than SP). That's worth a sentence in my methods ...it may reflect that composite sampling obscures real P spatial structure that single-point sampling captures, since P is often highly localized around application history.
# Kriging SE summary — OK
bind_rows(
  krig_var_SP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "SP", method = "OK kriging SE"),

  krig_var_COMP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "COMP", method = "OK kriging SE"),

  krig_var_SP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "SP", method = "RK kriging SE"),

  krig_var_COMP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "COMP", method = "RK kriging SE")
) %>%
  mutate(
    variable = str_extract(name, paste(soil_vars, collapse = "|")),
    stat     = if_else(str_detect(name, "_mean"), "mean", "median")
  ) %>%
  select(variable, stat, dataset, method, value) %>%
  arrange(variable, method, dataset, stat) 

# FINAL UNCERTAINTY SUMMARY:
# SP > COMP for most variables (SP higher uncertainty), except P (COMP slightly higher)
# OK > RK > RF in uncertainty magnitude
# HOWEVER — these are not apples-to-apples comparisons:
#   - OK SE is purely geometry-driven (data configuration)
#   - RK SE is residual kriging component only (incomplete — missing RF trend uncertainty)
#   - RF bootstrap SD captures training data sensitivity only
# RK total uncertainty = RF trend uncertainty + residual kriging variance (neither alone is complete)
# All three are complementary, not directly comparable on the same scale

```
```{r}
#uncertainty figure for raw maps:
# ── Unified uncertainty figure ─────────────────────────────
# Combines RF bootstrap SD, QRF PI width, and kriging SE
# for OK and RK into one comparable figure

soil_var_labels <- c(
  "lbc_1"  = "LBC-1",
  "lb_ceq" = "LBC Eq.",
  "p_h_2"  = "pH",
  "ca"     = "Calcium",
  "k"      = "Potassium",
  "mg"     = "Magnesium",
  "mn"     = "Manganese",
  "p"      = "Phosphorus",
  "zn"     = "Zinc"
)

# ── 1. RF uncertainty — bootstrap SD and QRF PI ───────────
rf_uncertainty <- uncertainty_summary %>%
  mutate(
    pred_method = "RF",
    metric_type = method
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc)

# ── 2. Kriging SE — OK and RK ─────────────────────────────
krig_se_long <- bind_rows(
  krig_var_SP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "SP", pred_method = "OK"),

  krig_var_COMP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "COMP", pred_method = "OK"),

  krig_var_SP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "SP", pred_method = "RK"),

  krig_var_COMP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "COMP", pred_method = "RK")
) %>%
  mutate(
    variable    = str_extract(name,
                               paste(soil_vars, collapse = "|")),
    metric_type = "Kriging SE"
  ) %>%
  filter(!is.na(variable), is.finite(median_unc)) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc)

# ── 3. Combine all uncertainty metrics ────────────────────
uncertainty_all <- bind_rows(
  rf_uncertainty %>%
    filter(metric_type == "Bootstrap SD"),
  krig_se_long
) %>%
  mutate(
    variable    = recode(variable, !!!soil_var_labels),
    variable    = factor(variable, levels = soil_var_labels),
    pred_method = factor(pred_method,
                         levels = c("RF", "OK", "RK")),
    dataset     = factor(dataset,
                         levels = c("SP", "COMP"),
                         labels = c("Single-Point",
                                    "Composite")),
    metric_type = factor(metric_type,
                         levels = c("Bootstrap SD",
                                    "Kriging SE"))
  )

# ── 4. Plot ───────────────────────────────────────────────
p_uncertainty <- uncertainty_all %>%
  ggplot(aes(x = variable,
             y = median_unc,
             fill = pred_method)) +
  geom_col(position = position_dodge(0.8),
           width = 0.7) +
  scale_fill_manual(
    values = c("RF" = "#2980b9",
               "OK" = "#8e44ad",
               "RK" = "#27ae60"),
    name = "Prediction method"
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1,
                                     size = 8),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 9),
    legend.position  = "bottom",
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title    = "Prediction uncertainty by method and sampling design",
    subtitle = paste0("Bootstrap SD (RF only) and kriging SE (OK, RK).",
                      " Median across all grid pixels.\n",
                      "NA = no convergent variogram model."),
    x        = NULL,
    y        = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty)

ggsave("../outputs/fig_uncertainty_by_method.png",
       p_uncertainty,
       width  = 12,
       height = 7,
       dpi    = 300)

#RK uncertainty:
# ── RK total uncertainty = RF trend SD + RK kriging SE ────
# Assumes independence between trend and residual uncertainty
# Total SE = sqrt(boot_SD² + krig_SE²)

rk_total_uncertainty <- bind_rows(
  # SP
  boot_uncertainty_SP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           dataset  = "SP") %>%
    left_join(
      krig_var_SP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "SP"),
      by = c("x", "y", "variable", "dataset")
    ),

  # COMP
  boot_uncertainty_COMP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           dataset  = "COMP") %>%
    left_join(
      krig_var_COMP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "COMP"),
      by = c("x", "y", "variable", "dataset")
    )
) %>%
  mutate(
    # Total RK uncertainty via quadrature
    # krig_var is variance so sqrt = SE
    rk_total_se = sqrt(boot_sd^2 + 
                       coalesce(krig_var, 0))
  )

# ── Summarise per variable and dataset ────────────────────
rk_uncertainty_summary <- rk_total_uncertainty %>%
  group_by(variable, dataset) %>%
  summarise(
    mean_rk_se   = round(mean(rk_total_se,
                               na.rm = TRUE), 3),
    median_rk_se = round(median(rk_total_se,
                                 na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(
    pred_method = "RK",
    metric_type = "Total SE (boot + krig)"
  )

print(rk_uncertainty_summary, n = 20)
write_csv(rk_uncertainty_summary,
          "../outputs/rk_total_uncertainty.csv")
#Taylor, J.R. (1997). An Introduction to Error Analysis: The Study of Uncertainties in Physical Measurements (2nd ed.). University Science Books. 

# ── Rebuild uncertainty_all with RK total SE added ────────
rk_krig_se <- krig_se_long %>%
  filter(pred_method == "RK") %>%
  mutate(metric_type = "Kriging SE (residual only)",
         pred_method_label = "RK (kriging SE only)")

rk_total_se <- rk_uncertainty_summary %>%
  mutate(
    median_unc        = median_rk_se,
    metric_type       = "Total SE (RF + kriging)",
    pred_method_label = "RK (total SE)"
  ) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable, levels = soil_var_labels),
    dataset  = factor(dataset,
                      levels = c("SP", "COMP"),
                      labels = c("Single-Point", "Composite"))
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc, pred_method_label)

uncertainty_all_v2 <- bind_rows(
  # RF bootstrap SD — fix dataset labels here
  rf_uncertainty %>%
    filter(metric_type == "Bootstrap SD") %>%
    mutate(
      pred_method_label = "RF",
      dataset = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      ),
      variable = recode(variable, !!!soil_var_labels),
      variable = factor(variable, levels = soil_var_labels)
    ),

  # OK kriging SE
  krig_se_long %>%
    filter(pred_method == "OK") %>%
    mutate(
      metric_type       = "Kriging SE",
      pred_method_label = "OK",
      variable          = recode(variable, !!!soil_var_labels),
      variable          = factor(variable,
                                  levels = soil_var_labels),
      dataset           = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ),

  # RK kriging SE — residual only
  rk_krig_se %>%
    mutate(
      metric_type       = "Kriging SE",
      variable          = recode(variable, !!!soil_var_labels),
      variable          = factor(variable,
                                  levels = soil_var_labels),
      dataset           = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ) %>%
    select(variable, dataset, pred_method,
           metric_type, median_unc, pred_method_label),

  # RK total SE — already has correct labels from earlier
  rk_total_se %>%
    mutate(metric_type = "Kriging SE")

) %>%
  mutate(
    pred_method_label = factor(
      pred_method_label,
      levels = c("RF",
                 "OK",
                 "RK (kriging SE only)",
                 "RK (total SE)")
    )
  )

# ── Plot ──────────────────────────────────────────────────
method_colors <- c(
  "RF"                   = "#8e44ad",  # purple
  "OK"                   = "#2980b9",  # blue
  "RK (kriging SE only)" = "#82e0aa",  # light green
  "RK (total SE)"        = "#1e8449"   # dark green
)

p_uncertainty_v2 <- uncertainty_all_v2 %>%
  filter(!is.na(median_unc), is.finite(median_unc)) %>%
  ggplot(aes(x       = variable,
             y       = median_unc,
             fill    = pred_method_label)) +
  geom_col(position = position_dodge(0.85),
           width     = 0.8) +
  scale_fill_manual(
    values = method_colors,
    name   = "Prediction method",
    labels = c(
      "RF"                   = "RF (bootstrap SD)",
      "OK"                   = "OK (kriging SE)",
      "RK (kriging SE only)" = "RK (residual kriging SE)",
      "RK (total SE)"        = "RK (total SE = RF + kriging)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    legend.position    = "bottom",
    legend.text        = element_text(size = 8),
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Prediction uncertainty by method and sampling design",
    subtitle = paste0(
      "Row 1: Bootstrap SD (RF only). ",
      "Row 2: Kriging SE — OK, RK residual component (light green),\n",
      "and RK total SE combining RF bootstrap SD and residual ",
      "kriging SE in quadrature (dark green). ",
      "Median across all grid pixels.\n ",
      "Missing bars = no convergent variogram model."
    ),
    x = NULL,
    y = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty_v2)

ggsave("../outputs/fig_uncertainty_by_method_v2.png",
       p_uncertainty_v2,
       width  = 13,
       height = 8,
       dpi    = 300)



# ── 1. RF uncertainty — QRF PI width (primary metric) ─────
rf_uncertainty_qrf <- uncertainty_summary %>%
  filter(method == "QRF 80% PI width") %>%
  mutate(
    pred_method       = "RF",
    metric_type       = "QRF 80% PI width",
    pred_method_label = "RF"
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc, pred_method_label)

# ── 2. RK total uncertainty using QRF PI width ────────────
# Total RK uncertainty = sqrt(QRF_PI² + krig_var)
# Using QRF PI as RF trend uncertainty instead of bootstrap SD
# since QRF PI is the more standard RF uncertainty metric

rk_total_uncertainty_qrf <- bind_rows(
  # SP
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "SP") %>%
    left_join(
      krig_var_SP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "SP"),
      by = c("x", "y", "variable", "dataset")
    ),

  # COMP
  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "COMP") %>%
    left_join(
      krig_var_COMP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "COMP"),
      by = c("x", "y", "variable", "dataset")
    )
) %>%
  mutate(
    rk_total_se = sqrt(qrf_pi^2 +
                       coalesce(krig_var, 0))
  )

# ── Summarise RK total SE ─────────────────────────────────
rk_uncertainty_summary_qrf <- rk_total_uncertainty_qrf %>%
  group_by(variable, dataset) %>%
  summarise(
    mean_rk_se   = round(mean(rk_total_se,   na.rm = TRUE), 3),
    median_rk_se = round(median(rk_total_se, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(
    pred_method = "RK",
    metric_type = "Total SE (QRF PI + kriging)"
  )

print(rk_uncertainty_summary_qrf, n = 20)
write_csv(rk_uncertainty_summary_qrf,
          "../outputs/rk_total_uncertainty_qrf.csv")

# ── 3. Rebuild RK objects ─────────────────────────────────
rk_krig_se <- krig_se_long %>%
  filter(pred_method == "RK") %>%
  mutate(
    metric_type       = "Kriging SE",
    pred_method_label = "RK (kriging SE only)"
  )

rk_total_se_qrf <- rk_uncertainty_summary_qrf %>%
  mutate(
    median_unc        = median_rk_se,
    metric_type       = "Kriging SE",
    pred_method_label = "RK (total SE)"
  ) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable, levels = soil_var_labels),
    dataset  = factor(dataset,
                      levels = c("SP", "COMP"),
                      labels = c("Single-Point", "Composite"))
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc, pred_method_label)

# ── 4. Combine all uncertainty metrics ────────────────────
uncertainty_all_v3 <- bind_rows(

  # RF QRF PI width
  rf_uncertainty_qrf %>%
    mutate(
      dataset  = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      ),
      variable = recode(variable, !!!soil_var_labels),
      variable = factor(variable, levels = soil_var_labels),
      metric_type = "QRF 80% PI width"
    ),

  # OK kriging SE
  krig_se_long %>%
    filter(pred_method == "OK") %>%
    mutate(
      metric_type       = "Kriging SE",
      pred_method_label = "OK",
      variable          = recode(variable, !!!soil_var_labels),
      variable          = factor(variable,
                                  levels = soil_var_labels),
      dataset           = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ),

  # RK kriging SE — residual only
  rk_krig_se %>%
    mutate(
      variable = recode(variable, !!!soil_var_labels),
      variable = factor(variable, levels = soil_var_labels),
      dataset  = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ) %>%
    select(variable, dataset, pred_method,
           metric_type, median_unc, pred_method_label),

  # RK total SE — QRF PI + kriging
  rk_total_se_qrf

) %>%
  mutate(
    pred_method_label = factor(
      pred_method_label,
      levels = c("RF",
                 "OK",
                 "RK (kriging SE only)",
                 "RK (total SE)")
    ),
    metric_type = factor(
      metric_type,
      levels = c("QRF 80% PI width", "Kriging SE")
    )
  )

# ── 5. Plot ───────────────────────────────────────────────
method_colors <- c(
  "RF"                   = "#8e44ad",  # purple
  "OK"                   = "#2980b9",  # blue
  "RK (kriging SE only)" = "#82e0aa",  # light green
  "RK (total SE)"        = "#1e8449"   # dark green
)

p_uncertainty_v3 <- uncertainty_all_v3 %>%
  filter(!is.na(median_unc), is.finite(median_unc)) %>%
  ggplot(aes(x    = variable,
             y    = median_unc,
             fill = pred_method_label)) +
  geom_col(position = position_dodge(0.85),
           width     = 0.8) +
  scale_fill_manual(
    values = method_colors,
    name   = "Prediction method",
    labels = c(
      "RF"                   = "RF (QRF 80% PI width)",
      "OK"                   = "OK (kriging SE)",
      "RK (kriging SE only)" = "RK (residual kriging SE only)",
      "RK (total SE)"        = "RK (total SE = QRF PI + kriging SE)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    legend.position    = "bottom",
    legend.text        = element_text(size = 8),
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Prediction uncertainty by method and sampling design",
    subtitle = paste0(
      "Row 1: QRF 80% prediction interval width (RF only).\n",
      "Row 2: Kriging SE — OK (blue), RK residual kriging SE (light green),\n",
      "and RK total SE combining QRF PI width and residual ",
      "kriging SE in quadrature (dark green).\n",
      "Median across all grid pixels. ",
      "Missing bars = no convergent variogram model."
    ),
    x = NULL,
    y = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty_v3)

ggsave("../outputs/fig_uncertainty_by_method_v3.png",
       p_uncertainty_v3,
       width  = 13,
       height = 8,
       dpi    = 300)
# Convert QRF 80% PI width to approximate SE
# 80% PI = ±1.28 SD on each side = 2.56 SD total
# Therefore SE ≈ PI_width / 2.56

rk_total_uncertainty_corrected <- bind_rows(
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "SP") %>%
    left_join(
      krig_var_SP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "SP"),
      by = c("x", "y", "variable", "dataset")
    ),

  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "COMP") %>%
    left_join(
      krig_var_COMP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "COMP"),
      by = c("x", "y", "variable", "dataset")
    )
) %>%
  mutate(
    # Convert QRF PI width to approximate SE
    # 80% PI = z(0.90) - z(0.10) = 1.282 - (-1.282) = 2.564 SD
    qrf_se      = qrf_pi / 2.564,

    # Combine in quadrature on SE scale
    rk_total_se = sqrt(qrf_se^2 + coalesce(krig_var, 0)),

    # Convert back to PI scale for interpretability
    # multiply by 2.564 to report as 80% PI equivalent
    rk_total_pi = rk_total_se * 2.564
  )

# ── Summarise ─────────────────────────────────────────────
rk_uncertainty_summary_corrected <- rk_total_uncertainty_corrected %>%
  group_by(variable, dataset) %>%
  summarise(
    median_rk_total_se = round(median(rk_total_se, na.rm = TRUE), 3),
    median_rk_total_pi = round(median(rk_total_pi, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(pred_method = "RK")

print(rk_uncertainty_summary_corrected, n = 20)
write_csv(rk_uncertainty_summary_corrected,
          "../outputs/rk_total_uncertainty_corrected.csv")

# ── Option C revised: RF + RK QRF PI in top row ───────────
# RK also uses RF as trend component so QRF PI applies
# Panel 1: QRF 80% PI width — RF and RK
# Panel 2: Kriging SE — OK and RK residual only

# ── 1. QRF PI width — RF and RK ───────────────────────────
qrf_panel <- uncertainty_summary %>%
  filter(method == "QRF 80% PI width") %>%
  mutate(
    # RF QRF PI is the trend uncertainty for both RF and RK
    pred_method_label = "RF / RK trend (QRF PI)",
    metric_type       = "RF Trend Uncertainty\n(QRF 80% Prediction Interval Width)",
    variable          = recode(variable, !!!soil_var_labels),
    variable          = factor(variable, levels = soil_var_labels),
    dataset           = factor(
      recode(dataset,
             "SP"   = "Single-Point",
             "COMP" = "Composite"),
      levels = c("Single-Point", "Composite")
    )
  ) %>%
  select(variable, dataset, pred_method_label,
         metric_type, median_unc)

# ── 2. Kriging SE — OK and RK residual ────────────────────
krig_se_panel <- bind_rows(

  # OK kriging SE
  krig_se_long %>%
    filter(pred_method == "OK") %>%
    mutate(pred_method_label = "OK (kriging SE)"),

  # RK residual kriging SE
  krig_se_long %>%
    filter(pred_method == "RK") %>%
    mutate(pred_method_label = "RK (residual kriging SE)")

) %>%
  mutate(
    metric_type = "Geostatistical Uncertainty\n(Kriging Standard Error)",
    variable    = recode(variable, !!!soil_var_labels),
    variable    = factor(variable, levels = soil_var_labels),
    dataset     = factor(
      recode(dataset,
             "SP"   = "Single-Point",
             "COMP" = "Composite"),
      levels = c("Single-Point", "Composite")
    )
  ) %>%
  select(variable, dataset, pred_method_label,
         metric_type, median_unc)

# ── 3. Combine ────────────────────────────────────────────
uncertainty_option_c2 <- bind_rows(
  qrf_panel,
  krig_se_panel
) %>%
  mutate(
    pred_method_label = factor(
      pred_method_label,
      levels = c(
        "RF / RK trend (QRF PI)",
        "OK (kriging SE)",
        "RK (residual kriging SE)"
      )
    ),
    metric_type = factor(
      metric_type,
      levels = c(
        "RF Trend Uncertainty\n(QRF 80% Prediction Interval Width)",
        "Geostatistical Uncertainty\n(Kriging Standard Error)"
      )
    )
  )

# ── 4. Colors ─────────────────────────────────────────────
method_colors_c2 <- c(
  "RF / RK trend (QRF PI)"   = "#8e44ad",  # purple
  "OK (kriging SE)"          = "#2980b9",  # blue
  "RK (residual kriging SE)" = "#27ae60"   # green
)

# ── 5. Plot ───────────────────────────────────────────────
p_uncertainty_c2 <- uncertainty_option_c2 %>%
  filter(!is.na(median_unc), is.finite(median_unc)) %>%
  ggplot(aes(x    = variable,
             y    = median_unc,
             fill = pred_method_label)) +
  geom_col(position = position_dodge(0.8),
           width     = 0.75) +
  scale_fill_manual(
    values = method_colors_c2,
    name   = "Prediction method / uncertainty component",
    labels = c(
      "RF / RK trend (QRF PI)"   = "RF and RK trend (QRF 80% PI width)",
      "OK (kriging SE)"          = "OK (kriging SE)",
      "RK (residual kriging SE)" = "RK (residual kriging SE)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    strip.text.y       = element_text(size = 8),
    legend.position    = "bottom",
    legend.text        = element_text(size = 9),
    legend.key.size    = unit(0.45, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(
    title    = "Spatial prediction uncertainty by method and sampling design",
    subtitle = paste0(
      "Top row: QRF 80% PI width — applies to RF predictions and ",
      "the RF trend component of RK (Meinshausen, 2006; ",
      "Vaysse & Lagacherie, 2017).\n",
      "Bottom row: Kriging SE — OK full kriging SE (blue) and ",
      "RK residual kriging SE after RF trend removal (green).\n",
      "Note: rows are not directly comparable — QRF PI captures ",
      "conditional distributional uncertainty; kriging SE captures ",
      "geometric data configuration uncertainty.\n",
      "Missing bars = no convergent variogram model. ",
      "Median across all grid pixels."
    ),
    x = NULL,
    y = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty_c2)

ggsave("../outputs/fig_uncertainty_option_c2.png",
       p_uncertainty_c2,
       width  = 13,
       height = 8,
       dpi    = 300)


# ── Compute mean per variable per dataset for scaling ─────
means_SP <- soil_prop_SP %>%
  st_drop_geometry() %>%
  select(all_of(soil_vars)) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "mean_val") %>%
  mutate(dataset = "Single-Point")

means_COMP <- soil_prop_COMP %>%
  st_drop_geometry() %>%
  select(all_of(soil_vars)) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "mean_val") %>%
  mutate(dataset = "Composite")

means_all <- bind_rows(means_SP, means_COMP) %>%
  mutate(variable = recode(variable, !!!soil_var_labels),
         variable = factor(variable, levels = soil_var_labels))

# ── Scale uncertainty as % of mean ────────────────────────
uncertainty_scaled <- uncertainty_option_c2 %>%
  left_join(means_all,
            by = c("variable", "dataset")) %>%
  mutate(
    uncertainty_pct = round(
      (median_unc / mean_val) * 100, 1)
  )

# ── Plot scaled version ───────────────────────────────────
p_uncertainty_scaled <- uncertainty_scaled %>%
  filter(!is.na(uncertainty_pct),
         is.finite(uncertainty_pct)) %>%
  ggplot(aes(x    = variable,
             y    = uncertainty_pct,
             fill = pred_method_label)) +
  geom_col(position = position_dodge(0.8),
           width     = 0.75) +
  scale_fill_manual(
    values = method_colors_c2,
    name   = "Prediction method / uncertainty component",
    labels = c(
      "RF / RK trend (QRF PI)"   = "RF and RK trend (QRF 80% PI width)",
      "OK (kriging SE)"          = "OK (kriging SE)",
      "RK (residual kriging SE)" = "RK (residual kriging SE)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    strip.text.y       = element_text(size = 8),
    legend.position    = "bottom",
    legend.text        = element_text(size = 9),
    legend.key.size    = unit(0.45, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(
    title    = "Spatial prediction uncertainty relative to property mean",
    subtitle = paste0(
      "Top row: QRF 80% PI width as % of mean — RF and RK trend uncertainty.\n",
      "Bottom row: Kriging SE as % of mean — OK and RK residual.\n",
      "Expressing uncertainty relative to the property mean allows ",
      "comparison across soil properties on different measurement scales.\n",
      "Note: rows are not directly comparable across uncertainty types. ",
      "Missing bars = no convergent variogram model."
    ),
    x = NULL,
    y = "Uncertainty as % of property mean"
  )

print(p_uncertainty_scaled)

ggsave("../outputs/fig_uncertainty_scaled.png",
       p_uncertainty_scaled,
       width  = 13,
       height = 8,
       dpi    = 300)
```


# ============================================================
# PHASE 6 — UMAP + HDBSCAN Management Zone Delineation
# ============================================================

```{r umap-hdbscan}
run_umap_hdbscan <- function(pred_df,
                              methods          = c("RF", "RK"),
                              soil_vars,
                              include_lags     = FALSE,
                              n_components     = 2,
                              min_dist         = 0.001,
                              n_neighbors      = 30,
                              minPts           = 50,
                              cluster_eps      = 0,
                              label            = "dataset") {

  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  if (include_lags) {
    lag_cols  <- paste0(col_names, "_lag")
    col_names <- c(col_names, lag_cols[lag_cols %in% names(pred_df)])
  }
  col_names <- col_names[col_names %in% names(pred_df)]

  mat           <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  mat_clean     <- mat[complete_rows, ]
  xy_clean      <- pred_df %>% select(x, y) %>% slice(which(complete_rows))

  cat(label, "— pixels:", nrow(mat_clean),
      "| cols:", ncol(mat_clean),
      "| lags:", include_lags, "\n")

  # Zero-variance guard
  col_ranges <- apply(mat_clean, 2, function(x)
                      diff(range(x, na.rm = TRUE)))
  zero_var   <- col_ranges < 1e-10

  if (any(zero_var)) {
    cat("Warning: dropping", sum(zero_var), "constant column(s):",
        paste(colnames(mat_clean)[zero_var], collapse = ", "), "\n")
    mat_clean <- mat_clean[, !zero_var, drop = FALSE]
  }

  if (ncol(mat_clean) < 2) {
    cat("ERROR: fewer than 2 valid columns for", label, "\n")
    return(NULL)
  }

  mat_scaled <- scale(mat_clean)
  mat_scaled[!is.finite(mat_scaled)] <- 0
  stopifnot("NA values remain" = !anyNA(mat_scaled))

  cat("Fitting UMAP...\n")
  umap_fit <- uwot::umap(
    mat_scaled,
    n_components = n_components,
    min_dist     = min_dist,
    n_neighbors  = n_neighbors,
    ret_model    = TRUE,
    seed         = 20260401,
    verbose      = TRUE
  )

  emb_df        <- as.data.frame(umap_fit$embedding)
  names(emb_df) <- paste0("UMAP", seq_len(n_components))

  hdbscan_sample_size <- min(nrow(emb_df), 10000)
  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df), hdbscan_sample_size))
  emb_sample <- emb_df[sample_idx, ]

  cat("Running HDBSCAN on subsample of", hdbscan_sample_size,
      "points...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts,
    cluster_selection_epsilon = cluster_eps
  )

  cat("Clusters on subsample:", max(hdb_fit$cluster),
      "| noise:", round(mean(hdb_fit$cluster == 0) * 100, 1), "%\n")

  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  cat("Full grid — zones:", max(full_clusters),
      "| noise:", round(mean(full_clusters == 0) * 100, 1), "%\n")

  bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard    = full_clusters,
           membership_prob = full_memb)
  ) %>%
    mutate(
      dataset      = label,
      spatial_lags = include_lags,
      zone         = if_else(cluster_hard == 0L, NA_integer_,
                             as.integer(cluster_hard))
    )
}

wrap_mz_result <- function(df) {
  if (is.data.frame(df)) {
    zones <- df$zone[!is.na(df$zone)]
    return(list(result_df   = df,
                n_zones     = length(unique(zones)),
                pct_noise   = mean(is.na(df$zone)) * 100,
                umap_fit    = NULL,
                hdbscan_fit = NULL))
  }
  zones <- df$result_df$zone[!is.na(df$result_df$zone)]
  df$n_zones   <- length(unique(zones))
  df$pct_noise <- mean(is.na(df$result_df$zone)) * 100
  df
}

# ── 10 UMAP+HDBSCAN runs ──────────────────────────────────
mz_SP_RF_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RF_nolags")

mz_SP_RF_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RF_lags")

mz_SP_RK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RK_nolags")

mz_SP_RK_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RK_lags")

mz_SP_OK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_OK_nolags")

mz_COMP_RF_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RF_nolags")

mz_COMP_RF_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RF_lags")

mz_COMP_RK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RK_nolags")

mz_COMP_RK_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RK_lags")

mz_COMP_OK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_OK_nolags")

save.image("../outputs/workspace_phase6_complete.RData")
beep("mario")

mz_named_tuned <- list(
  SP_RF_nolags   = wrap_mz_result(mz_SP_RF_nolags),
  SP_RF_lags     = wrap_mz_result(mz_SP_RF_lags),
  SP_RK_nolags   = wrap_mz_result(mz_SP_RK_nolags),
  SP_RK_lags     = wrap_mz_result(mz_SP_RK_lags),
  SP_OK_nolags   = wrap_mz_result(mz_SP_OK_nolags),
  COMP_RF_nolags = wrap_mz_result(mz_COMP_RF_nolags),
  COMP_RF_lags   = wrap_mz_result(mz_COMP_RF_lags),
  COMP_RK_nolags = wrap_mz_result(mz_COMP_RK_nolags),
  COMP_RK_lags   = wrap_mz_result(mz_COMP_RK_lags),
  COMP_OK_nolags = wrap_mz_result(mz_COMP_OK_nolags)
)

purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})
```

# ============================================================
# PHASE 6c — HDBSCAN Retuning
# ============================================================

```{r hdbscan-retune}
retune_hdbscan <- function(mz_obj, minPts_new,
                            cluster_eps = 0,
                            hdbscan_n   = 10000,
                            label       = NULL) {

  emb_df       <- mz_obj$result_df %>% select(UMAP1, UMAP2)
  xy_clean     <- mz_obj$result_df %>% select(x, y)
  spatial_lags <- mz_obj$result_df$spatial_lags[1]
  lbl          <- if (!is.null(label)) label else
                    mz_obj$result_df$dataset[1]

  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df),
                             min(hdbscan_n, nrow(emb_df))))
  emb_sample <- emb_df[sample_idx, ]

  cat("Retuning:", lbl, "| minPts =", minPts_new, "...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts_new,
    cluster_selection_epsilon = cluster_eps
  )

  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  result_df <- bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard = full_clusters,
           membership_prob = full_memb)
  ) %>%
    mutate(dataset      = lbl,
           spatial_lags = spatial_lags,
           zone         = if_else(cluster_hard == 0L,
                                  NA_integer_,
                                  as.integer(cluster_hard)))

  list(result_df   = result_df,
       hdbscan_fit = hdb_fit,
       n_zones     = length(unique(full_clusters[full_clusters != 0])),
       pct_noise   = mean(full_clusters == 0) * 100)
}

# ── Sensitivity scan ──────────────────────────────────────
minpts_candidates <- c(50, 100, 200, 400, 600, 800, 1000)

sensitivity_results <- purrr::map_dfr(minpts_candidates, function(mp) {
  purrr::map_dfr(names(mz_named_tuned), function(nm) {
    r <- retune_hdbscan(mz_named_tuned[[nm]], mp, label = nm)
    tibble(label = nm, minPts = mp,
           n_zones = r$n_zones, pct_noise = round(r$pct_noise, 1))
  })
})

sensitivity_results %>%
  select(label, minPts, n_zones) %>%
  pivot_wider(names_from = label, values_from = n_zones) %>%
  arrange(minPts) %>%
  print(width = Inf)

# Set minPts after reviewing sensitivity table
minpts_final <- list(
  SP_RF_nolags   = 200,
  SP_RF_lags     = 200,
  SP_RK_nolags   = 200,
  SP_RK_lags     = 200,
  SP_OK_nolags   = 200,
  COMP_RF_nolags = 200,
  COMP_RF_lags   = 200,
  COMP_RK_nolags = 200,
  COMP_RK_lags   = 200,
  COMP_OK_nolags = 200
)

mz_named_source <- mz_named_tuned  # snapshot before overwriting

mz_named_tuned <- purrr::imap(minpts_final, function(mp, nm) {
  retune_hdbscan(mz_named_source[[nm]], mp, label = nm)
})

purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})

save.image("../outputs/workspace_phase6c_retuned.RData")
beep("coin")
```

# ============================================================
# PHASE 6b — Post-processing + Zone Uncertainty Summary
# ============================================================

```{r post-processing}
post_process_zones <- function(result_df, zone_rast_crs = "EPSG:26917") {
  n_total <- nrow(result_df)
  n_noise <- sum(is.na(result_df$zone))
  cat("Noise pixels before fill:", n_noise,
      "(", round(100 * n_noise / n_total, 1), "% of field)\n")

  if (n_noise == n_total) {
    cat("WARNING: all pixels are noise — no zones found.\n")
    return(list(result_df_filled = result_df,
                zone_rast_raw    = NULL,
                zone_rast_smooth = NULL))
  }

  assigned  <- result_df %>% filter(!is.na(zone))
  noise_pts <- result_df %>% filter( is.na(zone))

  if (nrow(noise_pts) > 0) {
    nn_idx <- FNN::get.knnx(
      data  = assigned  %>% select(UMAP1, UMAP2) %>% as.matrix(),
      query = noise_pts %>% select(UMAP1, UMAP2) %>% as.matrix(),
      k     = 1
    )$nn.index
    noise_pts$zone            <- assigned$zone[nn_idx]
    noise_pts$membership_prob <- 0
    result_df <- bind_rows(assigned, noise_pts)
  }

  cat("Noise pixels after fill: 0\n")

  zone_rast        <- result_df %>%
    select(x, y, zone) %>%
    rast(type = "xyz", crs = zone_rast_crs)
  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)

  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("Pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "% of field)\n")

  list(result_df_filled = result_df,
       zone_rast_raw    = zone_rast,
       zone_rast_smooth = zone_rast_smooth)
}

pp_named <- purrr::imap(mz_named_tuned, function(x, nm) {
  cat("\n---", nm, "---\n")
  post_process_zones(x$result_df)
})

write_zone_raster <- function(pp_obj, path) {
  if (is.null(pp_obj$zone_rast_smooth)) {
    cat("Skipping", path, "— no zones\n")
    return(invisible(NULL))
  }
  writeRaster(pp_obj$zone_rast_smooth, path, overwrite = TRUE)
  cat("Saved:", path, "\n")
}

purrr::iwalk(pp_named, function(pp, nm) {
  write_zone_raster(pp, paste0("../outputs/zones_", nm, "_smooth.tif"))
})

# ── UMAP / spatial zone / confidence plots ────────────────
plot_umap_zones <- function(res, title) {
  res$result_df %>%
    ggplot(aes(UMAP1, UMAP2, color = factor(zone),
               alpha = membership_prob)) +
    geom_point(size = 0.4) +
    scale_alpha_continuous(range = c(0.2, 1)) +
    theme_minimal() +
    labs(title = title, color = "Zone",
         alpha = "Membership\nconfidence") +
    guides(color = guide_legend(
      override.aes = list(size = 3, alpha = 1)))
}

plot_spatial_zones <- function(res, title) {
  res$result_df %>%
    filter(!is.na(zone)) %>%
    ggplot(aes(x, y, fill = factor(zone))) +
    geom_raster() + coord_equal() + theme_minimal() +
    labs(title = title, fill = "Zone",
         x = "Easting", y = "Northing")
}

plot_confidence <- function(res, title) {
  res$result_df %>%
    ggplot(aes(x, y, fill = membership_prob)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1) +
    coord_equal() + theme_minimal() +
    labs(title = title, fill = "Confidence",
         x = "Easting", y = "Northing")
}

purrr::iwalk(mz_named_tuned, function(x, nm) {
  print(plot_umap_zones(x,    paste("UMAP —", nm)))
  print(plot_spatial_zones(x, paste("Zones —", nm)))
  print(plot_confidence(x,    paste("Confidence —", nm)))
})

# ── Zone assignment uncertainty summary per config ────────
# Summarises membership probability distribution per zone
# to flag zones with high proportions of low-confidence
# (noise-filled) pixels.

zone_membership_summary <- purrr::imap_dfr(mz_named_tuned,
  function(x, nm) {
    x$result_df %>%
      mutate(zone = if_else(is.na(zone),
                            as.integer(-1L), zone)) %>%
      group_by(zone) %>%
      summarise(
        n_pixels          = n(),
        mean_memb_prob    = round(mean(membership_prob,
                                       na.rm = TRUE), 3),
        pct_low_conf      = round(
          100 * mean(membership_prob < 0.3,
                     na.rm = TRUE), 1),
        .groups = "drop"
      ) %>%
      mutate(configuration = nm)
  }
)

print(zone_membership_summary, n = 40)
write_csv(zone_membership_summary,
          "../outputs/zone_membership_uncertainty.csv")
```
```{r}
# ── Print smoothed UMAP zone maps ─────────────────────────
# Uses zone_rast_smooth regenerated from result_df_filled
# avoids stale raster pointer issue

purrr::iwalk(pp_named, function(pp, nm) {

  if (is.null(pp$result_df_filled)) {
    cat(nm, "— skipped (NULL)\n")
    return(invisible(NULL))
  }

  tryCatch({

    # Regenerate smoothed raster from result_df_filled
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")

    zone_smooth <- focal(zone_raw, w = 5,
                         fun    = "modal",
                         na.rm  = TRUE)

    zone_df_smooth <- as.data.frame(
      zone_smooth,
      xy = TRUE, na.rm = TRUE
    ) %>%
      rename(zone = 3) %>%
      filter(!is.na(zone))

    # Number of zones for title
    n_zones <- n_distinct(zone_df_smooth$zone)

    p_smooth <- ggplot(zone_df_smooth,
                       aes(x = x, y = y,
                           fill = factor(zone))) +
      geom_raster() +
      coord_equal() +
      scale_fill_brewer(palette = "Set2",
                        name    = "Zone") +
      theme_bw(base_size = 9) +
      theme(
        axis.text       = element_text(size = 6),
        legend.position = "right",
        plot.title      = element_text(size = 9,
                                        face = "bold")
      ) +
      labs(
        title    = paste0("UMAP + HDBSCAN (k = ",
                          n_zones, ") — ", nm),
        subtitle = "Post-processed: noise fill + 5×5 majority filter",
        x        = "Easting",
        y        = "Northing"
      )

    print(p_smooth)

  }, error = function(e) {
    cat(nm, "— error:", conditionMessage(e), "\n")
  })
})
```
```{r}
# ── Check UMAP embedding structure for 2-zone configs ─────
two_zone_configs <- c("COMP_RF_nolags", 
                       "COMP_RK_nolags",
                       "COMP_RK_lags",
                       "COMP_OK_nolags")

purrr::walk(two_zone_configs, function(nm) {
  
  emb_df <- mz_named_tuned[[nm]]$result_df %>%
    select(UMAP1, UMAP2, zone)
  
  p <- ggplot(emb_df,
              aes(x     = UMAP1,
                  y     = UMAP2,
                  color = factor(zone))) +
    geom_point(size = 0.3, alpha = 0.3) +
    scale_color_brewer(palette = "Set2",
                       name    = "Zone",
                       na.value = "grey70") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("UMAP embedding — ", nm),
      subtitle = paste0(
        "n zones = ",
        mz_named_tuned[[nm]]$n_zones,
        " | noise = ",
        round(mz_named_tuned[[nm]]$pct_noise, 1), "%"
      ),
      x = "UMAP1", y = "UMAP2"
    )
  
  print(p)
})

# ── Compare against a well-structured config ──────────────
p_good <- mz_named_tuned[["COMP_RF_lags"]]$result_df %>%
  select(UMAP1, UMAP2, zone) %>%
  ggplot(aes(x     = UMAP1,
             y     = UMAP2,
             color = factor(zone))) +
  geom_point(size = 0.3, alpha = 0.3) +
  scale_color_brewer(palette = "Set2",
                     name    = "Zone",
                     na.value = "grey70") +
  theme_bw(base_size = 9) +
  labs(
    title    = "UMAP embedding — COMP_RF_lags (reference)",
    subtitle = paste0(
      "n zones = ",
      mz_named_tuned[["COMP_RF_lags"]]$n_zones,
      " | noise = ",
      round(mz_named_tuned[["COMP_RF_lags"]]$pct_noise, 1),
      "%"
    ),
    x = "UMAP1", y = "UMAP2"
  )

print(p_good)

# Plot all 4 problematic configs side by side
purrr::walk(two_zone_configs, function(nm) {
  p <- mz_named_tuned[[nm]]$result_df %>%
    select(UMAP1, UMAP2, zone) %>%
    ggplot(aes(x     = UMAP1,
               y     = UMAP2,
               color = factor(zone))) +
    geom_point(size = 0.3, alpha = 0.3) +
    scale_color_brewer(palette = "Set2",
                       name     = "Zone",
                       na.value = "grey70") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("UMAP embedding — ", nm),
      subtitle = paste0(
        "n zones = ", mz_named_tuned[[nm]]$n_zones,
        " | noise = ",
        round(mz_named_tuned[[nm]]$pct_noise, 1), "%"
      ),
      x = "UMAP1", y = "UMAP2"
    )
  print(p)
})

# Test lower minPts for COMP_RK_lags specifically
purrr::walk(c(50, 75, 100, 150), function(mp) {
  r <- retune_hdbscan(
    mz_named_source[["COMP_RK_lags"]], 
    mp, 
    label = paste0("COMP_RK_lags_mp", mp))
  cat("minPts =", mp, 
      "| zones:", r$n_zones,
      "| noise:", round(r$pct_noise, 1), "%\n")
})
```

```{r}
# ── Print smoothed UMAP zone maps ─────────────────────────
# Uses zone_rast_smooth regenerated from result_df_filled
# avoids stale raster pointer issue

purrr::iwalk(pp_named, function(pp, nm) {

  if (is.null(pp$result_df_filled)) {
    cat(nm, "— skipped (NULL)\n")
    return(invisible(NULL))
  }

  tryCatch({

    # Regenerate smoothed raster from result_df_filled
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")

    zone_smooth <- focal(zone_raw, w = 5,
                         fun    = "modal",
                         na.rm  = TRUE)

    zone_df_smooth <- as.data.frame(
      zone_smooth,
      xy = TRUE, na.rm = TRUE
    ) %>%
      rename(zone = 3) %>%
      filter(!is.na(zone))

    # Number of zones for title
    n_zones <- n_distinct(zone_df_smooth$zone)

    p_smooth <- ggplot(zone_df_smooth,
                       aes(x = x, y = y,
                           fill = factor(zone))) +
      geom_raster() +
      coord_equal() +
      scale_fill_brewer(palette = "Set2",
                        name    = "Zone") +
      theme_bw(base_size = 9) +
      theme(
        axis.text       = element_text(size = 6),
        legend.position = "right",
        plot.title      = element_text(size = 9,
                                        face = "bold")
      ) +
      labs(
        title    = paste0("UMAP + HDBSCAN (k = ",
                          n_zones, ") — ", nm),
        subtitle = "Post-processed: noise fill + 5×5 majority filter",
        x        = "Easting",
        y        = "Northing"
      )

    print(p_smooth)

  }, error = function(e) {
    cat(nm, "— error:", conditionMessage(e), "\n")
  })
})
```

# ============================================================
# PHASE 7 — Baseline Comparison: PCA + k-means (elbow k)
# ============================================================

```{r pca-kmeans-baseline}
# ── Define run_pca_kmeans_wcss ─────────────────────────────
run_pca_kmeans_wcss <- function(pred_df,
                                 methods   = c("RF", "RK"),
                                 soil_vars,
                                 include_lags = FALSE,  # new arg
                                 k_range   = 2:15,
                                 label     = "dataset") {
  col_names <- as.vector(outer(soil_vars, methods,
                                paste, sep = "_"))
  # Add lag columns if requested
  if (include_lags) {
    lag_cols  <- paste0(col_names, "_lag")
    col_names <- c(col_names,
                   lag_cols[lag_cols %in% names(pred_df)])
  }
  col_names <- col_names[col_names %in% names(pred_df)]
  mat           <- pred_df %>%
    select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  xy_clean      <- pred_df %>%
    select(x, y) %>% slice(which(complete_rows))

  mat_scaled <- scale(mat[complete_rows, ])
  col_ranges <- apply(mat_scaled, 2, function(x)
    diff(range(x, na.rm = TRUE)))
  const_cols <- !is.finite(mat_scaled[1, ]) |
    col_ranges < 1e-10
  if (any(const_cols)) {
    mat_scaled <- mat_scaled[, !const_cols, drop = FALSE]
  }
  mat_scaled[!is.finite(mat_scaled)] <- 0

  pca_out    <- prcomp(mat_scaled,
                        center = FALSE, scale. = FALSE)
  cum_var    <- cumsum(pca_out$sdev^2) /
    sum(pca_out$sdev^2)
  n_comp     <- which(cum_var >= 0.95)[1]
  pca_scores <- pca_out$x[, seq_len(n_comp), drop = FALSE]

  set.seed(20260401)
  sil_idx <- sample(nrow(pca_scores),
                    min(5000, nrow(pca_scores)))
  pca_sub <- pca_scores[sil_idx, ]

  # ── Compute WCSS across k range ───────────────────────
  wcss_vals <- purrr::map_dbl(k_range, function(k) {
    set.seed(20260401)
    km <- kmeans(pca_sub, centers = k,
                 nstart = 25, iter.max = 100)
    km$tot.withinss
  })

  # ── Kneedle algorithm ─────────────────────────────────
  x      <- k_range
  y      <- wcss_vals
  x_norm <- (x - min(x)) / (max(x) - min(x))
  y_norm <- (y - min(y)) / (max(y) - min(y))
  n      <- length(x_norm)
  x1 <- x_norm[1]; y1 <- y_norm[1]
  x2 <- x_norm[n]; y2 <- y_norm[n]
  dist_to_line <- abs(
    (y2 - y1) * x_norm -
      (x2 - x1) * y_norm +
      x2 * y1 - y2 * x1
  ) / sqrt((y2 - y1)^2 + (x2 - x1)^2)
  best_k <- x[which.max(dist_to_line)]

  cat(label, "— elbow k:", best_k, "\n")

  # ── Fit final kmeans with elbow k ─────────────────────
  set.seed(20260401)
  km_best <- kmeans(pca_scores,
                    centers  = best_k,
                    nstart   = 25,
                    iter.max = 100)

  result_df <- bind_cols(
    xy_clean,
    tibble(zone_kmeans = km_best$cluster)
  ) %>% mutate(dataset = label, method = "PCA_kmeans")

  list(
    result_df    = result_df,
    pca          = pca_out,
    km           = km_best,
    wcss_profile = tibble(k = k_range,
                           wcss = wcss_vals),
    best_k       = best_k,
    elbow_k      = best_k
  )
}

# ── Run all 10 configurations with elbow k detection ──────
cat("Running PCA+kmeans with elbow-detected k (kneedle algorithm)...\n")

# ── No-lag configurations ─────────────────────────────────
baseline_SP_RF_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP,   "RF", soil_vars,
  include_lags = FALSE, label = "SP_RF")
baseline_SP_RK_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP,   "RK", soil_vars,
  include_lags = FALSE, label = "SP_RK")
baseline_SP_OK_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP,   "OK", soil_vars,
  include_lags = FALSE, label = "SP_OK")
baseline_COMP_RF_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP, "RF", soil_vars,
  include_lags = FALSE, label = "COMP_RF")
baseline_COMP_RK_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP, "RK", soil_vars,
  include_lags = FALSE, label = "COMP_RK")
baseline_COMP_OK_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP, "OK", soil_vars,
  include_lags = FALSE, label = "COMP_OK")

# ── Lag configurations — include_lags = TRUE ──────────────
baseline_SP_RF_lags_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP_lag,   "RF", soil_vars,
  include_lags = TRUE,                    # <-- fixed
  label = "SP_RF_lags_kmeans")
baseline_SP_RK_lags_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP_lag,   "RK", soil_vars,
  include_lags = TRUE,                    # <-- fixed
  label = "SP_RK_lags_kmeans")
baseline_COMP_RF_lags_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP_lag, "RF", soil_vars,
  include_lags = TRUE,                    # <-- fixed
  label = "COMP_RF_lags_kmeans")
baseline_COMP_RK_lags_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP_lag, "RK", soil_vars,
  include_lags = TRUE,                    # <-- fixed
  label = "COMP_RK_lags_kmeans")

beep("coin")

# ── Verify lag configs now differ from no-lag configs ─────
cat("SP_RF no-lag zones:\n")
print(table(baseline_SP_RF_elbow$result_df$zone_kmeans))
cat("SP_RF lag zones:\n")
print(table(baseline_SP_RF_lags_elbow$result_df$zone_kmeans))
cat("Zone assignments identical?",
    identical(baseline_SP_RF_elbow$result_df$zone_kmeans,
              baseline_SP_RF_lags_elbow$result_df$zone_kmeans),
    "\n")

# ── Elbow k summary ───────────────────────────────────────
elbow_summary <- tibble(
  config  = c("SP_RF", "SP_RK", "SP_OK",
              "COMP_RF", "COMP_RK", "COMP_OK",
              "SP_RF_lags", "SP_RK_lags",
              "COMP_RF_lags", "COMP_RK_lags"),
  elbow_k = c(
    baseline_SP_RF_elbow$best_k,
    baseline_SP_RK_elbow$best_k,
    baseline_SP_OK_elbow$best_k,
    baseline_COMP_RF_elbow$best_k,
    baseline_COMP_RK_elbow$best_k,
    baseline_COMP_OK_elbow$best_k,
    baseline_SP_RF_lags_elbow$best_k,
    baseline_SP_RK_lags_elbow$best_k,
    baseline_COMP_RF_lags_elbow$best_k,
    baseline_COMP_RK_lags_elbow$best_k
  )
)
print(elbow_summary)

# ── WCSS elbow plot with detected k marked ────────────────
bind_rows(
  baseline_SP_RF_elbow$wcss_profile        %>% mutate(config = "SP_RF"),
  baseline_SP_RK_elbow$wcss_profile        %>% mutate(config = "SP_RK"),
  baseline_SP_OK_elbow$wcss_profile        %>% mutate(config = "SP_OK"),
  baseline_COMP_RF_elbow$wcss_profile      %>% mutate(config = "COMP_RF"),
  baseline_COMP_RK_elbow$wcss_profile      %>% mutate(config = "COMP_RK"),
  baseline_COMP_OK_elbow$wcss_profile      %>% mutate(config = "COMP_OK"),
  baseline_SP_RF_lags_elbow$wcss_profile   %>% mutate(config = "SP_RF_lags"),
  baseline_SP_RK_lags_elbow$wcss_profile   %>% mutate(config = "SP_RK_lags"),
  baseline_COMP_RF_lags_elbow$wcss_profile %>% mutate(config = "COMP_RF_lags"),
  baseline_COMP_RK_lags_elbow$wcss_profile %>% mutate(config = "COMP_RK_lags")
) %>%
  left_join(elbow_summary, by = "config") %>%
  ggplot(aes(x = k, y = wcss, color = config)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  geom_vline(data = elbow_summary,
             aes(xintercept = elbow_k, color = config),
             linetype = "dashed", linewidth = 0.4, alpha = 0.7) +
  scale_color_brewer(palette = "Paired", name = "Configuration") +
  theme_minimal(base_size = 10) +
  theme(legend.position  = "right",
        legend.text      = element_text(size = 8),
        legend.key.size  = unit(0.4, "cm")) +
  labs(
    title    = "Elbow method — PCA + k-means (kneedle detection)",
    subtitle = "Dashed lines = automatically detected elbow k (Satopaa et al., 2011)",
    x = "Number of zones (k)", y = "Total within-cluster SS"
  )

ggsave("../outputs/fig_elbow_wcss.png", width = 10, height = 6, dpi = 300)
```

```{r pca-kmeans-postprocess}
post_process_kmeans <- function(result_df,
                                 zone_rast_crs = "EPSG:26917") {
  n_total   <- nrow(result_df)
  zone_rast <- result_df %>%
    select(x, y, zone = zone_kmeans) %>%
    rast(type = "xyz", crs = zone_rast_crs)

  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)
  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("PCA+kmeans: pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "%)\n")

  list(result_df_filled = result_df %>% mutate(zone = zone_kmeans),
       zone_rast_raw    = zone_rast,
       zone_rast_smooth = zone_rast_smooth)
}

# ── Post-process all elbow configurations ─────────────────
pp_baseline <- list(
  SP_RF        = post_process_kmeans(baseline_SP_RF_elbow$result_df),
  SP_RK        = post_process_kmeans(baseline_SP_RK_elbow$result_df),
  SP_OK        = post_process_kmeans(baseline_SP_OK_elbow$result_df),
  COMP_RF      = post_process_kmeans(baseline_COMP_RF_elbow$result_df),
  COMP_RK      = post_process_kmeans(baseline_COMP_RK_elbow$result_df),
  COMP_OK      = post_process_kmeans(baseline_COMP_OK_elbow$result_df),
  SP_RF_lags   = post_process_kmeans(baseline_SP_RF_lags_elbow$result_df),
  SP_RK_lags   = post_process_kmeans(baseline_SP_RK_lags_elbow$result_df),
  COMP_RF_lags = post_process_kmeans(baseline_COMP_RF_lags_elbow$result_df),
  COMP_RK_lags = post_process_kmeans(baseline_COMP_RK_lags_elbow$result_df)
)

purrr::iwalk(pp_baseline, function(pp, nm) {
  writeRaster(pp$zone_rast_smooth,
              paste0("../outputs/zones_", nm, "_kmeans_elbow_smooth.tif"),
              overwrite = TRUE)
  cat("Saved:", nm, "\n")
})

# ── Print smoothed zone maps ───────────────────────────────
purrr::iwalk(pp_baseline, function(pp, nm) {
  zone_df_smooth <- as.data.frame(
    pp$zone_rast_smooth, xy = TRUE, na.rm = TRUE
  ) %>%
    rename(zone = 3) %>%
    filter(!is.na(zone))

  p <- ggplot(zone_df_smooth,
              aes(x = x, y = y, fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    scale_fill_brewer(palette = "Set2", name = "Zone") +
    theme_bw(base_size = 9) +
    theme(axis.text = element_text(size = 6),
          legend.position = "right",
          plot.title = element_text(size = 9, face = "bold")) +
    labs(
      title    = paste0("PCA + k-means (elbow k = ",
                        n_distinct(zone_df_smooth$zone),
                        ") — ", nm),
      subtitle = "Post-processed: 5×5 majority filter applied",
      x = "Easting", y = "Northing"
    )
  print(p)
})

save.image("../outputs/workspace_pca_elbow.RData")
beep("complete")
```

# ============================================================
# PHASE 8 — Zone Validation Metrics
# ============================================================

```{r zone-validation}
# ── Load UMAP workspace ────────────────────────────────────
load("../outputs/workspace_phase6c_retuned.RData")

cat("mz_named_tuned configs:", names(mz_named_tuned), "\n")
cat("pp_named configs:", names(pp_named), "\n")

# ── Rebuild pp_baseline from elbow objects ─────────────────
# Must happen before zone_validation_pca is computed
pp_baseline <- list(
  SP_RF        = post_process_kmeans(baseline_SP_RF_elbow$result_df),
  SP_RK        = post_process_kmeans(baseline_SP_RK_elbow$result_df),
  SP_OK        = post_process_kmeans(baseline_SP_OK_elbow$result_df),
  COMP_RF      = post_process_kmeans(baseline_COMP_RF_elbow$result_df),
  COMP_RK      = post_process_kmeans(baseline_COMP_RK_elbow$result_df),
  COMP_OK      = post_process_kmeans(baseline_COMP_OK_elbow$result_df),
  SP_RF_lags   = post_process_kmeans(baseline_SP_RF_lags_elbow$result_df),
  SP_RK_lags   = post_process_kmeans(baseline_SP_RK_lags_elbow$result_df),
  COMP_RF_lags = post_process_kmeans(baseline_COMP_RF_lags_elbow$result_df),
  COMP_RK_lags = post_process_kmeans(baseline_COMP_RK_lags_elbow$result_df)
)

# Verify fix
cat("pp_baseline COMP_OK zones after fix:\n")
print(table(pp_baseline[["COMP_OK"]]$result_df_filled$zone))

# Regenerate rasters fresh
pp_baseline <- purrr::imap(pp_baseline, function(pp, nm) {
  tryCatch({
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")
    zone_smooth <- focal(zone_raw, w = 5,
                         fun = "modal", na.rm = TRUE)
    pp$zone_rast_raw    <- zone_raw
    pp$zone_rast_smooth <- zone_smooth
    cat(nm, "— raster regenerated, k =",
        n_distinct(pp$result_df_filled$zone), "\n")
    pp
  }, error = function(e) {
    cat(nm, "— error:", conditionMessage(e), "\n"); pp
  })
})

# ── Metric functions ──────────────────────────────────────
calc_variance_reduction <- function(pred_df, zone_vec,
                                     soil_vars, methods) {
  col_names  <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names  <- col_names[col_names %in% names(pred_df)]
  mat        <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  col_ranges <- apply(mat, 2, function(x) diff(range(x, na.rm = TRUE)))
  mat        <- mat[, col_ranges > 1e-10, drop = FALSE]
  keep       <- !is.na(zone_vec) & complete.cases(mat)
  mat_k      <- mat[keep, ]
  zones_k    <- zone_vec[keep]
  total_var  <- mean(apply(mat_k, 2, var, na.rm = TRUE))
  within_var <- mean(sapply(unique(zones_k), function(z) {
    mean(apply(mat_k[zones_k == z, , drop = FALSE],
               2, var, na.rm = TRUE))
  }))
  1 - (within_var / total_var)
}

calc_silhouette_sub <- function(emb_df, zone_vec, n_sub = 5000) {
  keep  <- !is.na(zone_vec)
  emb_k <- emb_df[keep, ]
  z_k   <- zone_vec[keep]
  set.seed(20260401)
  idx   <- sample(nrow(emb_k), min(n_sub, nrow(emb_k)))
  sil   <- cluster::silhouette(z_k[idx], dist(emb_k[idx, ]))
  mean(sil[, 3])
}

# ── UMAP VR + silhouette ──────────────────────────────────
zone_validation_umap <- purrr::imap_dfr(mz_named_tuned,
  function(x, nm) {
    parts  <- strsplit(nm, "_")[[1]]
    samp   <- parts[1]; method <- parts[2]
    lag_yn <- length(parts) >= 3 && parts[3] == "lags"
    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = pred_stack_SP,   SP_RK   = pred_stack_SP,
      SP_OK   = pred_stack_SP,   COMP_RF = pred_stack_COMP,
      COMP_RK = pred_stack_COMP, COMP_OK = pred_stack_COMP)
    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = lag_yn,
      approach      = "UMAP+HDBSCAN",
      VR = calc_variance_reduction(
        pred_df, x$result_df$zone, soil_vars, method),
      silhouette = calc_silhouette_sub(
        x$result_df %>% select(UMAP1, UMAP2),
        x$result_df$zone)
    )
  })

# ── PCA+kmeans VR + silhouette — elbow objects ────────────
baseline_list <- list(
  SP_RF               = baseline_SP_RF_elbow,
  SP_RK               = baseline_SP_RK_elbow,
  SP_OK               = baseline_SP_OK_elbow,
  COMP_RF             = baseline_COMP_RF_elbow,
  COMP_RK             = baseline_COMP_RK_elbow,
  COMP_OK             = baseline_COMP_OK_elbow,
  SP_RF_lags_kmeans   = baseline_SP_RF_lags_elbow,
  SP_RK_lags_kmeans   = baseline_SP_RK_lags_elbow,
  COMP_RF_lags_kmeans = baseline_COMP_RF_lags_elbow,
  COMP_RK_lags_kmeans = baseline_COMP_RK_lags_elbow
)

zone_validation_pca <- purrr::imap_dfr(baseline_list,
  function(b, nm) {
    clean   <- str_remove(nm, "_kmeans")
    has_lag <- str_detect(clean, "_lags")
    clean   <- str_remove(clean, "_lags")
    parts   <- strsplit(clean, "_")[[1]]
    samp    <- parts[1]; method <- parts[2]
    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_OK   = pred_stack_SP,
      COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_OK = pred_stack_COMP)
    ok_cols  <- pred_df %>%
      select(any_of(paste0(soil_vars, "_", method))) %>% na.omit()
    col_vars <- apply(ok_cols, 2, var, na.rm = TRUE)
    ok_cols  <- ok_cols[, col_vars > 1e-10, drop = FALSE]
    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = has_lag,
      approach      = "PCA+kmeans",
      VR = calc_variance_reduction(
        pred_df, b$result_df$zone_kmeans, soil_vars, method),
      silhouette = calc_silhouette_sub(
        prcomp(scale(ok_cols), center = FALSE)$x[, 1:2],
        b$result_df$zone_kmeans)
    )
  })

# ── Combine ───────────────────────────────────────────────
zone_validation_all <- bind_rows(zone_validation_umap,
                                   zone_validation_pca) %>%
  arrange(sampling, pred_method, lags, approach)

print(zone_validation_all)
write_csv(zone_validation_all,
          "../outputs/zone_validation_metrics.csv")
```



# ============================================================
# PHASE 8b — Agronomic Validation (held-out ap_val_sf)
# ============================================================

```{r agronomic-validation}
# ── Regenerate pp_named rasters fresh from result_df ──────
# Avoids stale terra pointer issue from loaded workspace

pp_named <- purrr::imap(pp_named, function(pp, nm) {
  
  if (is.null(pp$result_df_filled)) {
    cat(nm, "— skipped (NULL result_df)\n")
    return(pp)
  }
  
  tryCatch({
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")
    
    zone_smooth <- focal(zone_raw, w = 5,
                         fun = "modal", na.rm = TRUE)
    
    pp$zone_rast_raw    <- zone_raw
    pp$zone_rast_smooth <- zone_smooth
    cat(nm, "— raster regenerated\n")
    pp
    
  }, error = function(e) {
    cat(nm, "— error regenerating raster:", 
        conditionMessage(e), "\n")
    pp
  })
})

# ── Do the same for pp_baseline ───────────────────────────
pp_baseline <- purrr::imap(pp_baseline, function(pp, nm) {
  
  if (is.null(pp$result_df_filled)) {
    cat(nm, "— skipped (NULL result_df)\n")
    return(pp)
  }
  
  tryCatch({
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")
    
    zone_smooth <- focal(zone_raw, w = 5,
                         fun = "modal", na.rm = TRUE)
    
    pp$zone_rast_raw    <- zone_raw
    pp$zone_rast_smooth <- zone_smooth
    cat(nm, "— raster regenerated\n")
    pp
    
  }, error = function(e) {
    cat(nm, "— error regenerating raster:",
        conditionMessage(e), "\n")
    pp
  })
})
validate_zones_with_holdout <- function(zone_rast_smooth, val_sf,
                                        pred_stack, soil_vars,
                                        methods = c("RF", "RK"),
                                        label = "dataset") {
  val_vect    <- terra::vect(val_sf)
  val_zone    <- terra::extract(zone_rast_smooth, val_vect)
  val_sf$zone <- val_zone[, 2]

  n_assigned <- sum(!is.na(val_sf$zone))
  cat(label, "— validation points with zone assignment:",
      n_assigned, "of", nrow(val_sf), "\n")

  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_stack)]

  pred_pts  <- terra::vect(
    pred_stack %>% select(x, y),
    geom = c("x", "y"), crs = "EPSG:26917"
  )
  pred_zone_extract <- terra::extract(zone_rast_smooth, pred_pts, ID = FALSE)
  pred_stack_zoned  <- pred_stack %>%
    mutate(zone = pred_zone_extract[[1]])

  zone_profiles <- pred_stack_zoned %>%
    filter(!is.na(zone)) %>%
    group_by(zone) %>%
    summarise(across(all_of(col_names), mean, na.rm = TRUE),
              .groups = "drop")

  val_long <- val_sf %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    pivot_longer(-zone, names_to = "variable", values_to = "obs")

  method_suffix <- paste0("_", methods[1])

  pred_long <- zone_profiles %>%
    select(zone, ends_with(method_suffix)) %>%
    pivot_longer(-zone, names_to = "col", values_to = "pred_zone_mean") %>%
    mutate(variable = str_remove(col, method_suffix)) %>%
    select(zone, variable, pred_zone_mean)

  comparison <- val_long %>%
    left_join(pred_long, by = c("zone", "variable")) %>%
    group_by(zone, variable) %>%
    summarise(
      n_val_pts      = n(),
      mean_obs       = mean(obs,                            na.rm = TRUE),
      pred_zone_mean = first(pred_zone_mean),
      bias           = mean(obs - pred_zone_mean,           na.rm = TRUE),
      rmse_vs_zone   = sqrt(mean((obs - pred_zone_mean)^2, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(dataset = label)

  kw_results <- purrr::map_dfr(soil_vars, function(v) {
    df <- val_sf %>%
      st_drop_geometry() %>%
      filter(!is.na(zone), !is.na(.data[[v]])) %>%
      select(zone, value = all_of(v))

    if (length(unique(df$zone)) < 2 || nrow(df) < 5) {
      return(tibble(variable = v, kw_p = NA_real_, dataset = label))
    }
    kt <- kruskal.test(value ~ factor(zone), data = df)
    tibble(variable = v, kw_p = kt$p.value, dataset = label)
  })

  list(comparison     = comparison,
       kw_results     = kw_results,
       val_with_zones = val_sf)
}

get_pred_stack <- function(nm) {
  clean   <- nm %>%
    str_remove("_kmeans") %>%
    str_remove("_nolags") %>%
    str_remove("_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]; method <- parts[2]
  has_lag <- str_detect(nm, "_lags")

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_OK = pred_stack_COMP)

  list(pred_df = pred_df, method = method,
       samp = samp, has_lag = has_lag)
}

# ── UMAP validation ───────────────────────────────────────
val_umap <- purrr::imap(pp_named, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- UMAP+HDBSCAN:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method, label = nm)
})
val_umap <- val_umap[!sapply(val_umap, is.null)]

# ── PCA+kmeans validation — elbow pp_baseline ─────────────
val_kmeans <- purrr::imap(pp_baseline, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- PCA+kmeans (elbow k):", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method,
    label   = paste0(nm, "_kmeans"))
})
val_kmeans <- val_kmeans[!sapply(val_kmeans, is.null)]

val_named <- val_umap
all_val   <- c(val_named, val_kmeans)

all_kw <- purrr::map_dfr(all_val, ~ .x$kw_results) %>%
  mutate(
    significant = kw_p < 0.05,
    approach    = if_else(str_detect(dataset, "_kmeans"),
                          "PCA+kmeans", "UMAP+HDBSCAN"),
    clean       = dataset %>%
                    str_remove("_kmeans") %>%
                    str_remove("_nolags") %>%
                    str_remove("_lags"),
    sampling    = str_extract(clean, "^(SP|COMP)"),
    pred_method = str_extract(clean, "(RF|RK|OK)"),
    lags        = str_detect(dataset, "_lags")
  ) %>%
  select(-clean)

all_comparisons      <- purrr::map_dfr(val_umap,   ~ .x$comparison)
all_comparisons_pca  <- purrr::map_dfr(val_kmeans, ~ .x$comparison)
all_comparisons_full <- bind_rows(all_comparisons, all_comparisons_pca)

kw_summary <- all_kw %>%
  group_by(dataset, approach, sampling, pred_method, lags) %>%
  summarise(
    n_significant   = sum(significant, na.rm = TRUE),
    n_total         = n(),
    pct_significant = round(100 * mean(significant, na.rm = TRUE)),
    sig_vars        = paste(variable[significant], collapse = ", "),
    .groups         = "drop"
  ) %>%
  arrange(approach, sampling, pred_method, lags)

print(kw_summary, n = 30)
write_csv(all_kw,     "../outputs/zone_kruskal_wallis_all.csv")
write_csv(kw_summary, "../outputs/kw_summary_all.csv")
```

# ============================================================
# PHASE 8c — Per-Variable Observed VR + Effect Sizes
# ============================================================

```{r observed-vr-and-effect-sizes}
calc_zone_consistency <- function(val_with_zones, soil_vars, label) {
  df <- val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    na.omit()

  if (nrow(df) < 5 || length(unique(df$zone)) < 2) {
    return(tibble(configuration    = label,
                  observed_VR_mean = NA_real_,
                  observed_VR_med  = NA_real_,
                  n_pos_VR         = NA_integer_))
  }

  total_var_vec <- df %>%
    select(all_of(soil_vars)) %>%
    summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
    unlist()

  per_var_vr <- purrr::map_dbl(soil_vars, function(v) {
    within_var <- df %>%
      group_by(zone) %>%
      summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
      pull(wv) %>%
      mean(na.rm = TRUE)
    1 - (within_var / total_var_vec[[v]])
  })
  names(per_var_vr) <- soil_vars

  tibble(configuration    = label,
         observed_VR_mean = round(mean(per_var_vr,   na.rm = TRUE), 3),
         observed_VR_med  = round(median(per_var_vr, na.rm = TRUE), 3),
         n_pos_VR         = sum(per_var_vr > 0, na.rm = TRUE))
}

obs_vr_all <- bind_rows(
  purrr::imap_dfr(val_named,  ~ calc_zone_consistency(
    .x$val_with_zones, soil_vars, .y)) %>%
    rename(config_join = configuration),
  purrr::imap_dfr(val_kmeans, ~ calc_zone_consistency(
    .x$val_with_zones, soil_vars, paste0(.y, "_kmeans"))) %>%
    rename(config_join = configuration)
)

zone_pred_rmse <- all_comparisons_full %>%
  filter(!is.na(rmse_vs_zone)) %>%
  group_by(dataset) %>%
  summarise(
    mean_zone_rmse   = round(mean(rmse_vs_zone,   na.rm = TRUE), 3),
    median_zone_rmse = round(median(rmse_vs_zone, na.rm = TRUE), 3),
    .groups          = "drop"
  ) %>%
  rename(config_join = dataset)

zone_r2 <- all_comparisons_full %>%
  filter(!is.na(pred_zone_mean), !is.na(mean_obs)) %>%
  group_by(dataset) %>%
  summarise(
    r2_obs_pred = round(
      cor(mean_obs, pred_zone_mean, use = "complete.obs")^2, 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

obs_vr_by_var_all <- bind_rows(
  purrr::imap_dfr(val_named, function(x, nm) {
    df <- x$val_with_zones %>% st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>% na.omit()
    if (nrow(df) < 5 || length(unique(df$zone)) < 2) return(NULL)
    tv <- df %>% select(all_of(soil_vars)) %>%
      summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>% unlist()
    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>% group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
        pull(wv) %>% mean(na.rm = TRUE)
      tibble(configuration = nm, variable = v,
             obs_VR = round(1 - wv / tv[[v]], 3))
    })
  }),
  purrr::imap_dfr(val_kmeans, function(x, nm) {
    df <- x$val_with_zones %>% st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>% na.omit()
    if (nrow(df) < 5 || length(unique(df$zone)) < 2) return(NULL)
    tv <- df %>% select(all_of(soil_vars)) %>%
      summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>% unlist()
    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>% group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
        pull(wv) %>% mean(na.rm = TRUE)
      tibble(configuration = paste0(nm, "_kmeans"), variable = v,
             obs_VR = round(1 - wv / tv[[v]], 3))
    })
  })
)

# Supplementary VR heatmap
p_obs_vr <- obs_vr_by_var_all %>%
  left_join(
    bind_rows(
      zone_validation_umap %>%
        select(configuration, clustering = approach),
      zone_validation_pca %>%
        mutate(configuration = if_else(
          !str_detect(configuration, "_kmeans"),
          paste0(configuration, "_kmeans"),
          configuration)) %>%
        select(configuration, clustering = approach)
    ),
    by = "configuration"
  ) %>%
  mutate(
    config_label = str_replace_all(
      configuration,
      c("_nolags" = "", "_lags" = "+lags",
        "_kmeans" = "", "COMP_" = "COMP/", "SP_" = "SP/")),
    clustering = if_else(
      is.na(clustering),
      if_else(str_detect(configuration, "_kmeans"),
              "PCA+kmeans", "UMAP+HDBSCAN"), clustering),
    approach = factor(clustering,
                      levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = config_label, y = variable, fill = obs_VR)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", obs_VR)), size = 2.2) +
  facet_wrap(~ approach, scales = "free_x") +
  scale_fill_gradient2(
    low = "#d73027", mid = "#ffffbf", high = "#1a9850",
    midpoint = 0, limits = c(-1, 1),
    oob = scales::squish, name = "Observed VR"
  ) +
  theme_bw(base_size = 9) +
  theme(axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
        strip.background = element_rect(fill = "grey92"),
        strip.text       = element_text(face = "bold"),
        panel.grid       = element_blank()) +
  labs(title    = "Supplementary Figure S1. Per-variable observed variance reduction",
       subtitle = "Held-out validation points (n = 59), fully independent of prediction surfaces",
       x = NULL, y = "Soil variable")

print(p_obs_vr)
ggsave("../outputs/figS1_observed_vr_heatmap.png",
       p_obs_vr, width = 12, height = 4, dpi = 300)
write_csv(obs_vr_by_var_all, "../outputs/observed_vr_per_variable.csv")
```

# ============================================================
# Final Summary + Results Table
# ============================================================

```{r final-summary-and-table}
# ── Build summary tables ───────────────────────────────────
summary_umap <- purrr::imap_dfr(mz_named_tuned, function(x, nm) {
  parts  <- strsplit(nm, "_")[[1]]
  samp   <- parts[1]; method <- parts[2]
  lag_yn <- length(parts) >= 3 && parts[3] == "lags"
  tibble(configuration     = nm,
         clustering        = "UMAP+HDBSCAN",
         sampling_design   = ifelse(samp == "SP", "Single-point", "Composite"),
         pred_method       = method,
         spatial_lags      = lag_yn,
         minPts_used       = minpts_final[[nm]],
         n_zones           = x$n_zones,
         pct_noise_prefill = round(x$pct_noise, 1))
}) %>%
  left_join(
    all_kw %>% filter(approach == "UMAP+HDBSCAN") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                sig_vars   = paste(variable[significant], collapse = ", "),
                .groups    = "drop") %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(
    n_sig_vars = replace_na(n_sig_vars, 0L),
    sig_vars   = replace_na(sig_vars,   "none"),
    sig_vars   = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(
    zone_validation_all %>% filter(approach == "UMAP+HDBSCAN") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

# ── PCA summary — elbow baseline_list ─────────────────────
summary_pca <- purrr::imap_dfr(baseline_list, function(b, nm) {
  clean   <- str_remove(nm, "_kmeans")
  has_lag <- str_detect(clean, "_lags")
  clean   <- str_remove(clean, "_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]; method <- parts[2]
  tibble(configuration     = nm,
         clustering        = "PCA+kmeans",
         sampling_design   = ifelse(samp == "SP", "Single-point", "Composite"),
         pred_method       = method,
         spatial_lags      = has_lag,
         minPts_used       = NA_real_,
         n_zones           = b$best_k,   # elbow k stored as best_k
         pct_noise_prefill = 0)
}) %>%
  left_join(
    all_kw %>% filter(approach == "PCA+kmeans") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                sig_vars   = paste(variable[significant], collapse = ", "),
                .groups    = "drop") %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(
    n_sig_vars = replace_na(n_sig_vars, 0L),
    sig_vars   = replace_na(sig_vars,   "none"),
    sig_vars   = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(
    zone_validation_all %>% filter(approach == "PCA+kmeans") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

# ── Join all metrics ───────────────────────────────────────
final_summary <- bind_rows(summary_umap, summary_pca) %>%
  mutate(
    config_join = case_when(
      clustering == "PCA+kmeans" &
        !str_detect(configuration, "_kmeans") ~
        paste0(configuration, "_kmeans"),
      TRUE ~ configuration
    ),
    sig_vars = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(zone_pred_rmse, by = "config_join") %>%
  left_join(obs_vr_all,     by = "config_join") %>%
  left_join(zone_r2 %>% select(config_join, r2_obs_pred),
            by = "config_join") %>%
  select(-config_join)

# ── Composite score: normalized 0–1, higher = better ──────
normalize_higher <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (diff(rng) == 0) return(rep(0.5, length(x)))
  (x - rng[1]) / diff(rng)
}

normalize_lower <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (diff(rng) == 0) return(rep(0.5, length(x)))
  (rng[2] - x) / diff(rng)
}

final_ranked <- final_summary %>%
  mutate(
    s_vr_pred = normalize_higher(VR),
    s_sil     = normalize_higher(silhouette),
    s_nsig    = normalize_higher(n_sig_vars),
    s_rmse    = normalize_lower(median_zone_rmse),
    s_vr_obs  = normalize_higher(observed_VR_med),
    s_npos    = normalize_higher(n_pos_VR),
    s_r2      = normalize_higher(r2_obs_pred),
    Score     = round(
      rowMeans(cbind(s_vr_pred, s_sil, s_nsig,
                     s_rmse,    s_vr_obs, s_npos,
                     s_r2),
               na.rm = TRUE), 3)
  ) %>%
  select(configuration, Score)

# ── Print ranking ──────────────────────────────────────────
cat("\n=== Composite score — all configurations (higher = better) ===\n")
final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  arrange(desc(Score)) %>%
  select(configuration, clustering, VR, silhouette,
         n_sig_vars, median_zone_rmse,
         observed_VR_med, n_pos_VR, r2_obs_pred, Score) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  print(n = 20)

write_csv(final_summary, "../outputs/final_results_summary_complete.csv")

# ── Build table_data ──────────────────────────────────────
table_data <- final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  mutate(
    Approach  = clustering,
    Sampling  = ifelse(sampling_design == "Single-point", "SP", "COMP"),
    Method    = pred_method,
    Lags      = ifelse(spatial_lags, "Yes", "No"),
    Zones     = n_zones,
    VR_pred   = round(VR,               3),
    Sil       = round(silhouette,       3),
    N_sig_pct = round(100 * n_sig_vars / length(soil_vars), 1),
    RMSE_zone = round(median_zone_rmse, 1),
    VR_obs    = round(observed_VR_med,  3),
    N_pos     = n_pos_VR,
    R2        = round(r2_obs_pred,      3)
  ) %>%
  select(Approach, Sampling, Method, Lags, Zones,
         VR_pred, Sil, N_sig_pct, RMSE_zone,
         VR_obs, N_pos, R2, Score) %>%
  arrange(Approach, Sampling, Method, Lags)

# ── Best values per column ────────────────────────────────
best <- list(
  VR_pred   = max(table_data$VR_pred,   na.rm = TRUE),
  Sil       = max(table_data$Sil,       na.rm = TRUE),
  N_sig_pct = max(table_data$N_sig_pct, na.rm = TRUE),
  RMSE_zone = min(table_data$RMSE_zone, na.rm = TRUE),
  VR_obs    = max(table_data$VR_obs,    na.rm = TRUE),
  N_pos     = max(table_data$N_pos,     na.rm = TRUE),
  R2        = max(table_data$R2,        na.rm = TRUE),
  Score     = max(table_data$Score,     na.rm = TRUE)
)

# ── gt table ──────────────────────────────────────────────
tbl <- table_data %>%
  gt(groupname_col = "Approach") %>%
  cols_label(
    Sampling  = "Sampling",
    Method    = "Pred. method",
    Lags      = "Lags",
    Zones     = md("*k*"),
    VR_pred   = md("VR~pred~"),
    Sil       = "Silhouette",
    N_sig_pct = md("*N*~sig~ (%)"),
    RMSE_zone = md("RMSE~zone~"),
    VR_obs    = md("VR~obs~ (median)"),
    N_pos     = md("*N*~pos~"),
    R2        = md("*R*²"),
    Score     = "Score"
  ) %>%
  tab_spanner(
    label   = md("**Configuration**"),
    columns = c(Sampling, Method, Lags, Zones)
  ) %>%
  tab_spanner(
    label   = md("**Zone structure** *(all grid pixels)*"),
    columns = c(VR_pred, Sil)
  ) %>%
  tab_spanner(
    label   = md("**Agronomic validation** *(held-out, n = 59)*"),
    columns = c(N_sig_pct, RMSE_zone, VR_obs, N_pos, R2)
  ) %>%
  tab_spanner(
    label   = md("**Overall**"),
    columns = Score
  ) %>%
  # Blue highlight — best overall row
  tab_style(
    style     = list(cell_fill(color = "#d6eaf8"),
                     cell_text(weight = "bold")),
    locations = cells_body(
      rows = Sampling == "COMP" & Method == "RF" &
               Lags == "Yes" & Approach == "UMAP+HDBSCAN")
  ) %>%
  # Red shading — negative VR_obs
  tab_style(
    style     = cell_fill(color = "#fde8e8"),
    locations = cells_body(
      columns = VR_obs,
      rows    = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  tab_style(
    style     = cell_text(color = "#c0392b"),
    locations = cells_body(
      columns = VR_obs,
      rows    = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  # Bold best Score
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = Score,
      rows    = !is.na(Score) & Score == best$Score)
  ) %>%
  # Bold best per metric column
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = VR_pred,
      rows    = !is.na(VR_pred) & VR_pred == best$VR_pred)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = Sil,
      rows    = !is.na(Sil) & Sil == best$Sil)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = N_sig_pct,
      rows    = !is.na(N_sig_pct) & N_sig_pct == best$N_sig_pct)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = RMSE_zone,
      rows    = !is.na(RMSE_zone) & RMSE_zone == best$RMSE_zone)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = VR_obs,
      rows    = !is.na(VR_obs) & VR_obs == best$VR_obs)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = N_pos,
      rows    = !is.na(N_pos) & N_pos == best$N_pos)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = R2,
      rows    = !is.na(R2) & R2 == best$R2)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—") %>%
  fmt_number(columns = c(VR_pred, Sil, VR_obs, R2, Score),
             decimals = 3) %>%
  fmt_number(columns = c(N_sig_pct, RMSE_zone),
             decimals = 1) %>%
  fmt_integer(columns = c(Zones, N_pos)) %>%
  tab_caption(
    md("**Table 1.** Zone quality metrics for all 20 management zone configurations (all 9 soil variables). PCA+k-means zone counts selected via kneedle elbow detection (Satopaa et al., 2011). *N*~sig~ (%) = percentage of soil variables (of 9) significantly differentiated across zones (Kruskal–Wallis, p < 0.05). Score = normalized composite index (0–1, higher = better) computed as the mean of min–max normalized scores across seven metrics, where higher-is-better metrics use (*x* − min)/( max − min) and RMSE~zone~ uses (max − *x*)/(max − min) (Hwang & Yoon, 1981; Roudier et al., 2011). SP = single-point; COMP = composite; RF = random forest; RK = regression kriging; OK = ordinary kriging. Blue row = best overall configuration. Red = negative VR~obs~.")
  ) %>%
  tab_options(
    table.font.size                   = px(11),
    column_labels.font.size           = px(11),
    table.width                       = pct(100),
    data_row.padding                  = px(4),
    table.border.top.style            = "solid",
    table.border.top.width            = px(2),
    table.border.top.color            = "#2c3e50",
    column_labels.border.bottom.style = "solid",
    column_labels.border.bottom.width = px(1),
    column_labels.border.bottom.color = "#7f8c8d"
  )

tbl
gt::gtsave(tbl, "../outputs/table1_zone_metrics.html")
cat("Table saved.\n")

save.image("../outputs/workspace_final.RData")
beep("complete")
```



<!-- The three key changes from the original are: -->

<!-- **Phase 7** — `run_pca_kmeans` (silhouette) is dropped entirely and replaced with `run_pca_kmeans_wcss` (elbow) calls. All `baseline_*` objects are now `baseline_*_elbow`. `pp_baseline` is rebuilt from elbow result_dfs. The silhouette profile plot is replaced by the WCSS elbow plot with detected k marked. -->
  
  <!-- **Phase 8** — `baseline_list` now points to all `*_elbow` objects. Everything downstream (VR, silhouette score, KW, RMSE, R², composite rank) is identical in structure since `result_df` and `zone_rast_smooth` have the same format regardless of k selection method. -->
  
  <!-- **Table caption** — updated to note elbow k selection and all 9 variables. -->
  
# ============================================================
# SENSITIVITY ANALYSIS — Fixed k=6 for all configurations
# ============================================================
  # Controls for zone count across all methods and sampling designs
  # to enable equal-comparison of clustering approaches
  # All objects use _k6 suffix — nothing is overwritten

```{r fixed-k6-sensitivity}

# ── 1. Helper: fit kmeans at fixed k on existing PCA scores ─
fit_kmeans_fixed_k <- function(elbow_obj, k_fixed = 6,
                                label = "dataset") {
  # Extract PCA scores from existing elbow object
  cum_var    <- cumsum(elbow_obj$pca$sdev^2) /
                  sum(elbow_obj$pca$sdev^2)
  n_comp     <- which(cum_var >= 0.95)[1]
  pca_scores <- elbow_obj$pca$x[,
                  seq_len(n_comp), drop = FALSE]

  set.seed(20260401)
  km <- kmeans(pca_scores,
               centers  = k_fixed,
               nstart   = 25,
               iter.max = 100)

  # Rebuild result_df with same xy as original
  result_df <- elbow_obj$result_df %>%
    select(x, y, dataset, method) %>%
    mutate(zone_kmeans = km$cluster)

  cat(label, "— fixed k =", k_fixed, "\n")

  list(
    result_df = result_df,
    pca       = elbow_obj$pca,
    km        = km,
    best_k    = k_fixed
  )
}

# ── 2. Fit all 10 PCA configs at k=6 ──────────────────────
cat("Fitting PCA+kmeans at fixed k=6...\n")

baseline_SP_RF_k6        <- fit_kmeans_fixed_k(
  baseline_SP_RF_elbow,        6, "SP_RF")
baseline_SP_RK_k6        <- fit_kmeans_fixed_k(
  baseline_SP_RK_elbow,        6, "SP_RK")
baseline_SP_OK_k6        <- fit_kmeans_fixed_k(
  baseline_SP_OK_elbow,        6, "SP_OK")
baseline_COMP_RF_k6      <- fit_kmeans_fixed_k(
  baseline_COMP_RF_elbow,      6, "COMP_RF")
baseline_COMP_RK_k6      <- fit_kmeans_fixed_k(
  baseline_COMP_RK_elbow,      6, "COMP_RK")
baseline_COMP_OK_k6      <- fit_kmeans_fixed_k(
  baseline_COMP_OK_elbow,      6, "COMP_OK")
baseline_SP_RF_lags_k6   <- fit_kmeans_fixed_k(
  baseline_SP_RF_lags_elbow,   6, "SP_RF_lags")
baseline_SP_RK_lags_k6   <- fit_kmeans_fixed_k(
  baseline_SP_RK_lags_elbow,   6, "SP_RK_lags")
baseline_COMP_RF_lags_k6 <- fit_kmeans_fixed_k(
  baseline_COMP_RF_lags_elbow, 6, "COMP_RF_lags")
baseline_COMP_RK_lags_k6 <- fit_kmeans_fixed_k(
  baseline_COMP_RK_lags_elbow, 6, "COMP_RK_lags")

# ── 3. Post-process PCA k=6 zones ─────────────────────────
pp_baseline_k6 <- list(
  SP_RF        = post_process_kmeans(
    baseline_SP_RF_k6$result_df),
  SP_RK        = post_process_kmeans(
    baseline_SP_RK_k6$result_df),
  SP_OK        = post_process_kmeans(
    baseline_SP_OK_k6$result_df),
  COMP_RF      = post_process_kmeans(
    baseline_COMP_RF_k6$result_df),
  COMP_RK      = post_process_kmeans(
    baseline_COMP_RK_k6$result_df),
  COMP_OK      = post_process_kmeans(
    baseline_COMP_OK_k6$result_df),
  SP_RF_lags   = post_process_kmeans(
    baseline_SP_RF_lags_k6$result_df),
  SP_RK_lags   = post_process_kmeans(
    baseline_SP_RK_lags_k6$result_df),
  COMP_RF_lags = post_process_kmeans(
    baseline_COMP_RF_lags_k6$result_df),
  COMP_RK_lags = post_process_kmeans(
    baseline_COMP_RK_lags_k6$result_df)
)

# ── 4. Retune UMAP+HDBSCAN to k=6 via cluster_eps ────────
# For UMAP we cannot directly set k — instead we retune
# minPts until we get as close to k=6 as possible per config
# Use retune_hdbscan which is already defined

cat("\nRetuning UMAP+HDBSCAN toward k=6...\n")

# Helper: find minPts that gives closest to target k
find_minpts_for_k <- function(mz_source_obj, target_k = 6,
                               nm = "config",
                               candidates = c(50, 75, 100,
                                              150, 200, 300,
                                              400, 600, 800)) {
  results <- purrr::map_dfr(candidates, function(mp) {
    r <- retune_hdbscan(mz_source_obj, mp, label = nm)
    tibble(minPts = mp, n_zones = r$n_zones,
           pct_noise = r$pct_noise,
           diff = abs(r$n_zones - target_k))
  })
  best_row <- results %>%
    arrange(diff, desc(minPts)) %>%   # prefer larger minPts if tie
    slice(1)
  cat(nm, "— best minPts:", best_row$minPts,
      "| k =", best_row$n_zones,
      "| noise:", round(best_row$pct_noise, 1), "%\n")
  best_row$minPts
}

# Find best minPts per config
cat("\nSearching for minPts closest to k=6...\n")
minpts_k6 <- purrr::imap(
  mz_named_tuned,                         # <-- CORRECT
  ~ find_minpts_for_k(.x, target_k = 6, nm = .y)
)


print(tibble(config  = names(minpts_k6),
             minPts  = unlist(minpts_k6)))

# Retune all UMAP configs with best minPts
mz_named_k6 <- purrr::imap(minpts_k6, function(mp, nm) {
  retune_hdbscan(mz_named_source[[nm]], mp, label = nm)
})

mz_named_k6 <- purrr::imap(minpts_k6, function(mp, nm) {
  retune_hdbscan(mz_named_tuned[[nm]],    # <-- CORRECT
                 mp, label = nm)
})
# ── 5. Post-process UMAP k≈6 zones ────────────────────────
pp_named_k6 <- purrr::imap(mz_named_k6, function(x, nm) {
  cat("\n---", nm, "---\n")
  post_process_zones(x$result_df)
})

# Regenerate rasters fresh to avoid pointer issues
pp_named_k6 <- purrr::imap(pp_named_k6, function(pp, nm) {
  if (is.null(pp$result_df_filled)) return(pp)
  tryCatch({
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")
    zone_smooth <- focal(zone_raw, w = 5,
                         fun = "modal", na.rm = TRUE)
    pp$zone_rast_raw    <- zone_raw
    pp$zone_rast_smooth <- zone_smooth
    cat(nm, "— raster regenerated\n")
    pp
  }, error = function(e) {
    cat(nm, "— error:", conditionMessage(e), "\n"); pp
  })
})

pp_baseline_k6 <- purrr::imap(pp_baseline_k6, function(pp, nm) {
  if (is.null(pp$result_df_filled)) return(pp)
  tryCatch({
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")
    zone_smooth <- focal(zone_raw, w = 5,
                         fun = "modal", na.rm = TRUE)
    pp$zone_rast_raw    <- zone_raw
    pp$zone_rast_smooth <- zone_smooth
    pp
  }, error = function(e) {
    cat(nm, "— error:", conditionMessage(e), "\n"); pp
  })
})

# ── 6. Zone structure metrics ─────────────────────────────
# VR_pred and Silhouette for k=6 configs

baseline_list_k6 <- list(
  SP_RF               = baseline_SP_RF_k6,
  SP_RK               = baseline_SP_RK_k6,
  SP_OK               = baseline_SP_OK_k6,
  COMP_RF             = baseline_COMP_RF_k6,
  COMP_RK             = baseline_COMP_RK_k6,
  COMP_OK             = baseline_COMP_OK_k6,
  SP_RF_lags_kmeans   = baseline_SP_RF_lags_k6,
  SP_RK_lags_kmeans   = baseline_SP_RK_lags_k6,
  COMP_RF_lags_kmeans = baseline_COMP_RF_lags_k6,
  COMP_RK_lags_kmeans = baseline_COMP_RK_lags_k6
)

zone_validation_umap_k6 <- purrr::imap_dfr(
  mz_named_k6, function(x, nm) {
    parts  <- strsplit(nm, "_")[[1]]
    samp   <- parts[1]; method <- parts[2]
    lag_yn <- length(parts) >= 3 && parts[3] == "lags"
    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = pred_stack_SP,
      SP_RK   = pred_stack_SP,
      SP_OK   = pred_stack_SP,
      COMP_RF = pred_stack_COMP,
      COMP_RK = pred_stack_COMP,
      COMP_OK = pred_stack_COMP)
    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = lag_yn,
      approach      = "UMAP+HDBSCAN",
      actual_k      = x$n_zones,
      VR = calc_variance_reduction(
        pred_df, x$result_df$zone, soil_vars, method),
      silhouette = calc_silhouette_sub(
        x$result_df %>% select(UMAP1, UMAP2),
        x$result_df$zone)
    )
  })

zone_validation_pca_k6 <- purrr::imap_dfr(
  baseline_list_k6, function(b, nm) {
    clean   <- str_remove(nm, "_kmeans")
    has_lag <- str_detect(clean, "_lags")
    clean   <- str_remove(clean, "_lags")
    parts   <- strsplit(clean, "_")[[1]]
    samp    <- parts[1]; method <- parts[2]
    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_OK   = pred_stack_SP,
      COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_OK = pred_stack_COMP)
    ok_cols  <- pred_df %>%
      select(any_of(paste0(soil_vars, "_", method))) %>% na.omit()
    col_vars <- apply(ok_cols, 2, var, na.rm = TRUE)
    ok_cols  <- ok_cols[, col_vars > 1e-10, drop = FALSE]
    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = has_lag,
      approach      = "PCA+kmeans",
      actual_k      = b$best_k,
      VR = calc_variance_reduction(
        pred_df, b$result_df$zone_kmeans, soil_vars, method),
      silhouette = calc_silhouette_sub(
        prcomp(scale(ok_cols), center = FALSE)$x[, 1:2],
        b$result_df$zone_kmeans)
    )
  })

zone_validation_all_k6 <- bind_rows(
  zone_validation_umap_k6,
  zone_validation_pca_k6)

# ── 7. Agronomic validation ───────────────────────────────
val_umap_k6 <- purrr::imap(pp_named_k6, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method,
    label   = paste0(nm, "_k6"))
})
val_umap_k6 <- val_umap_k6[!sapply(val_umap_k6, is.null)]

# ── Fix: standardise all k6 labels ────────────────────────
# The problem is val_kmeans_k6 lags configs get label
# "SP_RF_lags_kmeans" instead of "SP_RF_lags_kmeans_k6"
# because get_pred_stack strips "_lags" before "_k6" is added

# Check what labels actually exist
cat("val_umap_k6 labels:\n")
print(names(val_umap_k6))
cat("\nval_kmeans_k6 labels:\n")
print(names(val_kmeans_k6))

# Rename any missing _k6 suffixes in val_kmeans_k6
val_kmeans_k6 <- purrr::imap(val_kmeans_k6, function(x, nm) {
  # Fix comparison dataset label if missing _k6
  x$comparison <- x$comparison %>%
    mutate(dataset = if_else(
      !str_detect(dataset, "_k6"),
      paste0(dataset, "_k6"),
      dataset
    ))
  x$kw_results <- x$kw_results %>%
    mutate(dataset = if_else(
      !str_detect(dataset, "_k6"),
      paste0(dataset, "_k6"),
      dataset
    ))
  x
})

# Rename list names too
names(val_kmeans_k6) <- if_else(
  !str_detect(names(val_kmeans_k6), "_k6"),
  paste0(names(val_kmeans_k6), "_k6"),
  names(val_kmeans_k6)
)

cat("\nFixed val_kmeans_k6 labels:\n")
print(names(val_kmeans_k6))

# Now rebuild all_val_k6 with corrected labels
all_val_k6 <- c(val_umap_k6, val_kmeans_k6)

# Rebuild KW with corrected labels
all_kw_k6 <- purrr::map_dfr(all_val_k6, ~ .x$kw_results) %>%
  mutate(
    significant = kw_p < 0.05,
    approach    = if_else(str_detect(dataset, "_kmeans"),
                          "PCA+kmeans", "UMAP+HDBSCAN"),
    clean       = dataset %>%
                    str_remove("_kmeans_k6") %>%
                    str_remove("_k6") %>%
                    str_remove("_nolags") %>%
                    str_remove("_lags"),
    sampling    = str_extract(clean, "^(SP|COMP)"),
    pred_method = str_extract(clean, "(RF|RK|OK)"),
    lags        = str_detect(dataset, "_lags")
  ) %>%
  select(-clean)

all_comparisons_k6      <- purrr::map_dfr(
  val_umap_k6,   ~ .x$comparison)
all_comparisons_pca_k6  <- purrr::map_dfr(
  val_kmeans_k6, ~ .x$comparison)
all_comparisons_full_k6 <- bind_rows(
  all_comparisons_k6, all_comparisons_pca_k6)

cat("\nUnique dataset labels in all_kw_k6:\n")
print(sort(unique(all_kw_k6$dataset)))

# ── Fix: rebuild obs_vr, rmse, r2 with correct join keys ──

# The internal labels in val_kmeans_k6 are like
# "SP_RF_kmeans_k6", "SP_RF_lags_kmeans_k6" etc.
# The internal labels in val_umap_k6 are like
# "SP_RF_nolags_k6", "COMP_RF_lags_k6" etc.
# Use these directly as config_join keys

obs_vr_all_k6 <- bind_rows(

  # UMAP — use list name + _k6 as key
  purrr::imap_dfr(val_umap_k6, function(x, nm) {
    calc_zone_consistency(
      x$val_with_zones, soil_vars,
      label = unique(x$kw_results$dataset)  # use internal label
    )
  }) %>% rename(config_join = configuration),

  # PCA kmeans — use internal dataset label as key
  purrr::imap_dfr(val_kmeans_k6, function(x, nm) {
    calc_zone_consistency(
      x$val_with_zones, soil_vars,
      label = unique(x$kw_results$dataset)  # use internal label
    )
  }) %>% rename(config_join = configuration)
)

zone_pred_rmse_k6 <- all_comparisons_full_k6 %>%
  filter(!is.na(rmse_vs_zone)) %>%
  group_by(dataset) %>%
  summarise(
    median_zone_rmse = round(median(rmse_vs_zone,
                                    na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

zone_r2_k6 <- all_comparisons_full_k6 %>%
  filter(!is.na(pred_zone_mean), !is.na(mean_obs)) %>%
  group_by(dataset) %>%
  summarise(
    r2_obs_pred = round(
      cor(mean_obs, pred_zone_mean,
          use = "complete.obs")^2, 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

# Verify all three now have matching keys
cat("obs_vr keys:\n")
print(sort(unique(obs_vr_all_k6$config_join)))
cat("\nrmse keys:\n")
print(sort(unique(zone_pred_rmse_k6$config_join)))
cat("\nr2 keys:\n")
print(sort(unique(zone_r2_k6$config_join)))

# ── Rebuild final_summary_k6 with corrected join keys ─────
summary_umap_k6 <- purrr::imap_dfr(mz_named_k6, function(x, nm) {
  parts  <- strsplit(nm, "_")[[1]]
  samp   <- parts[1]; method <- parts[2]
  lag_yn <- length(parts) >= 3 && parts[3] == "lags"
  # internal label is nm + _k6
  int_label <- paste0(nm, "_k6")
  tibble(configuration     = nm,
         config_join       = int_label,
         clustering        = "UMAP+HDBSCAN",
         sampling_design   = ifelse(samp == "SP",
                                    "Single-point", "Composite"),
         pred_method       = method,
         spatial_lags      = lag_yn,
         n_zones           = x$n_zones,
         pct_noise_prefill = round(x$pct_noise, 1))
}) %>%
  left_join(
    all_kw_k6 %>%
      filter(approach == "UMAP+HDBSCAN") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                .groups    = "drop") %>%
      rename(config_join = dataset),
    by = "config_join"
  ) %>%
  mutate(n_sig_vars = replace_na(n_sig_vars, 0L)) %>%
  left_join(
    zone_validation_all_k6 %>%
      filter(approach == "UMAP+HDBSCAN") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

summary_pca_k6 <- purrr::imap_dfr(
  baseline_list_k6, function(b, nm) {
    clean   <- str_remove(nm, "_kmeans")
    has_lag <- str_detect(clean, "_lags")
    clean   <- str_remove(clean, "_lags")
    parts   <- strsplit(clean, "_")[[1]]
    samp    <- parts[1]; method <- parts[2]
    # internal label matches val_kmeans_k6 internal labels
    int_label <- if_else(
      has_lag,
      paste0(samp, "_", method, "_lags_kmeans_k6"),
      paste0(samp, "_", method, "_kmeans_k6")
    )
    tibble(configuration     = nm,
           config_join       = int_label,
           clustering        = "PCA+kmeans",
           sampling_design   = ifelse(samp == "SP",
                                      "Single-point", "Composite"),
           pred_method       = method,
           spatial_lags      = has_lag,
           n_zones           = b$best_k,
           pct_noise_prefill = 0)
  }) %>%
  left_join(
    all_kw_k6 %>%
      filter(approach == "PCA+kmeans") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                .groups    = "drop") %>%
      rename(config_join = dataset),
    by = "config_join"
  ) %>%
  mutate(n_sig_vars = replace_na(n_sig_vars, 0L)) %>%
  left_join(
    zone_validation_all_k6 %>%
      filter(approach == "PCA+kmeans") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

final_summary_k6 <- bind_rows(summary_umap_k6,
                                summary_pca_k6) %>%
  left_join(zone_pred_rmse_k6, by = "config_join") %>%
  left_join(obs_vr_all_k6,     by = "config_join") %>%
  left_join(zone_r2_k6 %>%
              select(config_join, r2_obs_pred),
            by = "config_join") %>%
  select(-config_join)

# Verify joins worked
cat("\nfinal_summary_k6 — missing VR_obs count:",
    sum(is.na(final_summary_k6$observed_VR_med)), "\n")
cat("final_summary_k6 — missing RMSE count:",
    sum(is.na(final_summary_k6$median_zone_rmse)), "\n")

print(final_summary_k6 %>%
        select(configuration, clustering, n_zones,
               VR, n_sig_vars, median_zone_rmse,
               observed_VR_med, n_pos_VR, r2_obs_pred) %>%
        mutate(across(where(is.double), ~ round(.x, 3))),
      n = 20)
# ── 12. Print ranking ─────────────────────────────────────
cat("\n=== Fixed k=6 sensitivity — composite scores ===\n")
final_summary_k6 %>%
  left_join(final_ranked_k6, by = "configuration") %>%
  arrange(desc(Score)) %>%
  select(configuration, clustering, n_zones,
         VR, silhouette, n_sig_vars,
         median_zone_rmse, observed_VR_med,
         n_pos_VR, r2_obs_pred, Score) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  print(n = 20)

# ── 13. gt table ──────────────────────────────────────────
table_data_k6 <- final_summary_k6 %>%
  left_join(final_ranked_k6, by = "configuration") %>%
  mutate(
    Approach  = clustering,
    Sampling  = ifelse(sampling_design == "Single-point",
                       "SP", "COMP"),
    Method    = pred_method,
    Lags      = ifelse(spatial_lags, "Yes", "No"),
    Zones     = n_zones,
    VR_pred   = round(VR,               3),
    Sil       = round(silhouette,       3),
    N_sig_pct = round(100 * n_sig_vars /
                        length(soil_vars), 1),
    RMSE_zone = round(median_zone_rmse, 1),
    VR_obs    = round(observed_VR_med,  3),
    N_pos     = n_pos_VR,
    R2        = round(r2_obs_pred,      3)
  ) %>%
  select(Approach, Sampling, Method, Lags, Zones,
         VR_pred, Sil, N_sig_pct, RMSE_zone,
         VR_obs, N_pos, R2, Score) %>%
  arrange(Approach, Sampling, Method, Lags)

best_k6 <- list(
  VR_pred   = max(table_data_k6$VR_pred,   na.rm = TRUE),
  Sil       = max(table_data_k6$Sil,       na.rm = TRUE),
  N_sig_pct = max(table_data_k6$N_sig_pct, na.rm = TRUE),
  RMSE_zone = min(table_data_k6$RMSE_zone, na.rm = TRUE),
  VR_obs    = max(table_data_k6$VR_obs,    na.rm = TRUE),
  N_pos     = max(table_data_k6$N_pos,     na.rm = TRUE),
  R2        = max(table_data_k6$R2,        na.rm = TRUE),
  Score     = max(table_data_k6$Score,     na.rm = TRUE)
)

tbl_k6 <- table_data_k6 %>%
  gt(groupname_col = "Approach") %>%
  cols_label(
    Sampling  = "Sampling",
    Method    = "Pred. method",
    Lags      = "Lags",
    Zones     = md("*k*"),
    VR_pred   = md("VR~pred~"),
    Sil       = "Silhouette",
    N_sig_pct = md("*N*~sig~ (%)"),
    RMSE_zone = md("RMSE~zone~"),
    VR_obs    = md("VR~obs~ (median)"),
    N_pos     = md("*N*~pos~"),
    R2        = md("*R*²"),
    Score     = "Score"
  ) %>%
  tab_spanner(
    label   = md("**Configuration**"),
    columns = c(Sampling, Method, Lags, Zones)
  ) %>%
  tab_spanner(
    label   = md("**Zone structure** *(all grid pixels)*"),
    columns = c(VR_pred, Sil)
  ) %>%
  tab_spanner(
    label   = md("**Agronomic validation** *(held-out, n = 59)*"),
    columns = c(N_sig_pct, RMSE_zone, VR_obs, N_pos, R2)
  ) %>%
  tab_spanner(
    label   = md("**Overall**"),
    columns = Score
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fde8e8"),
    locations = cells_body(
      columns = VR_obs,
      rows    = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  tab_style(
    style     = cell_text(color = "#c0392b"),
    locations = cells_body(
      columns = VR_obs,
      rows    = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = Score,
      rows    = !is.na(Score) & Score == best_k6$Score)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = VR_pred,
      rows    = !is.na(VR_pred) & VR_pred == best_k6$VR_pred)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = Sil,
      rows    = !is.na(Sil) & Sil == best_k6$Sil)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = N_sig_pct,
      rows    = !is.na(N_sig_pct) &
                  N_sig_pct == best_k6$N_sig_pct)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = RMSE_zone,
      rows    = !is.na(RMSE_zone) &
                  RMSE_zone == best_k6$RMSE_zone)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = VR_obs,
      rows    = !is.na(VR_obs) &
                  VR_obs == best_k6$VR_obs)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = N_pos,
      rows    = !is.na(N_pos) & N_pos == best_k6$N_pos)
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold", color = "#1a5276"),
    locations = cells_body(
      columns = R2,
      rows    = !is.na(R2) & R2 == best_k6$R2)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—") %>%
  fmt_number(columns = c(VR_pred, Sil, VR_obs, R2, Score),
             decimals = 3) %>%
  fmt_number(columns = c(N_sig_pct, RMSE_zone),
             decimals = 1) %>%
  fmt_integer(columns = c(Zones, N_pos)) %>%
  tab_caption(md(paste0(
    "**Table S1.** Sensitivity analysis: zone quality metrics ",
    "with all configurations fixed at k = 6 zones. ",
    "UMAP+HDBSCAN configurations were retuned to the minPts ",
    "value producing the closest zone count to k = 6; ",
    "actual zone counts shown in *k* column. ",
    "PCA+k-means configurations were refitted at k = 6 using ",
    "existing PCA scores. All other methods and metrics ",
    "identical to Table 1. ",
    "Score = normalized composite index (0–1, higher = better). ",
    "Red = negative VR~obs~."
  ))) %>%
  tab_options(
    table.font.size                   = px(11),
    column_labels.font.size           = px(11),
    table.width                       = pct(100),
    data_row.padding                  = px(4),
    table.border.top.style            = "solid",
    table.border.top.width            = px(2),
    table.border.top.color            = "#2c3e50",
    column_labels.border.bottom.style = "solid",
    column_labels.border.bottom.width = px(1),
    column_labels.border.bottom.color = "#7f8c8d"
  )

tbl_k6
gt::gtsave(tbl_k6, "../outputs/tableS1_fixed_k6.html")
cat("Sensitivity table saved.\n")

save.image("../outputs/workspace_k6_sensitivity.RData")
beep("complete")
```
```{r}
# ── Score vs k scatter — Table 1 vs Table S1 ──────────────
bind_rows(
  # Table 1 results
  final_summary %>%
    left_join(final_ranked, by = "configuration") %>%
    mutate(analysis = "Data-adaptive k (Table 1)",
           k = n_zones),
  
  # Table S1 results  
  final_summary_k6 %>%
    left_join(final_ranked_k6, by = "configuration") %>%
    mutate(analysis = "Toward k=6 (Table S1)",
           k = n_zones)
) %>%
  mutate(
    clustering = if_else(str_detect(configuration, "nolags|lags") &
                           !str_detect(configuration, "kmeans"),
                         "UMAP+HDBSCAN", "PCA+kmeans"),
    method_label = str_extract(configuration, "RF|RK|OK")
  ) %>%
  ggplot(aes(x = k, y = Score,
             color = clustering,
             shape = analysis)) +

  
  geom_point(size = 3, alpha = 0.8) +
  geom_line(aes(group = configuration),
            alpha = 0.3, linewidth = 0.5) +
  scale_color_manual(
    values = c("UMAP+HDBSCAN" = "#E05C3A",
               "PCA+kmeans"   = "#4E9AF1"),
    name = "Clustering approach"
  ) +
  scale_shape_manual(
    values = c("Data-adaptive k (Table 1)" = 16,
               "Toward k=6 (Table S1)"     = 17),
    name = "Analysis"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        legend.box = "vertical") +
  labs(
    title    = "Composite score vs zone count — sensitivity analysis",
    subtitle = paste0(
      "Lines connect the same configuration across analyses. ",
      "UMAP actual k shown — not guaranteed to equal 6."
    ),
    x = "Number of zones (k)",
    y = "Composite score (higher = better)"
  )


#same thing but jitter points:
bind_rows(
  # Table 1 results
  final_summary %>%
    left_join(final_ranked, by = "configuration") %>%
    mutate(analysis = "Data-adaptive k (Table 1)",
           k = n_zones),
  
  # Table S1 results  
  final_summary_k6 %>%
    left_join(final_ranked_k6, by = "configuration") %>%
    mutate(analysis = "Toward k=6 (Table S1)",
           k = n_zones)
) %>%
  mutate(
    clustering = if_else(str_detect(configuration, "nolags|lags") &
                           !str_detect(configuration, "kmeans"),
                         "UMAP+HDBSCAN", "PCA+kmeans"),
    method_label = str_extract(configuration, "RF|RK|OK")
  ) %>%
  ggplot(aes(x = k, y = Score,
             color = clustering,
             shape = analysis)) +
  
  
  geom_point(size = 3, alpha = 0.8,
             position = position_jitter(width = 0.2, height = 0)) +
  geom_line(aes(group = configuration),
            alpha = 0.3, linewidth = 0.5) +
  scale_color_manual(
    values = c("UMAP+HDBSCAN" = "#E05C3A",
               "PCA+kmeans"   = "#4E9AF1"),
    name = "Clustering approach"
  ) +
  scale_shape_manual(
    values = c("Data-adaptive k (Table 1)" = 16,
               "Toward k=6 (Table S1)"     = 17),
    name = "Analysis"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom",
        legend.box = "vertical") +
  labs(
    title    = "Composite score vs zone count — sensitivity analysis",
    subtitle = paste0(
      "Lines connect the same configuration across analyses. ",
      "UMAP actual k shown — not guaranteed to equal 6."
    ),
    x = "Number of zones (k)",
    y = "Composite score (higher = better)"
  )
ggsave("../outputs/fig_score_vs_k_sensitivity.png",
       width = 8, height = 6, dpi = 300)
```

