---
title: "Perry_MZ_umap_04_20_2026"
format: html
---

# ============================================================
# PHASE 0 — Libraries
# ============================================================

```{r libraries, message=FALSE, warning=FALSE}
library(terra)
library(tidyverse)
library(janitor)
library(sf)
library(sp)
library(gstat)
library(randomForest)
library(quantregForest)   # QRF prediction intervals
library(CAST)
library(caret)
library(umap)
library(dbscan)
library(beepr)
library(broom)
library(FNN)
library(uwot)
library(patchwork)
library(ggtext)
library(gt)
library(gtExtras)
library(ggrepel)
library(cowplot)
library(cluster)
library(scales)
```

# ============================================================
# PHASE 1 — Data Loading & Preparation
# ============================================================

```{r data-load}
# ── Raw datasets ───────────────────────────────────────────
soil_points_nutrients <- read.csv(
  "../LOCAL DATA DO NOT PUSH/soil/perry_soil_points_surface_03_2026.csv"
) %>% clean_names()

ap_val <- read_csv(
  "../LOCAL DATA DO NOT PUSH/Ap_hz_data/hz_0_15_wide_03_30_2026.csv"
)

covariate_stack <- rast(
  "../LOCAL DATA DO NOT PUSH/covariate_stack/covariate_stack_3m_all_with_SR_indices.tif"
)

# ── Agronomic soil variable subset ────────────────────────

soil_vars <- c("lbc_1", "lb_ceq", "p_h_2", "ca", "k", "mg", "mn", "p", "zn")
cov_names <- names(covariate_stack)

# ── Helper functions ───────────────────────────────────────
add_xy <- function(dat) {
  coords <- st_coordinates(dat)
  dat %>% mutate(X = coords[, 1], Y = coords[, 2])
}

extract_covariates <- function(dat_sf, stack) {
  vect_pts <- terra::vect(dat_sf)
  cov_vals <- terra::extract(stack, vect_pts)
  bind_cols(
    dat_sf %>% st_drop_geometry(),
    cov_vals %>% dplyr::select(-ID)
  ) %>%
    bind_cols(geometry = st_geometry(dat_sf)) %>%
    st_as_sf()
}

# ── SP and COMP modelling datasets ────────────────────────
soil_prop_SP <- soil_points_nutrients %>%
  filter(type == "SPID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

soil_prop_COMP <- soil_points_nutrients %>%
  filter(type == "CID") %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(26917) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Validation dataset — HELD OUT, never used in CV ───────
ap_val_sf <- ap_val %>%
  st_as_sf(coords = c("POINT_X", "POINT_Y"), crs = 3857) %>%
  st_transform(26917) %>%
  transmute(
    point_label = paste0("VAL_", Point),
    type        = "VAL",
    lb_ceq = LBCeq_0_15,
    p_h_2  = pH.2_0_15,
    ca     = Ca_0_15,
    k      = K_0_15,
    p      = P_0_15,
     mg     = Mg_0_15,
    mn     = Mn_0_15,
    zn     = Zn_0_15,
    lbc_1  = LBC.1_0_15,
    geometry = geometry
  ) %>%
  add_xy() %>%
  extract_covariates(covariate_stack)

# ── Spatial distribution check ────────────────────────────
ggplot() +
  geom_sf(data = soil_prop_SP,   color = "steelblue",   size = 2) +
  geom_sf(data = soil_prop_COMP, color = "tomato",      size = 2, shape = 17) +
  geom_sf(data = ap_val_sf,      color = "forestgreen", size = 2, shape = 15) +
  theme_minimal() +
  labs(title = "Spatial distribution: SP (blue), COMP (red), VAL (green)")
```

# ============================================================
# PHASE 2 — Spatial Cross-Validation Setup
# ============================================================

```{r spatial-cv-setup}
make_spatial_folds <- function(sf_data, k = 5, seed = 20260401) {
  coords <- st_coordinates(sf_data)
  set.seed(seed)
  km <- kmeans(coords, centers = k, nstart = 25)
  sf_data$spatial_block <- as.character(km$cluster)

  folds <- CAST::CreateSpacetimeFolds(
    x        = sf_data %>% st_drop_geometry(),
    spacevar = "spatial_block",
    k        = k,
    seed     = seed
  )

  purrr::map(seq_along(folds$index), function(i) {
    list(
      iteration = i,
      seed      = seed,
      train_ids = folds$index[[i]],
      test_ids  = folds$indexOut[[i]]
    )
  })
}

resamples_SP   <- make_spatial_folds(soil_prop_SP,   k = 5)
resamples_COMP <- make_spatial_folds(soil_prop_COMP, k = 5)

make_cutoff <- function(sf_data) max(dist(st_coordinates(sf_data))) / 2
width_lag   <- 75

cutoff_SP   <- make_cutoff(soil_prop_SP)
cutoff_COMP <- make_cutoff(soil_prop_COMP)

# Visualise spatial blocks
soil_prop_SP$spatial_block <- as.character(
  kmeans(st_coordinates(soil_prop_SP), centers = 5, nstart = 25)$cluster
)
ggplot() +
  geom_sf(data = soil_prop_SP, aes(color = spatial_block), size = 3) +
  theme_minimal() +
  labs(title = "Spatial CV blocks — SP dataset", color = "Block")
```

# ============================================================
# PHASE 3 — Model Fitting Helpers (RF, OK, RK)
# ============================================================

```{r model-helpers}
calc_metrics <- function(obs, pred) {
  tibble(
    n          = length(obs),
    RMSE       = sqrt(mean((obs - pred)^2, na.rm = TRUE)),
    MAE        = mean(abs(obs - pred),      na.rm = TRUE),
    R2         = cor(obs, pred, use = "complete.obs")^2,
    mean_error = mean(obs - pred,           na.rm = TRUE)
  )
}

tune_rf_oob <- function(train_df, response, predictors, ntree = 500) {
  x_train   <- train_df %>% select(all_of(predictors))
  y_train   <- train_df[[response]]
  p         <- length(predictors)
  mtry_grid <- unique(pmax(1, c(
    floor(sqrt(p)), floor(p / 4), floor(p / 3), floor(p / 2)
  )))

  fits <- purrr::map(mtry_grid, function(m) {
    set.seed(sample.int(1e6, 1))
    rf <- randomForest(x = x_train, y = y_train,
                       ntree = ntree, mtry = m, importance = TRUE)
    tibble(mtry = m, oob_mse = tail(rf$mse, 1))
  }) %>% bind_rows()

  list(
    best_mtry    = fits %>% slice_min(oob_mse, n = 1) %>% pull(mtry),
    tuning_table = fits
  )
}

fit_variogram_models <- function(vg) {
  vg_df       <- as.data.frame(vg)
  init_nugget <- max(0, min(vg_df$gamma, na.rm = TRUE) * 0.5)
  init_sill   <- max(vg_df$gamma, na.rm = TRUE)
  init_psill  <- max(init_sill - init_nugget, 1e-6)
  init_range  <- max(vg_df$dist, na.rm = TRUE) / 3

  fits <- purrr::map(c("Sph", "Exp"), function(m) {
    tryCatch({
      fit <- fit.variogram(
        vg,
        model = vgm(psill = init_psill, model = m,
                    range = init_range,  nugget = init_nugget)
      )
      if (any(!is.finite(fit$psill)))               return(NULL)
      if (any(!is.finite(fit$range)))               return(NULL)
      if (any(fit$range[fit$model != "Nug"] <= 0))  return(NULL)
      fit
    }, error = function(e) NULL)
  })

  fits <- fits[!sapply(fits, is.null)]
  if (length(fits) == 0) return(NULL)

  sse      <- sapply(fits, function(f) attr(f, "SSErr"))
  best_fit <- fits[[which.min(sse)]]

  # Degeneracy guard — pure nugget = constant surface
  str_row <- best_fit %>% filter(model != "Nug")
  if (nrow(str_row) > 0 && str_row$psill < 1e-6) {
    warning("Degenerate variogram fit: psill ≈ 0. Returning NULL.")
    return(NULL)
  }
  best_fit
}

# ── RF ─────────────────────────────────────────────────────
run_rf_model <- function(train_sf, test_sf, response,
                          predictors, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  tuning  <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  preds <- predict(rf_fit, newdata = test_df %>% select(all_of(predictors)))
  list(
    model       = rf_fit,
    predictions = tibble(obs = test_df[[response]], pred = preds),
    metrics     = calc_metrics(test_df[[response]], preds),
    best_mtry   = tuning$best_mtry
  )
}

# ── OK ─────────────────────────────────────────────────────
run_ok_model <- function(train_sf, test_sf, response, cutoff, width) {
  train_sp <- as(train_sf, "Spatial")
  test_sp  <- as(test_sf,  "Spatial")
  f        <- as.formula(paste(response, "~ 1"))
  vg       <- variogram(f, data = train_sp, cutoff = cutoff, width = width)
  fit      <- fit_variogram_models(vg)
  if (is.null(fit)) return(NULL)

  kr   <- krige(f, locations = train_sp, newdata = test_sp, model = fit)
  pred <- kr$var1.pred
  obs  <- test_sf %>% st_drop_geometry() %>% pull(.data[[response]])
  list(
    variogram   = vg,
    fit         = fit,
    predictions = tibble(obs = obs, pred = pred),
    metrics     = calc_metrics(obs, pred)
  )
}

# ── RK ─────────────────────────────────────────────────────
run_rk_model <- function(train_sf, test_sf, response, predictors,
                          cutoff, width, ntree = 500) {
  train_df <- train_sf %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  test_df  <- test_sf  %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()

  train_sf2 <- train_sf %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))
  test_sf2  <- test_sf  %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))

  tuning <- tune_rf_oob(train_df, response, predictors, ntree)
  rf_fit <- randomForest(
    x = train_df %>% select(all_of(predictors)),
    y = train_df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )

  trend_train        <- predict(rf_fit, newdata = train_df %>%
                                  select(all_of(predictors)))
  trend_test         <- predict(rf_fit, newdata = test_df  %>%
                                  select(all_of(predictors)))
  train_sf2$residual <- train_df[[response]] - trend_train

  train_sp  <- as(train_sf2, "Spatial")
  test_sp   <- as(test_sf2,  "Spatial")
  vg_resid  <- variogram(residual ~ 1, data = train_sp,
                          cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  if (is.null(fit_resid)) {
    return(list(
      metrics               = calc_metrics(test_df[[response]], trend_test),
      best_mtry             = tuning$best_mtry,
      used_residual_kriging = FALSE
    ))
  }

  kr_resid <- krige(residual ~ 1, locations = train_sp,
                    newdata = test_sp, model = fit_resid)
  pred     <- trend_test + kr_resid$var1.pred
  obs      <- test_df[[response]]

  list(
    metrics               = calc_metrics(obs, pred),
    best_mtry             = tuning$best_mtry,
    used_residual_kriging = TRUE
  )
}

# ── Combined iteration runner ──────────────────────────────
run_one_iteration <- function(sf_data, split, response, predictors,
                               dataset_name, cutoff, width, ntree = 500) {
  train_sf <- sf_data[split$train_ids, ]
  test_sf  <- sf_data[split$test_ids,  ]

  rf_res <- run_rf_model(train_sf, test_sf, response, predictors, ntree)
  ok_res <- run_ok_model(train_sf, test_sf, response, cutoff, width)
  rk_res <- run_rk_model(train_sf, test_sf, response, predictors,
                          cutoff, width, ntree)

  bind_rows(
    rf_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RF",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rf_res$best_mtry),
    if (!is.null(ok_res))
      ok_res$metrics %>%
        mutate(dataset = dataset_name, variable = response, method = "OK",
               iteration = split$iteration, seed = split$seed,
               best_mtry = NA_real_),
    rk_res$metrics %>%
      mutate(dataset = dataset_name, variable = response, method = "RK",
             iteration = split$iteration, seed = split$seed,
             best_mtry = rk_res$best_mtry,
             used_residual_kriging = rk_res$used_residual_kriging)
  )
}
```

```{r run-cv, warning=FALSE}
results_SP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_SP, function(split_obj) {
    run_one_iteration(soil_prop_SP, split_obj, v, cov_names,
                      "SP", cutoff_SP, width_lag)
  })
})

results_COMP <- purrr::map_dfr(soil_vars, function(v) {
  purrr::map_dfr(resamples_COMP, function(split_obj) {
    run_one_iteration(soil_prop_COMP, split_obj, v, cov_names,
                      "COMP", cutoff_COMP, width_lag)
  })
})

results_all <- bind_rows(results_SP, results_COMP)
beep("wilhelm")

# ── CV summary ────────────────────────────────────────────
cv_summary <- results_all %>%
  group_by(dataset, variable, method) %>%
  summarise(
    n_folds     = n(),
    median_R2   = median(R2,   na.rm = TRUE),
    q25_R2      = quantile(R2,   0.25, na.rm = TRUE),
    q75_R2      = quantile(R2,   0.75, na.rm = TRUE),
    median_RMSE = median(RMSE, na.rm = TRUE),
    median_MAE  = median(MAE,  na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  arrange(variable, dataset, desc(median_R2))

print(cv_summary)
write_csv(cv_summary, "../outputs/cv_summary_spatial_folds.csv")

# ── CV violin plots ───────────────────────────────────────
results_all %>%
  mutate(dataset = factor(dataset, c("SP", "COMP")),
         method  = factor(method,  c("OK", "RF", "RK"))) %>%
  ggplot(aes(x = method, y = R2, fill = method)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
  facet_grid(variable ~ dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "Spatial CV: R² by dataset and method",
       x = "Method", y = expression(R^2)) +
  theme(legend.position = "none")

library(rstatix)

# Reshape CV results for statistical comparison
cv_long <- cv_summary %>%
  mutate(method_design = paste(method, dataset, sep = "_"))

# Friedman test — does method × sampling design affect R²?
friedman_result <- cv_long %>%
  friedman_test(median_R2 ~ method_design | variable)

print(friedman_result)

# Post-hoc pairwise Wilcoxon signed-rank tests
posthoc_result <- cv_long %>%
  wilcox_test(median_R2 ~ method_design,
              paired = TRUE,
              p.adjust.method = "bonferroni") %>%
  add_significance()

print(posthoc_result, n = 30)
# SP vs COMP comparison within each method
sp_comp_comparison <- cv_summary %>%
  select(variable, method, dataset, median_R2) %>%
  pivot_wider(names_from = dataset,
              values_from = median_R2) %>%
  group_by(method) %>%
  summarise(
    wilcox_p = wilcox.test(SP, COMP,
                            paired = TRUE,
                            exact = FALSE)$p.value,
    median_diff = median(SP - COMP, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(significant = wilcox_p < 0.05)

print(sp_comp_comparison)
# Method comparison within SP and COMP
method_comparison <- cv_summary %>%
  select(variable, method, dataset, median_R2) %>%
  group_by(dataset) %>%
  summarise(
    # RF vs OK
    rf_ok_p = wilcox.test(
      median_R2[method == "RF"],
      median_R2[method == "OK"],
      paired = TRUE, exact = FALSE)$p.value,
    # RF vs RK
    rf_rk_p = wilcox.test(
      median_R2[method == "RF"],
      median_R2[method == "RK"],
      paired = TRUE, exact = FALSE)$p.value,
    # RK vs OK
    rk_ok_p = wilcox.test(
      median_R2[method == "RK"],
      median_R2[method == "OK"],
      paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  )

print(method_comparison)

# SP outperforms COMP for RF
#SP outperforms COMP for RK
#RF_SP vs RK_COMP (p = 0.055) — borderline

#RF: significant (p = 0.033, median diff = 0.098) — SP outperforms COMP by about 0.10 R² units for RF, statistically significant
# RK: significant (p = 0.044, median diff = 0.050) — SP outperforms COMP for RK, statistically significant
# OK: not significant (p = 0.477) — no meaningful SP vs COMP difference for OK
#COMP: RF vs RK significant (p = 0.044) — RK outperforms RF under COMP sampling
```

# ============================================================
# PHASE 4 — Final Full-Dataset Models + Held-Out Validation
# ============================================================

```{r final-models}
fit_final_rf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)
  randomForest(
    x          = df %>% select(all_of(predictors)),
    y          = df[[response]],
    ntree      = ntree,
    mtry       = tuning$best_mtry,
    importance = TRUE
  )
}

fit_final_ok <- function(sf_data, response, cutoff, width) {
  sp_data <- as(sf_data, "Spatial")
  f       <- as.formula(paste(response, "~ 1"))
  vg      <- variogram(f, data = sp_data, cutoff = cutoff, width = width)
  fit     <- fit_variogram_models(vg)
  list(variogram = vg, fit = fit, formula = f, sp_data = sp_data)
}

fit_final_rk <- function(sf_data, response, predictors,
                          cutoff, width, ntree = 500) {
  df      <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  sf_data2 <- sf_data %>%
    filter(complete.cases(st_drop_geometry(
      select(., all_of(c(response, predictors))))))

  tuning  <- tune_rf_oob(df, response, predictors, ntree)
  rf_fit  <- randomForest(
    x = df %>% select(all_of(predictors)),
    y = df[[response]],
    ntree = ntree, mtry = tuning$best_mtry, importance = TRUE
  )
  trend             <- predict(rf_fit, newdata = df %>%
                                 select(all_of(predictors)))
  sf_data2$residual <- df[[response]] - trend

  sp_data  <- as(sf_data2, "Spatial")
  vg_resid <- variogram(residual ~ 1, data = sp_data,
                         cutoff = cutoff, width = width)
  fit_resid <- fit_variogram_models(vg_resid)

  list(rf_fit = rf_fit, fit_resid = fit_resid, sp_data = sp_data)
}

final_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_SP, v, cov_names),
    ok = fit_final_ok(soil_prop_SP, v, cutoff_SP, width_lag),
    rk = fit_final_rk(soil_prop_SP, v, cov_names, cutoff_SP, width_lag)
  )
)

final_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  function(v) list(
    rf = fit_final_rf(soil_prop_COMP, v, cov_names),
    ok = fit_final_ok(soil_prop_COMP, v, cutoff_COMP, width_lag),
    rk = fit_final_rk(soil_prop_COMP, v, cov_names, cutoff_COMP, width_lag)
  )
)

beep("fanfare")
```

```{r variogram-summary}
# ── Phase 4b: Variogram model summary ─────────────────────
extract_variogram_summary <- function(final_models, soil_vars, label) {
  purrr::map_dfr(soil_vars, function(v) {
    fit <- final_models[[v]]$ok$fit
    if (is.null(fit)) {
      return(tibble(variable = v, dataset = label,
                    model = NA, nugget = NA, psill = NA,
                    total_sill = NA, range_m = NA,
                    nugget_sill_ratio = NA))
    }
    nug_row  <- fit %>% filter(model == "Nug")
    str_row  <- fit %>% filter(model != "Nug")
    nugget   <- if (nrow(nug_row) > 0) nug_row$psill else 0
    psill    <- if (nrow(str_row) > 0) str_row$psill else NA
    rng      <- if (nrow(str_row) > 0) str_row$range else NA
    mod_type <- if (nrow(str_row) > 0) as.character(str_row$model) else NA
    tibble(variable = v, dataset = label, model = mod_type,
           nugget   = round(nugget, 4), psill = round(psill, 4),
           total_sill = round(nugget + psill, 4),
           range_m    = round(rng, 1),
           nugget_sill_ratio = round(nugget / (nugget + psill), 3))
  })
}

vg_summary_all <- bind_rows(
  extract_variogram_summary(final_models_SP,   soil_vars, "SP"),
  extract_variogram_summary(final_models_COMP, soil_vars, "COMP")
)

vg_summary_all %>%
  mutate(spatial_dependence = case_when(
    nugget_sill_ratio < 0.25 ~ "Strong",
    nugget_sill_ratio < 0.75 ~ "Moderate",
    TRUE                     ~ "Weak — kriging limited"
  )) %>%
  arrange(dataset, variable) %>%
  print(n = 20)

write_csv(vg_summary_all, "../outputs/variogram_summary_table.csv")
```

```{r held-out-validation}
# ── Predict to held-out validation set ────────────────────
predict_to_points <- function(models_list, response, new_sf,
                               predictors, cutoff, width) {
  new_df <- new_sf %>% st_drop_geometry() %>%
    select(any_of(c(response, predictors))) %>% na.omit()
  obs    <- new_df[[response]]

  rf_pred <- predict(models_list$rf,
                     newdata = new_df %>% select(all_of(predictors)))

  ok_pred <- tryCatch({
    kr <- krige(models_list$ok$formula,
                locations = models_list$ok$sp_data,
                newdata   = as(new_sf, "Spatial"),
                model     = models_list$ok$fit)
    kr$var1.pred
  }, error = function(e) rep(NA_real_, nrow(new_df)))

  rk_trend <- predict(models_list$rk$rf_fit,
                      newdata = new_df %>% select(all_of(predictors)))
  rk_resid <- tryCatch({
    new_sf2 <- new_sf %>%
      filter(complete.cases(st_drop_geometry(
        select(., any_of(predictors)))))
    kr2 <- krige(residual ~ 1,
                 locations = models_list$rk$sp_data,
                 newdata   = as(new_sf2, "Spatial"),
                 model     = models_list$rk$fit_resid)
    kr2$var1.pred
  }, error = function(e) rep(0, length(rk_trend)))
  rk_pred <- rk_trend + rk_resid

  bind_rows(
    calc_metrics(obs, rf_pred) %>%
      mutate(variable = response, method = "RF"),
    calc_metrics(obs, ok_pred) %>%
      mutate(variable = response, method = "OK"),
    calc_metrics(obs, rk_pred) %>%
      mutate(variable = response, method = "RK")
  )
}

val_results_all <- bind_rows(
  purrr::map_dfr(soil_vars, function(v) {
    predict_to_points(final_models_SP[[v]], v, ap_val_sf,
                      cov_names, cutoff_SP, width_lag) %>%
      mutate(dataset = "SP")
  }),
  purrr::map_dfr(soil_vars, function(v) {
    predict_to_points(final_models_COMP[[v]], v, ap_val_sf,
                      cov_names, cutoff_COMP, width_lag) %>%
      mutate(dataset = "COMP")
  })
)

write_csv(val_results_all, "../outputs/validation_results_held_out.csv")
val_results_all %>% arrange(variable, dataset, desc(R2))

#Unlike internal CV where OK occasionally outperformed RF/RK (e.g. K COMP, pH), the held-out validation tells a cleaner story — RF and RK almost universally produce higher R² than OK across all variables and both sampling designs. This is the more reliable finding since it comes from truly independent data.
```

# ============================================================
# PHASE 4c — RF Variable Importance Analysis
# ============================================================

```{r rf-importance}
# ── Extract importance from final RF models ────────────────
# Uses %IncMSE (permutation importance) — more reliable
# than node impurity (%IncNodePurity) for correlated predictors.
# Top-20 covariates shown per variable × dataset.

extract_importance <- function(rf_model, variable, dataset) {
  imp <- importance(rf_model, type = 1) # type=1 = %IncMSE
  tibble(
    covariate = rownames(imp),
    pct_inc_mse = imp[, 1],
    variable  = variable,
    dataset   = dataset
  ) %>%
    arrange(desc(pct_inc_mse))
}

importance_SP <- purrr::map_dfr(soil_vars, function(v) {
  extract_importance(final_models_SP[[v]]$rf, v, "SP")
})

importance_COMP <- purrr::map_dfr(soil_vars, function(v) {
  extract_importance(final_models_COMP[[v]]$rf, v, "COMP")
})

importance_all <- bind_rows(importance_SP, importance_COMP)

# ── Per-variable importance plots (top 20 covariates) ─────
plot_importance_per_var <- function(imp_df, dataset_label) {
  imp_df %>%
    filter(dataset == dataset_label) %>%
    group_by(variable) %>%
    slice_max(order_by = pct_inc_mse, n = 20) %>%
    ungroup() %>%
    mutate(covariate = tidytext::reorder_within(
      covariate, pct_inc_mse, variable)) %>%
    ggplot(aes(x = covariate, y = pct_inc_mse,
               fill = pct_inc_mse)) +
    geom_col(show.legend = FALSE) +
    tidytext::scale_x_reordered() +
    scale_fill_viridis_c(option = "plasma", direction = -1) +
    coord_flip() +
    facet_wrap(~ variable, scales = "free") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("RF Variable Importance — ",
                        dataset_label, " dataset"),
      subtitle = "Top 20 covariates by %IncMSE (permutation importance)",
      x        = NULL,
      y        = "% Increase in MSE"
    )
}

# Note: tidytext is used for reorder_within().
# If not installed: install.packages("tidytext")
# Alternative without tidytext: use fct_reorder within facets.

print(plot_importance_per_var(importance_all, "SP"))
print(plot_importance_per_var(importance_all, "COMP"))

# ── Aggregate importance: mean %IncMSE across all soil vars ─
# Identifies covariates that are broadly useful across
# the full agronomic variable set, not just one variable.

agg_importance <- importance_all %>%
  group_by(covariate, dataset) %>%
  summarise(
    mean_pct_inc_mse   = mean(pct_inc_mse,   na.rm = TRUE),
    median_pct_inc_mse = median(pct_inc_mse, na.rm = TRUE),
    n_vars_top20 = sum(rank(desc(pct_inc_mse)) <= 20,
                       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(dataset, desc(mean_pct_inc_mse))

# Top-30 aggregate importance
p_agg_imp <- agg_importance %>%
  group_by(dataset) %>%
  slice_max(mean_pct_inc_mse, n = 30) %>%
  ungroup() %>%
  mutate(
    covariate = tidytext::reorder_within(
      covariate, mean_pct_inc_mse, dataset)
  ) %>%
  ggplot(aes(x = covariate, y = mean_pct_inc_mse,
             fill = dataset)) +
  geom_col(show.legend = FALSE) +
  tidytext::scale_x_reordered() +
  scale_fill_manual(values = c("SP"   = "#2980b9",
                                "COMP" = "#e74c3c")) +
  coord_flip() +
  facet_wrap(~ dataset, scales = "free") +
  theme_bw(base_size = 9) +
  labs(
    title    = "Aggregate RF Variable Importance",
    subtitle = paste0("Mean %IncMSE across all ",
                      length(soil_vars),
                      " agronomic soil variables"),
    x        = NULL,
    y        = "Mean % Increase in MSE"
  )

print(p_agg_imp)

# ── Covariate overlap: top-15 per variable heatmap ────────
top15_per_var <- importance_all %>%
  group_by(dataset, variable) %>%
  slice_max(pct_inc_mse, n = 15) %>%
  ungroup() %>%
  mutate(in_top15 = TRUE)

# Which covariates appear in top-15 for multiple variables?
overlap_counts <- top15_per_var %>%
  group_by(covariate, dataset) %>%
  summarise(n_vars_in_top15 = n(), .groups = "drop") %>%
  filter(n_vars_in_top15 >= 2)

p_overlap <- top15_per_var %>%
  filter(covariate %in% overlap_counts$covariate) %>%
  ggplot(aes(x = variable, y = covariate,
             fill = pct_inc_mse)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                        name = "%IncMSE") +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title    = "Covariate importance overlap across soil variables",
    subtitle = "Covariates appearing in top-15 for ≥2 variables",
    x = "Soil variable", y = "Covariate"
  )

print(p_overlap)
single_bands <- c("coastal_blue", "blue", "green_i", "green",
                  "yellow", "red", "rededge", "nir")

# Soil variable display names
soil_var_labels <- c(
  "lbc_1"  = "Lime Buffer Capacity",
  "lb_ceq" = "LBC Equivalent",
  "p_h_2"  = "pH",
  "ca"     = "Calcium",
  "k"      = "Potassium",
  "mg"     = "Magnesium",
  "mn"     = "Manganese",
  "p"      = "Phosphorus",
  "zn"     = "Zinc"
)

# ── Covariate overlap: top-15 per variable heatmap ────────
top15_per_var <- importance_all %>%
  filter(!covariate %in% single_bands) %>%
  group_by(dataset, variable) %>%
  slice_max(pct_inc_mse, n = 15) %>%
  ungroup() %>%
  mutate(in_top15 = TRUE)

overlap_counts <- top15_per_var %>%
  group_by(covariate, dataset) %>%
  summarise(n_vars_in_top15 = n(), .groups = "drop") %>%
  filter(n_vars_in_top15 >= 2)

p_overlap2 <- top15_per_var %>%
  filter(covariate %in% overlap_counts$covariate) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable,
                      levels = soil_var_labels)
  ) %>%
  ggplot(aes(x = variable, y = covariate,
             fill = pct_inc_mse)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                        name = "%IncMSE") +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title    = "Covariate importance overlap across soil variables",
    subtitle = "Covariates appearing in top-15 for ≥2 variables (single bands excluded)",
    x = "Soil variable", y = "Covariate"
  )

print(p_overlap2)
write_csv(importance_all,
          "../outputs/rf_importance_per_variable.csv")
write_csv(agg_importance,
          "../outputs/rf_importance_aggregate.csv")

soil_var_labels <- c(
  "lbc_1"  = "Lime Buffer Capacity",
  "lb_ceq" = "Lime Buffer Capacity Eq.",
  "p_h_2"  = "pH",
  "ca"     = "Calcium (Ca)",
  "k"      = "Potassium (K)",
  "mg"     = "Magnesium (Mg)",
  "mn"     = "Manganese (Mn)",
  "p"      = "Phosphorus (P)",
  "zn"     = "Zinc (Zn)"
)

# ── Soil variable × covariate correlation heatmap ─────────
# Pearson correlation between each soil variable and each
# covariate at SP and COMP sampling locations combined.
# Helps interpret which environmental drivers are most
# associated with each soil property.

cor_data <- bind_rows(
  soil_prop_SP   %>% st_drop_geometry() %>% mutate(dataset = "SP"),
  soil_prop_COMP %>% st_drop_geometry() %>% mutate(dataset = "COMP")
) %>%
  select(all_of(soil_vars), all_of(cov_names)) %>%
  na.omit()

cor_mat <- cor(
  cor_data %>% select(all_of(soil_vars)),
  cor_data %>% select(all_of(cov_names)),
  use = "complete.obs"
) %>%
  as.data.frame() %>%
  rownames_to_column("soil_var") %>%
  pivot_longer(-soil_var,
               names_to  = "covariate",
               values_to = "r") %>%
    mutate(
        soil_var  = factor(recode(soil_var, !!!soil_var_labels),
                           levels = soil_var_labels),
        covariate = factor(covariate, levels = cov_names)
      )

p_cor_heatmap <- cor_mat %>%
  ggplot(aes(x = soil_var, y = covariate, fill = r)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = ifelse(abs(r) >= 0.30,
                                sprintf("%.2f", r), "")),
            size = 2.0, color = "grey10") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9850",
    midpoint = 0,
    limits   = c(-1, 1),
    oob      = scales::squish,
    name     = "Pearson r"
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1,
                                     size = 8, face = "bold"),
    axis.text.y      = element_text(size = 7),
    panel.grid       = element_blank(),
    legend.position  = "right"
  ) +
  labs(
    title    = "Soil property × covariate correlations",
    subtitle = "SP and COMP sampling locations combined (n = 272). Values shown where |r| ≥ 0.30",
    x        = "Soil variable",
    y        = "Covariate"
  )

print(p_cor_heatmap)
ggsave("../outputs/fig_soil_covariate_correlations.png",
       p_cor_heatmap,
       width  = 10,
       height = 14,
       dpi    = 300)
write_csv(cor_mat, "../outputs/soil_covariate_correlations.csv")
```

# ============================================================
# PHASE 5 — Full-Grid Spatial Prediction
# ============================================================

```{r grid-prediction}
grid_df <- as.data.frame(covariate_stack, xy = TRUE, na.rm = TRUE)
grid_sf <- st_as_sf(grid_df, coords = c("x", "y"), crs = 26917)
grid_sp <- as(grid_sf, "Spatial")

predict_grid_rf <- function(rf_model, grid_df, predictors) {
  predict(rf_model, newdata = grid_df %>% select(all_of(predictors)))
}

predict_grid_ok <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    kr$var1.pred
  }, error = function(e) {
    warning("OK prediction failed; returning NA")
    rep(NA_real_, nrow(grid_sp@data))
  })
}

predict_grid_rk <- function(rk_obj, grid_df, grid_sp, predictors) {
  trend <- predict(rk_obj$rf_fit,
                   newdata = grid_df %>% select(all_of(predictors)))
  resid <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    kr2$var1.pred
  }, error = function(e) rep(0, length(trend)))
  trend + resid
}

build_prediction_stack <- function(final_models, grid_df, grid_sf,
                                    grid_sp, soil_vars, predictors,
                                    label) {
  preds <- purrr::map(setNames(soil_vars, soil_vars), function(v) {
    m <- final_models[[v]]
    tibble(
      !!paste0(v, "_RF") := predict_grid_rf(m$rf, grid_df, predictors),
      !!paste0(v, "_RK") := predict_grid_rk(m$rk, grid_df, grid_sp, predictors),
      !!paste0(v, "_OK") := predict_grid_ok(m$ok, grid_sp)
    )
  }) %>% bind_cols()

  bind_cols(grid_df %>% select(x, y), preds, dataset = label)
}

pred_stack_SP   <- build_prediction_stack(
  final_models_SP,   grid_df, grid_sf, grid_sp,
  soil_vars, cov_names, "SP")
pred_stack_COMP <- build_prediction_stack(
  final_models_COMP, grid_df, grid_sf, grid_sp,
  soil_vars, cov_names, "COMP")

beep("mario")
save.image("../outputs/workspace_phase4b.RData")
beep("coin")
```

# ============================================================
# PHASE 5b — Spatial Lag Features
# ============================================================

```{r spatial-lags}
add_spatial_lags_focal <- function(pred_df, soil_vars, methods,
                                    crs = "EPSG:26917") {
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_df)]

  cat("Computing spatial lags via terra::focal for",
      length(col_names), "layers...\n")

  pred_rast <- pred_df %>%
    select(x, y, all_of(col_names)) %>%
    rast(type = "xyz", crs = crs)

  w       <- matrix(1, 3, 3)
  w[2, 2] <- 0
  lag_rast <- terra::focal(pred_rast, w = w,
                            fun = "mean", na.rm = TRUE)
  names(lag_rast) <- paste0(col_names, "_lag")

  pts    <- terra::vect(pred_df %>% select(x, y),
                        geom = c("x", "y"), crs = crs)
  lag_df <- terra::extract(lag_rast, pts, ID = FALSE)

  cat("pred_df rows:", nrow(pred_df),
      "| lag rows:", nrow(lag_df), "\n")
  bind_cols(pred_df, lag_df)
}

pred_stack_SP_lag   <- add_spatial_lags_focal(
  pred_stack_SP,   soil_vars, c("RF", "RK"))
pred_stack_COMP_lag <- add_spatial_lags_focal(
  pred_stack_COMP, soil_vars, c("RF", "RK"))

cat("SP lag stack columns:",   ncol(pred_stack_SP_lag),   "\n")
cat("COMP lag stack columns:", ncol(pred_stack_COMP_lag), "\n")
```

# ============================================================
# PHASE 5c — Uncertainty Analysis
# ============================================================

```{r uncertainty-analysis}
# ── 1. Quantile Regression Forest (RF Prediction Intervals) ─
# QRF extends RF to estimate conditional quantiles.
# We compute the 10th and 90th percentiles to produce
# an 80% prediction interval for each grid pixel.
# Interval width = spatial map of RF prediction uncertainty.

fit_qrf <- function(sf_data, response, predictors, ntree = 500) {
  df     <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(response, predictors))) %>% na.omit()
  tuning <- tune_rf_oob(df, response, predictors, ntree)

  set.seed(20260401)
  quantregForest(
    x    = as.matrix(df %>% select(all_of(predictors))),
    y    = df[[response]],
    ntree = ntree,
    mtry  = tuning$best_mtry
  )
}

cat("Fitting QRF models for SP dataset...\n")
qrf_models_SP <- purrr::map(
  setNames(soil_vars, soil_vars),
  ~ fit_qrf(soil_prop_SP, .x, cov_names)
)

cat("Fitting QRF models for COMP dataset...\n")
qrf_models_COMP <- purrr::map(
  setNames(soil_vars, soil_vars),
  ~ fit_qrf(soil_prop_COMP, .x, cov_names)
)

# Predict 80% PI to grid
grid_mat <- as.matrix(grid_df %>% select(all_of(cov_names)))

predict_qrf_grid <- function(qrf_model, grid_mat,
                              quantiles = c(0.10, 0.90)) {
  preds <- predict(qrf_model, newdata = grid_mat,
                   what = quantiles)
  tibble(
    pi_lower = preds[, 1],
    pi_upper = preds[, 2],
    pi_width = preds[, 2] - preds[, 1]
  )
}

qrf_uncertainty_SP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    p <- predict_qrf_grid(qrf_models_SP[[v]], grid_mat)
    setNames(p, paste0(v, "_", names(p)))
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

qrf_uncertainty_COMP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    p <- predict_qrf_grid(qrf_models_COMP[[v]], grid_mat)
    setNames(p, paste0(v, "_", names(p)))
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

# ── QRF uncertainty maps — one panel per variable ─────────
plot_qrf_width <- function(qrf_df, dataset_label) {
  width_cols <- paste0(soil_vars, "_pi_width")
  width_cols <- width_cols[width_cols %in% names(qrf_df)]

  qrf_df %>%
    select(x, y, all_of(width_cols)) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "pi_width") %>%
    mutate(variable = str_remove(variable, "_pi_width")) %>%
    ggplot(aes(x = x, y = y, fill = pi_width)) +
    geom_raster() +
    scale_fill_viridis_c(option = "inferno", direction = -1,
                          name = "PI width\n(80%)") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("QRF 80% Prediction Interval Width — ",
                        dataset_label),
      subtitle = "Wider = higher RF prediction uncertainty",
      x = "Easting", y = "Northing"
    )
}

print(plot_qrf_width(qrf_uncertainty_SP,   "SP"))
print(plot_qrf_width(qrf_uncertainty_COMP, "COMP"))

write_csv(qrf_uncertainty_SP,
          "../outputs/qrf_uncertainty_SP.csv")
write_csv(qrf_uncertainty_COMP,
          "../outputs/qrf_uncertainty_COMP.csv")

# ── 2. Kriging Variance Surfaces (OK and RK) ──────────────
# krige() returns var1.var — the kriging variance at each
# prediction location. This is the geostatistical uncertainty
# estimate; sqrt(var1.var) = standard error of the kriging
# prediction. For RK, this is the residual kriging variance.

predict_grid_ok_with_var <- function(ok_obj, grid_sp) {
  tryCatch({
    kr <- krige(ok_obj$formula,
                locations = ok_obj$sp_data,
                newdata   = grid_sp,
                model     = ok_obj$fit)
    tibble(
      pred = kr$var1.pred,
      krig_var = kr$var1.var
    )
  }, error = function(e) {
    tibble(pred    = rep(NA_real_, nrow(grid_sp@data)),
           krig_var = rep(NA_real_, nrow(grid_sp@data)))
  })
}

predict_grid_rk_with_var <- function(rk_obj, grid_df,
                                      grid_sp, predictors) {
  trend <- predict(rk_obj$rf_fit,
                   newdata = grid_df %>% select(all_of(predictors)))
  resid_pred <- tryCatch({
    kr2 <- krige(residual ~ 1,
                 locations = rk_obj$sp_data,
                 newdata   = grid_sp,
                 model     = rk_obj$fit_resid)
    tibble(resid = kr2$var1.pred,
           krig_var = kr2$var1.var)
  }, error = function(e) {
    tibble(resid    = rep(0, length(trend)),
           krig_var = rep(NA_real_, length(trend)))
  })
  tibble(
    pred     = trend + resid_pred$resid,
    krig_var = resid_pred$krig_var
  )
}

krig_var_SP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    m <- final_models_SP[[v]]
    ok <- predict_grid_ok_with_var(m$ok, grid_sp)
    rk <- predict_grid_rk_with_var(m$rk, grid_df, grid_sp, cov_names)
    tibble(
      !!paste0(v, "_ok_krig_var") := ok$krig_var,
      !!paste0(v, "_rk_krig_var") := rk$krig_var
    )
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

krig_var_COMP <- purrr::map_dfc(
  setNames(soil_vars, soil_vars), function(v) {
    m <- final_models_COMP[[v]]
    ok <- predict_grid_ok_with_var(m$ok, grid_sp)
    rk <- predict_grid_rk_with_var(m$rk, grid_df, grid_sp, cov_names)
    tibble(
      !!paste0(v, "_ok_krig_var") := ok$krig_var,
      !!paste0(v, "_rk_krig_var") := rk$krig_var
    )
  }
) %>% bind_cols(grid_df %>% select(x, y), .)

# ── Kriging SE maps ───────────────────────────────────────
plot_krig_se <- function(kv_df, dataset_label, method_suffix) {
  se_cols <- paste0(soil_vars, method_suffix)
  se_cols <- se_cols[se_cols %in% names(kv_df)]

  kv_df %>%
    select(x, y, all_of(se_cols)) %>%
    mutate(across(all_of(se_cols), sqrt)) %>%  # var → SE
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "krig_se") %>%
    mutate(variable = str_remove(
      variable, str_escape(method_suffix))) %>%
    ggplot(aes(x = x, y = y, fill = krig_se)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1,
                          name = "Kriging SE",
                          na.value = "grey80") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("Kriging Standard Error — ",
                        dataset_label, " (",
                        str_remove(method_suffix,
                                   "_krig_var"), ")"),
      subtitle = "sqrt(kriging variance) at each grid pixel",
      x = "Easting", y = "Northing"
    )
}

print(plot_krig_se(krig_var_SP,   "SP",   "_ok_krig_var"))
print(plot_krig_se(krig_var_SP,   "SP",   "_rk_krig_var"))
print(plot_krig_se(krig_var_COMP, "COMP", "_ok_krig_var"))
print(plot_krig_se(krig_var_COMP, "COMP", "_rk_krig_var"))

write_csv(krig_var_SP,   "../outputs/kriging_variance_SP.csv")
write_csv(krig_var_COMP, "../outputs/kriging_variance_COMP.csv")

# ── 3. Bootstrap Uncertainty for RF Predictions ───────────
# Bootstrap resampling of training data (B=100 replicates)
# to estimate prediction SD at each grid pixel.
# SD captures the sensitivity of RF predictions to the
# specific set of training observations used.

bootstrap_rf_uncertainty <- function(sf_data, soil_vars,
                                      predictors, grid_df,
                                      B = 100, ntree = 500,
                                      seed = 20260401) {
  cat("Bootstrap RF uncertainty (B =", B, "replicates)...\n")

  df_full <- sf_data %>% st_drop_geometry() %>%
    select(all_of(c(soil_vars, predictors))) %>% na.omit()

  set.seed(seed)
  boot_preds <- purrr::map(seq_len(B), function(b) {
    if (b %% 10 == 0) cat(" Bootstrap rep:", b, "/", B, "\n")
    boot_idx <- sample(nrow(df_full), replace = TRUE)
    boot_df  <- df_full[boot_idx, ]

    purrr::map_dfc(soil_vars, function(v) {
      rf_b <- tryCatch({
        randomForest(
          x = boot_df %>% select(all_of(predictors)),
          y = boot_df[[v]],
          ntree = ntree, mtry = floor(sqrt(length(predictors)))
        )
      }, error = function(e) NULL)

      if (is.null(rf_b)) {
        return(tibble(!!v := rep(NA_real_, nrow(grid_df))))
      }

      tibble(
        !!v := predict(rf_b,
                       newdata = grid_df %>%
                         select(all_of(predictors)))
      )
    })
  })

  # Compute SD across bootstrap replicates per pixel
  purrr::map_dfc(soil_vars, function(v) {
    boot_mat <- do.call(cbind, purrr::map(boot_preds, ~ .x[[v]]))
    tibble(
      !!paste0(v, "_boot_sd")   := apply(boot_mat, 1,
                                          sd, na.rm = TRUE),
      !!paste0(v, "_boot_mean") := apply(boot_mat, 1,
                                          mean, na.rm = TRUE)
    )
  }) %>%
    bind_cols(grid_df %>% select(x, y), .)
}

# B=100 replicates — increase to 500 for publication
cat("Running bootstrap for SP dataset...\n")
boot_uncertainty_SP   <- bootstrap_rf_uncertainty(
  soil_prop_SP,   soil_vars, cov_names, grid_df, B = 100)

cat("Running bootstrap for COMP dataset...\n")
boot_uncertainty_COMP <- bootstrap_rf_uncertainty(
  soil_prop_COMP, soil_vars, cov_names, grid_df, B = 100)

# ── Bootstrap SD maps ─────────────────────────────────────
plot_boot_sd <- function(boot_df, dataset_label) {
  sd_cols <- paste0(soil_vars, "_boot_sd")
  sd_cols <- sd_cols[sd_cols %in% names(boot_df)]

  boot_df %>%
    select(x, y, all_of(sd_cols)) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd")) %>%
    ggplot(aes(x = x, y = y, fill = boot_sd)) +
    geom_raster() +
    scale_fill_viridis_c(option = "rocket", direction = -1,
                          name = "Bootstrap SD") +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal(base_size = 9) +
    labs(
      title    = paste0("RF Bootstrap Uncertainty (B=100) — ",
                        dataset_label),
      subtitle = "SD of predictions across 100 bootstrap replicates",
      x = "Easting", y = "Northing"
    )
}

print(plot_boot_sd(boot_uncertainty_SP,   "SP"))
print(plot_boot_sd(boot_uncertainty_COMP, "COMP"))

write_csv(boot_uncertainty_SP,
          "../outputs/bootstrap_uncertainty_SP.csv")
write_csv(boot_uncertainty_COMP,
          "../outputs/bootstrap_uncertainty_COMP.csv")

# ── 4. Zone Assignment Uncertainty (membership probability) ─
# Already computed within HDBSCAN — handled in Phase 6b.
# Membership probability ∈ [0,1]: 1 = core cluster member,
# 0 = peripherally assigned (borderline pixel).
# Noise pixels assigned by NN have membership_prob = 0.
# Reported spatially and summarised per zone below.
# See plot_confidence() in Phase 6b for spatial maps.

# ── Uncertainty comparison summary ────────────────────────
# Side-by-side comparison of uncertainty magnitude by method
uncertainty_summary <- bind_rows(
  # QRF mean PI width
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable  = str_remove(variable, "_pi_width"),
           method    = "QRF 80% PI width",
           dataset   = "SP"),

  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           method   = "QRF 80% PI width",
           dataset  = "COMP"),

  # Bootstrap SD
  boot_uncertainty_SP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           method   = "Bootstrap SD",
           dataset  = "SP"),

  boot_uncertainty_COMP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable", values_to = "value") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           method   = "Bootstrap SD",
           dataset  = "COMP")
) %>%
  group_by(variable, method, dataset) %>%
  summarise(
    mean_unc   = round(mean(value,   na.rm = TRUE), 3),
    median_unc = round(median(value, na.rm = TRUE), 3),
    .groups    = "drop"
  )

print(uncertainty_summary, n = 40)
write_csv(uncertainty_summary,
          "../outputs/uncertainty_summary.csv")

beep("wilhelm")
save.image("../outputs/workspace_uncertainty.RData")
#ERROR:
#Per-variable interpretation:

  # pH — bootstrap SD ~0.07–0.10, PI width ~0.9–1.1. Very tight, makes sense — pH varies narrowly at field scale and RF predicts it well
  # P — bootstrap SD ~2–3, PI width ~20–23. Low uncertainty relative to the other nutrients, consistent with P often being spatially structured by management history
  # K — moderate uncertainty, COMP tighter than SP which is expected given composite sampling smooths local variability
  # Ca and lb_ceq — largest absolute uncertainty, which makes sense as these are highly variable and correlated with soil texture and organic matter gradients
#The SP > COMP pattern for most variables is also exactly what I expect w/ composite samples average over a larger area so the RF trained on them sees less local noise, producing tighter predictions.

#P flips the pattern (COMP SD slightly higher than SP). That's worth a sentence in my methods ...it may reflect that composite sampling obscures real P spatial structure that single-point sampling captures, since P is often highly localized around application history.
# Kriging SE summary — OK
bind_rows(
  krig_var_SP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "SP", method = "OK kriging SE"),

  krig_var_COMP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "COMP", method = "OK kriging SE"),

  krig_var_SP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "SP", method = "RK kriging SE"),

  krig_var_COMP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(), list(
      mean   = ~ mean(sqrt(.), na.rm = TRUE),
      median = ~ median(sqrt(.), na.rm = TRUE)))) %>%
    pivot_longer(everything()) %>%
    mutate(dataset = "COMP", method = "RK kriging SE")
) %>%
  mutate(
    variable = str_extract(name, paste(soil_vars, collapse = "|")),
    stat     = if_else(str_detect(name, "_mean"), "mean", "median")
  ) %>%
  select(variable, stat, dataset, method, value) %>%
  arrange(variable, method, dataset, stat) 

# FINAL UNCERTAINTY SUMMARY:
# SP > COMP for most variables (SP higher uncertainty), except P (COMP slightly higher)
# OK > RK > RF in uncertainty magnitude
# HOWEVER — these are not apples-to-apples comparisons:
#   - OK SE is purely geometry-driven (data configuration)
#   - RK SE is residual kriging component only (incomplete — missing RF trend uncertainty)
#   - RF bootstrap SD captures training data sensitivity only
# RK total uncertainty = RF trend uncertainty + residual kriging variance (neither alone is complete)
# All three are complementary, not directly comparable on the same scale

```
```{r}
#uncertainty figure for raw maps:
# ── Unified uncertainty figure ─────────────────────────────
# Combines RF bootstrap SD, QRF PI width, and kriging SE
# for OK and RK into one comparable figure

soil_var_labels <- c(
  "lbc_1"  = "LBC-1",
  "lb_ceq" = "LBC Eq.",
  "p_h_2"  = "pH",
  "ca"     = "Calcium",
  "k"      = "Potassium",
  "mg"     = "Magnesium",
  "mn"     = "Manganese",
  "p"      = "Phosphorus",
  "zn"     = "Zinc"
)

# ── 1. RF uncertainty — bootstrap SD and QRF PI ───────────
rf_uncertainty <- uncertainty_summary %>%
  mutate(
    pred_method = "RF",
    metric_type = method
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc)

# ── 2. Kriging SE — OK and RK ─────────────────────────────
krig_se_long <- bind_rows(
  krig_var_SP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "SP", pred_method = "OK"),

  krig_var_COMP %>%
    select(ends_with("_ok_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "COMP", pred_method = "OK"),

  krig_var_SP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "SP", pred_method = "RK"),

  krig_var_COMP %>%
    select(ends_with("_rk_krig_var")) %>%
    summarise(across(everything(),
                     ~ median(sqrt(.), na.rm = TRUE))) %>%
    pivot_longer(everything(),
                 names_to = "name",
                 values_to = "median_unc") %>%
    mutate(dataset = "COMP", pred_method = "RK")
) %>%
  mutate(
    variable    = str_extract(name,
                               paste(soil_vars, collapse = "|")),
    metric_type = "Kriging SE"
  ) %>%
  filter(!is.na(variable), is.finite(median_unc)) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc)

# ── 3. Combine all uncertainty metrics ────────────────────
uncertainty_all <- bind_rows(
  rf_uncertainty %>%
    filter(metric_type == "Bootstrap SD"),
  krig_se_long
) %>%
  mutate(
    variable    = recode(variable, !!!soil_var_labels),
    variable    = factor(variable, levels = soil_var_labels),
    pred_method = factor(pred_method,
                         levels = c("RF", "OK", "RK")),
    dataset     = factor(dataset,
                         levels = c("SP", "COMP"),
                         labels = c("Single-Point",
                                    "Composite")),
    metric_type = factor(metric_type,
                         levels = c("Bootstrap SD",
                                    "Kriging SE"))
  )

# ── 4. Plot ───────────────────────────────────────────────
p_uncertainty <- uncertainty_all %>%
  ggplot(aes(x = variable,
             y = median_unc,
             fill = pred_method)) +
  geom_col(position = position_dodge(0.8),
           width = 0.7) +
  scale_fill_manual(
    values = c("RF" = "#2980b9",
               "OK" = "#8e44ad",
               "RK" = "#27ae60"),
    name = "Prediction method"
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 35, hjust = 1,
                                     size = 8),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 9),
    legend.position  = "bottom",
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title    = "Prediction uncertainty by method and sampling design",
    subtitle = paste0("Bootstrap SD (RF only) and kriging SE (OK, RK).",
                      " Median across all grid pixels.\n",
                      "NA = no convergent variogram model."),
    x        = NULL,
    y        = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty)

ggsave("../outputs/fig_uncertainty_by_method.png",
       p_uncertainty,
       width  = 12,
       height = 7,
       dpi    = 300)

#RK uncertainty:
# ── RK total uncertainty = RF trend SD + RK kriging SE ────
# Assumes independence between trend and residual uncertainty
# Total SE = sqrt(boot_SD² + krig_SE²)

rk_total_uncertainty <- bind_rows(
  # SP
  boot_uncertainty_SP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           dataset  = "SP") %>%
    left_join(
      krig_var_SP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "SP"),
      by = c("x", "y", "variable", "dataset")
    ),

  # COMP
  boot_uncertainty_COMP %>%
    select(x, y, ends_with("_boot_sd")) %>%
    pivot_longer(-c(x, y),
                 names_to = "variable",
                 values_to = "boot_sd") %>%
    mutate(variable = str_remove(variable, "_boot_sd"),
           dataset  = "COMP") %>%
    left_join(
      krig_var_COMP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "COMP"),
      by = c("x", "y", "variable", "dataset")
    )
) %>%
  mutate(
    # Total RK uncertainty via quadrature
    # krig_var is variance so sqrt = SE
    rk_total_se = sqrt(boot_sd^2 + 
                       coalesce(krig_var, 0))
  )

# ── Summarise per variable and dataset ────────────────────
rk_uncertainty_summary <- rk_total_uncertainty %>%
  group_by(variable, dataset) %>%
  summarise(
    mean_rk_se   = round(mean(rk_total_se,
                               na.rm = TRUE), 3),
    median_rk_se = round(median(rk_total_se,
                                 na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(
    pred_method = "RK",
    metric_type = "Total SE (boot + krig)"
  )

print(rk_uncertainty_summary, n = 20)
write_csv(rk_uncertainty_summary,
          "../outputs/rk_total_uncertainty.csv")
#Taylor, J.R. (1997). An Introduction to Error Analysis: The Study of Uncertainties in Physical Measurements (2nd ed.). University Science Books. 

# ── Rebuild uncertainty_all with RK total SE added ────────
rk_krig_se <- krig_se_long %>%
  filter(pred_method == "RK") %>%
  mutate(metric_type = "Kriging SE (residual only)",
         pred_method_label = "RK (kriging SE only)")

rk_total_se <- rk_uncertainty_summary %>%
  mutate(
    median_unc        = median_rk_se,
    metric_type       = "Total SE (RF + kriging)",
    pred_method_label = "RK (total SE)"
  ) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable, levels = soil_var_labels),
    dataset  = factor(dataset,
                      levels = c("SP", "COMP"),
                      labels = c("Single-Point", "Composite"))
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc, pred_method_label)

uncertainty_all_v2 <- bind_rows(
  # RF bootstrap SD — fix dataset labels here
  rf_uncertainty %>%
    filter(metric_type == "Bootstrap SD") %>%
    mutate(
      pred_method_label = "RF",
      dataset = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      ),
      variable = recode(variable, !!!soil_var_labels),
      variable = factor(variable, levels = soil_var_labels)
    ),

  # OK kriging SE
  krig_se_long %>%
    filter(pred_method == "OK") %>%
    mutate(
      metric_type       = "Kriging SE",
      pred_method_label = "OK",
      variable          = recode(variable, !!!soil_var_labels),
      variable          = factor(variable,
                                  levels = soil_var_labels),
      dataset           = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ),

  # RK kriging SE — residual only
  rk_krig_se %>%
    mutate(
      metric_type       = "Kriging SE",
      variable          = recode(variable, !!!soil_var_labels),
      variable          = factor(variable,
                                  levels = soil_var_labels),
      dataset           = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ) %>%
    select(variable, dataset, pred_method,
           metric_type, median_unc, pred_method_label),

  # RK total SE — already has correct labels from earlier
  rk_total_se %>%
    mutate(metric_type = "Kriging SE")

) %>%
  mutate(
    pred_method_label = factor(
      pred_method_label,
      levels = c("RF",
                 "OK",
                 "RK (kriging SE only)",
                 "RK (total SE)")
    )
  )

# ── Plot ──────────────────────────────────────────────────
method_colors <- c(
  "RF"                   = "#8e44ad",  # purple
  "OK"                   = "#2980b9",  # blue
  "RK (kriging SE only)" = "#82e0aa",  # light green
  "RK (total SE)"        = "#1e8449"   # dark green
)

p_uncertainty_v2 <- uncertainty_all_v2 %>%
  filter(!is.na(median_unc), is.finite(median_unc)) %>%
  ggplot(aes(x       = variable,
             y       = median_unc,
             fill    = pred_method_label)) +
  geom_col(position = position_dodge(0.85),
           width     = 0.8) +
  scale_fill_manual(
    values = method_colors,
    name   = "Prediction method",
    labels = c(
      "RF"                   = "RF (bootstrap SD)",
      "OK"                   = "OK (kriging SE)",
      "RK (kriging SE only)" = "RK (residual kriging SE)",
      "RK (total SE)"        = "RK (total SE = RF + kriging)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    legend.position    = "bottom",
    legend.text        = element_text(size = 8),
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Prediction uncertainty by method and sampling design",
    subtitle = paste0(
      "Row 1: Bootstrap SD (RF only). ",
      "Row 2: Kriging SE — OK, RK residual component (light green),\n",
      "and RK total SE combining RF bootstrap SD and residual ",
      "kriging SE in quadrature (dark green). ",
      "Median across all grid pixels.\n ",
      "Missing bars = no convergent variogram model."
    ),
    x = NULL,
    y = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty_v2)

ggsave("../outputs/fig_uncertainty_by_method_v2.png",
       p_uncertainty_v2,
       width  = 13,
       height = 8,
       dpi    = 300)



# ── 1. RF uncertainty — QRF PI width (primary metric) ─────
rf_uncertainty_qrf <- uncertainty_summary %>%
  filter(method == "QRF 80% PI width") %>%
  mutate(
    pred_method       = "RF",
    metric_type       = "QRF 80% PI width",
    pred_method_label = "RF"
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc, pred_method_label)

# ── 2. RK total uncertainty using QRF PI width ────────────
# Total RK uncertainty = sqrt(QRF_PI² + krig_var)
# Using QRF PI as RF trend uncertainty instead of bootstrap SD
# since QRF PI is the more standard RF uncertainty metric

rk_total_uncertainty_qrf <- bind_rows(
  # SP
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "SP") %>%
    left_join(
      krig_var_SP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "SP"),
      by = c("x", "y", "variable", "dataset")
    ),

  # COMP
  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "COMP") %>%
    left_join(
      krig_var_COMP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "COMP"),
      by = c("x", "y", "variable", "dataset")
    )
) %>%
  mutate(
    rk_total_se = sqrt(qrf_pi^2 +
                       coalesce(krig_var, 0))
  )

# ── Summarise RK total SE ─────────────────────────────────
rk_uncertainty_summary_qrf <- rk_total_uncertainty_qrf %>%
  group_by(variable, dataset) %>%
  summarise(
    mean_rk_se   = round(mean(rk_total_se,   na.rm = TRUE), 3),
    median_rk_se = round(median(rk_total_se, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(
    pred_method = "RK",
    metric_type = "Total SE (QRF PI + kriging)"
  )

print(rk_uncertainty_summary_qrf, n = 20)
write_csv(rk_uncertainty_summary_qrf,
          "../outputs/rk_total_uncertainty_qrf.csv")

# ── 3. Rebuild RK objects ─────────────────────────────────
rk_krig_se <- krig_se_long %>%
  filter(pred_method == "RK") %>%
  mutate(
    metric_type       = "Kriging SE",
    pred_method_label = "RK (kriging SE only)"
  )

rk_total_se_qrf <- rk_uncertainty_summary_qrf %>%
  mutate(
    median_unc        = median_rk_se,
    metric_type       = "Kriging SE",
    pred_method_label = "RK (total SE)"
  ) %>%
  mutate(
    variable = recode(variable, !!!soil_var_labels),
    variable = factor(variable, levels = soil_var_labels),
    dataset  = factor(dataset,
                      levels = c("SP", "COMP"),
                      labels = c("Single-Point", "Composite"))
  ) %>%
  select(variable, dataset, pred_method,
         metric_type, median_unc, pred_method_label)

# ── 4. Combine all uncertainty metrics ────────────────────
uncertainty_all_v3 <- bind_rows(

  # RF QRF PI width
  rf_uncertainty_qrf %>%
    mutate(
      dataset  = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      ),
      variable = recode(variable, !!!soil_var_labels),
      variable = factor(variable, levels = soil_var_labels),
      metric_type = "QRF 80% PI width"
    ),

  # OK kriging SE
  krig_se_long %>%
    filter(pred_method == "OK") %>%
    mutate(
      metric_type       = "Kriging SE",
      pred_method_label = "OK",
      variable          = recode(variable, !!!soil_var_labels),
      variable          = factor(variable,
                                  levels = soil_var_labels),
      dataset           = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ),

  # RK kriging SE — residual only
  rk_krig_se %>%
    mutate(
      variable = recode(variable, !!!soil_var_labels),
      variable = factor(variable, levels = soil_var_labels),
      dataset  = factor(
        recode(dataset,
               "SP"   = "Single-Point",
               "COMP" = "Composite"),
        levels = c("Single-Point", "Composite")
      )
    ) %>%
    select(variable, dataset, pred_method,
           metric_type, median_unc, pred_method_label),

  # RK total SE — QRF PI + kriging
  rk_total_se_qrf

) %>%
  mutate(
    pred_method_label = factor(
      pred_method_label,
      levels = c("RF",
                 "OK",
                 "RK (kriging SE only)",
                 "RK (total SE)")
    ),
    metric_type = factor(
      metric_type,
      levels = c("QRF 80% PI width", "Kriging SE")
    )
  )

# ── 5. Plot ───────────────────────────────────────────────
method_colors <- c(
  "RF"                   = "#8e44ad",  # purple
  "OK"                   = "#2980b9",  # blue
  "RK (kriging SE only)" = "#82e0aa",  # light green
  "RK (total SE)"        = "#1e8449"   # dark green
)

p_uncertainty_v3 <- uncertainty_all_v3 %>%
  filter(!is.na(median_unc), is.finite(median_unc)) %>%
  ggplot(aes(x    = variable,
             y    = median_unc,
             fill = pred_method_label)) +
  geom_col(position = position_dodge(0.85),
           width     = 0.8) +
  scale_fill_manual(
    values = method_colors,
    name   = "Prediction method",
    labels = c(
      "RF"                   = "RF (QRF 80% PI width)",
      "OK"                   = "OK (kriging SE)",
      "RK (kriging SE only)" = "RK (residual kriging SE only)",
      "RK (total SE)"        = "RK (total SE = QRF PI + kriging SE)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    legend.position    = "bottom",
    legend.text        = element_text(size = 8),
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Prediction uncertainty by method and sampling design",
    subtitle = paste0(
      "Row 1: QRF 80% prediction interval width (RF only).\n",
      "Row 2: Kriging SE — OK (blue), RK residual kriging SE (light green),\n",
      "and RK total SE combining QRF PI width and residual ",
      "kriging SE in quadrature (dark green).\n",
      "Median across all grid pixels. ",
      "Missing bars = no convergent variogram model."
    ),
    x = NULL,
    y = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty_v3)

ggsave("../outputs/fig_uncertainty_by_method_v3.png",
       p_uncertainty_v3,
       width  = 13,
       height = 8,
       dpi    = 300)
# Convert QRF 80% PI width to approximate SE
# 80% PI = ±1.28 SD on each side = 2.56 SD total
# Therefore SE ≈ PI_width / 2.56

rk_total_uncertainty_corrected <- bind_rows(
  qrf_uncertainty_SP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "SP") %>%
    left_join(
      krig_var_SP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "SP"),
      by = c("x", "y", "variable", "dataset")
    ),

  qrf_uncertainty_COMP %>%
    select(x, y, ends_with("_pi_width")) %>%
    pivot_longer(-c(x, y),
                 names_to  = "variable",
                 values_to = "qrf_pi") %>%
    mutate(variable = str_remove(variable, "_pi_width"),
           dataset  = "COMP") %>%
    left_join(
      krig_var_COMP %>%
        select(x, y, ends_with("_rk_krig_var")) %>%
        pivot_longer(-c(x, y),
                     names_to  = "variable",
                     values_to = "krig_var") %>%
        mutate(variable = str_remove(variable,
                                      "_rk_krig_var"),
               dataset  = "COMP"),
      by = c("x", "y", "variable", "dataset")
    )
) %>%
  mutate(
    # Convert QRF PI width to approximate SE
    # 80% PI = z(0.90) - z(0.10) = 1.282 - (-1.282) = 2.564 SD
    qrf_se      = qrf_pi / 2.564,

    # Combine in quadrature on SE scale
    rk_total_se = sqrt(qrf_se^2 + coalesce(krig_var, 0)),

    # Convert back to PI scale for interpretability
    # multiply by 2.564 to report as 80% PI equivalent
    rk_total_pi = rk_total_se * 2.564
  )

# ── Summarise ─────────────────────────────────────────────
rk_uncertainty_summary_corrected <- rk_total_uncertainty_corrected %>%
  group_by(variable, dataset) %>%
  summarise(
    median_rk_total_se = round(median(rk_total_se, na.rm = TRUE), 3),
    median_rk_total_pi = round(median(rk_total_pi, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  mutate(pred_method = "RK")

print(rk_uncertainty_summary_corrected, n = 20)
write_csv(rk_uncertainty_summary_corrected,
          "../outputs/rk_total_uncertainty_corrected.csv")

# ── Option C revised: RF + RK QRF PI in top row ───────────
# RK also uses RF as trend component so QRF PI applies
# Panel 1: QRF 80% PI width — RF and RK
# Panel 2: Kriging SE — OK and RK residual only

# ── 1. QRF PI width — RF and RK ───────────────────────────
qrf_panel <- uncertainty_summary %>%
  filter(method == "QRF 80% PI width") %>%
  mutate(
    # RF QRF PI is the trend uncertainty for both RF and RK
    pred_method_label = "RF / RK trend (QRF PI)",
    metric_type       = "RF Trend Uncertainty\n(QRF 80% Prediction Interval Width)",
    variable          = recode(variable, !!!soil_var_labels),
    variable          = factor(variable, levels = soil_var_labels),
    dataset           = factor(
      recode(dataset,
             "SP"   = "Single-Point",
             "COMP" = "Composite"),
      levels = c("Single-Point", "Composite")
    )
  ) %>%
  select(variable, dataset, pred_method_label,
         metric_type, median_unc)

# ── 2. Kriging SE — OK and RK residual ────────────────────
krig_se_panel <- bind_rows(

  # OK kriging SE
  krig_se_long %>%
    filter(pred_method == "OK") %>%
    mutate(pred_method_label = "OK (kriging SE)"),

  # RK residual kriging SE
  krig_se_long %>%
    filter(pred_method == "RK") %>%
    mutate(pred_method_label = "RK (residual kriging SE)")

) %>%
  mutate(
    metric_type = "Geostatistical Uncertainty\n(Kriging Standard Error)",
    variable    = recode(variable, !!!soil_var_labels),
    variable    = factor(variable, levels = soil_var_labels),
    dataset     = factor(
      recode(dataset,
             "SP"   = "Single-Point",
             "COMP" = "Composite"),
      levels = c("Single-Point", "Composite")
    )
  ) %>%
  select(variable, dataset, pred_method_label,
         metric_type, median_unc)

# ── 3. Combine ────────────────────────────────────────────
uncertainty_option_c2 <- bind_rows(
  qrf_panel,
  krig_se_panel
) %>%
  mutate(
    pred_method_label = factor(
      pred_method_label,
      levels = c(
        "RF / RK trend (QRF PI)",
        "OK (kriging SE)",
        "RK (residual kriging SE)"
      )
    ),
    metric_type = factor(
      metric_type,
      levels = c(
        "RF Trend Uncertainty\n(QRF 80% Prediction Interval Width)",
        "Geostatistical Uncertainty\n(Kriging Standard Error)"
      )
    )
  )

# ── 4. Colors ─────────────────────────────────────────────
method_colors_c2 <- c(
  "RF / RK trend (QRF PI)"   = "#8e44ad",  # purple
  "OK (kriging SE)"          = "#2980b9",  # blue
  "RK (residual kriging SE)" = "#27ae60"   # green
)

# ── 5. Plot ───────────────────────────────────────────────
p_uncertainty_c2 <- uncertainty_option_c2 %>%
  filter(!is.na(median_unc), is.finite(median_unc)) %>%
  ggplot(aes(x    = variable,
             y    = median_unc,
             fill = pred_method_label)) +
  geom_col(position = position_dodge(0.8),
           width     = 0.75) +
  scale_fill_manual(
    values = method_colors_c2,
    name   = "Prediction method / uncertainty component",
    labels = c(
      "RF / RK trend (QRF PI)"   = "RF and RK trend (QRF 80% PI width)",
      "OK (kriging SE)"          = "OK (kriging SE)",
      "RK (residual kriging SE)" = "RK (residual kriging SE)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    strip.text.y       = element_text(size = 8),
    legend.position    = "bottom",
    legend.text        = element_text(size = 9),
    legend.key.size    = unit(0.45, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(
    title    = "Spatial prediction uncertainty by method and sampling design",
    subtitle = paste0(
      "Top row: QRF 80% PI width — applies to RF predictions and ",
      "the RF trend component of RK (Meinshausen, 2006; ",
      "Vaysse & Lagacherie, 2017).\n",
      "Bottom row: Kriging SE — OK full kriging SE (blue) and ",
      "RK residual kriging SE after RF trend removal (green).\n",
      "Note: rows are not directly comparable — QRF PI captures ",
      "conditional distributional uncertainty; kriging SE captures ",
      "geometric data configuration uncertainty.\n",
      "Missing bars = no convergent variogram model. ",
      "Median across all grid pixels."
    ),
    x = NULL,
    y = "Median uncertainty (ppm; pH = dimensionless)"
  )

print(p_uncertainty_c2)

ggsave("../outputs/fig_uncertainty_option_c2.png",
       p_uncertainty_c2,
       width  = 13,
       height = 8,
       dpi    = 300)


# ── Compute mean per variable per dataset for scaling ─────
means_SP <- soil_prop_SP %>%
  st_drop_geometry() %>%
  select(all_of(soil_vars)) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "mean_val") %>%
  mutate(dataset = "Single-Point")

means_COMP <- soil_prop_COMP %>%
  st_drop_geometry() %>%
  select(all_of(soil_vars)) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  pivot_longer(everything(),
               names_to  = "variable",
               values_to = "mean_val") %>%
  mutate(dataset = "Composite")

means_all <- bind_rows(means_SP, means_COMP) %>%
  mutate(variable = recode(variable, !!!soil_var_labels),
         variable = factor(variable, levels = soil_var_labels))

# ── Scale uncertainty as % of mean ────────────────────────
uncertainty_scaled <- uncertainty_option_c2 %>%
  left_join(means_all,
            by = c("variable", "dataset")) %>%
  mutate(
    uncertainty_pct = round(
      (median_unc / mean_val) * 100, 1)
  )

# ── Plot scaled version ───────────────────────────────────
p_uncertainty_scaled <- uncertainty_scaled %>%
  filter(!is.na(uncertainty_pct),
         is.finite(uncertainty_pct)) %>%
  ggplot(aes(x    = variable,
             y    = uncertainty_pct,
             fill = pred_method_label)) +
  geom_col(position = position_dodge(0.8),
           width     = 0.75) +
  scale_fill_manual(
    values = method_colors_c2,
    name   = "Prediction method / uncertainty component",
    labels = c(
      "RF / RK trend (QRF PI)"   = "RF and RK trend (QRF 80% PI width)",
      "OK (kriging SE)"          = "OK (kriging SE)",
      "RK (residual kriging SE)" = "RK (residual kriging SE)"
    )
  ) +
  facet_grid(metric_type ~ dataset, scales = "free_y") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x        = element_text(angle = 35, hjust = 1,
                                       size = 8),
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(face = "bold", size = 9),
    strip.text.y       = element_text(size = 8),
    legend.position    = "bottom",
    legend.text        = element_text(size = 9),
    legend.key.size    = unit(0.45, "cm"),
    panel.grid.major.x = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 1)) +
  labs(
    title    = "Spatial prediction uncertainty relative to property mean",
    subtitle = paste0(
      "Top row: QRF 80% PI width as % of mean — RF and RK trend uncertainty.\n",
      "Bottom row: Kriging SE as % of mean — OK and RK residual.\n",
      "Expressing uncertainty relative to the property mean allows ",
      "comparison across soil properties on different measurement scales.\n",
      "Note: rows are not directly comparable across uncertainty types. ",
      "Missing bars = no convergent variogram model."
    ),
    x = NULL,
    y = "Uncertainty as % of property mean"
  )

print(p_uncertainty_scaled)

ggsave("../outputs/fig_uncertainty_scaled.png",
       p_uncertainty_scaled,
       width  = 13,
       height = 8,
       dpi    = 300)
```


# ============================================================
# PHASE 6 — UMAP + HDBSCAN Management Zone Delineation
# ============================================================

```{r umap-hdbscan}
run_umap_hdbscan <- function(pred_df,
                              methods          = c("RF", "RK"),
                              soil_vars,
                              include_lags     = FALSE,
                              n_components     = 2,
                              min_dist         = 0.001,
                              n_neighbors      = 30,
                              minPts           = 50,
                              cluster_eps      = 0,
                              label            = "dataset") {

  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  if (include_lags) {
    lag_cols  <- paste0(col_names, "_lag")
    col_names <- c(col_names, lag_cols[lag_cols %in% names(pred_df)])
  }
  col_names <- col_names[col_names %in% names(pred_df)]

  mat           <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  mat_clean     <- mat[complete_rows, ]
  xy_clean      <- pred_df %>% select(x, y) %>% slice(which(complete_rows))

  cat(label, "— pixels:", nrow(mat_clean),
      "| cols:", ncol(mat_clean),
      "| lags:", include_lags, "\n")

  # Zero-variance guard
  col_ranges <- apply(mat_clean, 2, function(x)
                      diff(range(x, na.rm = TRUE)))
  zero_var   <- col_ranges < 1e-10

  if (any(zero_var)) {
    cat("Warning: dropping", sum(zero_var), "constant column(s):",
        paste(colnames(mat_clean)[zero_var], collapse = ", "), "\n")
    mat_clean <- mat_clean[, !zero_var, drop = FALSE]
  }

  if (ncol(mat_clean) < 2) {
    cat("ERROR: fewer than 2 valid columns for", label, "\n")
    return(NULL)
  }

  mat_scaled <- scale(mat_clean)
  mat_scaled[!is.finite(mat_scaled)] <- 0
  stopifnot("NA values remain" = !anyNA(mat_scaled))

  cat("Fitting UMAP...\n")
  umap_fit <- uwot::umap(
    mat_scaled,
    n_components = n_components,
    min_dist     = min_dist,
    n_neighbors  = n_neighbors,
    ret_model    = TRUE,
    seed         = 20260401,
    verbose      = TRUE
  )

  emb_df        <- as.data.frame(umap_fit$embedding)
  names(emb_df) <- paste0("UMAP", seq_len(n_components))

  hdbscan_sample_size <- min(nrow(emb_df), 10000)
  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df), hdbscan_sample_size))
  emb_sample <- emb_df[sample_idx, ]

  cat("Running HDBSCAN on subsample of", hdbscan_sample_size,
      "points...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts,
    cluster_selection_epsilon = cluster_eps
  )

  cat("Clusters on subsample:", max(hdb_fit$cluster),
      "| noise:", round(mean(hdb_fit$cluster == 0) * 100, 1), "%\n")

  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  cat("Full grid — zones:", max(full_clusters),
      "| noise:", round(mean(full_clusters == 0) * 100, 1), "%\n")

  bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard    = full_clusters,
           membership_prob = full_memb)
  ) %>%
    mutate(
      dataset      = label,
      spatial_lags = include_lags,
      zone         = if_else(cluster_hard == 0L, NA_integer_,
                             as.integer(cluster_hard))
    )
}

wrap_mz_result <- function(df) {
  if (is.data.frame(df)) {
    zones <- df$zone[!is.na(df$zone)]
    return(list(result_df   = df,
                n_zones     = length(unique(zones)),
                pct_noise   = mean(is.na(df$zone)) * 100,
                umap_fit    = NULL,
                hdbscan_fit = NULL))
  }
  zones <- df$result_df$zone[!is.na(df$result_df$zone)]
  df$n_zones   <- length(unique(zones))
  df$pct_noise <- mean(is.na(df$result_df$zone)) * 100
  df
}

# ── 10 UMAP+HDBSCAN runs ──────────────────────────────────
mz_SP_RF_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RF_nolags")

mz_SP_RF_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RF_lags")

mz_SP_RK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_RK_nolags")

mz_SP_RK_lags <- run_umap_hdbscan(
  pred_stack_SP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "SP_RK_lags")

mz_SP_OK_nolags <- run_umap_hdbscan(
  pred_stack_SP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "SP_OK_nolags")

mz_COMP_RF_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RF", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RF_nolags")

mz_COMP_RF_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RF", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RF_lags")

mz_COMP_RK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "RK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_RK_nolags")

mz_COMP_RK_lags <- run_umap_hdbscan(
  pred_stack_COMP_lag, methods = "RK", soil_vars = soil_vars,
  include_lags = TRUE, label = "COMP_RK_lags")

mz_COMP_OK_nolags <- run_umap_hdbscan(
  pred_stack_COMP, methods = "OK", soil_vars = soil_vars,
  include_lags = FALSE, label = "COMP_OK_nolags")

save.image("../outputs/workspace_phase6_complete.RData")
beep("mario")

mz_named_tuned <- list(
  SP_RF_nolags   = wrap_mz_result(mz_SP_RF_nolags),
  SP_RF_lags     = wrap_mz_result(mz_SP_RF_lags),
  SP_RK_nolags   = wrap_mz_result(mz_SP_RK_nolags),
  SP_RK_lags     = wrap_mz_result(mz_SP_RK_lags),
  SP_OK_nolags   = wrap_mz_result(mz_SP_OK_nolags),
  COMP_RF_nolags = wrap_mz_result(mz_COMP_RF_nolags),
  COMP_RF_lags   = wrap_mz_result(mz_COMP_RF_lags),
  COMP_RK_nolags = wrap_mz_result(mz_COMP_RK_nolags),
  COMP_RK_lags   = wrap_mz_result(mz_COMP_RK_lags),
  COMP_OK_nolags = wrap_mz_result(mz_COMP_OK_nolags)
)

purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})
```

# ============================================================
# PHASE 6c — HDBSCAN Retuning
# ============================================================

```{r hdbscan-retune}
retune_hdbscan <- function(mz_obj, minPts_new,
                            cluster_eps = 0,
                            hdbscan_n   = 10000,
                            label       = NULL) {

  emb_df       <- mz_obj$result_df %>% select(UMAP1, UMAP2)
  xy_clean     <- mz_obj$result_df %>% select(x, y)
  spatial_lags <- mz_obj$result_df$spatial_lags[1]
  lbl          <- if (!is.null(label)) label else
                    mz_obj$result_df$dataset[1]

  set.seed(20260401)
  sample_idx <- sort(sample(nrow(emb_df),
                             min(hdbscan_n, nrow(emb_df))))
  emb_sample <- emb_df[sample_idx, ]

  cat("Retuning:", lbl, "| minPts =", minPts_new, "...\n")
  hdb_fit <- dbscan::hdbscan(
    as.matrix(emb_sample),
    minPts                    = minPts_new,
    cluster_selection_epsilon = cluster_eps
  )

  nn_assign     <- FNN::get.knnx(
    data  = as.matrix(emb_sample),
    query = as.matrix(emb_df),
    k     = 1
  )
  full_clusters <- hdb_fit$cluster[nn_assign$nn.index[, 1]]
  full_memb     <- hdb_fit$membership_prob[nn_assign$nn.index[, 1]]

  result_df <- bind_cols(
    xy_clean, emb_df,
    tibble(cluster_hard = full_clusters,
           membership_prob = full_memb)
  ) %>%
    mutate(dataset      = lbl,
           spatial_lags = spatial_lags,
           zone         = if_else(cluster_hard == 0L,
                                  NA_integer_,
                                  as.integer(cluster_hard)))

  list(result_df   = result_df,
       hdbscan_fit = hdb_fit,
       n_zones     = length(unique(full_clusters[full_clusters != 0])),
       pct_noise   = mean(full_clusters == 0) * 100)
}

# ── Sensitivity scan ──────────────────────────────────────
minpts_candidates <- c(50, 100, 200, 400, 600, 800, 1000)

sensitivity_results <- purrr::map_dfr(minpts_candidates, function(mp) {
  purrr::map_dfr(names(mz_named_tuned), function(nm) {
    r <- retune_hdbscan(mz_named_tuned[[nm]], mp, label = nm)
    tibble(label = nm, minPts = mp,
           n_zones = r$n_zones, pct_noise = round(r$pct_noise, 1))
  })
})

sensitivity_results %>%
  select(label, minPts, n_zones) %>%
  pivot_wider(names_from = label, values_from = n_zones) %>%
  arrange(minPts) %>%
  print(width = Inf)

# Set minPts after reviewing sensitivity table
minpts_final <- list(
  SP_RF_nolags   = 200,
  SP_RF_lags     = 200,
  SP_RK_nolags   = 200,
  SP_RK_lags     = 200,
  SP_OK_nolags   = 200,
  COMP_RF_nolags = 200,
  COMP_RF_lags   = 200,
  COMP_RK_nolags = 200,
  COMP_RK_lags   = 200,
  COMP_OK_nolags = 200
)

mz_named_source <- mz_named_tuned  # snapshot before overwriting

mz_named_tuned <- purrr::imap(minpts_final, function(mp, nm) {
  retune_hdbscan(mz_named_source[[nm]], mp, label = nm)
})

purrr::imap(mz_named_tuned, function(x, nm) {
  cat(nm, "| zones:", x$n_zones,
      "| noise:", round(x$pct_noise, 1), "%\n")
})

save.image("../outputs/workspace_phase6c_retuned.RData")
beep("coin")
```

# ============================================================
# PHASE 6b — Post-processing + Zone Uncertainty Summary
# ============================================================

```{r post-processing}
post_process_zones <- function(result_df, zone_rast_crs = "EPSG:26917") {
  n_total <- nrow(result_df)
  n_noise <- sum(is.na(result_df$zone))
  cat("Noise pixels before fill:", n_noise,
      "(", round(100 * n_noise / n_total, 1), "% of field)\n")

  if (n_noise == n_total) {
    cat("WARNING: all pixels are noise — no zones found.\n")
    return(list(result_df_filled = result_df,
                zone_rast_raw    = NULL,
                zone_rast_smooth = NULL))
  }

  assigned  <- result_df %>% filter(!is.na(zone))
  noise_pts <- result_df %>% filter( is.na(zone))

  if (nrow(noise_pts) > 0) {
    nn_idx <- FNN::get.knnx(
      data  = assigned  %>% select(UMAP1, UMAP2) %>% as.matrix(),
      query = noise_pts %>% select(UMAP1, UMAP2) %>% as.matrix(),
      k     = 1
    )$nn.index
    noise_pts$zone            <- assigned$zone[nn_idx]
    noise_pts$membership_prob <- 0
    result_df <- bind_rows(assigned, noise_pts)
  }

  cat("Noise pixels after fill: 0\n")

  zone_rast        <- result_df %>%
    select(x, y, zone) %>%
    rast(type = "xyz", crs = zone_rast_crs)
  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)

  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("Pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "% of field)\n")

  list(result_df_filled = result_df,
       zone_rast_raw    = zone_rast,
       zone_rast_smooth = zone_rast_smooth)
}

pp_named <- purrr::imap(mz_named_tuned, function(x, nm) {
  cat("\n---", nm, "---\n")
  post_process_zones(x$result_df)
})

write_zone_raster <- function(pp_obj, path) {
  if (is.null(pp_obj$zone_rast_smooth)) {
    cat("Skipping", path, "— no zones\n")
    return(invisible(NULL))
  }
  writeRaster(pp_obj$zone_rast_smooth, path, overwrite = TRUE)
  cat("Saved:", path, "\n")
}

purrr::iwalk(pp_named, function(pp, nm) {
  write_zone_raster(pp, paste0("../outputs/zones_", nm, "_smooth.tif"))
})

# ── UMAP / spatial zone / confidence plots ────────────────
plot_umap_zones <- function(res, title) {
  res$result_df %>%
    ggplot(aes(UMAP1, UMAP2, color = factor(zone),
               alpha = membership_prob)) +
    geom_point(size = 0.4) +
    scale_alpha_continuous(range = c(0.2, 1)) +
    theme_minimal() +
    labs(title = title, color = "Zone",
         alpha = "Membership\nconfidence") +
    guides(color = guide_legend(
      override.aes = list(size = 3, alpha = 1)))
}

plot_spatial_zones <- function(res, title) {
  res$result_df %>%
    filter(!is.na(zone)) %>%
    ggplot(aes(x, y, fill = factor(zone))) +
    geom_raster() + coord_equal() + theme_minimal() +
    labs(title = title, fill = "Zone",
         x = "Easting", y = "Northing")
}

plot_confidence <- function(res, title) {
  res$result_df %>%
    ggplot(aes(x, y, fill = membership_prob)) +
    geom_raster() +
    scale_fill_viridis_c(option = "magma", direction = -1) +
    coord_equal() + theme_minimal() +
    labs(title = title, fill = "Confidence",
         x = "Easting", y = "Northing")
}

purrr::iwalk(mz_named_tuned, function(x, nm) {
  print(plot_umap_zones(x,    paste("UMAP —", nm)))
  print(plot_spatial_zones(x, paste("Zones —", nm)))
  print(plot_confidence(x,    paste("Confidence —", nm)))
})

# ── Zone assignment uncertainty summary per config ────────
# Summarises membership probability distribution per zone
# to flag zones with high proportions of low-confidence
# (noise-filled) pixels.

zone_membership_summary <- purrr::imap_dfr(mz_named_tuned,
  function(x, nm) {
    x$result_df %>%
      mutate(zone = if_else(is.na(zone),
                            as.integer(-1L), zone)) %>%
      group_by(zone) %>%
      summarise(
        n_pixels          = n(),
        mean_memb_prob    = round(mean(membership_prob,
                                       na.rm = TRUE), 3),
        pct_low_conf      = round(
          100 * mean(membership_prob < 0.3,
                     na.rm = TRUE), 1),
        .groups = "drop"
      ) %>%
      mutate(configuration = nm)
  }
)

print(zone_membership_summary, n = 40)
write_csv(zone_membership_summary,
          "../outputs/zone_membership_uncertainty.csv")
```
```{r}
# ── Print smoothed UMAP zone maps ─────────────────────────
# Uses zone_rast_smooth regenerated from result_df_filled
# avoids stale raster pointer issue

purrr::iwalk(pp_named, function(pp, nm) {

  if (is.null(pp$result_df_filled)) {
    cat(nm, "— skipped (NULL)\n")
    return(invisible(NULL))
  }

  tryCatch({

    # Regenerate smoothed raster from result_df_filled
    zone_raw <- pp$result_df_filled %>%
      select(x, y, zone) %>%
      filter(!is.na(zone)) %>%
      rast(type = "xyz", crs = "EPSG:26917")

    zone_smooth <- focal(zone_raw, w = 5,
                         fun    = "modal",
                         na.rm  = TRUE)

    zone_df_smooth <- as.data.frame(
      zone_smooth,
      xy = TRUE, na.rm = TRUE
    ) %>%
      rename(zone = 3) %>%
      filter(!is.na(zone))

    # Number of zones for title
    n_zones <- n_distinct(zone_df_smooth$zone)

    p_smooth <- ggplot(zone_df_smooth,
                       aes(x = x, y = y,
                           fill = factor(zone))) +
      geom_raster() +
      coord_equal() +
      scale_fill_brewer(palette = "Set2",
                        name    = "Zone") +
      theme_bw(base_size = 9) +
      theme(
        axis.text       = element_text(size = 6),
        legend.position = "right",
        plot.title      = element_text(size = 9,
                                        face = "bold")
      ) +
      labs(
        title    = paste0("UMAP + HDBSCAN (k = ",
                          n_zones, ") — ", nm),
        subtitle = "Post-processed: noise fill + 5×5 majority filter",
        x        = "Easting",
        y        = "Northing"
      )

    print(p_smooth)

  }, error = function(e) {
    cat(nm, "— error:", conditionMessage(e), "\n")
  })
})
```
```{r}
# ── Check UMAP embedding structure for 2-zone configs ─────
two_zone_configs <- c("COMP_RF_nolags", 
                       "COMP_RK_nolags",
                       "COMP_RK_lags",
                       "COMP_OK_nolags")

purrr::walk(two_zone_configs, function(nm) {
  
  emb_df <- mz_named_tuned[[nm]]$result_df %>%
    select(UMAP1, UMAP2, zone)
  
  p <- ggplot(emb_df,
              aes(x     = UMAP1,
                  y     = UMAP2,
                  color = factor(zone))) +
    geom_point(size = 0.3, alpha = 0.3) +
    scale_color_brewer(palette = "Set2",
                       name    = "Zone",
                       na.value = "grey70") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("UMAP embedding — ", nm),
      subtitle = paste0(
        "n zones = ",
        mz_named_tuned[[nm]]$n_zones,
        " | noise = ",
        round(mz_named_tuned[[nm]]$pct_noise, 1), "%"
      ),
      x = "UMAP1", y = "UMAP2"
    )
  
  print(p)
})

# ── Compare against a well-structured config ──────────────
p_good <- mz_named_tuned[["COMP_RF_lags"]]$result_df %>%
  select(UMAP1, UMAP2, zone) %>%
  ggplot(aes(x     = UMAP1,
             y     = UMAP2,
             color = factor(zone))) +
  geom_point(size = 0.3, alpha = 0.3) +
  scale_color_brewer(palette = "Set2",
                     name    = "Zone",
                     na.value = "grey70") +
  theme_bw(base_size = 9) +
  labs(
    title    = "UMAP embedding — COMP_RF_lags (reference)",
    subtitle = paste0(
      "n zones = ",
      mz_named_tuned[["COMP_RF_lags"]]$n_zones,
      " | noise = ",
      round(mz_named_tuned[["COMP_RF_lags"]]$pct_noise, 1),
      "%"
    ),
    x = "UMAP1", y = "UMAP2"
  )

print(p_good)

# Plot all 4 problematic configs side by side
purrr::walk(two_zone_configs, function(nm) {
  p <- mz_named_tuned[[nm]]$result_df %>%
    select(UMAP1, UMAP2, zone) %>%
    ggplot(aes(x     = UMAP1,
               y     = UMAP2,
               color = factor(zone))) +
    geom_point(size = 0.3, alpha = 0.3) +
    scale_color_brewer(palette = "Set2",
                       name     = "Zone",
                       na.value = "grey70") +
    theme_bw(base_size = 9) +
    labs(
      title    = paste0("UMAP embedding — ", nm),
      subtitle = paste0(
        "n zones = ", mz_named_tuned[[nm]]$n_zones,
        " | noise = ",
        round(mz_named_tuned[[nm]]$pct_noise, 1), "%"
      ),
      x = "UMAP1", y = "UMAP2"
    )
  print(p)
})

# Test lower minPts for COMP_RK_lags specifically
purrr::walk(c(50, 75, 100, 150), function(mp) {
  r <- retune_hdbscan(
    mz_named_source[["COMP_RK_lags"]], 
    mp, 
    label = paste0("COMP_RK_lags_mp", mp))
  cat("minPts =", mp, 
      "| zones:", r$n_zones,
      "| noise:", round(r$pct_noise, 1), "%\n")
})
```

# ============================================================
# PHASE 7 — Baseline Comparison: PCA + k-means
# ============================================================

```{r pca-kmeans-baseline}
run_pca_kmeans <- function(pred_df, methods = c("RF", "RK"),
                            soil_vars, k_range = 2:8,
                            label = "dataset") {
  col_names     <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names     <- col_names[col_names %in% names(pred_df)]
  mat           <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  xy_clean      <- pred_df %>% select(x, y) %>%
    slice(which(complete_rows))

  mat_scaled <- scale(mat[complete_rows, ])
  col_ranges <- apply(mat_scaled, 2, function(x)
                       diff(range(x, na.rm = TRUE)))
  const_cols <- !is.finite(mat_scaled[1, ]) | col_ranges < 1e-10
  if (any(const_cols)) {
    cat("Dropping", sum(const_cols), "constant column(s):",
        paste(colnames(mat_scaled)[const_cols], collapse = ", "), "\n")
    mat_scaled <- mat_scaled[, !const_cols, drop = FALSE]
  }
  mat_scaled[!is.finite(mat_scaled)] <- 0

  pca_out    <- prcomp(mat_scaled, center = FALSE, scale. = FALSE)
  cum_var    <- cumsum(pca_out$sdev^2) / sum(pca_out$sdev^2)
  n_comp     <- which(cum_var >= 0.95)[1]
  pca_scores <- pca_out$x[, seq_len(n_comp), drop = FALSE]

  set.seed(20260401)
  sil_idx <- sample(nrow(pca_scores), min(5000, nrow(pca_scores)))
  pca_sub <- pca_scores[sil_idx, ]

  sil_scores <- purrr::map_dbl(k_range, function(k) {
    set.seed(20260401)
    km  <- kmeans(pca_sub, centers = k,
                  nstart = 25, iter.max = 100)
    sil <- cluster::silhouette(km$cluster, dist(pca_sub))
    mean(sil[, 3])
  })
  best_k <- k_range[which.max(sil_scores)]
  cat(label, "— best k:", best_k, "\n")

  set.seed(20260401)
  km_best <- kmeans(pca_scores, centers = best_k,
                    nstart = 25, iter.max = 100)

  result_df <- bind_cols(
    xy_clean,
    tibble(zone_kmeans = km_best$cluster)
  ) %>% mutate(dataset = label, method = "PCA_kmeans")

  list(result_df   = result_df,
       pca         = pca_out,
       km          = km_best,
       sil_profile = tibble(k = k_range, silhouette = sil_scores),
       best_k      = best_k)
}

baseline_SP_RF        <- run_pca_kmeans(pred_stack_SP,   "RF", soil_vars, label = "SP_RF")
baseline_SP_RK        <- run_pca_kmeans(pred_stack_SP,   "RK", soil_vars, label = "SP_RK")
baseline_SP_OK        <- run_pca_kmeans(pred_stack_SP,   "OK", soil_vars, label = "SP_OK")
baseline_COMP_RF      <- run_pca_kmeans(pred_stack_COMP, "RF", soil_vars, label = "COMP_RF")
baseline_COMP_RK      <- run_pca_kmeans(pred_stack_COMP, "RK", soil_vars, label = "COMP_RK")
baseline_COMP_OK      <- run_pca_kmeans(pred_stack_COMP, "OK", soil_vars, label = "COMP_OK")
baseline_SP_RF_lags   <- run_pca_kmeans(pred_stack_SP_lag,   "RF", soil_vars, label = "SP_RF_lags_kmeans")
baseline_SP_RK_lags   <- run_pca_kmeans(pred_stack_SP_lag,   "RK", soil_vars, label = "SP_RK_lags_kmeans")
baseline_COMP_RF_lags <- run_pca_kmeans(pred_stack_COMP_lag, "RF", soil_vars, label = "COMP_RF_lags_kmeans")
baseline_COMP_RK_lags <- run_pca_kmeans(pred_stack_COMP_lag, "RK", soil_vars, label = "COMP_RK_lags_kmeans")

# Silhouette profiles
bind_rows(
  baseline_SP_RF$sil_profile        %>% mutate(config = "SP_RF"),
  baseline_SP_RK$sil_profile        %>% mutate(config = "SP_RK"),
  baseline_SP_OK$sil_profile        %>% mutate(config = "SP_OK"),
  baseline_COMP_RF$sil_profile      %>% mutate(config = "COMP_RF"),
  baseline_COMP_RK$sil_profile      %>% mutate(config = "COMP_RK"),
  baseline_COMP_OK$sil_profile      %>% mutate(config = "COMP_OK"),
  baseline_SP_RF_lags$sil_profile   %>% mutate(config = "SP_RF_lags"),
  baseline_SP_RK_lags$sil_profile   %>% mutate(config = "SP_RK_lags"),
  baseline_COMP_RF_lags$sil_profile %>% mutate(config = "COMP_RF_lags"),
  baseline_COMP_RK_lags$sil_profile %>% mutate(config = "COMP_RK_lags")
) %>%
  ggplot(aes(x = k, y = silhouette, color = config)) +
  geom_line() + geom_point() +
  theme_minimal() +
  labs(title = "PCA+k-means silhouette profiles",
       x = "k", y = "Mean silhouette")



#elbow plots:
# ── Elbow plot — within-cluster SS ────────────────────────
wcss_list <- list(
  SP_RF          = baseline_SP_RF,
  SP_RK          = baseline_SP_RK,
  SP_OK          = baseline_SP_OK,
  COMP_RF        = baseline_COMP_RF,
  COMP_RK        = baseline_COMP_RK,
  COMP_OK        = baseline_COMP_OK,
  SP_RF_lags     = baseline_SP_RF_lags,
  SP_RK_lags     = baseline_SP_RK_lags,
  COMP_RF_lags   = baseline_COMP_RF_lags,
  COMP_RK_lags   = baseline_COMP_RK_lags
)

# Recompute WCSS for each config across k range
wcss_results <- purrr::imap_dfr(wcss_list, function(b, nm) {
  
  # Get the same PCA scores used for clustering
  cum_var    <- cumsum(b$pca$sdev^2) / sum(b$pca$sdev^2)
  n_comp     <- which(cum_var >= 0.95)[1]
  pca_scores <- b$pca$x[, seq_len(n_comp), drop = FALSE]
  
  set.seed(20260401)
  sil_idx <- sample(nrow(pca_scores),
                    min(5000, nrow(pca_scores)))
  pca_sub <- pca_scores[sil_idx, ]
  
  purrr::map_dfr(2:15, function(k) {
    set.seed(20260401)
    km <- kmeans(pca_sub, centers = k,
                 nstart = 25, iter.max = 100)
    tibble(
      config = nm,
      k      = k,
      wcss   = km$tot.withinss
    )
  })
})

# ── Plot — same style as silhouette ───────────────────────
ggplot(wcss_results,
       aes(x = k, y = wcss,
           color = config)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Paired",
                     name    = "Configuration") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 8),
    legend.key.size  = unit(0.4, "cm")
  ) +
  labs(
    title    = "Elbow method — PCA + k-means",
    subtitle = paste0(
      "Within-cluster sum of squares across k = 2:8. ",
      "Look for elbow where rate of decrease flattens."
    ),
    x = "Number of zones (k)",
    y = "Total within-cluster SS"
  )

#This actually strengthens the argument for UMAP+HDBSCAN over PCA+kmeans — HDBSCAN found 4-7 zones without requiring k to be specified, and those zone counts fall within the elbow-suggested range of 6-8 while being more agronomically interpretable than the silhouette-optimal k=2. The zone counts identified by HDBSCAN (k=2-7 depending on configuration) were broadly consistent with elbow analysis of PCA+k-means within-cluster variance, suggesting that HDBSCAN's data-driven zone count selection produced agronomically reasonable partitions without requiring a priori specification of k.


library(inflection)

# ── Automated elbow detection via kneedle ─────────────────
elbow_k <- wcss_results %>%
  group_by(config) %>%
  summarise(
    # finds index of maximum curvature
    elbow_k = {
      x <- k
      y <- wcss
      # second difference method — finds where
      # rate of change decreases most rapidly
      d1 <- diff(y) / diff(x)        # first derivative
      d2 <- diff(d1) / diff(x[-1])   # second derivative
      # elbow = where second derivative is maximized
      # (most negative = sharpest bend)
      x[-c(1,2)][which.min(d2)]
    },
    .groups = "drop"
  )

print(elbow_k, n = 10)

# ── No package needed — maximum second difference ─────────
elbow_k_simple <- wcss_results %>%
  group_by(config) %>%
  arrange(k) %>%
  summarise(
    elbow_k = {
      y    <- wcss
      x    <- k
      # normalize to 0-1 to make comparable across configs
      y_norm <- (y - min(y)) / (max(y) - min(y))
      x_norm <- (x - min(x)) / (max(x) - min(x))
      # distance from each point to the line
      # connecting first and last point
      # (standard kneedle approach)
      n     <- length(x_norm)
      x1    <- x_norm[1];   y1 <- y_norm[1]
      x2    <- x_norm[n];   y2 <- y_norm[n]
      # perpendicular distance to line
      dist_to_line <- abs(
        (y2 - y1) * x_norm -
        (x2 - x1) * y_norm +
        x2 * y1 - y2 * x1
      ) / sqrt((y2 - y1)^2 + (x2 - x1)^2)
      # elbow = point furthest from the line
      x[which.max(dist_to_line)]
    },
    .groups = "drop"
  ) %>%
  arrange(config)

print(elbow_k_simple, n = 10)
```

```{r pca-kmeans-postprocess}
post_process_kmeans <- function(result_df,
                                 zone_rast_crs = "EPSG:26917") {
  n_total   <- nrow(result_df)
  zone_rast <- result_df %>%
    select(x, y, zone = zone_kmeans) %>%
    rast(type = "xyz", crs = zone_rast_crs)

  zone_rast_smooth <- focal(zone_rast, w = 5,
                             fun = "modal", na.rm = TRUE)
  changed <- sum(values(zone_rast) != values(zone_rast_smooth),
                 na.rm = TRUE)
  cat("PCA+kmeans: pixels reassigned by majority filter:", changed,
      "(", round(100 * changed / n_total, 1), "%)\n")

  list(result_df_filled = result_df %>% mutate(zone = zone_kmeans),
       zone_rast_raw    = zone_rast,
       zone_rast_smooth = zone_rast_smooth)
}

pp_baseline <- list(
  SP_RF          = post_process_kmeans(baseline_SP_RF$result_df),
  SP_RK          = post_process_kmeans(baseline_SP_RK$result_df),
  SP_OK          = post_process_kmeans(baseline_SP_OK$result_df),
  COMP_RF        = post_process_kmeans(baseline_COMP_RF$result_df),
  COMP_RK        = post_process_kmeans(baseline_COMP_RK$result_df),
  COMP_OK        = post_process_kmeans(baseline_COMP_OK$result_df),
  SP_RF_lags     = post_process_kmeans(baseline_SP_RF_lags$result_df),
  SP_RK_lags     = post_process_kmeans(baseline_SP_RK_lags$result_df),
  COMP_RF_lags   = post_process_kmeans(baseline_COMP_RF_lags$result_df),
  COMP_RK_lags   = post_process_kmeans(baseline_COMP_RK_lags$result_df)
)

purrr::iwalk(pp_baseline, function(pp, nm) {
  writeRaster(pp$zone_rast_smooth,
              paste0("../outputs/zones_", nm, "_kmeans_smooth.tif"),
              overwrite = TRUE)
  cat("Saved:", nm, "\n")
})
```

```{r}
# ── Simple spatial zone maps for all PCA+kmeans configs ───
purrr::iwalk(pp_baseline, function(pp, nm) {
  
  # Extract zone values from result_df directly
  # avoids the raster NULL issue
  zone_df <- pp$result_df_filled %>%
    select(x, y, zone) %>%
    filter(!is.na(zone))
  
  p <- ggplot(zone_df,
              aes(x = x, y = y,
                  fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    scale_fill_brewer(palette = "Set2",
                      name    = "Zone") +
    theme_bw(base_size = 10) +
    theme(
      axis.text        = element_text(size = 7),
      legend.position  = "right",
      strip.background = element_rect(fill = "grey92")
    ) +
    labs(
      title = paste0("PCA + k-means zones — ", nm),
      x     = "Easting",
      y     = "Northing"
    )
  
  print(p)
})

# ── PCA biplot overlay for best config ────────────────────
# Shows PCA vectors + zone assignments together
# Standard visualization in MZ papers

best_nm  <- "COMP_RF_lags"
best_b   <- baseline_COMP_RF_lags
best_pp  <- pp_baseline[[best_nm]]

# PCA scores + zone assignments
pca_df <- as.data.frame(best_b$pca$x[, 1:2]) %>%
  bind_cols(
    tibble(zone = factor(best_b$result_df$zone_kmeans))
  )

# PCA loadings for biplot vectors
loadings_df <- as.data.frame(
  best_b$pca$rotation[, 1:2]
) %>%
  rownames_to_column("variable") %>%
  mutate(
    # Clean variable names
    variable = str_remove(variable, "_RF|_RK|_OK"),
    variable = recode(variable, !!!soil_var_labels)
  ) %>%
  # Scale vectors for visibility
  mutate(
    PC1 = PC1 * max(abs(pca_df$PC1)) * 0.8,
    PC2 = PC2 * max(abs(pca_df$PC2)) * 0.8
  )

# ── Biplot ────────────────────────────────────────────────
p_biplot <- ggplot() +
  # Points colored by zone
  geom_point(data = pca_df,
             aes(x = PC1, y = PC2,
                 color = zone),
             size = 0.3, alpha = 0.3) +
  # Loading vectors
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0,
                   xend = PC1, yend = PC2),
               arrow     = arrow(length = unit(0.2, "cm"),
                                  type   = "closed"),
               color     = "grey20",
               linewidth = 0.5) +
  # Variable labels
  ggrepel::geom_text_repel(
    data        = loadings_df,
    aes(x = PC1, y = PC2, label = variable),
    size        = 2.8,
    color       = "grey10",
    fontface    = "bold",
    box.padding = 0.3,
    max.overlaps = 20
  ) +
  # Zone centroids
  stat_summary(data = pca_df,
               aes(x = PC1, y = PC2,
                   color = zone,
                   label = zone),
               fun.x   = mean,
               fun.y   = mean,
               geom    = "text",
               size    = 4,
               fontface = "bold") +
  scale_color_brewer(palette = "Set2",
                     name    = "Zone") +
  theme_bw(base_size = 10) +
  theme(legend.position = "right") +
  labs(
    title    = paste0("PCA biplot — ", best_nm),
    subtitle = paste0(
      "Points = grid pixels colored by zone. ",
      "Vectors = soil property loadings on PC1 and PC2."
    ),
    x = paste0("PC1 (",
               round(summary(best_b$pca)$importance[2, 1] * 100, 1),
               "% variance)"),
    y = paste0("PC2 (",
               round(summary(best_b$pca)$importance[2, 2] * 100, 1),
               "% variance)")
  )

print(p_biplot)

# ── Side by side: spatial map + biplot ────────────────────
zone_df_best <- best_pp$result_df_filled %>%
  select(x, y, zone) %>%
  filter(!is.na(zone))

p_spatial_best <- ggplot(zone_df_best,
                          aes(x = x, y = y,
                              fill = factor(zone))) +
  geom_raster() +
  coord_equal() +
  scale_fill_brewer(palette = "Set2",
                    name    = "Zone") +
  theme_bw(base_size = 10) +
  theme(legend.position = "right") +
  labs(
    title = "Spatial zone map — COMP RF + lags",
    x     = "Easting",
    y     = "Northing"
  )

p_pca_final <- p_spatial_best + p_biplot +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title    = "PCA + k-means management zone delineation",
    subtitle = paste0(
      "Left: spatial zone map. ",
      "Right: PCA biplot showing zone separation ",
      "and soil property loading vectors."
    ),
    theme = theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, color = "grey30")
    )
  )

print(p_pca_final)

ggsave("../outputs/fig_pca_zones_biplot.png",
       p_pca_final,
       width  = 12,
       height = 5,
       dpi    = 300)


#do for comp_OK as well
# ── COMP_OK biplot + spatial ───────────────────────────────
best_nm_ok  <- "COMP_OK"
best_b_ok   <- baseline_COMP_OK
best_pp_ok  <- pp_baseline[["COMP_OK"]]

# PCA scores + zone assignments
pca_df_ok <- as.data.frame(best_b_ok$pca$x[, 1:2]) %>%
  bind_cols(
    tibble(zone = factor(best_b_ok$result_df$zone_kmeans))
  )

# PCA loadings
loadings_df_ok <- as.data.frame(
  best_b_ok$pca$rotation[, 1:2]
) %>%
  rownames_to_column("variable") %>%
  mutate(
    variable = str_remove(variable, "_RF|_RK|_OK"),
    variable = recode(variable, !!!soil_var_labels)
  ) %>%
  mutate(
    PC1 = PC1 * max(abs(pca_df_ok$PC1)) * 0.8,
    PC2 = PC2 * max(abs(pca_df_ok$PC2)) * 0.8
  )

# ── Biplot ────────────────────────────────────────────────
p_biplot_ok <- ggplot() +
  geom_point(data  = pca_df_ok,
             aes(x = PC1, y = PC2,
                 color = zone),
             size  = 0.3, alpha = 0.3) +
  geom_segment(data = loadings_df_ok,
               aes(x = 0, y = 0,
                   xend = PC1, yend = PC2),
               arrow     = arrow(
                 length = unit(0.2, "cm"),
                 type   = "closed"),
               color     = "grey20",
               linewidth = 0.5) +
  ggrepel::geom_text_repel(
    data         = loadings_df_ok,
    aes(x = PC1, y = PC2, label = variable),
    size         = 2.8,
    color        = "grey10",
    fontface     = "bold",
    box.padding  = 0.3,
    max.overlaps = 20
  ) +
  scale_color_brewer(palette = "Set2",
                     name    = "Zone") +
  theme_bw(base_size = 10) +
  theme(legend.position = "right") +
  labs(
    title    = "PCA biplot — COMP OK",
    subtitle = paste0(
      "Points = grid pixels colored by zone. ",
      "Vectors = soil property loadings."
    ),
    x = paste0("PC1 (",
               round(summary(best_b_ok$pca)$importance[2, 1] * 100, 1),
               "% variance)"),
    y = paste0("PC2 (",
               round(summary(best_b_ok$pca)$importance[2, 2] * 100, 1),
               "% variance)")
  )

# ── Spatial map ───────────────────────────────────────────
zone_df_ok <- best_pp_ok$result_df_filled %>%
  select(x, y, zone) %>%
  filter(!is.na(zone))

p_spatial_ok <- ggplot(zone_df_ok,
                        aes(x = x, y = y,
                            fill = factor(zone))) +
  geom_raster() +
  coord_equal() +
  scale_fill_brewer(palette = "Set2",
                    name    = "Zone") +
  theme_bw(base_size = 10) +
  theme(legend.position = "right") +
  labs(
    title = "Spatial zone map — COMP OK",
    x     = "Easting",
    y     = "Northing"
  )

# ── Combined ──────────────────────────────────────────────
p_ok_final <- p_spatial_ok + p_biplot_ok +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title    = "PCA + k-means — COMP OK",
    subtitle = paste0(
      "k = ", best_b_ok$best_k, " zones. ",
      "Left: spatial map. Right: PCA biplot."
    ),
    theme = theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9, color = "grey30")
    )
  )

print(p_ok_final)

ggsave("../outputs/fig_pca_zones_COMP_OK.png",
       p_ok_final,
       width  = 12,
       height = 5,
       dpi    = 300)

```
```{r}
#non silhouette method:

# ── Step 1: Compute WCSS and detect elbow k ───────────────
# Run this first to get elbow_k values

run_pca_kmeans_wcss <- function(pred_df, 
                                 methods   = c("RF", "RK"),
                                 soil_vars,
                                 k_range   = 2:15,
                                 label     = "dataset") {
  col_names     <- as.vector(outer(soil_vars, methods,
                                    paste, sep = "_"))
  col_names     <- col_names[col_names %in% names(pred_df)]
  mat           <- pred_df %>%
    select(all_of(col_names)) %>% as.matrix()
  complete_rows <- complete.cases(mat)
  xy_clean      <- pred_df %>%
    select(x, y) %>% slice(which(complete_rows))

  mat_scaled <- scale(mat[complete_rows, ])
  col_ranges <- apply(mat_scaled, 2, function(x)
    diff(range(x, na.rm = TRUE)))
  const_cols <- !is.finite(mat_scaled[1, ]) |
    col_ranges < 1e-10
  if (any(const_cols)) {
    mat_scaled <- mat_scaled[, !const_cols, drop = FALSE]
  }
  mat_scaled[!is.finite(mat_scaled)] <- 0

  pca_out    <- prcomp(mat_scaled,
                        center = FALSE, scale. = FALSE)
  cum_var    <- cumsum(pca_out$sdev^2) /
    sum(pca_out$sdev^2)
  n_comp     <- which(cum_var >= 0.95)[1]
  pca_scores <- pca_out$x[, seq_len(n_comp), drop = FALSE]

  set.seed(20260401)
  sil_idx <- sample(nrow(pca_scores),
                    min(5000, nrow(pca_scores)))
  pca_sub <- pca_scores[sil_idx, ]

  # ── Compute WCSS across k range ───────────────────────
  wcss_vals <- purrr::map_dbl(k_range, function(k) {
    set.seed(20260401)
    km <- kmeans(pca_sub, centers = k,
                 nstart = 25, iter.max = 100)
    km$tot.withinss
  })

  # ── Kneedle algorithm ─────────────────────────────────
  x      <- k_range
  y      <- wcss_vals
  x_norm <- (x - min(x)) / (max(x) - min(x))
  y_norm <- (y - min(y)) / (max(y) - min(y))
  n      <- length(x_norm)
  x1 <- x_norm[1]; y1 <- y_norm[1]
  x2 <- x_norm[n]; y2 <- y_norm[n]
  dist_to_line <- abs(
    (y2 - y1) * x_norm -
      (x2 - x1) * y_norm +
      x2 * y1 - y2 * x1
  ) / sqrt((y2 - y1)^2 + (x2 - x1)^2)
  best_k <- x[which.max(dist_to_line)]

  cat(label, "— elbow k:", best_k, "\n")

  # ── Fit final kmeans with elbow k ─────────────────────
  set.seed(20260401)
  km_best <- kmeans(pca_scores,
                    centers  = best_k,
                    nstart   = 25,
                    iter.max = 100)

  result_df <- bind_cols(
    xy_clean,
    tibble(zone_kmeans = km_best$cluster)
  ) %>% mutate(dataset = label, method = "PCA_kmeans")

  list(
    result_df    = result_df,
    pca          = pca_out,
    km           = km_best,
    wcss_profile = tibble(k = k_range,
                           wcss = wcss_vals),
    best_k       = best_k,
    elbow_k      = best_k
  )
}

# ── Step 2: Rerun all 10 configurations ───────────────────
cat("Rerunning PCA+kmeans with elbow-detected k...\n")

baseline_SP_RF_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP,   "RF", soil_vars,
  label = "SP_RF")

baseline_SP_RK_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP,   "RK", soil_vars,
  label = "SP_RK")

baseline_SP_OK_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP,   "OK", soil_vars,
  label = "SP_OK")

baseline_COMP_RF_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP, "RF", soil_vars,
  label = "COMP_RF")

baseline_COMP_RK_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP, "RK", soil_vars,
  label = "COMP_RK")

baseline_COMP_OK_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP, "OK", soil_vars,
  label = "COMP_OK")

baseline_SP_RF_lags_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP_lag,   "RF", soil_vars,
  label = "SP_RF_lags_kmeans")

baseline_SP_RK_lags_elbow   <- run_pca_kmeans_wcss(
  pred_stack_SP_lag,   "RK", soil_vars,
  label = "SP_RK_lags_kmeans")

baseline_COMP_RF_lags_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP_lag, "RF", soil_vars,
  label = "COMP_RF_lags_kmeans")

baseline_COMP_RK_lags_elbow <- run_pca_kmeans_wcss(
  pred_stack_COMP_lag, "RK", soil_vars,
  label = "COMP_RK_lags_kmeans")

beep("coin")

# ── Step 3: Print elbow k summary ─────────────────────────
elbow_summary <- tibble(
  config  = c("SP_RF", "SP_RK", "SP_OK",
              "COMP_RF", "COMP_RK", "COMP_OK",
              "SP_RF_lags", "SP_RK_lags",
              "COMP_RF_lags", "COMP_RK_lags"),
  elbow_k = c(
    baseline_SP_RF_elbow$best_k,
    baseline_SP_RK_elbow$best_k,
    baseline_SP_OK_elbow$best_k,
    baseline_COMP_RF_elbow$best_k,
    baseline_COMP_RK_elbow$best_k,
    baseline_COMP_OK_elbow$best_k,
    baseline_SP_RF_lags_elbow$best_k,
    baseline_SP_RK_lags_elbow$best_k,
    baseline_COMP_RF_lags_elbow$best_k,
    baseline_COMP_RK_lags_elbow$best_k
  )
)

print(elbow_summary)

# ── Step 4: Post-process elbow zones ──────────────────────
pp_baseline_elbow <- list(
  SP_RF        = post_process_kmeans(
    baseline_SP_RF_elbow$result_df),
  SP_RK        = post_process_kmeans(
    baseline_SP_RK_elbow$result_df),
  SP_OK        = post_process_kmeans(
    baseline_SP_OK_elbow$result_df),
  COMP_RF      = post_process_kmeans(
    baseline_COMP_RF_elbow$result_df),
  COMP_RK      = post_process_kmeans(
    baseline_COMP_RK_elbow$result_df),
  COMP_OK      = post_process_kmeans(
    baseline_COMP_OK_elbow$result_df),
  SP_RF_lags   = post_process_kmeans(
    baseline_SP_RF_lags_elbow$result_df),
  SP_RK_lags   = post_process_kmeans(
    baseline_SP_RK_lags_elbow$result_df),
  COMP_RF_lags = post_process_kmeans(
    baseline_COMP_RF_lags_elbow$result_df),
  COMP_RK_lags = post_process_kmeans(
    baseline_COMP_RK_lags_elbow$result_df)
)

# ── Step 5: Save rasters ───────────────────────────────────
purrr::iwalk(pp_baseline_elbow, function(pp, nm) {
  writeRaster(
    pp$zone_rast_smooth,
    paste0("../outputs/zones_", nm,
           "_kmeans_elbow_smooth.tif"),
    overwrite = TRUE)
  cat("Saved:", nm, "\n")
})

# ── Step 6: Quick zone maps ────────────────────────────────
purrr::iwalk(pp_baseline_elbow, function(pp, nm) {
  zone_df <- pp$result_df_filled %>%
    select(x, y, zone) %>%
    filter(!is.na(zone))

  p <- ggplot(zone_df,
              aes(x = x, y = y,
                  fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    scale_fill_brewer(palette = "Set2",
                      name    = "Zone") +
    theme_bw(base_size = 10) +
    theme(
      axis.text       = element_text(size = 7),
      legend.position = "right"
    ) +
    labs(
      title = paste0("PCA + k-means (elbow k) — ", nm),
      x     = "Easting",
      y     = "Northing"
    )

  print(p)
})

# ── Step 7: WCSS elbow plots ───────────────────────────────
bind_rows(
  baseline_SP_RF_elbow$wcss_profile        %>%
    mutate(config = "SP_RF"),
  baseline_SP_RK_elbow$wcss_profile        %>%
    mutate(config = "SP_RK"),
  baseline_SP_OK_elbow$wcss_profile        %>%
    mutate(config = "SP_OK"),
  baseline_COMP_RF_elbow$wcss_profile      %>%
    mutate(config = "COMP_RF"),
  baseline_COMP_RK_elbow$wcss_profile      %>%
    mutate(config = "COMP_RK"),
  baseline_COMP_OK_elbow$wcss_profile      %>%
    mutate(config = "COMP_OK"),
  baseline_SP_RF_lags_elbow$wcss_profile   %>%
    mutate(config = "SP_RF_lags"),
  baseline_SP_RK_lags_elbow$wcss_profile   %>%
    mutate(config = "SP_RK_lags"),
  baseline_COMP_RF_lags_elbow$wcss_profile %>%
    mutate(config = "COMP_RF_lags"),
  baseline_COMP_RK_lags_elbow$wcss_profile %>%
    mutate(config = "COMP_RK_lags")
) %>%
  left_join(elbow_summary, by = "config") %>%
  ggplot(aes(x = k, y = wcss, color = config)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  geom_vline(data = elbow_summary,
             aes(xintercept = elbow_k,
                 color      = config),
             linetype  = "dashed",
             linewidth = 0.4,
             alpha     = 0.7) +
  scale_color_brewer(palette = "Paired",
                     name    = "Configuration") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "right",
    legend.text     = element_text(size = 8),
    legend.key.size = unit(0.4, "cm")
  ) +
  labs(
    title    = "Elbow method — PCA + k-means (kneedle detection)",
    subtitle = paste0(
      "Dashed vertical lines = automatically detected elbow k ",
      "per configuration (Satopaa et al., 2011)."
    ),
    x = "Number of zones (k)",
    y = "Total within-cluster SS"
  )

ggsave("../outputs/fig_elbow_wcss.png",
       width = 10, height = 6, dpi = 300)

save.image("../outputs/workspace_pca_elbow.RData")
beep("complete")


# ── Plot POST-PROCESSED smoothed zones ────────────────────
# Uses zone_rast_smooth — after majority filter
# Equivalent to what UMAP plots show

purrr::iwalk(pp_baseline_elbow, function(pp, nm) {

  # Convert smoothed raster to data frame for plotting
  zone_df_smooth <- as.data.frame(
    pp$zone_rast_smooth,
    xy = TRUE, na.rm = TRUE
  ) %>%
    rename(zone = 3)

  # Raw (pre-filter) from result_df
  zone_df_raw <- pp$result_df_filled %>%
    select(x, y, zone) %>%
    filter(!is.na(zone))

  # ── Raw zones ───────────────────────────────────────────
  p_raw <- ggplot(zone_df_raw,
                  aes(x = x, y = y,
                      fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    scale_fill_brewer(palette = "Set2",
                      name    = "Zone") +
    theme_bw(base_size = 9) +
    theme(
      axis.text       = element_text(size = 6),
      legend.position = "right",
      plot.title      = element_text(size = 9,
                                      face = "bold")
    ) +
    labs(
      title = paste0("Raw — ", nm),
      x     = "Easting",
      y     = "Northing"
    )

  # ── Smoothed zones ──────────────────────────────────────
  p_smooth <- ggplot(zone_df_smooth,
                     aes(x = x, y = y,
                         fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    scale_fill_brewer(palette = "Set2",
                      name    = "Zone") +
    theme_bw(base_size = 9) +
    theme(
      axis.text       = element_text(size = 6),
      legend.position = "right",
      plot.title      = element_text(size = 9,
                                      face = "bold")
    ) +
    labs(
      title = paste0("Smoothed (5×5 modal) — ", nm),
      x     = "Easting",
      y     = "Northing"
    )

  # ── Side by side ────────────────────────────────────────
  p_combined <- p_raw + p_smooth +
    plot_layout(ncol = 2, guides = "collect") +
    plot_annotation(
      title    = paste0("PCA + k-means post-processing — ",
                        nm),
      subtitle = paste0(
        "Left: raw k-means zones. ",
        "Right: after 5×5 majority filter. ",
        "k = ", pp$result_df_filled %>%
          pull(zone) %>% n_distinct(), " zones."
      ),
      theme = theme(
        plot.title    = element_text(face = "bold",
                                      size = 10),
        plot.subtitle = element_text(size = 8,
                                      color = "grey30")
      )
    )

  print(p_combined)

  # Save each config
  ggsave(
    paste0("../outputs/fig_pca_elbow_postprocess_",
           nm, ".png"),
    p_combined,
    width  = 10,
    height = 4.5,
    dpi    = 300
  )

  cat("Plotted:", nm, "\n")
})

# ── Best config side by side with UMAP ────────────────────
# Most useful thesis figure — direct visual comparison
# after post-processing for both methods

# COMP RF lags — best configuration
comp_rf_lags_smooth <- as.data.frame(
  pp_baseline_elbow[["COMP_RF_lags"]]$zone_rast_smooth,
  xy = TRUE, na.rm = TRUE
) %>% rename(zone = 3)

umap_comp_rf_lags <- pp_named[["COMP_RF_lags"]]$result_df_filled %>%
  select(x, y, zone) %>%
  filter(!is.na(zone))

p_pca_best <- ggplot(comp_rf_lags_smooth,
                      aes(x = x, y = y,
                          fill = factor(zone))) +
  geom_raster() +
  coord_equal() +
  scale_fill_brewer(palette = "Set2", name = "Zone") +
  theme_bw(base_size = 9) +
  theme(legend.position = "right") +
  labs(
    title    = paste0("PCA + k-means (elbow k = ",
                      pp_baseline_elbow[["COMP_RF_lags"]]$result_df_filled %>%
                        pull(zone) %>% n_distinct(),
                      ") — COMP RF + lags"),
    subtitle = "After 5×5 majority filter",
    x = "Easting", y = "Northing"
  )

p_umap_best <- ggplot(umap_comp_rf_lags,
                       aes(x = x, y = y,
                           fill = factor(zone))) +
  geom_raster() +
  coord_equal() +
  scale_fill_brewer(palette = "Set2", name = "Zone") +
  theme_bw(base_size = 9) +
  theme(legend.position = "right") +
  labs(
    title    = paste0("UMAP + HDBSCAN (k = ",
                      umap_comp_rf_lags %>%
                        pull(zone) %>%
                        n_distinct(),
                      ") — COMP RF + lags"),
    subtitle = "After noise fill + 5×5 majority filter",
    x = "Easting", y = "Northing"
  )

p_method_comparison <- p_pca_best + p_umap_best +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title    = "Zone map comparison after post-processing",
    subtitle = paste0(
      "Best configuration: COMP RF + spatial lags. ",
      "Left: PCA + k-means (elbow k). ",
      "Right: UMAP + HDBSCAN."
    ),
    theme = theme(
      plot.title    = element_text(face = "bold", size = 11),
      plot.subtitle = element_text(size = 9,
                                    color = "grey30")
    )
  )

print(p_method_comparison)
#plot just all of the smoothed ones:


ggsave("../outputs/fig_method_comparison_postprocess.png",
       p_method_comparison,
       width  = 10,
       height = 5,
       dpi    = 300)



# ── Print smoothed zones only ─────────────────────────────
purrr::iwalk(pp_baseline_elbow, function(pp, nm) {

  zone_df_smooth <- as.data.frame(
    pp$zone_rast_smooth,
    xy = TRUE, na.rm = TRUE
  ) %>%
    rename(zone = 3) %>%
  filter(!is.na(zone)) 

  p_smooth <- ggplot(zone_df_smooth,
                     aes(x = x, y = y,
                         fill = factor(zone))) +
    geom_raster() +
    coord_equal() +
    scale_fill_brewer(palette = "Set2",
                      name    = "Zone") +
    theme_bw(base_size = 9) +
    theme(
      axis.text       = element_text(size = 6),
      legend.position = "right",
      plot.title      = element_text(size = 9,
                                      face = "bold")
    ) +
    labs(
      title    = paste0("PCA + k-means (elbow k = ",
                        n_distinct(zone_df_smooth$zone),
                        ") — ", nm),
      subtitle = "Post-processed: 5×5 majority filter applied",
      x        = "Easting",
      y        = "Northing"
    )

  print(p_smooth)
})
```


# ============================================================
# PHASE 8 — Zone Validation Metrics
# ============================================================

```{r zone-validation}
calc_variance_reduction <- function(pred_df, zone_vec,
                                     soil_vars, methods) {
  col_names  <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names  <- col_names[col_names %in% names(pred_df)]
  mat        <- pred_df %>% select(all_of(col_names)) %>% as.matrix()
  col_ranges <- apply(mat, 2, function(x)
                       diff(range(x, na.rm = TRUE)))
  mat        <- mat[, col_ranges > 1e-10, drop = FALSE]

  keep      <- !is.na(zone_vec) & complete.cases(mat)
  mat_k     <- mat[keep, ]
  zones_k   <- zone_vec[keep]
  total_var <- mean(apply(mat_k, 2, var, na.rm = TRUE))
  within_var <- mean(sapply(unique(zones_k), function(z) {
    mean(apply(mat_k[zones_k == z, , drop = FALSE],
               2, var, na.rm = TRUE))
  }))
  1 - (within_var / total_var)
}

calc_silhouette_sub <- function(emb_df, zone_vec, n_sub = 5000) {
  keep  <- !is.na(zone_vec)
  emb_k <- emb_df[keep, ]
  z_k   <- zone_vec[keep]
  set.seed(20260401)
  idx   <- sample(nrow(emb_k), min(n_sub, nrow(emb_k)))
  sil   <- cluster::silhouette(z_k[idx], dist(emb_k[idx, ]))
  mean(sil[, 3])
}

# UMAP VR + silhouette
zone_validation_umap <- purrr::imap_dfr(mz_named_tuned,
  function(x, nm) {
    parts  <- strsplit(nm, "_")[[1]]
    samp   <- parts[1]; method <- parts[2]
    lag_yn <- length(parts) >= 3 && parts[3] == "lags"

    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = pred_stack_SP,   SP_RK = pred_stack_SP,
      SP_OK   = pred_stack_SP,   COMP_RF = pred_stack_COMP,
      COMP_RK = pred_stack_COMP, COMP_OK = pred_stack_COMP)

    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = lag_yn,
      approach      = "UMAP+HDBSCAN",
      VR = calc_variance_reduction(
        pred_df, x$result_df$zone, soil_vars, method),
      silhouette = calc_silhouette_sub(
        x$result_df %>% select(UMAP1, UMAP2),
        x$result_df$zone)
    )
  })

# PCA+kmeans VR + silhouette
baseline_list <- list(
  SP_RF               = baseline_SP_RF,
  SP_RK               = baseline_SP_RK,
  SP_OK               = baseline_SP_OK,
  COMP_RF             = baseline_COMP_RF,
  COMP_RK             = baseline_COMP_RK,
  COMP_OK             = baseline_COMP_OK,
  SP_RF_lags_kmeans   = baseline_SP_RF_lags,
  SP_RK_lags_kmeans   = baseline_SP_RK_lags,
  COMP_RF_lags_kmeans = baseline_COMP_RF_lags,
  COMP_RK_lags_kmeans = baseline_COMP_RK_lags
)

zone_validation_pca <- purrr::imap_dfr(baseline_list,
  function(b, nm) {
    clean   <- str_remove(nm, "_kmeans")
    has_lag <- str_detect(clean, "_lags")
    clean   <- str_remove(clean, "_lags")
    parts   <- strsplit(clean, "_")[[1]]
    samp    <- parts[1]; method <- parts[2]

    pred_df <- switch(paste0(samp, "_", method),
      SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
      SP_OK   = pred_stack_SP,
      COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
      COMP_OK = pred_stack_COMP)

    ok_cols <- pred_df %>%
      select(any_of(paste0(soil_vars, "_", method))) %>% na.omit()
    col_vars <- apply(ok_cols, 2, var, na.rm = TRUE)
    ok_cols  <- ok_cols[, col_vars > 1e-10, drop = FALSE]

    tibble(
      configuration = nm, sampling = samp,
      pred_method   = method, lags = has_lag,
      approach      = "PCA+kmeans",
      VR = calc_variance_reduction(
        pred_df, b$result_df$zone_kmeans, soil_vars, method),
      silhouette = calc_silhouette_sub(
        prcomp(scale(ok_cols), center = FALSE)$x[, 1:2],
        b$result_df$zone_kmeans)
    )
  })

zone_validation_all <- bind_rows(zone_validation_umap,
                                   zone_validation_pca) %>%
  arrange(sampling, pred_method, lags, approach)

print(zone_validation_all)
write_csv(zone_validation_all,
          "../outputs/zone_validation_metrics.csv")
```

# ============================================================
# PHASE 8b — Agronomic Validation (held-out ap_val_sf)
# ============================================================

```{r agronomic-validation}
# Load validate_zones_with_holdout from Chapter 1 workspace
validate_zones_with_holdout <- function(zone_rast_smooth, val_sf,
                                        pred_stack, soil_vars,
                                        methods = c("RF", "RK"),
                                        label = "dataset") {

  # ── 1. Extract zone for each validation point ─────────────
  val_vect    <- terra::vect(val_sf)
  val_zone    <- terra::extract(zone_rast_smooth, val_vect)
  val_sf$zone <- val_zone[, 2]

  n_assigned  <- sum(!is.na(val_sf$zone))
  cat(label, "— validation points with zone assignment:",
      n_assigned, "of", nrow(val_sf), "\n")

# Replace step 2 inside validate_zones_with_holdout

  # ── 2. Per-zone summary of PREDICTED values ───────────────
  col_names <- as.vector(outer(soil_vars, methods, paste, sep = "_"))
  col_names <- col_names[col_names %in% names(pred_stack)]

  # Extract zone from smoothed raster at pred_stack coordinates
  # rather than joining by x/y — avoids floating point mismatch
  pred_pts  <- terra::vect(
    pred_stack %>% select(x, y),
    geom = c("x", "y"), crs = "EPSG:26917"
  )
  pred_zone_extract <- terra::extract(zone_rast_smooth, pred_pts, ID = FALSE)
  pred_stack_zoned  <- pred_stack %>%
    mutate(zone = pred_zone_extract[[1]])

  zone_profiles <- pred_stack_zoned %>%
    filter(!is.na(zone)) %>%
    group_by(zone) %>%
    summarise(across(all_of(col_names), mean, na.rm = TRUE),
              .groups = "drop")

  # ── 3. Compare observed val points to zone predicted means ─
  val_long <- val_sf %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    pivot_longer(-zone, names_to = "variable", values_to = "obs")

  # FIX: use the actual method suffix, not hardcoded "_RF"
  method_suffix <- paste0("_", methods[1])   # "_RF" or "_RK"

  pred_long <- zone_profiles %>%
    select(zone, ends_with(method_suffix)) %>%
    pivot_longer(-zone, names_to = "col", values_to = "pred_zone_mean") %>%
    mutate(variable = str_remove(col, method_suffix)) %>%
    select(zone, variable, pred_zone_mean)

  comparison <- val_long %>%
    left_join(pred_long, by = c("zone", "variable")) %>%
    group_by(zone, variable) %>%
    summarise(
      n_val_pts      = n(),
      mean_obs       = mean(obs,                        na.rm = TRUE),
      pred_zone_mean = first(pred_zone_mean),
      bias           = mean(obs - pred_zone_mean,       na.rm = TRUE),
      rmse_vs_zone   = sqrt(mean((obs - pred_zone_mean)^2, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(dataset = label)

  # ── 4. Kruskal-Wallis: do zones differ in observed values? ─
  kw_results <- purrr::map_dfr(soil_vars, function(v) {
    df <- val_sf %>%
      st_drop_geometry() %>%
      filter(!is.na(zone), !is.na(.data[[v]])) %>%
      select(zone, value = all_of(v))

    if (length(unique(df$zone)) < 2 || nrow(df) < 5) {
      return(tibble(variable = v, kw_p = NA_real_, dataset = label))
    }

    kt <- kruskal.test(value ~ factor(zone), data = df)
    tibble(variable = v, kw_p = kt$p.value, dataset = label)
  })

  list(
    comparison     = comparison,
    kw_results     = kw_results,
    val_with_zones = val_sf
  )
}
get_pred_stack <- function(nm) {
  clean   <- nm %>%
    str_remove("_kmeans") %>%
    str_remove("_nolags") %>%
    str_remove("_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]; method <- parts[2]
  has_lag <- str_detect(nm, "_lags")

  pred_df <- switch(paste0(samp, "_", method),
    SP_RF   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_RK   = if (has_lag) pred_stack_SP_lag   else pred_stack_SP,
    SP_OK   = pred_stack_SP,
    COMP_RF = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_RK = if (has_lag) pred_stack_COMP_lag else pred_stack_COMP,
    COMP_OK = pred_stack_COMP)

  list(pred_df = pred_df, method = method,
       samp = samp, has_lag = has_lag)
}

# validate_zones_with_holdout() — must already be defined
# in workspace or sourced separately.

val_umap <- purrr::imap(pp_named, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- UMAP+HDBSCAN:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method, label = nm)
})
val_umap <- val_umap[!sapply(val_umap, is.null)]

val_kmeans <- purrr::imap(pp_baseline, function(pp, nm) {
  if (is.null(pp$zone_rast_smooth)) {
    cat(nm, "— skipped (no zones)\n"); return(NULL)
  }
  cfg <- get_pred_stack(nm)
  cat("\n--- PCA+kmeans:", nm, "---\n")
  validate_zones_with_holdout(
    pp$zone_rast_smooth, ap_val_sf,
    cfg$pred_df, soil_vars,
    methods = cfg$method,
    label   = paste0(nm, "_kmeans"))
})
val_kmeans <- val_kmeans[!sapply(val_kmeans, is.null)]

# Add OK configs missed in original loop
# val_ok_missing <- list(
#   SP_OK_nolags   = validate_zones_with_holdout(
#     pp_named[["SP_OK_nolags"]]$zone_rast_smooth,
#     ap_val_sf, pred_stack_SP, soil_vars,
#     methods = "OK", label = "SP_OK_nolags"),
#   COMP_OK_nolags = validate_zones_with_holdout(
#     pp_named[["COMP_OK_nolags"]]$zone_rast_smooth,
#     ap_val_sf, pred_stack_COMP, soil_vars,
#     methods = "OK", label = "COMP_OK_nolags")
# )

val_named <- val_umap  # all 10 configs including OK
all_val   <- c(val_named, val_kmeans)

all_kw <- purrr::map_dfr(all_val, ~ .x$kw_results) %>%
  mutate(
    significant = kw_p < 0.05,
    approach    = if_else(str_detect(dataset, "_kmeans"),
                          "PCA+kmeans", "UMAP+HDBSCAN"),
    clean       = dataset %>%
                    str_remove("_kmeans") %>%
                    str_remove("_nolags") %>%
                    str_remove("_lags"),
    sampling    = str_extract(clean, "^(SP|COMP)"),
    pred_method = str_extract(clean, "(RF|RK|OK)"),
    lags        = str_detect(dataset, "_lags")
  ) %>%
  select(-clean)

all_comparisons      <- purrr::map_dfr(val_umap,   ~ .x$comparison)
all_comparisons_pca  <- purrr::map_dfr(val_kmeans, ~ .x$comparison)
all_comparisons_full <- bind_rows(all_comparisons, all_comparisons_pca)

kw_summary <- all_kw %>%
  group_by(dataset, approach, sampling, pred_method, lags) %>%
  summarise(
    n_significant   = sum(significant, na.rm = TRUE),
    n_total         = n(),
    pct_significant = round(100 * mean(significant, na.rm = TRUE)),
    sig_vars        = paste(variable[significant], collapse = ", "),
    .groups         = "drop"
  ) %>%
  arrange(approach, sampling, pred_method, lags)

print(kw_summary, n = 30)
write_csv(all_kw,      "../outputs/zone_kruskal_wallis_all.csv")
write_csv(kw_summary,  "../outputs/kw_summary_all.csv")
```

# ============================================================
# PHASE 8c — Per-Variable Observed VR + Effect Sizes
# ============================================================

```{r observed-vr-and-effect-sizes}
calc_zone_consistency <- function(val_with_zones,
                                   soil_vars, label) {
  df <- val_with_zones %>%
    st_drop_geometry() %>%
    filter(!is.na(zone)) %>%
    select(zone, all_of(soil_vars)) %>%
    na.omit()

  if (nrow(df) < 5 || length(unique(df$zone)) < 2) {
    return(tibble(configuration = label,
                  observed_VR_mean = NA_real_,
                  observed_VR_med  = NA_real_,
                  n_pos_VR         = NA_integer_))
  }

  total_var_vec <- df %>%
    select(all_of(soil_vars)) %>%
    summarise(across(everything(),
                     ~ var(.x, na.rm = TRUE))) %>%
    unlist()

  per_var_vr <- purrr::map_dbl(soil_vars, function(v) {
    within_var <- df %>%
      group_by(zone) %>%
      summarise(wv = var(.data[[v]], na.rm = TRUE),
                .groups = "drop") %>%
      pull(wv) %>%
      mean(na.rm = TRUE)
    1 - (within_var / total_var_vec[[v]])
  })
  names(per_var_vr) <- soil_vars

  tibble(configuration    = label,
         observed_VR_mean = round(mean(per_var_vr,   na.rm = TRUE), 3),
         observed_VR_med  = round(median(per_var_vr, na.rm = TRUE), 3),
         n_pos_VR         = sum(per_var_vr > 0, na.rm = TRUE))
}

obs_vr_all <- bind_rows(
  purrr::imap_dfr(val_named,  ~ calc_zone_consistency(
    .x$val_with_zones, soil_vars, .y)) %>%
    rename(config_join = configuration),
  purrr::imap_dfr(val_kmeans, ~ calc_zone_consistency(
    .x$val_with_zones, soil_vars, paste0(.y, "_kmeans"))) %>%
    rename(config_join = configuration)
)

# Zone prediction RMSE
zone_pred_rmse <- all_comparisons_full %>%
  filter(!is.na(rmse_vs_zone)) %>%
  group_by(dataset) %>%
  summarise(
    mean_zone_rmse   = round(mean(rmse_vs_zone,   na.rm = TRUE), 3),
    median_zone_rmse = round(median(rmse_vs_zone, na.rm = TRUE), 3),
    .groups          = "drop"
  ) %>%
  rename(config_join = dataset)

# R²
zone_r2 <- all_comparisons_full %>%
  filter(!is.na(pred_zone_mean), !is.na(mean_obs)) %>%
  group_by(dataset) %>%
  summarise(
    r2_obs_pred = round(
      cor(mean_obs, pred_zone_mean, use = "complete.obs")^2, 3),
    .groups = "drop"
  ) %>%
  rename(config_join = dataset)

# Per-variable observed VR for heatmap
obs_vr_by_var_all <- bind_rows(
  purrr::imap_dfr(val_named, function(x, nm) {
    df <- x$val_with_zones %>% st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>% na.omit()
    if (nrow(df) < 5 || length(unique(df$zone)) < 2) return(NULL)
    tv <- df %>% select(all_of(soil_vars)) %>%
      summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
      unlist()
    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>% group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
        pull(wv) %>% mean(na.rm = TRUE)
      tibble(configuration = nm, variable = v,
             obs_VR = round(1 - wv / tv[[v]], 3))
    })
  }),
  purrr::imap_dfr(val_kmeans, function(x, nm) {
    df <- x$val_with_zones %>% st_drop_geometry() %>%
      filter(!is.na(zone)) %>%
      select(zone, all_of(soil_vars)) %>% na.omit()
    if (nrow(df) < 5 || length(unique(df$zone)) < 2) return(NULL)
    tv <- df %>% select(all_of(soil_vars)) %>%
      summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
      unlist()
    purrr::map_dfr(soil_vars, function(v) {
      wv <- df %>% group_by(zone) %>%
        summarise(wv = var(.data[[v]], na.rm = TRUE), .groups = "drop") %>%
        pull(wv) %>% mean(na.rm = TRUE)
      tibble(configuration = paste0(nm, "_kmeans"), variable = v,
             obs_VR = round(1 - wv / tv[[v]], 3))
    })
  })
)

# Supplementary heatmap
p_obs_vr <- obs_vr_by_var_all %>%
  left_join(
    bind_rows(
      zone_validation_umap %>%
        select(configuration, clustering = approach),
      zone_validation_pca  %>%
        mutate(configuration = if_else(
          !str_detect(configuration, "_kmeans"),
          paste0(configuration, "_kmeans"),
          configuration)) %>%
        select(configuration, clustering = approach)
    ),
    by = "configuration"
  ) %>%
  mutate(
    config_label = str_replace_all(
      configuration,
      c("_nolags" = "", "_lags" = "+lags",
            "_kmeans" = "",  "COMP_" = "COMP/", "SP_" = "SP/")),
              clustering = if_else(
                is.na(clustering),
                if_else(str_detect(configuration, "_kmeans"),
                        "PCA+kmeans",
                        "UMAP+HDBSCAN"), clustering),
    approach   = factor(clustering,
                        levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  ) %>%
  ggplot(aes(x = config_label, y = variable, fill = obs_VR)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", obs_VR)),
            size = 2.2) +
  facet_wrap(~ approach, scales = "free_x") +
  scale_fill_gradient2(
    low = "#d73027", mid = "#ffffbf", high = "#1a9850",
    midpoint = 0, limits = c(-1, 1),
    oob  = scales::squish, name = "Observed VR"
  ) +
  theme_bw(base_size = 9) +
  theme(axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
        strip.background = element_rect(fill = "grey92"),
        strip.text       = element_text(face = "bold"),
        panel.grid       = element_blank()) +
  labs(
    title    = "Supplementary Figure S1. Per-variable observed variance reduction",
    subtitle = "Held-out validation points (n = 59), fully independent of prediction surfaces",
    x = NULL, y = "Soil variable"
  )

print(p_obs_vr)
ggsave("../outputs/figS1_observed_vr_heatmap.png",
       p_obs_vr, width = 12, height = 4, dpi = 300)
write_csv(obs_vr_by_var_all,
          "../outputs/observed_vr_per_variable.csv")
```

# ============================================================
# Final Summary + Results Table
# ============================================================

```{r final-summary-and-table}
# ── Build summary tables ───────────────────────────────────
summary_umap <- purrr::imap_dfr(mz_named_tuned, function(x, nm) {
  parts  <- strsplit(nm, "_")[[1]]
  samp   <- parts[1]; method <- parts[2]
  lag_yn <- length(parts) >= 3 && parts[3] == "lags"
  tibble(configuration     = nm,
         clustering        = "UMAP+HDBSCAN",
         sampling_design   = ifelse(samp == "SP",
                                    "Single-point", "Composite"),
         pred_method       = method,
         spatial_lags      = lag_yn,
         minPts_used       = minpts_final[[nm]],
         n_zones           = x$n_zones,
         pct_noise_prefill = round(x$pct_noise, 1))
}) %>%
  left_join(
    all_kw %>% filter(approach == "UMAP+HDBSCAN") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                sig_vars   = paste(variable[significant], collapse = ", "),
                .groups    = "drop") %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(
    n_sig_vars = replace_na(n_sig_vars, 0L),
    sig_vars   = replace_na(sig_vars,   "none"),
    sig_vars   = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(
    zone_validation_all %>% filter(approach == "UMAP+HDBSCAN") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

summary_pca <- purrr::imap_dfr(baseline_list, function(b, nm) {
  clean   <- str_remove(nm, "_kmeans")
  has_lag <- str_detect(clean, "_lags")
  clean   <- str_remove(clean, "_lags")
  parts   <- strsplit(clean, "_")[[1]]
  samp    <- parts[1]; method <- parts[2]
  tibble(configuration     = nm,
         clustering        = "PCA+kmeans",
         sampling_design   = ifelse(samp == "SP",
                                    "Single-point", "Composite"),
         pred_method       = method,
         spatial_lags      = has_lag,
         minPts_used       = NA_real_,
         n_zones           = b$best_k,
         pct_noise_prefill = 0)
}) %>%
  left_join(
    all_kw %>% filter(approach == "PCA+kmeans") %>%
      group_by(dataset) %>%
      summarise(n_sig_vars = sum(significant, na.rm = TRUE),
                sig_vars   = paste(variable[significant], collapse = ", "),
                .groups    = "drop") %>%
      rename(configuration = dataset),
    by = "configuration"
  ) %>%
  mutate(
    n_sig_vars = replace_na(n_sig_vars, 0L),
    sig_vars   = replace_na(sig_vars,   "none"),
    sig_vars   = ifelse(sig_vars == "" | is.na(sig_vars), "none", sig_vars)
  ) %>%
  left_join(
    zone_validation_all %>% filter(approach == "PCA+kmeans") %>%
      select(configuration, VR, silhouette),
    by = "configuration"
  )

# ── Join all metrics ───────────────────────────────────────
final_summary <- bind_rows(summary_umap, summary_pca) %>%
  mutate(
    config_join = case_when(
      clustering == "PCA+kmeans" &
        !str_detect(configuration, "_kmeans") ~
        paste0(configuration, "_kmeans"),
      TRUE ~ configuration
    ),
    sig_vars = ifelse(sig_vars == "" | is.na(sig_vars),
                      "none", sig_vars)
  ) %>%
  left_join(zone_pred_rmse, by = "config_join") %>%
  left_join(obs_vr_all,     by = "config_join") %>%
  left_join(zone_r2 %>% select(config_join, r2_obs_pred),
            by = "config_join") %>%
  select(-config_join)

# ── Composite ranking ─────────────────────────────────────
final_ranked <- final_summary %>%
  mutate(
    vr_rank     = rank(-VR,              na.last = "keep"),
    sil_rank    = rank(-silhouette,      na.last = "keep"),
    kw_rank     = rank(-n_sig_vars,      na.last = "keep"),
    rmse_rank   = rank(median_zone_rmse, na.last = "keep"),
    obs_vr_rank = rank(-observed_VR_med, na.last = "keep"),
    npos_rank   = rank(-n_pos_VR,        na.last = "keep"),
    r2_rank     = rank(-r2_obs_pred,     na.last = "keep"),
    Rank        = round(vr_rank + sil_rank + kw_rank +
                          rmse_rank + obs_vr_rank +
                          npos_rank + r2_rank, 1)
  ) %>%
  select(configuration, Rank)

cat("\n=== Composite ranking — all configurations ===\n")
final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  arrange(Rank) %>%
  select(configuration, clustering, VR, silhouette,
         n_sig_vars, median_zone_rmse,
         observed_VR_med, n_pos_VR, r2_obs_pred, Rank) %>%
  mutate(across(where(is.double), ~ round(.x, 3))) %>%
  print(n = 20)

write_csv(final_summary,
          "../outputs/final_results_summary_complete.csv")

# ── gt Results Table ───────────────────────────────────────
table_data <- final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  mutate(
    Approach  = clustering,
    Sampling  = ifelse(sampling_design == "Single-point",
                       "SP", "COMP"),
    Method    = pred_method,
    Lags      = ifelse(spatial_lags, "Yes", "No"),
    Zones     = n_zones,
    VR_pred   = round(VR,               3),
    Sil       = round(silhouette,       3),
    N_sig     = n_sig_vars,
    RMSE_zone = round(median_zone_rmse, 1),
    VR_obs    = round(observed_VR_med,  3),
    N_pos     = n_pos_VR,
    R2        = round(r2_obs_pred,      3)
  ) %>%
  select(Approach, Sampling, Method, Lags, Zones,
         VR_pred, Sil, N_sig, RMSE_zone,
         VR_obs, N_pos, R2, Rank) %>%
  arrange(Approach, Sampling, Method, Lags)

best <- list(
  VR_pred   = max(table_data$VR_pred,   na.rm = TRUE),
  Sil       = max(table_data$Sil,       na.rm = TRUE),
  N_sig     = max(table_data$N_sig,     na.rm = TRUE),
  RMSE_zone = min(table_data$RMSE_zone, na.rm = TRUE),
  VR_obs    = max(table_data$VR_obs,    na.rm = TRUE),
  N_pos     = max(table_data$N_pos,     na.rm = TRUE),
  R2        = max(table_data$R2,        na.rm = TRUE),
  Rank      = min(table_data$Rank,      na.rm = TRUE)
)

tbl <- table_data %>%
  gt(groupname_col = "Approach") %>%
  cols_label(
    Sampling  = "Sampling",   Method    = "Pred. method",
    Lags      = "Lags",       Zones     = md("*k*"),
    VR_pred   = md("VR~pred~"), Sil     = "Silhouette",
    N_sig     = md("*N*~sig~"), RMSE_zone = md("RMSE~zone~"),
    VR_obs    = md("VR~obs~ (median)"),
    N_pos     = md("*N*~pos~"), R2      = md("*R*²"),
    Rank      = "Rank"
  ) %>%
  tab_spanner(label = md("**Configuration**"),
              columns = c(Sampling, Method, Lags, Zones)) %>%
  tab_spanner(
    label   = md("**Zone structure** *(all grid pixels)*"),
    columns = c(VR_pred, Sil)) %>%
  tab_spanner(
    label   = md("**Agronomic validation** *(held-out, n = 59)*"),
    columns = c(N_sig, RMSE_zone, VR_obs, N_pos, R2)) %>%
  tab_spanner(label = md("**Overall**"), columns = Rank) %>%
  tab_style(
    style     = list(cell_fill(color = "#d6eaf8"),
                     cell_text(weight = "bold")),
    locations = cells_body(
      rows = Sampling == "COMP" & Method == "RF" &
               Lags == "Yes" & Approach == "UMAP+HDBSCAN")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fde8e8"),
    locations = cells_body(columns = VR_obs,
                           rows = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  tab_style(
    style     = cell_text(color = "#c0392b"),
    locations = cells_body(columns = VR_obs,
                           rows = !is.na(VR_obs) & VR_obs < 0)
  ) %>%
  fmt_missing(columns = everything(), missing_text = "—") %>%
  fmt_number(columns = c(VR_pred, Sil, VR_obs, R2), decimals = 3) %>%
  fmt_number(columns = RMSE_zone, decimals = 1) %>%
  fmt_number(columns = Rank,      decimals = 1) %>%
  fmt_integer(columns = c(Zones, N_sig, N_pos)) %>%
  tab_caption(
    md("**Table 1.** Zone quality metrics for all 20 management zone configurations (5-variable agronomic subset: LB-Ceq, pH, Ca, K, P). SP = single-point; COMP = composite; RF = random forest; RK = regression kriging; OK = ordinary kriging. *k* = zones after post-processing. Blue row = best overall configuration. Red = negative VR~obs~.")
  ) %>%
  tab_options(
    table.font.size                   = px(11),
    column_labels.font.size           = px(11),
    table.width                       = pct(100),
    data_row.padding                  = px(4),
    table.border.top.style            = "solid",
    table.border.top.width            = px(2),
    table.border.top.color            = "#2c3e50",
    column_labels.border.bottom.style = "solid",
    column_labels.border.bottom.width = px(1),
    column_labels.border.bottom.color = "#7f8c8d"
  )

tbl
gt::gtsave(tbl, "../outputs/table1_zone_metrics.html")
cat("Table saved.\n")

save.image("../outputs/workspace_final.RData")
beep("complete")

#UMAP is more sensitive to variable reduction than PCA+kmeans — PCA+kmeans performance is relatively stable across both variable sets, while UMAP configurations shift substantially. This makes sense because UMAP's manifold learning is more dependent on the full multivariate structure of the input space
# Lag augmentation loses its benefit — COMP_RF_lags was your Chapter 1 winner but drops to 0 sig vars here. The spatial lag features may have been capturing structure in the dropped variables (lbc_1, mn, zn) more than in the agronomic subset
# SP configurations become more competitive — SP_RF_nolags and SP_RK_nolags now match or beat COMP configs on n_sig, suggesting composite sampling's advantage diminishes when the variable set is reduced
# The best agronomic config is SP_RF_nolags or SP_RK_nolags with 3 significant variables each



```

# ============================================================
# PHASE 9 — Three-panel thesis figure
# ============================================================

```{r thesis-figure}
fig_data <- final_summary %>%
  left_join(final_ranked, by = "configuration") %>%
  mutate(
    x_group = factor(
      paste0(pred_method, ifelse(spatial_lags, "+lags", "")),
      levels = c("RF", "RF+lags", "RK", "RK+lags", "OK")
    ),
    sampling_f   = factor(sampling_design,
                          levels = c("Single-point", "Composite")),
    clustering_f = factor(clustering,
                          levels = c("PCA+kmeans", "UMAP+HDBSCAN"))
  )

clust_colors <- c("PCA+kmeans"   = "#4E9AF1",
                  "UMAP+HDBSCAN" = "#E05C3A")
lag_alpha    <- c("No lags" = 0.5, "With lags" = 1.0)

lag_aes <- aes(
  x     = x_group,
  fill  = clustering_f,
  alpha = factor(spatial_lags,
                 levels = c(FALSE, TRUE),
                 labels = c("No lags", "With lags"))
)

base_theme <- theme_bw(base_size = 9) +
  theme(
    legend.position    = "none",
    strip.background   = element_rect(fill = "grey92"),
    strip.text         = element_text(size = 8, face = "bold"),
    axis.text.x        = element_text(size = 7, angle = 30, hjust = 1),
    panel.grid.major.x = element_blank()
  )

pA <- fig_data %>%
  ggplot(lag_aes) + aes(y = VR) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.40, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha, name = "Spatial lags") +
  scale_y_continuous(limits = c(0, 0.85), breaks = seq(0, 0.8, 0.2)) +
  labs(title = "A).  Variance Reduction",
       subtitle = "Higher = zones more internally homogeneous",
       x = NULL, y = "VR") +
  base_theme

pB <- fig_data %>%
  ggplot(lag_aes) + aes(y = silhouette) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.50, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha, name = "Spatial lags") +
  scale_y_continuous(limits = c(0, 0.65), breaks = seq(0, 0.6, 0.2)) +
  labs(title = "B).  Silhouette Coefficient",
       subtitle = "Higher = zones more spatially distinct",
       x = NULL, y = "Silhouette") +
  base_theme

pC <- fig_data %>%
  ggplot(lag_aes) + aes(y = n_sig_vars) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_text(aes(label = ifelse(n_sig_vars > 0,
                                as.character(n_sig_vars), "")),
            position = position_dodge(0.8),
            vjust = -0.4, size = 2.5, color = "grey30") +
  facet_wrap(~ sampling_f) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha, name = "Spatial lags") +
  scale_y_continuous(breaks = 0:5,
                     limits = c(0, length(soil_vars) + 0.5)) +
  labs(title = paste0("C). Significant Soil Properties (of ",
                       length(soil_vars), ") (KW, p < 0.05)"),
       subtitle = "Validation points: n = 59",
       x = "Prediction method",
       y = "# of properties significant") +
  base_theme

# Shared legend
legend_plot <- fig_data %>%
  ggplot(lag_aes) + aes(y = VR) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  scale_fill_manual(values = clust_colors, name = "Clustering") +
  scale_alpha_manual(values = lag_alpha,   name = "Spatial lags") +
  theme_bw(base_size = 9) +
  theme(legend.position  = "bottom",
        legend.box       = "horizontal",
        legend.title     = element_text(size = 8, face = "bold"),
        legend.text      = element_text(size = 8),
        legend.key.size  = unit(0.4, "cm"))

shared_legend <- cowplot::get_legend(legend_plot)

fig_combined <- cowplot::plot_grid(
  pA / pB / pC,
  shared_legend,
  ncol    = 1,
  rel_heights = c(10, 1)
)

print(fig_combined)
ggsave("../outputs/fig_zone_quality_metrics.png",
       fig_combined, width = 9, height = 11, dpi = 300)
cat("Figure saved.\n")
```

```{r}
# For a given config, correlate each soil var prediction with UMAP coords
umap_contributions <- purrr::map_dfr(soil_vars, function(v) {
  pred_col <- paste0(v, "_RF")  # or _RK, _OK
  umap_df  <- mz_named_tuned[["COMP_RF_lags"]]$result_df
  
  pred_vals <- pred_stack_COMP_lag %>%
    filter(complete.cases(.)) %>%
    pull(pred_col)
  
  tibble(
    variable = v,
    cor_UMAP1 = cor(pred_vals, umap_df$UMAP1, use = "complete.obs"),
    cor_UMAP2 = cor(pred_vals, umap_df$UMAP2, use = "complete.obs")
  )
})
print(umap_contributions)

# Drop one variable at a time, rerun UMAP, record VR
loo_results <- purrr::map_dfr(soil_vars, function(drop_var) {
  vars_subset <- soil_vars[soil_vars != drop_var]
  
  mz <- run_umap_hdbscan(
    pred_stack_COMP_lag,
    methods      = "RF",
    soil_vars    = vars_subset,
    include_lags = TRUE,
    label        = paste0("drop_", drop_var)
  )
  
  pp <- post_process_zones(mz)
  vr <- calc_variance_reduction(
    pred_stack_COMP_lag, mz$zone, vars_subset, "RF")
  
  tibble(dropped = drop_var, n_vars = length(vars_subset), VR = vr)
})
print(loo_results)

pred_cols <- paste0(soil_vars, "_RF")

pca_contributions <- pred_stack_COMP %>%
  select(all_of(pred_cols)) %>%
  na.omit() %>%
  scale() %>%
  prcomp()

# Loadings on first 2 PCs
loadings <- as.data.frame(pca_contributions$rotation[, 1:2]) %>%
  rownames_to_column("variable") %>%
  mutate(variable = str_remove(variable, "_RF"),
         loading_magnitude = sqrt(PC1^2 + PC2^2)) %>%
  arrange(desc(loading_magnitude))

print(umap_contributions)
print(loo_results)
print(loadings)
# pH and K are the primary drivers of zone structure (UMAP1), while the broad fertility variables (Ca, Mg, P, lb_ceq) collectively define UMAP2. Zn and Ca may be candidates for exclusion in future work as they appear to add noise rather than structure — but removing the full agronomic subset simultaneously collapsed both axes, explaining the Chapter 2 performance drop.
```
```{r}
#pH vs P availability:
bind_rows(
  soil_prop_SP   %>% st_drop_geometry() %>% mutate(dataset = "Single-Point"),
  soil_prop_COMP %>% st_drop_geometry() %>% mutate(dataset = "Composite")
) %>%
  mutate(dataset = factor(dataset, levels = c("Single-Point", "Composite"))) %>%
  ggplot(aes(x = p_h_2, y = p, color = dataset)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Single-Point" = "#2980b9",
                                 "Composite"    = "#e74c3c")) +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10)
  ) +
  labs(
    title    = "Phosphorus vs pH by sampling design",
    subtitle = "Linear trend shown with 95% confidence interval",
    x        = "pH",
    y        = expression("Phosphorus (mg (ppm))")
  )

# ── Correlation coefficients ──────────────────────────────
bind_rows(
  soil_prop_SP   %>% st_drop_geometry() %>% mutate(dataset = "Single-Point"),
  soil_prop_COMP %>% st_drop_geometry() %>% mutate(dataset = "Composite")
) %>%
  group_by(dataset) %>%
  summarise(
    r       = round(cor(p_h_2, p, use = "complete.obs"), 3),
    r2      = round(cor(p_h_2, p, use = "complete.obs")^2, 3),
    p_value = round(cor.test(p_h_2, p)$p.value, 4),
    n       = n(),
    .groups = "drop"
  ) %>%
  print()

# ── Plot with polynomial curve ────────────────────────────
# Second-degree polynomial captures the bell-shaped
# availability curve — P availability peaks around pH 6–7
# and declines in strongly acid or alkaline conditions

plot_dat <- bind_rows(
  soil_prop_SP   %>% st_drop_geometry() %>% mutate(dataset = "Single-Point"),
  soil_prop_COMP %>% st_drop_geometry() %>% mutate(dataset = "Composite")
) %>%
  mutate(dataset = factor(dataset,
                          levels = c("Single-Point", "Composite")))

# Compute r per dataset for annotation
cor_dat <- plot_dat %>%
  group_by(dataset) %>%
  summarise(
    r       = round(cor(p_h_2, p, use = "complete.obs"), 3),
    p_value = cor.test(p_h_2, p)$p.value,
    label   = paste0("r = ", r,
                     ifelse(p_value < 0.001, "***",
                     ifelse(p_value < 0.01,  "**",
                     ifelse(p_value < 0.05,  "*", "")))),
    x       = max(plot_dat$p_h_2, na.rm = TRUE) - 0.15,
    y       = max(plot_dat$p,     na.rm = TRUE) - 2,
    .groups = "drop"
  )

ggplot(plot_dat, aes(x = p_h_2, y = p, color = dataset)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method  = "lm",
              formula = y ~ poly(x, 2),
              se      = TRUE,
              linewidth = 0.9) +
  geom_text(data = cor_dat,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            size = 3.5, hjust = 1, fontface = "italic") +
  scale_color_manual(values = c("Single-Point" = "#2980b9",
                                 "Composite"    = "#e74c3c")) +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10)
  ) +
  labs(
    title    = "Phosphorus availability vs pH by sampling design",
    subtitle = "Polynomial (degree 2) fit — captures bell-shaped P availability curve",
    x        = "pH",
    y        = expression("Phosphorus (mg kg"^{-1}*")"),
    caption  = "P availability peaks near pH 6–7; declines under acidic or alkaline conditions (Brady & Weil, 2008)"
  )
```
```{r}
#all nutrient correlation

# ── All soil property pairwise correlations ───────────────
soil_var_labels <- c(
  "lbc_1"  = "LBC-1",
  "lb_ceq" = "LBC Eq.",
  "p_h_2"  = "pH",
  "ca"     = "Calcium",
  "k"      = "Potassium",
  "mg"     = "Magnesium",
  "mn"     = "Manganese",
  "p"      = "Phosphorus",
  "zn"     = "Zinc"
)

# ── Correlation matrix + p-values ─────────────────────────
cor_pval <- function(dat, vars) {
  n <- length(vars)
  r_mat <- matrix(NA, n, n, dimnames = list(vars, vars))
  p_mat <- matrix(NA, n, n, dimnames = list(vars, vars))
  
  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      if (i == j) {
        r_mat[i, j] <- 1
        p_mat[i, j] <- NA
      } else {
        ct <- cor.test(dat[[vars[i]]], dat[[vars[j]]],
                       use = "complete.obs")
        r_mat[i, j] <- ct$estimate
        p_mat[i, j] <- ct$p.value
      }
    }
  }
  list(r = r_mat, p = p_mat)
}

cor_SP   <- cor_pval(soil_prop_SP   %>% st_drop_geometry(), soil_vars)
cor_COMP <- cor_pval(soil_prop_COMP %>% st_drop_geometry(), soil_vars)

# ── Tidy for ggplot ───────────────────────────────────────
tidy_cor <- function(cor_list, dataset_label) {
  as.data.frame(cor_list$r) %>%
    rownames_to_column("var1") %>%
    pivot_longer(-var1, names_to = "var2", values_to = "r") %>%
    left_join(
      as.data.frame(cor_list$p) %>%
        rownames_to_column("var1") %>%
        pivot_longer(-var1, names_to = "var2", values_to = "p_value"),
      by = c("var1", "var2")
    ) %>%
    mutate(
      dataset  = dataset_label,
      sig      = case_when(
        is.na(p_value)  ~ "",
        p_value < 0.001 ~ "***",
        p_value < 0.01  ~ "**",
        p_value < 0.05  ~ "*",
        TRUE            ~ ""
      ),
      label    = ifelse(var1 == var2, "",
                        paste0(sprintf("%.2f", r), sig)),
      var1     = recode(var1, !!!soil_var_labels),
      var2     = recode(var2, !!!soil_var_labels),
      var1     = factor(var1, levels = soil_var_labels),
      var2     = factor(var2, levels = soil_var_labels)
    )
}

cor_tidy <- bind_rows(
  tidy_cor(cor_SP,   "Single-Point"),
  tidy_cor(cor_COMP, "Composite")
) %>%
  mutate(dataset = factor(dataset,
                          levels = c("Single-Point", "Composite")))

# ── Plot ──────────────────────────────────────────────────
ggplot(cor_tidy, aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = label),
            size = 2.5, color = "grey10") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9850",
    midpoint = 0,
    limits   = c(-1, 1),
    oob      = scales::squish,
    name     = "Pearson r",
    na.value = "grey90"
  ) +
  facet_wrap(~ dataset) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1,
                                     size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid       = element_blank(),
    legend.position  = "right"
  ) +
  labs(
    title    = "Pairwise soil property correlations by sampling design",
    subtitle = "Values shown with significance: *p<0.05  **p<0.01  ***p<0.001",
    x        = NULL,
    y        = NULL
  )

ggsave("../outputs/fig_soil_pairwise_correlations.png",
       width = 12, height = 6, dpi = 300)

#save all results in csv
write_csv(cor_tidy, "../outputs/soil_property_correlations.csv")
```


```{r}
#RK residual variograms + fits
# ── RK Residual Variogram Summary ─────────────────────────
# Extract fitted residual variogram parameters from
# final_models_SP and final_models_COMP for all soil vars

extract_rk_vg_metrics <- function(final_models, soil_vars, label) {
  purrr::map_dfr(soil_vars, function(v) {
    fit <- final_models[[v]]$rk$fit_resid
    if (is.null(fit)) {
      return(tibble(
        variable = v, dataset = label,
        model = NA, nugget = NA, psill = NA,
        total_sill = NA, range_m = NA,
        nugget_sill_ratio = NA
      ))
    }
    nug_row  <- fit %>% filter(model == "Nug")
    str_row  <- fit %>% filter(model != "Nug")
    nugget   <- if (nrow(nug_row) > 0) nug_row$psill else 0
    psill    <- if (nrow(str_row) > 0) str_row$psill else NA
    rng      <- if (nrow(str_row) > 0) str_row$range else NA
    mod_type <- if (nrow(str_row) > 0) as.character(str_row$model) else NA
    tibble(
      variable          = v,
      dataset           = label,
      model             = mod_type,
      nugget            = round(nugget, 4),
      psill             = round(psill,  4),
      total_sill        = round(nugget + psill, 4),
      range_m           = round(rng, 1),
      nugget_sill_ratio = round(nugget / (nugget + psill), 3)
    )
  })
}

rk_vg_summary <- bind_rows(
  extract_rk_vg_metrics(final_models_SP,   soil_vars, "SP"),
  extract_rk_vg_metrics(final_models_COMP, soil_vars, "COMP")
) %>%
  mutate(
    spatial_dependence = case_when(
      is.na(nugget_sill_ratio)      ~ "No convergent model",
      nugget_sill_ratio < 0.25      ~ "Strong",
      nugget_sill_ratio < 0.75      ~ "Moderate",
      TRUE                          ~ "Weak — kriging limited"
    ),
    variable = recode(variable,
      "lbc_1"  = "LBC-1",
      "lb_ceq" = "LBC Equivalent",
      "p_h_2"  = "pH",
      "ca"     = "Calcium",
      "k"      = "Potassium",
      "mg"     = "Magnesium",
      "mn"     = "Manganese",
      "p"      = "Phosphorus",
      "zn"     = "Zinc"
    ),
    dataset = factor(dataset, levels = c("SP", "COMP"))
  )

print(rk_vg_summary, n = 20)

# ── Plot 1: Nugget:Sill ratio comparison SP vs COMP ───────
p_rk_nugget <- rk_vg_summary %>%
  filter(!is.na(nugget_sill_ratio)) %>%
  ggplot(aes(x = reorder(variable, nugget_sill_ratio),
             y = nugget_sill_ratio,
             fill = dataset)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.75, linetype = "dashed",
             color = "grey40", linewidth = 0.5) +
  geom_hline(yintercept = 0.25, linetype = "dotted",
             color = "grey60", linewidth = 0.5) +
  annotate("text", x = 0.6, y = 0.77,
           label = "Weak threshold (0.75)",
           size = 2.8, color = "grey40", hjust = 0) +
  annotate("text", x = 0.6, y = 0.27,
           label = "Strong threshold (0.25)",
           size = 2.8, color = "grey60", hjust = 0) +
  scale_fill_manual(values = c("SP"   = "#2980b9",
                                "COMP" = "#e74c3c"),
                    name = "Sampling design") +
  scale_y_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, 0.25)) +
  coord_flip() +
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    legend.position  = "bottom"
  ) +
  labs(
    title    = "RK Residual Variogram — Nugget:Sill Ratio",
    subtitle = "After RF trend removal. Cambardella et al. (1994) thresholds shown.",
    x        = NULL,
    y        = "Nugget:Sill ratio"
  )

print(p_rk_nugget)

# ── Plot 2: Range comparison SP vs COMP ───────────────────
p_rk_range <- rk_vg_summary %>%
  filter(!is.na(range_m)) %>%
  ggplot(aes(x = reorder(variable, range_m),
             y = range_m,
             fill = dataset)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  scale_fill_manual(values = c("SP"   = "#2980b9",
                                "COMP" = "#e74c3c"),
                    name = "Sampling design") +
  coord_flip() +
  theme_bw(base_size = 10) +
  theme(
    legend.position  = "bottom"
  ) +
  labs(
    title    = "RK Residual Variogram — Spatial Range",
    subtitle = "Distance at which residual spatial dependence ceases",
    x        = NULL,
    y        = "Range (m)"
  )

print(p_rk_range)

# ── Plot 3: Side by side OK vs RK nugget:sill comparison ──
ok_vg <- vg_summary_all %>%
  mutate(
    variable = recode(variable,
      "lbc_1"  = "LBC-1",
      "lb_ceq" = "LBC Equivalent",
      "p_h_2"  = "pH",
      "ca"     = "Calcium",
      "k"      = "Potassium",
      "mg"     = "Magnesium",
      "mn"     = "Manganese",
      "p"      = "Phosphorus",
      "zn"     = "Zinc"
    ),
    method = "OK"
  ) %>%
  select(variable, dataset, nugget_sill_ratio, method)

rk_vg <- rk_vg_summary %>%
  select(variable, dataset, nugget_sill_ratio) %>%
  mutate(method = "RK residual")

bind_rows(ok_vg, rk_vg) %>%
  filter(!is.na(nugget_sill_ratio)) %>%
  mutate(
    method  = factor(method, levels = c("OK", "RK residual")),
    dataset = factor(dataset, levels = c("SP", "COMP"))
  ) %>%
  ggplot(aes(x = variable, y = nugget_sill_ratio,
             fill = method)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_hline(yintercept = 0.75, linetype = "dashed",
             color = "grey40", linewidth = 0.4) +
  scale_fill_manual(values = c("OK"          = "#8e44ad",
                                "RK residual" = "#27ae60"),
                    name = "Method") +
  facet_wrap(~ dataset) +
  coord_flip() +
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom"
  ) +
  labs(
    title    = "Nugget:Sill ratio — OK vs RK residual variograms",
    subtitle = "Lower = stronger spatial structure. Dashed = weak threshold (0.75).",
    x        = NULL,
    y        = "Nugget:Sill ratio"
  )

ggsave("../outputs/fig_rk_residual_variograms.png",
       width = 10, height = 6, dpi = 300)

write_csv(rk_vg_summary,
          "../outputs/rk_residual_variogram_summary.csv")


# ── 1. OK variogram parameters (already computed) ─────────
vg_summary_all %>%
  mutate(method = "OK") %>%
  select(variable, dataset, method, 
         nugget_sill_ratio, range_m) %>%
  print(n = 20)

# ── 2. RK residual — already have from CSV ────────────────
# Already sent
# Option 1 — paste output
print(rk_vg_summary, n = 20)

# Option 2 — save and upload
write_csv(rk_vg_summary, 
          "../outputs/rk_residual_variogram_summary.csv")
# ── 3. RF residual Moran's I (spatial autocorrelation) ────
# This is the closest RF equivalent — tells you how much
# spatial structure remains after RF prediction
bind_rows(
  rf_moran_SP_correct   %>% mutate(dataset = "SP"),
  rf_moran_COMP_correct %>% mutate(dataset = "COMP")
) %>%
  mutate(
    variable = recode(variable,
      "lbc_1"  = "LBC-1",
      "lb_ceq" = "LBC Equivalent",
      "p_h_2"  = "pH",
      "ca"     = "Calcium",
      "k"      = "Potassium",
      "mg"     = "Magnesium",
      "mn"     = "Manganese",
      "p"      = "Phosphorus",
      "zn"     = "Zinc"
    ),
    method = "RF residual Moran's I"
  ) %>%
  select(variable, dataset, method, moran_I, p_value) %>%
  print(n = 20)
```

